---
title: "Amyloid test"
format: pdf
editor: visual
---

# Amyloid version (instead of plasma biomarkers)

## Problemet här = 

-   Finns både SCANDATE och PROCESSDATE som tidsvariabler, jag vet inte vilken jag ska utgå ifrån (hittills har jag utgått ifrån PROCESSDATE)

-   Jag behövde merge by både RID och VISCODE2 för att få diagnosis-variabeln att se korrekt ut (annars fick jag bara MCI, NC och NA, INTE Dementia)

```{r}
# loading libraries
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
```

## 1. We load our datasets

```{r}
# We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM

# We load amyloid data 
load(file.path("ADNIMERGE2_unpacked/ADNIMERGE2/data/UCBERKELEY_AMY_6MM.rda"))
amyloid_data <- UCBERKELEY_AMY_6MM
```

## 2. We select only the required variables and merge them

```{r}
# selecting the relevant variables 
amyloid_data <- amyloid_data %>%
  select(
    "RID",        # Participant ID
    "ORIGPROT",   # Original protocol
    "VISCODE2",   # Visit code
    "SCANDATE",   # Date of amyloid-PET
    "PROCESSDATE",
    "CENTILOIDS",  # Main longitudinal variable (continuous amyloid)
    "AMYLOID_STATUS"
  )

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# We merge by RID 
# merged_data = amyloid_data %>%
  #merge(diagnosis_data, by = "RID")

# We merge by RID and VISCODE2
merged_data <- amyloid_data %>%
  merge(diagnosis_data, by = c("RID", "VISCODE2"))
```

## 3. We filter participants with at least 2 measurements

```{r}
keep = unique((amyloid_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 2 measurements 
length(keep)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(PROCESSDATE)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event)
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarize(
    # Event = 1 om individen någonsin fått "Dementia", annars 0 (censurerad)
    event = ifelse(any(DIAGNOSIS == "Dementia"), 1, 0),
    
    # Tid till första "Dementia", annars tid till sista observation
    time_to_event = as.numeric(
      ifelse(
        any(DIAGNOSIS == "Dementia"),
        min(PROCESSDATE[DIAGNOSIS == "Dementia"]) - min(PROCESSDATE),  # dagar till event
        max(PROCESSDATE) - min(PROCESSDATE)                               # censurerad tid
      )
    ) / 365.25  # konvertera till år
  ) %>%
  ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(PROCESSDATE - baseline_date) / 365.25)
```

## 6. We check for missing values 

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

## 7. We remove missing values 

```{r}
merged_data <- merged_data %>%
  filter(
    !is.na(CENTILOIDS),
    !is.na(AMYLOID_STATUS),
    !is.na(EXAMDATE),
    !is.na(DIAGNOSIS),
    !is.na(event),
    !is.na(time_to_event)
  )

# We check missing values again 
sapply(merged_data, function(x) sum(is.na(x)))
```

# Longitudinal model

```{r}
# Longitudinal model
long_model <- lme(
  CENTILOIDS ~ time_years, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model 

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)

# Cox survival-modell
surv_model <- coxph(
  surv_object ~ 1,   # ~1 = ingen covariat just nu
  data = surv_data)

# Kontrollera modellen
summary(surv_model)
```

# Joint model 

```{r}
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```
