---
title: "Untitled"
format: pdf
editor: visual
---

# Amyloid 2026

```{r}
# loading libraries
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
```

```{r}
library(readr)
amyloid_data <- read.csv("UCBERKELEY_AMY_6MM_30Mar2026.csv")

# We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM

# We load demographics data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
```

## 1. We select only the required variables and merge them

```{r}
# selecting the relevant variables -> why these ones???
amyloid_data <- amyloid_data %>%
  select(
    "RID",        # Participant ID
    "VISCODE2",   # Visit code
    "SCANDATE",   # Date of amyloid-PET
    "CENTILOIDS"  # Main longitudinal variable (continuous amyloid
    )

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = amyloid_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_amy", "_dx")) %>%
  mutate(dx = case_when(DIAGNOSIS == "Dementia" ~ 1, TRUE ~ 0))
```

## 2. We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# Merge demographics into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
  )

# Make sure gender is a factor
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# Convert SCANDATE to Date
merged_data$SCANDATE <- as.Date(merged_data$SCANDATE, format = "%Y-%m-%d")

# Create age at each visit
merged_data$age_at_visit <- as.numeric(format(merged_data$SCANDATE, "%Y")) - merged_data$PTDOBYY
```

## 3. We filter participants with at least 2 measurements

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((amyloid_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 2 measurements 
length(keep)

# We check exact number of measurements per individual
#table(merged_data$RID)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(SCANDATE)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event)
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(SCANDATE)],
    PTGENDER = PTGENDER[which.min(SCANDATE)],
    time_to_event = case_when(

    any(DIAGNOSIS == "Dementia") ~
      as.numeric(min(SCANDATE[DIAGNOSIS == "Dementia"]) -
                   min(c(SCANDATE, EXAMDATE))),

    TRUE ~ as.numeric(max(c(SCANDATE, EXAMDATE)) -
                    min(c(SCANDATE, EXAMDATE)))) / 365,

  event = as.integer(any(DIAGNOSIS == "Dementia"))
) %>% ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(SCANDATE - baseline_date) / 365.25)
```

## 6. We check the distribution of the main variable

```{r}
boxplot(merged_data$CENTILOIDS, main="CENTILOIDS")
```

```{r}
merged_data$log_CENTILOIDS <- log(merged_data$CENTILOIDS)
```

## 7. We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(merged_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## 8. We make sure we don't have any extreme values (for the longitudinal model)

```{r}
summary(merged_data$log_CENTILOIDS)
any(is.infinite(merged_data$log_CENTILOIDS)) # should be FALSE
# Inf = extreme values 
```

```{r}
# Remove rows with -Inf or Inf in log_CENTILOIDS
merged_data <- merged_data[is.finite(merged_data$log_CENTILOIDS), ]
any(is.infinite(merged_data$log_CENTILOIDS))  # should be FALSE
any(is.na(merged_data$log_CENTILOIDS))       # also FALSE
```

## 9. We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
length(unique(merged_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

```{r}
# Gör RID till character (säkerhet)
merged_data$RID <- as.character(merged_data$RID)
surv_data$RID   <- as.character(surv_data$RID)

# Hitta gemensamma ID:n
common_IDs <- intersect(unique(merged_data$RID), unique(surv_data$RID))

# Behåll endast gemensamma individer
merged_data <- merged_data %>% filter(RID %in% common_IDs)
surv_data   <- surv_data %>% filter(RID %in% common_IDs)

# Kontroll
length(unique(merged_data$RID))  # ska matcha
length(unique(surv_data$RID))    # ska matcha
```

# Longitudinal model

```{r}
# Longitudinal model
long_model <- lme(
  log_CENTILOIDS ~ time_years + age_at_visit + PTGENDER.y, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline amyloid", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

# Joint model

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```

Kan bero på att vi har individer med time_years = 0 - vissa individer har ingen variation alls i time_years (alla mätningar ligger på exakt samma tidpunkt)? ska kolla
