---
title: "test"
format: html
editor: visual
---

# CLOCK DRAWING TEST

## För att se hur många individer har 5+ mätningar

(se sedan joint modelling längre ner)

```{r}
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/NEUROBAT.rda")
cdt_data <- NEUROBAT
```

```{r}
library(dplyr)

# 1. Välj kolumner och ta bort dubbletter (RID + VISCODE + VISDATE)
cdt_unique <- cdt_data %>%
  select(RID, VISCODE, VISDATE, CLOCKSCOR) %>%  # välj de kolumner du vill behålla
  filter(!is.na(CLOCKSCOR)) %>%                # ta bort rader med NA i CLOCKSCOR
  distinct()                                   # tar bort exakta dubbletter

# 2. Räkna antal unika mätningar per individ
cdt_unique <- cdt_unique %>%
  group_by(RID) %>%
  mutate(n_measurements = n()) %>%
  ungroup()

# 4. Räkna antal individer med fler än 5 mätningar
num_individuals <- cdt_unique %>%
  filter(n_measurements > 5) %>%
  distinct(RID) %>%
  nrow()

num_individuals  # visar resultatet
```

# Vanlig analys

## 1. Loading and merging

```{r}
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/NEUROBAT.rda")
cdt_data <- NEUROBAT

load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM

load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
```

```{r}
library(dplyr)

# Välj CDT-variabler
cdt_data <- cdt_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, VISDATE,
         CLOCKSCOR, CLOCKCIRC, CLOCKNUM, CLOCKHAND, CLOCKTIME, CLOCKSYM)

# Välj diagnosdata
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# Slå ihop CDT-data med diagnosdata via RID
merged_data <- cdt_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_cdt", "_dx"))

# Ta bort rader med NA i CLOCKSCOR (longitudinal variabel)
merged_data <- merged_data %>%
  filter(!is.na(CLOCKSCOR))
```

## 2. We add age and gender

```{r}
# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# We merge age and gender into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
)

# We make sure that gender is categorical (factor)
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# We create age at each visit using the visit date (EXAMDATE.plasma)
merged_data$age_at_visit <- as.numeric(format(merged_data$VISDATE, "%Y")) - merged_data$PTDOBYY
```

## 3. We count and filter participants with at least 5 measurements

```{r}
keep = unique((cdt_data %>% group_by(RID) %>% filter(n() > 5))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 5 measurements 
length(keep)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(VISDATE)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event)
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarise(
    age_at_baseline = age_at_visit[which.min(VISDATE)],
    PTGENDER = PTGENDER[which.min(VISDATE)],
    time_to_event = case_when(
      any(DIAGNOSIS == "Dementia") ~
        as.numeric(min(VISDATE[DIAGNOSIS == "Dementia"]) -
                   min(c(VISDATE, EXAMDATE))),
      TRUE ~ as.numeric(max(c(VISDATE, EXAMDATE)) -
                        min(c(VISDATE, EXAMDATE)))
    ) / 365,
    event = as.integer(any(DIAGNOSIS == "Dementia"))
  ) %>%
  ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(VISDATE - baseline_date) / 365.25)
```

## 6. We check the distribution of the variables 

```{r}
boxplot(merged_data$CLOCKSCOR, main="CLOCKSCOR")
```

```{r}
merged_data$log_CLOCKSCOR <- log(merged_data$CLOCKSCOR)
```

## 7. We check for and remove missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data = merged_data %>%
  na.omit()

surv_data = surv_data %>%
  na.omit()

# We check again
sapply(merged_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## 8. We make sure we don't have any extreme values (for the longitudinal model)

```{r}
summary(merged_data$log_CLOCKSCOR)
any(is.infinite(merged_data$log_CLOCKSCOR)) # should be FALSE
# Inf = extreme values 
```

```{r}
# Remove rows with -Inf or Inf in log_CENTILOIDS
merged_data <- merged_data[is.finite(merged_data$log_CLOCKSCOR), ]
any(is.infinite(merged_data$log_CLOCKSCOR))  # should be FALSE
any(is.na(merged_data$log_CLOCKSCOR))       # also FALSE
```

## 9. We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
length(unique(merged_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

# Longitudinal model

```{r}
library(nlme)
# Longitudinal model
long_model <- lme(
  log_CLOCKSCOR ~ time_years + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model

```{r}
library(survival)
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline clock drawing test", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

# Joint-model

```{r}
library(JMbayes2)
# Joint model 
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```

# Multivariate version

We check what values the variables take

```{r}
table(merged_data$CLOCKCIRC, useNA = "ifany")
table(merged_data$CLOCKNUM, useNA = "ifany")
table(merged_data$CLOCKHAND, useNA = "ifany")
table(merged_data$CLOCKTIME, useNA = "ifany")
table(merged_data$CLOCKSYM, useNA = "ifany")
```

Vanlig survival modell, men flera olika LME modeller för varje klcock-variabel.

```{r}
merged_data$CLOCKCIRC <- ifelse(merged_data$CLOCKCIRC == "Correct", 1, 0)
merged_data$CLOCKNUM  <- ifelse(merged_data$CLOCKNUM  == "Correct", 1, 0)
merged_data$CLOCKHAND <- ifelse(merged_data$CLOCKHAND == "Correct", 1, 0)
merged_data$CLOCKTIME <- ifelse(merged_data$CLOCKTIME == "Correct", 1, 0)
merged_data$CLOCKSYM  <- ifelse(merged_data$CLOCKSYM  == "Correct", 1, 0)
```

```{r}
table(merged_data$CLOCKCIRC, useNA = "ifany")
table(merged_data$CLOCKNUM, useNA = "ifany")
table(merged_data$CLOCKHAND, useNA = "ifany")
table(merged_data$CLOCKTIME, useNA = "ifany")
table(merged_data$CLOCKSYM, useNA = "ifany")
```

```{r}
library(nlme)
library(JMbayes2)

# Modell för CLOCKSCOR
fm1 <- lme(
  log_CLOCKSCOR ~ time_years,
  random = ~ time_years | RID,
  data = merged_data
)

fm2 <- mixed_model(
  CLOCKTIME ~ time_years,
  random = ~ time_years | RID,
  family = binomial(),
  data = merged_data
)

fm3 <- mixed_model(
  CLOCKSYM ~ time_years,
  random = ~ time_years | RID,
  family = binomial(),
  data = merged_data
)

Mixed <- list(fm1, fm2, fm3)
```

```{r}
fForms <- list(
  "log_CLOCKSCOR" = ~ value(log_CLOCKSCOR) + slope(log_CLOCKSCOR),
  "CLOCKTIME" = ~ area(CLOCKTIME),
  "CLOCKSYM" = ~ value(CLOCKSYM) + area(CLOCKSYM)
)
```

```{r}
library(survival)
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)

# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ 1,
  data = surv_data
)

summary(surv_model)
```

```{r}
multi_joint_model <- jm(
  surv_model,
  Mixed,
  time_var = "time_years",
  functional_forms = fForms,
  n_chains = 3L,
  n_iter = 20000L,
  n_burnin = 2000L
)

summary(multi_joint_model)
```

# 3 separata joint models

```{r}
library(JMbayes2)
joint_model_scor <- jm(
  surv_model,
  fm1,
  time_var = "time_years"
)
summary(joint_model_scor)
```

```{r}
  
joint_model_time <- jm(
  surv_model,
  fm2,
  time_var = "time_years"
)
summary(joint_model_time)
```

```{r}
  
joint_model_sym <- jm(
  surv_model,
  fm3,
  time_var = "time_years"
)
summary(joint_model_sym)
```
