---
title: "Full and restricted model comparison"
format: pdf
editor: visual
---

# Comparing full and restricted joint-models

## Pre-processing

```{r}
# loading librarys
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggsurvfit)
```

-   Loading the data

```{r}
# loading the data sets
# Charley
library(ADNIMERGE2)

# dx data
diagnosis_data = DXSUM

# demographics
demographics = PTDEMOG

# read data Milou
library(readr)
plasma_data <- read.csv("UPENN_PLASMA_FUJIREBIO_QUANTERIX_27Mar2026.csv")

# # We load diagnosis data
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
# diagnosis_data <- DXSUM
# 
# # We load demographics data 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
# demographics <- PTDEMOG
```

-   Selecting relevant variables

```{r}
# selecting the relevant variables -> why these ones???
plasma_data = plasma_data %>%
  select("RID", "VISCODE2", "EXAMDATE", "pT217_F",
         "AB42_F", "AB40_F", "AB42_AB40_F", "pT217_AB42_F", "NfL_Q", "GFAP_Q")

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")
```

-   Checking number of individuals after removing missing values

```{r}
# PLASMA
# checking for missing values
sum(is.na(plasma_data))

# checking number of individuals
length(unique(plasma_data$RID))
```

-   Merging diagnosis data set with the plasma data set

```{r}
# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = plasma_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx")) %>%
  mutate(dx = case_when(DIAGNOSIS == "Dementia" ~ 1, TRUE ~ 0))
```

-   Demographic data: gender as factor, creating age variable, merging

```{r}
# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# We merge age and gender into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
)

# We make sure that gender is categorical (factor)
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# Converting EXAMDATE_plasma till Date
merged_data$EXAMDATE_plasma <- as.Date(merged_data$EXAMDATE_plasma, format="%Y-%m-%d")

# Age at visit
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_plasma, "%Y")) - merged_data$PTDOBYY
```

-   Creating baseline date, survival time and event variables

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_plasma)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event)
surv_data = merged_data %>%
  group_by(RID) %>%
  summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_plasma)],
    time_to_event = case_when(
      
# survial time for events
    any(DIAGNOSIS == "Dementia") ~
      as.numeric(min(EXAMDATE_plasma[DIAGNOSIS == "Dementia"]) -
                   min(c(EXAMDATE_plasma, EXAMDATE_dx))),
    
# survival time for censored
    TRUE ~ as.numeric(max(c(EXAMDATE_plasma, EXAMDATE_dx)) -
                    min(c(EXAMDATE_plasma, EXAMDATE_dx)))) / 365,
    event = as.integer(any(DIAGNOSIS == "Dementia"))
    ) %>%
  ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

```{r}
# Computing the event ratio
19/109
```

-   Time since baseline for the plasma exams

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_plasma - baseline_date) / 365.25)
```

-   Checking NfL_Q and GFAP_Q values: $618$ with $-4$ which indicates missing data, so remove these variables

```{r}
# NfL_Q and GFAP_Q have many -4 values, checking how many
nrow(merged_data %>% filter(NfL_Q < 0))
nrow(merged_data %>% filter(GFAP_Q < 0))

# checking how many obs have biomarker values less than 0 
nrow(merged_data %>% filter(pT217_F < 0 | AB42_F < 0 | AB40_F < 0 | AB42_AB40_F < 0 |
                             pT217_AB42_F < 0 | NfL_Q < 0 | GFAP_Q < 0))

# seems as if the only ones are NfL_Q and GFAP_Q, so we remove them from the data set
merged_data = merged_data %>% select(!NfL_Q & !GFAP_Q)
```

-   Checking outliers and transforming variables

```{r}
# Vi kollar efter outliers 
boxplot(merged_data$pT217_AB42_F, main="pT217_AB42_F")
boxplot(merged_data$AB42_AB40_F, main="AB42_AB40_F")
```

```{r}
# Vi transformerar dem
merged_data$log_pT217_AB42_F <- log(merged_data$pT217_AB42_F)
merged_data$log_AB42_AB40_F <- log(merged_data$AB42_AB40_F)
```

-   Checking missing vals: **why are there missing values?**

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
length(unique(merged_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

# Longitudinal model

-   Using both time_years + age_at_visit is redundant since we have age_at_baseline in the coxph model: if we omit age_at_visit variable from the lme model, the joint model is stable.

```{r}
# Longitudinal model
long_model <- lme(
  log_pT217_AB42_F ~ time_years + PTGENDER, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model

-   Creating the survival object

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

-   Adding PTGENDER into the surv_data data frame

```{r}
ptgender = merged_data %>% select(RID, PTGENDER)
surv_data = surv_data %>% merge(ptgender, by = "RID") 
surv_data = surv_data[!duplicated(surv_data$RID), ]
```

-   Visual assessment of Kaplan-Meier curves with no covariate, PTGENDER as sole covariate, and age at baseline as sole covariate: ggsurvfit:

```{r}
# Kaplan-Meier curve
km_fit = survfit(surv_object ~ 1, data = surv_data)

# Kaplan-Meier curve by gender
km_fit_gender = survfit(surv_object ~ PTGENDER, data = surv_data)

# Kaplan-Meier curve by age at baseline
km_fit_age = survfit(surv_object ~ age_at_baseline, data = surv_data)


# plotting the survival curves
ggsurvfit(km_fit) +
  labs(x = "Years since baseline measurement",
       title = "Kaplan Meier: Disease-free survival probability") +
  ylim(0, 1) +
  theme_classic()

# plotting the survival curves
ggsurvfit(km_fit_gender) +
  labs(x = "Years since baseline measurement",
       title = "Kaplan Meier: Disease-free survival probability (PTGENDER)") +
  ylim(0, 1) +
  theme_classic()

# plotting the survival curves
ggsurvfit(km_fit_age) +
  labs(x = "Years since baseline measurement",
       title = "Kaplan Meier: Disease-free survival probability (age_at_baseline)") +
  ylim(0, 1) +
  theme_classic()
```

-   Creating a categorical age variable: \< 70 , \> 70

```{r}
summary(surv_data$age_at_baseline)
boxplot(surv_data$age_at_baseline, main="age at baseline")
```

```{r}
surv_data = surv_data %>%
  mutate(age_range = as.factor(case_when(
    age_at_baseline < 70 ~ "< 70",
    age_at_baseline > 69 ~ "70+ + "
  )))
```

-   Checking Kaplan Meier curve with age range as sole covariate (since coefficients not significant for PTGENDER or age_at_baseline in the cox PH models WITH REDUCED DATA - \> 3 MEASUREMENTS - but no change, coefficient still not significant)

```{r}
# fitting the KM curve
km_fit_age_range = survfit(surv_object ~ age_range, data = surv_data)

# plotting the survival curves for age_range
ggsurvfit(km_fit_age_range) +
  labs(title = "Kaplan Meier curve for age_range",
       x = "Years since baseline measurement") +
  ylim(0, 1) +
  theme_classic()
```

```{r}
# Cox PH model with covariate gender
surv_model_gender = coxph(
  surv_object ~ PTGENDER,
  data = surv_data
)

# Cox PH model with covariate age at baseline
surv_model_age = coxph(
  surv_object ~ age_at_baseline,
  data = surv_data
)

# Cox PH model with covariate age_range at baseline
surv_model_age_range = coxph(
  surv_object ~ age_range,
  data = surv_data
)

# Cox PH model with covariate age at baseline
surv_model_age_gender = coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data,
  x = TRUE
)

summary(surv_model_gender)

summary(surv_model_age)

summary(surv_model_age_range)

summary(surv_model_age_gender)
```

```{r}
# Running proportionality tests for all models
cox.zph(surv_model_gender)
cox.zph(surv_model_age)
cox.zph(surv_model_age_range)
cox.zph(surv_model_age_gender)
```

# Joint model

-   The Rhat values are $intercept = 1.0007$, $time\_years = 1.0145$, $PTGENDER = 0.9999$, $age\_at\_baseline = 1.0005$, and $sigma = 1.0072$: these are acceptable values.

```{r}
# Joint model 
joint_model = jm(
  surv_model_age_gender,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```

# Restricted model: individuals with \> 3 measurements

-   Filtering for individuals with at least 3 measurements

```{r}
# Filtering individuals with three or more measurements
keep = unique((plasma_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data_restricted = merged_data %>% filter(RID %in% keep)
surv_data_restricted = surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(unique(merged_data_restricted$RID))
length(unique(surv_data_restricted$RID))
```

# Longitudinal Model

-   Using both time_years + age_at_visit is redundant since we have age_at_baseline in the coxph model: if we omit age_at_visit variable from the lme model, the joint model is stable.

```{r}
# Longitudinal model
long_model_restricted <- lme(
  log_pT217_AB42_F ~ time_years + PTGENDER, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data_restricted
  )

summary(long_model_restricted)
```

# Survival Model

-   From the full model, we see that PTGENDER and age_at_baseline are significant, so we include these in our restricted model.

```{r}
# creating a survival object with restricted data
surv_object_restricted <- Surv(
  time = surv_data_restricted$time_to_event, 
  event = surv_data_restricted$event
)

# fitting the coxph model
surv_model_restricted = coxph(
  surv_object_restricted ~ age_at_baseline + PTGENDER,
  surv_data_restricted,
  x = TRUE
)

summary(surv_model_restricted)

cox.zph(surv_model_restricted)
```

# Joint Model

-   Rhat values are $intercept = 1.0009$, $time\_years = 1.0004$, $age\_at\_baseline = 1.0008$, $PTGENDER = 1.0000$, and $sigma = 1.0012$: these are acceptable values.

```{r}
# Joint model 
joint_model_restricted = jm(
  surv_model_restricted,
  long_model_restricted,
  time_var = "time_years"
  )

summary(joint_model_restricted)
```
