---
title: "CSF_test"
format: pdf
editor: visual
---

# CSF-version

3 olika biomarkörer - valde TAU för att använda i modellerna (TAU changes continuously over time and reflects neurodegeneration, making it ideal for longitudinal modeling)

Problem = 2 olika tidvariabler för CSF (DRAWDTE och RUNDATE) - Jag valde DRAWDTE, men vet ej vilken som är rätt. Möjlig motivering:

-   DRAWDTE reflects the actual date the CSF sample was taken, capturing the participant’s biological timeline, whereas RUNDATE is just the lab analysis date and doesn’t reflect disease progression.

```{r}
# We load CSF data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UPENNBIOMK_MASTER.rda")
csf_data <- UPENNBIOMK_MASTER

# We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
```

## 1. We select only the required variables and merge them

```{r}
csf_data = csf_data %>%
  select("ORIGPROT", "RID", "VISCODE", "DRAWDTE", "RUNDATE",
         "ABETA", "TAU", "PTAU")

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = csf_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx"))
```

## 3. We filter participants with at least 2 measurements

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((csf_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 2 measurements 
length(keep)

# We check exact number of measurements per individual
#table(merged_data$RID)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Create baseline date based on CSF sample
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(DRAWDTE)) %>%  # baseline = first CSF draw
  ungroup()

# Create event + time_to_event (Dementia counts as event)
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarize(
    # Event = 1 if participant ever had "Dementia", else 0 (censored)
    event = ifelse(any(DIAGNOSIS == "Dementia"), 1, 0),
    
    # Time to first "Dementia", otherwise time to last observation
    time_to_event = as.numeric(
      ifelse(
        any(DIAGNOSIS == "Dementia"),
        min(DRAWDTE[DIAGNOSIS == "Dementia"]) - min(DRAWDTE),  # days to event
        max(DRAWDTE) - min(DRAWDTE)                             # censored time
      )
    ) / 365.25  # convert to years
  ) %>%
  ungroup()

# Merge survival data back into merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# Check the number of events
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# Create time_years (time from baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(DRAWDTE - baseline_date) / 365.25)  # time in years since baseline
```

## 6. We check for and remove missing values in both merged_data and long_data 

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We re-count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

# 3. Longitudinal model

## 3.1 Fitted LME (linear mixed-effects) model

```{r}
library(nlme)

# Linear mixed-effects model for TAU over time
long_model <- lme(
  TAU ~ time_years,
  random = ~ time_years | RID,        # random intercept + slope per participant
  data = merged_data
)

# Summary of the model
summary(long_model)
```

# 4. Survival model

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)

# Cox survival-model 
surv_model <- coxph(
  surv_object ~ 1,
  data = surv_data
)

summary(surv_model)
```

# 5. Joint-model

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```
