---
title: "CSF_test"
format: pdf
editor: visual
---

# CSF-version

3 olika biomarkörer - valde TAU för att använda i modellerna (TAU changes continuously over time and reflects neurodegeneration, making it ideal for longitudinal modeling)

Problem = 2 olika tidvariabler för CSF (DRAWDTE och RUNDATE) - Jag valde DRAWDTE, men vet ej vilken som är rätt. Möjlig motivering:

-   DRAWDTE reflects the actual date the CSF sample was taken, capturing the participant’s biological timeline, whereas RUNDATE is just the lab analysis date and doesn’t reflect disease progression.

```{r}
# We load CSF data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UPENNBIOMK_MASTER.rda")
csf_data <- UPENNBIOMK_MASTER

# We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM

# We load demographics data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
```

## 1. We select only the required variables and merge them

```{r}
csf_data = csf_data %>%
  select("ORIGPROT", "RID", "VISCODE", "DRAWDTE", "RUNDATE",
         "ABETA", "TAU", "PTAU")

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = csf_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx"))
```

## 2. We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# We merge age and gender into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
)

# We make sure that gender is categorical (factor)
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# We create age at each visit using the visit date (DRAWDTE)
merged_data$age_at_visit <- as.numeric(format(merged_data$DRAWDTE, "%Y")) - merged_data$PTDOBYY
```

## 3. We filter participants with at least 2 measurements

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((csf_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 2 measurements 
length(keep)

# We check exact number of measurements per individual
#table(merged_data$RID)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Create baseline date based on CSF sample
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(DRAWDTE)) %>%  # baseline = first CSF draw
  ungroup()

# Create event + time_to_event (Dementia counts as event)
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarize(
    # Baseline-variabler
    age_at_baseline = age_at_visit[which.min(DRAWDTE)],
    PTGENDER = PTGENDER[which.min(DRAWDTE)],
    
    # Event = 1 if participant ever had "Dementia", else 0 (censored)
    event = ifelse(any(DIAGNOSIS == "Dementia"), 1, 0),
    
    # Time to first "Dementia", otherwise time to last observation
    time_to_event = as.numeric(
      ifelse(
        any(DIAGNOSIS == "Dementia"),
        min(DRAWDTE[DIAGNOSIS == "Dementia"]) - min(DRAWDTE),  # days to event
        max(DRAWDTE) - min(DRAWDTE)                             # censored time
      )
    ) / 365.25  # convert to years
  ) %>%
  ungroup()

# Merge survival data back into merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# Check the number of events
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# Create time_years (time from baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(DRAWDTE - baseline_date) / 365.25)  # time in years since baseline
```

## 6. We check for and remove missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We re-count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

# 3. Longitudinal model

## 3.1 Fitted LME (linear mixed-effects) model

```{r}
library(nlme)

merged_data$time_c <- merged_data$time_years - mean(merged_data$time_years, na.rm = TRUE)

long_model <- lme(
  TAU ~ time_c + age_at_visit + PTGENDER.x,
  random = ~ time_c | RID,   # random intercept + slope
  data = merged_data
  )

# Summary of the model
summary(long_model)
```

# 4. Survival model

## 4.1 We create a survival object

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

## 4.2 Kaplan-Meier curve

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline plasma", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

## 4.3 Fitted Coxph model

```{r}
# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

## 4.4 Testing for the assumption of proportional hazards

```{r}
cox.zph(surv_model)
```

cox.zph() fungerar inte matematiskt när risk-setet vid någon eventtidpunkt saknar variation i covariater, eftersom den försöker lösa ett linjärt system med score-residualer som blir singulärt och därmed omöjligt att invertera.

–\> Alternativ:

```{r}
library(survival)
library(survminer)

fit <- survfit(surv_object ~ PTGENDER, data = surv_data)
ggsurvplot(fit, data = surv_data)
```

"Linjerna är nästan horisontella → PH-antagandet verkar rimligt för kön.

Spike vid tid 0 → orsakar singularitet i cox.zph().

Vårt dataset har helt enkelt tidpunkter med risk-set utan variation, vilket gör att cox.zph() inte kan lösa matrisen, men visuellt är PH-antagandet inte brutet."

# 5. Joint-model

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```
