---
title: "CSF_test"
format: pdf
editor: visual
---

# CSF-version

3 olika biomarkörer - valde TAU för att använda i modellerna (TAU changes continuously over time and reflects neurodegeneration, making it ideal for longitudinal modeling)

Problem = 2 olika tidvariabler för CSF (DRAWDTE och RUNDATE) - Jag valde DRAWDTE, men vet ej vilken som är rätt. Möjlig motivering:

-   DRAWDTE reflects the actual date the CSF sample was taken, capturing the participant’s biological timeline, whereas RUNDATE is just the lab analysis date and doesn’t reflect disease progression.

```{r}
library(dplyr)
```

```{r}
#HEAD
# We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM

# We load demographics data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
```

```{r}
# load data sets: Milou
# 
# # We load CSF data 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UPENNBIOMK_MASTER.rda")
# csf_data <- UPENNBIOMK_MASTER
# 
# # We load diagnosis data
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
# diagnosis_data <- DXSUM
```

```{r}
# load data sets: Charley

# loading adni library
library(ADNIMERGE2)

# biomarker data set
csf_data = UPENNBIOMK_MASTER

# dx data
diagnosis_data = DXSUM

```

## 1. We select only the required variables and merge them

```{r}
csf_data = csf_data %>%
  select("ORIGPROT", "RID", "VISCODE", "DRAWDTE", "RUNDATE",
         "ABETA", "TAU", "PTAU")

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = csf_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx"))
```

## 2. We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# We merge age and gender into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
)

# We make sure that gender is categorical (factor)
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# We create age at each visit using the visit date (DRAWDTE)
merged_data$age_at_visit <- as.numeric(format(merged_data$DRAWDTE, "%Y")) - merged_data$PTDOBYY
```

## 3. We filter participants with at least 2 measurements

First, we group the CSF data set by RID and DRAWDTE, create an average over different RUNDATE for each individual, remove the unwanted columns (RUNDATE, ABETA, TAU, PTAU). Then we remove duplicate rows and are left with a data frame with unique DRAWDTE for each individual and an average measurement for each biomarker. We check how many unique RID are in the data frame: $n_{RID} = 159$ but we have $n_{obs} = 613$.

```{r}
# grouping by RID and then DRAWDTE in order to compute means for ABETA, TAU and PTAU measurements over the different RUNDATES for each unique DRAWDTE
csf_data_reduced = csf_data %>%
#   grouping by RID and DRAWDTE
  group_by(RID, DRAWDTE) %>%
#   computing means for ABETA, TAU, PTAU for different RUNDATEs for each DRAWDAT
  mutate(
  ABETA_MEAN = mean(ABETA),
  TAU_MEAN = mean(TAU),
  PTAU_MEAN = mean(PTAU)
) %>%
#   removing unwanted columns
  select(!c(RUNDATE, ABETA, TAU, PTAU))

# removing duplicate rows with !duplicated()
csf_data_reduced = csf_data_reduced[!duplicated(csf_data_reduced), ]

# code that checks which individuals have more than 2 measurements AFTER we have gotten unique draw dates
csf_data_reduced = csf_data_reduced %>% group_by(RID) %>% filter(n() > 2)

# checking how many individuals we have in the reduced data set with only individuals with more than three CSF biomarker measurements 
length(unique(csf_data_reduced$RID))
```

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique(csf_data_reduced$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 2 measurements 
length(keep)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Create baseline date based on CSF sample
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(DRAWDTE)) %>%  # baseline = first CSF draw
  ungroup()

# Create event + time_to_event (Dementia counts as event)
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarize(
    # Baseline-variabler
    age_at_baseline = age_at_visit[which.min(DRAWDTE)],
    PTGENDER = PTGENDER[which.min(DRAWDTE)],
    
    # Event = 1 if participant ever had "Dementia", else 0 (censored)
    event = ifelse(any(DIAGNOSIS == "Dementia"), 1, 0),
    
    # Time to first "Dementia", otherwise time to last observation
    time_to_event = as.numeric(
      ifelse(
        any(DIAGNOSIS == "Dementia"),
        min(DRAWDTE[DIAGNOSIS == "Dementia"]) - min(DRAWDTE),  # days to event
        max(DRAWDTE) - min(DRAWDTE)                             # censored time
      )
    ) / 365.25  # convert to years
  ) %>%
  ungroup()

# Merge survival data back into merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# Check the number of events
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# Create time_years (time from baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(DRAWDTE - baseline_date) / 365.25)  # time in years since baseline
```

## 6. We check for and remove missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We re-count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

# 3. Longitudinal model

## 3.1 Fitted LME (linear mixed-effects) model

```{r}
library(nlme)

merged_data$time_c <- merged_data$time_years - mean(merged_data$time_years, na.rm = TRUE)

long_model <- lme(
  TAU ~ time_c + age_at_visit + PTGENDER.x,
  random = ~ time_c | RID,   # random intercept + slope
  data = merged_data
  )

# Summary of the model
summary(long_model)
```

# 4. Survival model

## 4.1 We create a survival object

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

## 4.2 Kaplan-Meier curve

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline plasma", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

## 4.3 Fitted Coxph model

```{r}
# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

## 4.4 Testing for the assumption of proportional hazards

```{r}
cox.zph(surv_model)
```

cox.zph() fungerar inte matematiskt när risk-setet vid någon eventtidpunkt saknar variation i covariater, eftersom den försöker lösa ett linjärt system med score-residualer som blir singulärt och därmed omöjligt att invertera.

–\> Alternativ:

```{r}
library(survival)
library(survminer)

fit <- survfit(surv_object ~ PTGENDER, data = surv_data)
ggsurvplot(fit, data = surv_data)
```

"Linjerna är nästan horisontella → PH-antagandet verkar rimligt för kön.

Spike vid tid 0 → orsakar singularitet i cox.zph().

Vårt dataset har helt enkelt tidpunkter med risk-set utan variation, vilket gör att cox.zph() inte kan lösa matrisen, men visuellt är PH-antagandet inte brutet."

# 5. Joint-model

```{r}
# # Joint model 
# joint_model <- jm(
#   surv_model,
#   long_model,
#   time_var = "time_years"
#   )
# 
# summary(joint_model)
```
