---
title: "data_cleaning"
format: pdf
editor: visual
---

## Data cleaning

```{r}
library(dplyr)
library(JMbayes2)
```

```{r}
# load data sets: Milou

# # Skapa en mapp för att packa upp
# outdir <- "ADNIMERGE2_unpacked"
# dir.create(outdir, showWarnings = FALSE)
# 
# # Packa upp tar.gz-filen
# untar("ADNIMERGE2.tar.gz", exdir = outdir)
# 
# # Plasma-biomarkörer
# load(file.path(outdir, "ADNIMERGE2/data/UPENN_PLASMA_FUJIREBIO_QUANTERIX.rda"))
# plasma_data <- UPENN_PLASMA_FUJIREBIO_QUANTERIX
# 
# # Amyloid-PET
# load(file.path(outdir, "ADNIMERGE2/data/UCBERKELEY_AMY_6MM.rda"))
# amyloid_data <- UCBERKELEY_AMY_6MM
# 
# # Diagnos
# load(file.path(outdir, "ADNIMERGE2/data/DXSUM.rda"))
# diagnosis_data <- DXSUM
```

```{r}
# load data sets: Charley

# loading adni library
library(ADNIMERGE2)

# biomarker data set
plasma_data = UPENN_PLASMA_FUJIREBIO_QUANTERIX

# PET scan data
amyloid_data = UCBERKELEY_AMY_6MM

# dx data
diagnosis_data = DXSUM
```

```{r}
cat("Plasma-data unika RID:", length(unique(plasma_data$RID)), "\n")
cat("Amyloid-PET-data unika RID:", length(unique(amyloid_data$RID)), "\n")
cat("Diagnos-data unika RID:", length(unique(diagnosis_data$RID)), "\n")
```

```{r}
# selecting the relevant variables -> why these ones???
plasma_data = plasma_data %>% select(
  "ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "pT217_F", "AB42_F",
  "AB40_F", "AB42_AB40_F", "pT217_AB42_F", "NfL_Q", "GFAP_Q")

diagnosis_data = diagnosis_data %>% select("ORIGPROT", "COLPROT", "RID", "VISCODE2",
                                           "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier)
merged = plasma_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx")) %>%
  mutate(dx = case_when(DIAGNOSIS == "Dementia" ~ 1, TRUE ~ 0))
  
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", merged$PTID)

# checking for missing values in EXAMDATE_dx and EXAMDATE_plasma
sum(as.numeric(is.na(merged$EXAMDATE_plasma)))
sum(as.numeric(is.na(merged$EXAMDATE_dx)))

# removing the obs with no EXAMDATE_dx
merged = merged %>% filter(as.character(EXAMDATE_dx) != "NA")

# checking they were removed
sum(as.numeric(is.na(merged$EXAMDATE_dx)))
```

### Linear mixed effects model:

We want:

-   **Several observations per patient which relates to measurements made on different dates**

-   **Years**: time between enrollment and the earlier of diagnosis (for events) or study analysis time (for censored)

-   Year: time between enrollment and current exam date

-   All other variables constant that don't change with different biomarker measurement visits

-   Event variable

```{r}
# compare to this data frame from the JMBayes documentation
lme_data = pbc2

?pbc2
```

```{r}
# test for LME data 
lme_test = merge(merged, surv_test, by = "RID") %>% 
  mutate(year =
           as.numeric(EXAMDATE_plasma - min(c(EXAMDATE_dx, EXAMDATE_plasma))) / 365.25)
```

### Survival model:

We want:

-   One observation per individual

-   Event variable

-   Baseline measurements at t = 0

```{r}
# compare with this data frame from the JMBayes documentation
surv_data = pbc2.id
```

```{r}
# test for survival data, fixes NA where there is EXAMDATE_dx with NA values, fixes survival_time = 0 where there is only one biomarker exam, by using both EXAMDATE_dx and EXAMDATE_plasma -> not useful for joint model as one biomarker measurement is not enough for our joint model, but could be useful for comparing the joint-model with normal survival model

surv_test = merged %>%
  group_by(RID) %>%
  summarise(
#     survival time is time between enrollment and the earlier of diagnosis (for events) or study analysis time (for censored)
    survival_time = case_when(

# survival time for event
    any(DIAGNOSIS == "Dementia") ~
      as.numeric(min(EXAMDATE_plasma[DIAGNOSIS == "Dementia"]) -
                   min(c(EXAMDATE_plasma, EXAMDATE_dx))),
    
# survival time for censored
    TRUE ~ as.numeric(max(c(EXAMDATE_plasma, EXAMDATE_dx)) -
                        min(c(EXAMDATE_plasma, EXAMDATE_dx)))) 

# converting to years
/ 365.25,

# creating event variable, 1 = AD, 0 = CN, MCI
    event = as.integer(any(DIAGNOSIS == "Dementia"))
    ) %>% ungroup()
```

Checking the number of observations with transitions from CN/MCI to AD, where $n_{event} = 173$ and $n_{censored} = 1148$ and $n_{total} = 1321$.

```{r}
merged %>% filter(as.character(EXAMDATE_dx) != "NA")
```

```{r}
# We want to check how many obs we have going from CN -> AD: if this is low we need to take in to account going from CN -> MCI, MCI -> AD, CN -> AD

# creating a data frame with only the baseline diagnosis and baseline plasma measurements for each patient
first_dx = merged %>% group_by(RID) %>%
  filter(EXAMDATE_dx == min(EXAMDATE_dx)) %>%
  filter(EXAMDATE_plasma == min(EXAMDATE_plasma))

# creating a data frame with only the last diagnosis and baseline plasma measurements for each patient
last_dx = merged %>% group_by(RID) %>%
  filter(EXAMDATE_dx == max(EXAMDATE_dx)) %>%
  filter(EXAMDATE_plasma == min(EXAMDATE_plasma))

# merging the two data frames with the first and last and recoding diagnosis as 1 = AD, 0 otherwise in order to compare the first and last values
first_last_dx = merge(
#   merging the data set of first and last diagnoses by RID 
  first_dx, last_dx,by = "RID", suffix = c("_first", "_last")) %>%
#   arranging by ascending RID to check format
  arrange(RID) %>%
#   adding a variable to check whether the diagnosis changed from CN/MCI to AD
  mutate(dx_change = dx_first != dx_last)

# count how many obs have change in diagnosis from CN to AD
table(first_last_dx$dx_change)

# checking if any NA exam dates
sum(as.numeric(is.na(first_last_dx$EXAMDATE_dx)))
```

```{r}
# the RID of the observations in the merged data set that have NA EXAMDATE_dx
RID_NA_examdate_dx = merged[which(is.na(merged$EXAMDATE_dx)), 1]

# looks like if a patient has several observations and one has EXAMDATE_dx = NA, min/max will not fetch any of the observations for the patient
merged %>% filter(RID %in% RID_NA_examdate_dx) %>% filter(EXAMDATE_dx == max(EXAMDATE_dx))
```
