---
title: "GAMMAL_PLASMA"
format: pdf
editor: visual
---

# GAMMAL PLASMA

```{r, warning=FALSE}
library(dplyr)
library(nlme)
library(survival)
library(JMbayes2)
```

```{r}
# loading data
# Milou
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
#
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
#
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
#
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG

load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

```{r}
# Charley
library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

apoe = APOERES
```

## 1. We select the required variables and merge them

```{r}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# We merge NFL, ptau och diagnosis
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))
```

Checking missing values

```{r}
# checking for missing values with NA

# NfL
sum(is.na(nfl_data))
nfl_data[which(is.na(nfl_data)), ]

# pTAU
sum(is.na(ptau_data))

# checking with ADNI code -4
# NfL
sum(nfl_data$PLASMA_NFL < 0, na.rm = TRUE)

# pTAU
sum(ptau_data$PLASMAPTAU181 < 0, na.rm = TRUE)
```

## 2. We add selected demographics variables (age + gender) and apoe4 status, to later use in survival model and longitudinal model

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We merge with merged_data
merged_data <- merged_data %>%
  left_join(demographics_small, by = "RID")

# We make sure that gender is a factor 
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# We convert EXAMDATE_ptau to date 
merged_data$EXAMDATE_ptau <- as.Date(as.character(merged_data$EXAMDATE_ptau), format="%Y-%m-%d")

# We create age_at_visit based on ptau date 
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_ptau, "%Y")) - merged_data$PTDOBYY

# We create a variable for apoe4 status
apoe$APOE4_status <- ifelse(apoe$GENOTYPE %in% c("2/4","3/4","4/4"), 1, 0)

# We merge apoe with merged_data 
merged_data <- merge(merged_data, apoe[, c("RID", "APOE4_status")], by="RID")
```

## 3. We filter participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# We create a baseline-date
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau)) %>%
  ungroup()

# We create event and time_to_event (where dementia is the event) and add age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_ptau)],
    PTGENDER = PTGENDER[which.min(EXAMDATE_ptau)],
    APOE4_status = APOE4_status[which.min(EXAMDATE_ptau)],
    
    time_to_event = case_when(

    any(DIAGNOSIS_dx == "Dementia") ~
      as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,

  event = as.integer(any(DIAGNOSIS_dx == "Dementia"))
) %>% ungroup()

# We put the survival data back into the merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# We create time_years (time from baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## 6. We check the distribution of each relevant variable for multivariate joint model and transform

```{r}
# We check the distribution of each biomarker
hist(merged_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(merged_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# We log transform the biomarkers 
merged_data$log_PLASMAPTAU181 <- log(merged_data$PLASMAPTAU181)
merged_data$log_PLASMA_NFL <- log(merged_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(merged_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(merged_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

Ptau still looks skewed, even after log transformation

--\> In our LME model, we can use a Student’s t-distribution instead of a normal distribution (lme() → normal distribution assumed by default); it is more robust to outliers and heavy-tailed residual distributions.

Gamma distribution on the original values could be possible (see if it works below, under Longitudinal model), BUT it's very sensitive to outliers, which we have (see below).

## **7. We check for outliers**

We still keep them, as we assume they are correct observations.

-   Checking for observations either side of the mean of the raw data shows there are no values 3 sds below the mean, only above.

```{r}
merged_data$ptau_z <- scale(merged_data$PLASMAPTAU181)

merged_data$nfl_z <- scale(merged_data$PLASMA_NFL)

# We identify observations that are 3 standard deviations above or below the mean (z > 3 or < -3)
large_outliers_ptau <- merged_data[merged_data$ptau_z > 3, ]
small_outliers_ptau <- merged_data[merged_data$ptau_z < -3, ]
large_outliers_ptau
small_outliers_ptau

large_outliers_nfl <- merged_data[merged_data$nfl_z > 3, ]
small_outliers_nfl <- merged_data[merged_data$nfl_z < -3, ]
large_outliers_nfl
small_outliers_nfl
```

## 8. We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# We remove missing values 
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(merged_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## 9. We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
# We check the length of each dataset 
length(unique(merged_data$RID))   # longitudinal data
length(unique(surv_data$RID))     # survival data
```

## 10. We check the correlation between the biomarkers

If the biomarkers are too correlated, the multivariate joint model won't converge, I think.

```{r}
# Correlation between the biomarkers 
cor(merged_data$PLASMAPTAU181, merged_data$PLASMA_NFL, use = "complete.obs")
```

# Longitudinal model

## 1. For ptau with T-distribution

```{r}
# Longitudinal model with t-distribution (identity link) 
library(GLMMadaptive)
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER.y + APOE4_status.y,
  random = ~ time_years | RID,                    
  data = merged_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

```{r}
# residuals on a subject-specific scale
res <- residuals(long_model_ptau, type = "subject_specific")

# fitted values
fitted_vals <- fitted(long_model_ptau, type = "subject_specific")

# plot
plot(fitted_vals, res,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)
```

```{r}
qqnorm(res)
qqline(res)
```

## 2. For ptau with gamma distribution

```{r}
# Longitudinal model with gamma distribution (log link) 
long_model_ptau_gamma <- mixed_model(
  fixed = PLASMAPTAU181 ~ time_years + PTGENDER.y + APOE4_status.y,  
  random = ~ time_years | RID,                      
  data = merged_data,
  family = Gamma(link = "log")
)

summary(long_model_ptau_gamma)
```

## 3. For nfl

```{r}
# Longitudinal model, with chosen covariates 
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER.y + APOE4_status.y, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model_nfl)
```

## 4. Removing gender from nfl LME model

We see that gender is not significant in the model for nfl, so we remove it:

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + APOE4_status.y, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model_nfl)
```

## 5. Removing APOE4 status from ptau LME model

Even though APOE4 status was significant in the LME model for ptau, including it in both the LME model and the survival model led to poor MCMC convergence (high Rhat) in the joint model, probably due to difficulties in separating its direct effect on survival from its indirect effect through the biomarker. Therefore, to ensure stable estimation, we keep the APOE4 status only in the survival model, removing it from the LME modell:

```{r}
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER.y,
  random = ~ time_years | RID,                    
  data = merged_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

## (6. Removing random slope)

To know whether to include a random slope in the LME model or not, we can examine the variance of the slope. If the slope variance is close to zero, this indicates that individual trajectories over time differ very little. Then, the slope contributes almost no information, and including a random slope may destabilize the joint model estimation and lead to high Rhat values:

summary(long_model_nfl) shows that the slope variance is 0.0011 (Std.Dev = 0.033), which is very small compared to the residual variance (0.047) –\> we include only a random intercept in the joint model to stabilize estimation.

BUT then we can't do separate joint models,

BUT we can still do a multivariate joint model!

```{r}
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER.y,
  random = ~ 1 | RID, # only random intercept
  data = merged_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + APOE4_status.y, 
  random = ~ 1 | RID, # only random intercept
  data = merged_data
  )

summary(long_model_nfl)
```

```{r}
# Joint model without random slope in the LME models
Mixed <- list(long_model_ptau, long_model_nfl)

fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )    

multi_joint_model <- jm(surv_model, Mixed,
                  time_var = "time_years",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 20000L,
                  n_burnin = 2000L
                  )
summary(multi_joint_model)
```

We get bad Rhat-values in the multivariate joint model when removing the slope from the LME models.

--\> So even though the slope variance is very small, including a minimal random slope helps stabilize MCMC in joint models. Without it, the longitudinal and survival parts become numerically decoupled, causing high Rhat, even if the slope theoretically adds little information.

## 7. What if we used interactions?

```{r}
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER.y + time_years:PTGENDER.y,
  random = ~ time_years | RID,                    
  data = merged_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + APOE4_status.y + time_years:APOE4_status.y, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model_nfl)
```

Both interaction effects are non-significant.

# Survival model

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline amyloid", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
# Cox survival-model, with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

apoe4 status uppfyller inte assumption om proportional hazards –\> testa med Time-varying effect istället? - tt(APOE4_status)

Gender is not significant, so we remove it:

```{r}
# Cox survival-model, without gender 
surv_model <- coxph(
  surv_object ~ age_at_baseline + APOE4_status,
  data = surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

# Joint model

Multivariate joint models generally require more iterations and a longer burn-in period than univariate models, but for a simple joint model, fewer iterations (like 11,000 iterations with 1,000-2,000 burn-in) are usually sufficient. And Rhat is calculated across multiple chains to assess convergence, so we choose 3 chains.

## 1. For ptau with T-distribution

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )
summary(joint_model)
```

## 2. For ptau with gamma distribution

```{r}
# Joint model 
joint_model_gamma <- jm(
  surv_model,
  long_model_ptau_gamma,
  time_var = "time_years",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

summary(joint_model_gamma)
```

Too high Rhat-values!! Probably because gamma is sensitive to outliers

## 3. For nfl

```{r}
# Joint model 
joint_model_nfl <- jm(
  surv_model,
  long_model_nfl,
  time_var = "time_years",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

summary(joint_model_nfl)
```

# Multivariate joint model

```{r}
Mixed <- list(long_model_ptau, long_model_nfl)
```

```{r}
# Har modellen stabila random slopes?
summary(long_model_ptau)
VarCorr(long_model_nfl)
# Om slope-variansen är runt 0 → den biomarkören bidrar nästan inget i en joint model
```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )      
```

```{r}
multi_joint_model <- jm(surv_model, Mixed,
                  time_var = "time_years",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 20000L,
                  n_burnin = 2000L
                  )
summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival
```

# Multistate survival version

## 1. We convert DIAGNOSIS to numeric state

```{r}
# We convert DIAGNOSIS to numeric state
merged_data <- merged_data %>%
  mutate(state = case_when(
    DIAGNOSIS_dx == "CN"       ~ 1, # CN = 1
    DIAGNOSIS_dx == "MCI"      ~ 2, # MCI = 2
    DIAGNOSIS_dx == "Dementia" ~ 3, # Dementia = 3
    TRUE ~ NA_real_ # missing/other
  ))
```

## 2. We create start-stop times and next_state

```{r}
# We arrange data by RID and visit date, create start-stop times and next_state
ms_data <- merged_data %>%
  arrange(RID, EXAMDATE_ptau) %>%
  group_by(RID) %>%
  mutate(
    baseline_date = first(EXAMDATE_ptau), # baseline exam date
    
    time_start = as.numeric(EXAMDATE_ptau - baseline_date) / 365, # time from baseline in years
    time_stop  = lead(time_start), # time to next visit
    
    next_state = lead(state)  # state at next visit
  ) %>%
  ungroup()

```

## 3. We define monotonic progression events

```{r}
ms_data <- ms_data %>%
  mutate(
    event = case_when(
      is.na(next_state) ~ 0, # censored / no progression
      
      state == 1 & next_state == 2 ~ 1, # CN → MCI
      state == 2 & next_state == 3 ~ 2, # MCI → Dementia
      state == 1 & next_state == 3 ~ 3, # CN → Dementia
      
      TRUE ~ 0 # all other transitions (including stays) censored
    )
  )
```

## 4. We add baseline covariates from merged_data

```{r}
baseline_covariates <- merged_data %>%
  group_by(RID) %>%
  summarise(
    age_at_baseline = first(age_at_visit),
    APOE4_status    = first(APOE4_status.y)
  )
ms_data <- ms_data %>%
  left_join(baseline_covariates, by = "RID")
```

## 5. We inspect the transition table

```{r}
table(ms_data$state, ms_data$next_state)
```

--\> most individuals remain in their current state (for example 1171 MCI → MCI), while some transitions (particularly CN → Dementia (2 events) and Dementia → other states (0-7 events)) are extremely sparse –\> A full multistate joint model would be unable to provide stable estimates.

## 6. We create an interval-censored survival object (time_start → time_stop)

```{r}
surv_object <- Surv(time = ms_data$time_start, 
                    time2 = ms_data$time_stop, 
                    event = ms_data$event != 0) # counts only progression events
```

## 7. We fit a multistate Cox model

```{r}
surv_model <- coxph(
  surv_object ~ age_at_baseline.y + APOE4_status + strata(event), # lets each transition have its own baseline hazard
  data = ms_data, 
  id = RID # handles repeated measures per subject
)

# We view summary of the model
summary(surv_model)
```

```{r}
cox.zph(fit)
```

## 8. Joint model

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )
summary(joint_model)
```
