---
title: "GAMMAL_PLASMA"
format: pdf
editor: visual
---

# GAMMAL PLASMA

```{r, warning=FALSE}
library(dplyr)
library(nlme)
library(survival)
library(JMbayes2)
```

```{r}
# loading data
# Milou
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
# 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
# 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
# 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG

load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

```{r}
# Charley
library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

APOE = APOERES
```

## 1. We select the required variables and merge them

```{r}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# We merge NFL, ptau och diagnosis
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))
```

Checking missing values

```{r}
# checking for missing values with NA

# NfL
sum(is.na(nfl_data))
nfl_data[which(is.na(nfl_data)), ]

# pTAU
sum(is.na(ptau_data))

# checking with ADNI code -4
# NfL
sum(nfl_data$PLASMA_NFL < 0, na.rm = TRUE)

# pTAU
sum(ptau_data$PLASMAPTAU181 < 0, na.rm = TRUE)
```

## 2. We add selected demographics variables (age + gender) and apoe4 status, to later use in survival model and longitudinal model

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We merge with merged_data
merged_data <- merged_data %>%
  left_join(demographics_small, by = "RID")

# We make sure that gender is a factor 
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# We convert EXAMDATE_ptau to date 
merged_data$EXAMDATE_ptau <- as.Date(as.character(merged_data$EXAMDATE_ptau), format="%Y-%m-%d")

# We create age_at_visit based on ptau date 
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_ptau, "%Y")) - merged_data$PTDOBYY

# We create a variable for apoe4 status
apoe$APOE4_status <- ifelse(apoe$GENOTYPE %in% c("2/4","3/4","4/4"), 1, 0)

# We merge apoe with merged_data 
merged_data <- merge(merged_data, apoe[, c("RID", "APOE4_status")], by="RID")
```

## 3. We filter participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

```{r}
# We create a baseline-date
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau)) %>%
  ungroup()

# We create event and time_to_event (where dementia is the event) and add age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_ptau)],
    APOE4_status = APOE4_status[which.min(EXAMDATE_ptau)],
    
    time_to_event = case_when(

    any(DIAGNOSIS_dx == "Dementia") ~
      as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,

  event = as.integer(any(DIAGNOSIS_dx == "Dementia"))
) %>% ungroup()

# We put the survival data back into the merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# We create time_years (time from baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## 6. We check the distribution of each relevant variable for multivariate joint model and transform

```{r}
# We check the distribution of each biomarker
hist(merged_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(merged_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# We log transform the biomarkers 
merged_data$log_PLASMAPTAU181 <- log(merged_data$PLASMAPTAU181)
merged_data$log_PLASMA_NFL <- log(merged_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(merged_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(merged_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

Ptau still looks skewed, even after log transformation --\> In our LME model, we can use a Student’s t-distribution instead of a normal distribution (lme() → normal distribution assumed by default); it is more robust to outliers and heavy-tailed residual distributions.

## 7. We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# We remove missing values 
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(merged_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## 8. We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
# We check the length of each dataset 
length(unique(merged_data$RID))   # longitudinal data
length(unique(surv_data$RID))     # survival data
```

## 9. We check the correlation between the biomarkers

```{r}
# Correlation between the biomarkers 
cor(merged_data$log_PLASMAPTAU181, merged_data$log_PLASMA_NFL, use = "complete.obs")
```

# Longitudinal model

## 1. For ptau

```{r}
# Longitudinal model, with chosen covariates 
library(GLMMadaptive)

# Robust mixed model med t-fördelade residualer
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER,
  random = ~ time_years | RID,                    
  data = merged_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

```{r}
res <- residuals(long_model_ptau, type = "pearson")
fitted_vals <- fitted(long_model_ptau)
plot(fitted_vals, res)
```

```{r}
qqnorm(res)
qqline(res)
```

## 2. For nfl

```{r}
# Longitudinal model, with chosen covariates 
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model_nfl)
```

# Survival model

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline amyloid", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
# Cox survival-model, with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

# Joint model

## 1. For ptau

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model_t,
  time_var = "time_years"
  )
summary(joint_model)
```

Om man inkluderar apoe4-status i LME-modellen:

```         
 Mean   StDev    2.5%    97.5%      P   Rhat (Intercept)    12.0186 10.8322 -1.7967  32.4320 0.2178 1.0109 time_years      1.8400  2.1714 -0.4811   6.2867 0.5400 1.2854 PTGENDER.yMale  6.1144  6.3903 -1.8632  21.1512 0.2422 1.0179 APOE4_status.y 22.4511 10.4932  0.5712  38.8105 0.0000 1.0570 sigma          93.6249 58.5292  1.9585 198.8798 0.0000 1.0461
```

--\> Rhat-värdet vid "time_years" för högt...

om man exkluderar apoe4-status i LME-modellen:

```         
                Mean   StDev    2.5%    97.5%      P   Rhat (Intercept)    15.0192  9.8059  0.3463  32.6337 0.0047 1.0126 time_years      2.2955  2.4611 -0.4832   7.1026 0.4838 1.0286 PTGENDER.yMale  8.5018  7.5652 -0.6200  25.1432 0.1222 1.0134 sigma          92.9033 59.1923  1.2973 196.7808 0.0000 1.0296
```

Variabeln är dock signifikant i LME-modellen (p-värde: 0e+00), men vi ser ju att Rhat-värdena blir för höga med den inkluderad...

--\> Ska vi exkludera den i LME-modellen? (fortfarande ett kovariat i survival-modellen, där den är signifikant

## 2. For NFL

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model_nfl,
  time_var = "time_years"
  )

summary(joint_model)
```

# Multivariate joint model

```{r}
Mixed <- list(long_model_ptau, long_model_nfl)
```

```{r}
# Har modellen stabila random slopes?
summary(long_model_ptau)
VarCorr(long_model_nfl)
# Om slope-variansen är runt 0 → den biomarkören bidrar nästan inget i en joint model
```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

```{r}
fForms <- list(
  "log_PLASMAPTAU181"   = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL"     = ~ value(log_PLASMA_NFL) 
  )      
```

```{r}
multi_joint_model <- jm(surv_model, Mixed,
                  time_var = "time_years",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 20000L,
                  n_burnin = 2000L
                  )
summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival
```
