---
title: "GAMMAL_PLASMA"
format: pdf
editor: visual
---

# GAMMAL PLASMA

```{r, warning=FALSE}
library(dplyr)
library(nlme)
library(survival)
library(JMbayes2)
```

```{r}
# loading data
# Milou
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
# nfl_data <- BLENNOWPLASMANFLLONG
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
# ptau_data <- UGOTPTAU181
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
# diagnosis_data <- DXSUM
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
# demographics <- PTDEMOG


# Charley
library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

APOE = APOERES
```

## 1. We select only the required variables and merge them

```{r}
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# 1. Lägg _dx på diagnos-kolumner
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# 2. Slå ihop NFL, ptau och diagnos
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))
```

Checking missing values

```{r}
# checking for missing values with NA

# NfL
sum(is.na(nfl_data))
nfl_data[which(is.na(nfl_data)), ]

# pTAU
sum(is.na(ptau_data))

# checking with ADNI code -4
# NfL
sum(nfl_data$PLASMA_NFL < 0, na.rm = TRUE)

# pTAU
sum(ptau_data$PLASMAPTAU181 < 0, na.rm = TRUE)
```

## 2. We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# 1. Håll bara demografi-variabler vi behöver och en rad per person
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# 2. Slå ihop med merged_data
merged_data <- merged_data %>%
  left_join(demographics_small, by = "RID")

# 3. Säkerställ att kön är faktor
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# 4. Konvertera EXAMDATE_ptau till Date
merged_data$EXAMDATE_ptau <- as.Date(as.character(merged_data$EXAMDATE_ptau), format="%Y-%m-%d")

# 5. Skapa age_at_visit baserat på ptau datum
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_ptau, "%Y")) - merged_data$PTDOBYY
```

## 3. We filter participants with at least 3 measurements

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event) + lägger till age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_ptau)],
    PTGENDER = PTGENDER[which.min(EXAMDATE_ptau)],
    
    time_to_event = case_when(

    any(DIAGNOSIS_dx == "Dementia") ~
      as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,

  event = as.integer(any(DIAGNOSIS_dx == "Dementia"))
) %>% ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## 6. We check the distribution of each relevant variable for multivariate joint model and transform

```{r}
# Vi kollar efter outliers 
boxplot(merged_data$PLASMAPTAU181, main="PLASMAPTAU181")
boxplot(merged_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
merged_data$log_PLASMAPTAU181 <- log(merged_data$PLASMAPTAU181)
merged_data$log_PLASMA_NFL <- log(merged_data$PLASMA_NFL)
```

## 7. We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(merged_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## 8. We make sure we don't have any extreme values (for the longitudinal model)

```{r}
summary(merged_data$log_PLASMAPTAU181)
any(is.infinite(merged_data$log_PLASMAPTAU181)) # should be FALSE
# Inf = extreme values 

summary(merged_data$log_PLASMA_NFL)
any(is.infinite(merged_data$log_PLASMA_NFL)) # should be FALSE
# Inf = extreme values 
```

```{r}
# Remove rows with -Inf or Inf in log_CENTILOIDS
merged_data <- merged_data[is.finite(merged_data$log_PLASMAPTAU181), ]
any(is.infinite(merged_data$log_PLASMAPTAU181))  # should be FALSE
any(is.na(merged_data$log_PLASMAPTAU181))       # also FALSE

merged_data <- merged_data[is.finite(merged_data$log_PLASMA_NFL), ]
any(is.infinite(merged_data$log_PLASMA_NFL))  # should be FALSE
any(is.na(merged_data$log_PLASMA_NFL))       # also FALSE
```

## 9. We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
length(unique(merged_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

# Longitudinal model

```{r}
# Longitudinal model
long_model <- lme(
  log_PLASMAPTAU181 ~ time_years + PTGENDER.y, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline amyloid", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data <- surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

# Joint model

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```

# Multivariate joint model

```{r}
fm_ptau <- lme(log_PLASMAPTAU181 ~ time_years + PTGENDER.x,
                random = ~ time_years | RID,
                data = merged_data)

fm_nfl <- lme(log_PLASMA_NFL ~ time_years + PTGENDER.x,
                  random = ~ time_years | RID,
                  data = merged_data)

Mixed <- list(fm_ptau, fm_nfl)
```

```{r}
# Har modellen stabila random slopes?
VarCorr(fm_ptau)
VarCorr(fm_nfl)
# Om slope-variansen är runt 0 → den biomarkören bidrar nästan inget i en joint model
```

```{r}
fForms <- list(
  "log_PLASMAPTAU181"   = ~ value(log_PLASMAPTAU181) + slope(log_PLASMAPTAU181),   # tau
  "log_PLASMA_NFL"     = ~ value(log_PLASMA_NFL) + slope(log_PLASMA_NFL)       # neurodegeneration
  )      
```

```{r}
multi_joint_model <- jm(surv_model, Mixed,
                  time_var = "time_years",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 20000L,
                  n_burnin = 2000L
                  )
summary(multi_joint_model)
```

# 

# 2 separate joint models

```{r}
lme_log_PLASMAPTAU181 <- lme(
  log_PLASMAPTAU181 ~ time_years + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

lme_log_PLASMA_NFL <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )
```

```{r}
# Joint model 1 
joint_model1 <- jm(
  surv_model,
  lme_log_PLASMAPTAU181,
  time_var = "time_years"
  )

summary(joint_model1)
```

```{r}
# Joint model 2
joint_model2 <- jm(
  surv_model,
  lme_log_PLASMA_NFL,
  time_var = "time_years"
  )

summary(joint_model2)
```
