---
title: "Joint Latent Class Mixed Models / Multistate Survival"
format: pdf
editor: visual
---

```{r}
# loading the packages
library(lcmm)
library(dplyr)
library(tidyr)
library(survival)
library(ADNIMERGE2)
library(nlme)
library(lcmm)
library(lubridate)
library(ggsurvfit)
library(Hmisc)
```

```{r}
# loading data
# NfL
nfl_data <- BLENNOWPLASMANFLLONG

# pTAU
ptau_data <- UGOTPTAU181

# dx data
diagnosis_data <- DXSUM

# demographic data
demographics <- PTDEMOG

# genetic data
apoe <- APOERES
```

# Pre-processing

-   Check how we merge with diagnosis data: we have more data than is present here. For example RID 2 changes from CN to MCI to CN again, but with the full diagnosis data we see that they convert again to MCI. Let's think about how to include this data correctly. Should we make a different data set with conversions from the full diagnosis data set?

```{r}
# selecting relevant variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# adding a suffix to the dx data
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))
```

```{r}
# removing missing values
nfl_data <- na.omit(nfl_data)
ptau_data <- na.omit(ptau_data)
diagnosis_data <- na.omit(diagnosis_data)
```

### Checking RID with more than one measurement recorded on the same exam date

-   There are 52 individuals with two measurements recorded on the same exam date.

-   How should we deal with these? Average the measurements? Remove the observations?

-   Some of the values are quite different, what do we assume by averaging them? What effect does it have to remove them?

-   For now, remove them:

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
# ptau_data %>%
#   filter(RID %in% ptau_double$RID) %>% 
#   filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

```{r}
# merging only nfl and ptau
nfl_ptau <- nfl_data %>%
  merge(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))
```

### Baseline date in nfl_ptau

```{r}
# checking which observations have the same examdate for nfl and ptau
common <- which(nfl_ptau$EXAMDATE_nfl %in% nfl_ptau$EXAMDATE_ptau)

# checking the observations with different exam dates
nfl_ptau[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
nfl_ptau <- nfl_ptau %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
  ungroup()
```

### Time since baseline in nfl ptau

```{r}
# creating a time since baseline visit variable (in years)
nfl_ptau <- nfl_ptau %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

### Selected demographics in nfl_ptau

-   Merging DOB, gender and creating an age at baseline variable

-   Converting gender to factor variable

```{r}
# selecting relevant variables (age, gender) and keeping only one row per person
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# merging the demographic data with the biomarker data
nfl_ptau <- nfl_ptau %>% merge(demographics_small, by = "RID")
```

```{r}
# converting the gender variable to a factor variable 
nfl_ptau <- nfl_ptau %>% mutate(PTGENDER = as.factor(PTGENDER))

# creating an age at baseline variable
nfl_ptau <- nfl_ptau %>% mutate(
  age_at_baseline = as.numeric(format(baseline_date, "%Y")) - PTDOBYY)
```

### Factor variable for genotype in apoe

-   Trying with a factor for each specific genotype

-   If there is not good distinction between these, since we do not have many individuals with some genotypes, we can collpase down to fewer factors later

```{r}
# creating a factor variable for ApoE genotype and selecting relevant variables
apoe = apoe %>% mutate(
  genotype = as.factor(case_when(
    GENOTYPE == "4/4" ~ "e4/e4",
    GENOTYPE %in% c("3/4", "2/4") ~ "e4",
    GENOTYPE == "3/3" ~ "e3/e3",
    GENOTYPE %in% c("2/3", "2/2") ~ "e2"
  ))
) %>%
  select(RID, genotype)
```

### We filter participants with at least 3 measurements in nfl_ptau

```{r}
# Filtering individuals with at least 3 measurements
keep <- unique((nfl_ptau %>% group_by(RID) %>% filter(n() > 2))$RID)
nfl_ptau <- nfl_ptau %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

## Survival data

```{r}
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)

surv_data <- diagnosis_data %>%
  group_by(RID) %>%
  
  # We create baseline date per patient (whichever EXAMDATE comes first, nfl or ptau)
  mutate(baseline_date = min(EXAMDATE),
  
  # We calculate event and time_to_event based on all diagnosis data
    # 1 if patient ever developed dementia
  event = as.integer(any(DIAGNOSIS == "Dementia")),
  time_to_event = case_when(
    event == 1 ~
      as.numeric(min(EXAMDATE[DIAGNOSIS == "Dementia"]) - baseline_date)/365,
    TRUE ~
      as.numeric(max(EXAMDATE) - baseline_date)/365)) %>%
  
      ungroup() %>%
  select(RID, event, time_to_event) %>%
  distinct()
  
surv_data <- surv_data %>% filter(RID %in% keep)

# We check the number of events 
table(surv_data$event)
```

### Merging genotype data

```{r}
surv_data <- surv_data %>% merge(apoe, by = "RID")
```

### Creating a single EXAMDATE variable

```{r}
nfl_ptau <- nfl_ptau %>%
  mutate(EXAMDATE = EXAMDATE_nfl + (EXAMDATE_nfl - EXAMDATE_ptau)) %>%
  select(!c(EXAMDATE_nfl, EXAMDATE_ptau)) %>%
  relocate(c(baseline_date, EXAMDATE, time_years), .after = VISCODE2)
```

# Joint Latent Class Mixed Models DATA

### Creating the survival data from the diagnosis data, and then merging the longitudinal data and survival data by RID

-   Keeps all diagnosis data

-   We can then use diagnoses which happen not on the same VISCODE as the biomarker data

-   Check VISCODE2 variable names -\> change to \_bio and \_dx

```{r}
# reference
# lcmm package data - example that works with the jlcmm
d = data_lcmm
```

```{r}
# ordering the data
nfl_ptau <- nfl_ptau %>% arrange(RID, time_years)
```

```{r}
# merging longitudinal and survival data by RID only not VISCODE
jlcmm_data <- nfl_ptau %>%
  merge(surv_data, by = "RID") %>%
  relocate(c(event, time_to_event), .after = time_years)
```

```{r}
# transforming the data
jlcmm_data <- jlcmm_data %>%
  mutate(ln_PLASMA_NFL = log(PLASMA_NFL))
```

# Data visualisation

## Plotting the relationships between biomarkers by RID

```{r}
par(mfrow = c(1, 2))
boxplot(PLASMA_NFL ~ RID, data = nfl_ptau)
boxplot(PLASMAPTAU181 ~ RID, data = nfl_ptau)
```

```{r}
ggplot(jlcmm_data, aes(x = time_years, y = PLASMA_NFL, colour = as.factor(event))) +
  geom_point() +
  theme_classic()

ggplot(jlcmm_data, aes(x = time_years, y = PLASMAPTAU181, colour = as.factor(event))) +
  geom_point() +
  theme_classic()
```

# ..........................................................................................................

# JLCMM fitting

## Specifying random components

-   random = \~ 1 (random intercept)

-   random = \~ time (random slope)

-   random = \~ 1 + time (random intercept + slope)

## Model summary output

-   Iteration process/convergence criteria

    -   

-   Goodness-of-fit statistics

    -   Information criteria are an example of model selection criteria. Some information criteria are better than others for a specific purpose, depending on, for example the data generating process, suggested models and sample size (pg 2). Zhang et al. (pg 4) explain how when the objective of model selection is for prediction, it may not be critical that the selected covariates change with sample size, yet when the focus is on inference, stable model covariates are necessary for a clear interpretation of the data genereating process. Model consistency: inference, model efficiency: prediction.

        Zhang, J., Yang, Y. and Ding, J. (2023). Information criteria for model selection. *Wiley interdisciplinary reviews. Computational statistics*, 15(5). doi:https://doi.org/10.1002/wics.1607.

    -   Akaike Information Criterion: Akaike (1974, 1998 -\> original sources)

        $$
        AIC = 2k - 2 ln(\widehat L)
        $$

        $$
        \begin{aligned}
        where \quad \widehat L =\text{the maximised value of the likelihood function} \\
        k = \text{the number of parameters estimated by the model}
        \end{aligned}
        $$

    -   Bayesian Information Criterion: Schwarz (1978): penalises complex models more than AIC

        $$
        BIC = kln(n) - 2ln(\widehat L)
        $$

    -   The score statistic for the score test of conditional independence (Jacqmin-Gadda) tests whether there is residual dependence betweent the longitudinal outcome and the survival outcome: given the latent class, the responses are independent of each other, there are no residual correlations after conditioning on class. In other words, the latent class is the sole cause of the observed variables, and if we knew the latent classes, there would be NO residual dependence between the outcomes (they are independent of each other) such that:

        $$
        P(Y_1, ... , Y_J|C) = \prod^J_{j = 1}P(Y_j|C)
        $$

        Pairs $(Y_i, Y_k)$ are chosen, and a residual association, $\delta_{i,k}$ is defined for each pair. Under conditional independence:

        $$
        H_0 : \delta_{i,k} = 0 \quad \text{for all item pairs}
        $$

        A restricted model assuming conditional independence is fit. An alternative model is posited, where $\delta_{i,k}$ is allowed to be non-zero. The score statistics analyses the slope of the log-likelihood of the restricted model, and if allowances for non-zero residual associations would improve fit, the conditional independence assumption is violated. If the score statistic is large (significant, p-value \< 0.05) then we cannot assume conditional independence.

-   Estimates, Wald and p-value

-   Variance-covariance matrix for the random effects

    -   If the

### Model fitting examples from the lcmm package

```{r}
# jlcmm1 <- Jointlcmm(Ydep1 ~ X1 * Time,
#                     random = ~ Time,
#                     subject = "ID",
#                     survival = Surv(Tevent, Event) ~ X1 + X2, hazard = "3-quant-splines",
#                     data = d)

# jlcmm2 <- Jointlcmm(Ydep1 ~ Time * X1,
#                     random = ~ Time,
#                     subject = "ID",
#                     mixture = ~ Time,
#                     survival = Surv(Tevent, Event) ~ X1 + mixture(X2),
#                     hazard = "3-quant-splines", hazardtype = "PH", ng = 2,
#                     data = data_lcmm, B = jlcmm1)
```

```{r}
# jlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years,
#   random = ~ time_years,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ genotype,
#   data = jlcmm_data,
#   maxiter = 500
# )
```

# JLCMM PROBLEMS

-   Infinite parameters/the program stopped abnormally: Hessian matrix does not converge

```{r}
# checking for slope variance = 0 or correlation = +-1
lme(ln_PLASMA_NFL ~ time_years, random = ~ time_years | RID, data = jlcmm_data)
```

```{r}
# checking for the event distribution by genotype
table(jlcmm_data$event, jlcmm_data$genotype)
```

```{r}
# checking the range for time_years and ln_PLASMA_NFL
summary(jlcmm_data$time_years)
summary(jlcmm_data$ln_PLASMA_NFL)
```

## Try again with time_years_z

```{r}
# adding a scaled time_years variable
jlcmm_data <- jlcmm_data %>%
  mutate(time_years_z = scale(time_years))
```

```{r}
# fitting a simple mixed model to check the random slope 
lme(ln_PLASMA_NFL ~ time_years_z, random = ~ time_years_z | RID, data = jlcmm_data)
```

```{r}
# jlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years_z,
#   random = ~ time_years_z,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ genotype,
#   data = jlcmm_data,
#   maxiter = 500
# )
```

## Try again with time_years_z and ln_NFL_z

```{r}
jlcmm_data <- jlcmm_data %>%
  mutate(ln_PLASMA_NFL_z = scale(ln_PLASMA_NFL))
```

```{r}
# fitting a simple mixed model to check the random slope 
lme(ln_PLASMA_NFL_z ~ time_years_z, random = ~ time_years_z | RID, data = jlcmm_data)
```

```{r}
# jlcmm(
#   fixed = ln_PLASMA_NFL_z ~ time_years_z,
#   random = ~ time_years_z,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ genotype,
#   data = jlcmm_data,
#   maxiter = 500
# )
```

## jlcmm_data_test: try again with only time_to_event \> 0

-   This works!

-   $AIC = 7419$

-   $BIC = 7470$

-   Non-significant intercept

-   Low slope variance

-   We see that genotype e3/e3 is not significantly different from genotypes containing e2 (maybe since we have lower number of individuals with e2 genotypes): we can try collapsing the genotype variable down to e2/e3 containing genotypes, e4 heterozygotes and e4 homozygotes (try later)

```{r}
# there are 83 individuals with a start dementia dx, so time_to_event is 0
# remove these and test the model again
jlcmm_data %>% filter(time_to_event <= 0)

# creating a test data frame with no time_to_event = 0
jlcmm_data_test <- jlcmm_data %>% filter(time_to_event > 0)
```

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL_z ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test,
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

## Try with normal time_years and ln_PLASMA_NFL

-   $AIC = 3237.74$

-   $BIC = 3288.37$

-   Lower information criteria, and significant intercept and slope

-   Even lower slope variance

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years,
  random = ~ 1 + time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

## Try with time_years_z and ln_PLASMA_NFL

-   $AIC = 3237$

-   $BIC = 3288$

-   Increased slope variance

# ..........................................................................................................

# NFL: Latent class model (intercept and slope)

-   Need to run latent class model first to check latent structure

-   hlme() fits linear mixed models (ng = 1) and latent class linear mixed models (ng \> 1). Fit hlme() with ng = 1 and use the model with one class as a start point for the model with ng = 2

### M0 (1 latent class)

-   $log-likelihood = 747.51$

-   $AIC = 1507$

-   $BIC = 1535$

```{r}
hlme_m0 <- hlme(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ time_years_z,
  subject = "RID",
  data = jlcmm_data_test
)
```

```{r}
summary(hlme_m0)
```

### M1 (2 latent classes)

-   Intercept and slope nearly identical between classes, log-likelihood identical between model with 1 latent class and model with 2 latent classes

-   $log-likelihood = -747.51$

-   $AIC = 1513$

-   $BIC = 1554$

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMA_NFL ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

-   Two classes show a nearly even split, shows no real class structure, random 50/50 split

-   Posterior classification table shows no distinction between classes

$$
P(C_1 |C_1) = 0.51 \quad P(C_1 |C_2) = 0.49 \quad P(C_2 |C_1) = 0.49 \quad P(C_2 |C_2) = 0.51
$$

-   Histograms of the posterior probabilities by class show values clustered around 0.5 for both class1 and class2

-   Scatterplot of class certainty again shows values clustered in the center

```{r}
postprob(hlme_m1)
```

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()

# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()

# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

### Comparing ng = 1 and ng = 2

-   $BIC$ lower in the 1 class model

-   Entropy very low in the 2 class model

```{r}
summarytable(hlme_m0, hlme_m1, which = c("G", "loglik", "BIC", "entropy"))
```

### Shannon Entropy

-   Uncertainty of the variable itself

-   Entropy is a quantification of the (average) level of uncertainty associated with a random variable's potential outcomes (similary to entropy in thermodynamics) <https://en.wikipedia.org/wiki/Entropy_(information_theory)>

-   The expected amount of information needed to describe the state of the random variable, given the distribution of probabilities across all possible outcomes

-   "Information content" of an outcome is related to how probable it is. If a highly likely outcome occurs, the information content is small. The information/surprisal, $I(E)$ of an event approaches $0$ as $p(E)$ approaches $1$:

$$
I(E) = log \left( \frac{1}{p(E)} \right) = -log(p(E))
$$

-   Entropy $H$ calculates the average expected amount of information across all possible outcomes of an random variable, $X$, such that:

    $$ 
    H(X) = - \sum_{x \in X} p(x)log(p(x))
    $$

-   why 1 - p(x)log(p(x))

-   <https://www.youtube.com/watch?v=LBOVhy3ITA4>

-   <https://www.youtube.com/watch?v=-4y4X8ELcEM>

### Assignment based entropy

-   Uncertainty of the class assignment!

-   Posterior probabilities:

$$
p_i(c_g) = P(C = c_g | \mathbf{X}_ii)
$$

-   For each individual we calculate shannon entropy, adn then the overall entropy is H = 1 - sum (H_i) /

```{r}
p <- as.matrix(pp[, grep("prob", names(pp))])  # extract prob columns
N <- nrow(p)
G <- ncol(p)

entropy <- 1 - ( -sum(p * log(p)) / (N * log(G)) )
entropy
```

## Adding covariates to the latent class structure

-   The lcmm without class membership covariates was not significant, meaning that the longitudinal biomarker trajectories alone do not predict latent class membership, the latent classes are predicted by covariates NOT biomarker trajectories

-   Biomarker trajectories alone do not distinguish clusters, hence there is no latent class structure based solely on NfL or ptau trajectories. Demographic covariates explain the heterogenity in disease progression.

### M0 (1 latent class) same as before

-   $AIC = 1640$

-   $BIC = 1663$

```{r}
hlme_m0 <- hlme(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ time_years_z,
  subject = "RID",
  data = jlcmm_data_test
)
```

```{r}
summary(hlme_m0)
```

### M1 (2 latent classes)

-   All genotype levels (with e2 allele collapsed into e3 variants), age at baseline, and gender are significant

-   $AIC = 1283$

-   $BIC = 1343$

-   $log-likelihood = -628.46$

-   Both information criteria are reduced and log-likelihood increased

-   Intercepts and slopes are much different for the two classes compared to the reduced model specified above

-   $19$ iterations vs. $5$ from the reduced model

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMA_NFL ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    classmb = ~ genotype + age_at_baseline + PTGENDER,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

-   Much better separation:

$$
P(C_1 |C_1) = 0.89 \quad P(C_1 |C_2) = 0.11 \quad P(C_2 |C_1) = 0.11 \quad P(C_2 |C_2) = 0.89
$$

-   $86\%$ have posterior probabilities over 0.70, $77\%$ over 0.80 and $65 - 67 \%$ have over 0.90.

-   plots of the posterior probabilities show most values clustered at either 0 or 1, lower density closer to 0.5

```{r}
postprob(hlme_m1)
```

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()

# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()

# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

### Entropy

-   $0.64$, moderate classification with two latent classes. Could this motivate testing more latent classes?

```{r}
p <- as.matrix(pp[, grep("prob", names(pp))])  # extract prob columns
N <- nrow(p)
G <- ncol(p)

entropy <- 1 - ( -sum(p * log(p)) / (N * log(G)) )
entropy
```

### Spaghetti plots

-   Can we do this? Individual level data... ?

<https://stats.oarc.ucla.edu/r/faq/how-can-i-visualize-longitudinal-data-in-ggplot2/>

<https://www.cararthompson.com/posts/2025-01-14-optimising-colours-in-a-spaghetti-plot/>

```{r}
# predictions by RID for each measurement time
pred_RID <- predictY(hlme_m1, jlcmm_data_test, var.time = "time_years_z")

# extracting the posterior probabilities from the 2 class model
pp <- hlme_m1$pprob

# selecting which class each individual is assigned to based on which class has the highest posterior probability
assigned_class <- apply(pp[ ,c("prob1","prob2")], 1, which.max)
```

```{r}
ggplot(data = jlcmm_data_test, aes(x = time_years_z, y = ln_PLASMA_NFL, group = genotype)) +
  geom_line() +
#   to visualise by gender level 
  facet_grid(. ~ PTGENDER) +
#   for loewess smoother
  # stat_smooth(aes(group = 1)) +
#   for avergae among group
  stat_summary(aes(group = 1), geom = "point", fun = mean) +
  theme_minimal()
```

### Plotting biomarker trajectories

```{r}
# creating a data frame for new data with the baseline levels of genotype and gender and average age: so a data frame over 100 time points, for females with e2 or e3 alleles and an average age at baseline
newdata <- data.frame(
  time_years_z = seq(min(jlcmm_data_test$time_years_z),
                     max(jlcmm_data_test$time_years_z),
                     length = 100),
  genotype = factor("e2", levels = c("e2", "e3/e3", "e4", "e4/e4")),
  PTGENDER = factor("Female", levels = c("Female", "Male")),
  age_at_baseline = mean(jlcmm_data_test$age_at_baseline)
)
```

```{r}
# checking which variables to include in new data
hlme_m1$Xnames2

# checkng how new data has been created
str(newdata)
```

```{r}
# predicting the longitudinal ln_PLASMA_NFL values given class for the new data
preds <- predictY(hlme_m1, newdata, var.time = "time_years_z")
```

```{r}
preds_df <- data.frame(
  time = newdata$time_years_z,
  class1 = preds$pred[,1],
  class2 = preds$pred[,2]
)

ggplot(preds_df, aes(x = time)) +
  geom_line(aes(y = class1, color = "Class 1")) +
  geom_line(aes(y = class2, color = "Class 2")) +
  labs(y = "Predicted ln(NFL)", x = "Standardized Time") +
  theme_minimal()
```

### M2 (3 latent classes)

```{r}
# hlme_m2 <- hlme(
#     fixed = ln_PLASMA_NFL ~ time_years_z,
#     mixture = ~ time_years_z,
#     random = ~ time_years_z,
#     classmb = ~ genotype + age_at_baseline + PTGENDER,
#     subject = "RID",
#     ng = 2,
#     data = jlcmm_data_test,
#     B = hlme_m1
#   )
```

# PTAU: Latent class model (intercept and slope)

```{r}
# raw data
ggplot(nfl_ptau, aes(x = PLASMAPTAU181)) +
  geom_histogram(colour = "white", fill = "cornflowerblue") + 
  theme_minimal()

# log transformed
ggplot(nfl_ptau, aes(x = log(PLASMAPTAU181))) +
  geom_histogram(colour = "white", fill = "cornflowerblue") + 
  theme_minimal()

# cube root transformed
ggplot(nfl_ptau, aes(x = sqrt(sqrt(PLASMAPTAU181)))) +
  geom_histogram(colour = "white", fill = "cornflowerblue") + 
  theme_minimal()
```

### M0 (1 latent class)

```{r}
jlcmm_data_test <- jlcmm_data_test %>% mutate(ln_PLASMAPTAU181 = log(PLASMAPTAU181))
```

```{r}
hlme_m0 <- hlme(
  fixed = ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  subject = "RID",
  data = jlcmm_data_test
)
```

```{r}
summary(hlme_m0)
```

### M1 (2 latent classes)

-   Same problem as with nfl

-   Converges in 5 iterations, poor classification

-   Posterior classification table shows values close to 0.50, poor classification

-   Fixed effects in the longitudinal model are not significant

-   Next: add covariates

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMAPTAU181 ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

```{r}
postprob(hlme_m1)
```

### M1 (2 latent classes) with covariates

-   Improvement in $AIC$, $BIC$ and $log-likelihood$ is weak

-   e3/e3 and e4/e4 not significant, collapse genotype down to "e2 e3", "e4" and "e4/e4"

-   Test with cube root transformed ptau, since better gaussian approximation

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMAPTAU181 ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    classmb = ~ genotype + age_at_baseline + PTGENDER,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

-   class 1 \~ $30\%$

-   class 2 \~ $70\%$

```{r}
postprob(hlme_m1)
```

### M1 (2 latent classes) with only genotype as covariate

-   Posterior probabilities good, but time_years class 2 is not significant

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMAPTAU181 ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    classmb = ~ genotype,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

```{r}
postprob(hlme_m1)
```

### M1 (2 latent classes) with only age_at_baseline as covariate

-   Posterior probabilities OK but intercept not significant for

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMAPTAU181 ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    classmb = ~ age_at_baseline,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

```{r}
postprob(hlme_m1)
```

### M1 (2 latent classes) with only PTGENDER as covariate

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMAPTAU181 ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    classmb = ~ PTGENDER,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

```{r}
postprob(hlme_m1)
```

### M1 (2 latent classes) with ln_PLASMANFL as covariate!

-   **RETHINK HERE: check that it is significant when genotype is collapsed down to 3 levels, i think it will be so this can be ignored...**

-   This is significant!

-   What does this tell us? NfL classifies better than ptau181. ptau improves only when nfl is included as a covariate. NfL may give a better indication of disease severity/progression than ptau181, despite ptau181 being the disease specific biomarker - in fact, i think this comes up in the literature, that NfL is better for tracking disease when disease after diagnosis **\*\*SOURCE\*\***

-   Other covariates not significant, but nfl is. Is this due to correlation? Could we reduce two dimensions to two uncorrelated PCs and test if there is still a relationship? The correlation between them is not extremely high, so PCA will not yield great results

-   Evidence for multivariate latent class mixed model?

```{r}
hlme_m1 <- hlme(
    fixed = ln_PLASMAPTAU181 ~ time_years_z,
    mixture = ~ time_years_z,
    random = ~ time_years_z,
    classmb = ~ ln_PLASMA_NFL,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

```{r}
postprob(hlme_m1)
```

### M0 (2 latent class) with cube root ptau

```{r}
jlcmm_data_test <- jlcmm_data_test %>%
  mutate(cubert_PLASMAPTAU181 = sqrt(sqrt(PLASMAPTAU181)))
```

```{r}
hlme_m0 <- hlme(
  fixed = cubert_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  subject = "RID",
  data = jlcmm_data_test
)
```

```{r}
summary(hlme_m0)
```

### M1 (2 latent classes) with cube root ptau and covariates

```{r}
hlme_m1 <- hlme(
  fixed = cubert_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  mixture = ~ time_years_z,
  classmb = ~ genotype + age_at_baseline + PTGENDER,
  subject = "RID",
  ng = 2,
  data = jlcmm_data_test,
  B = random(hlme_m0)
)
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

-   class 1 \~ $42\%$

-   class 2 \~ $58\%$

-   reduced classification accuracy for class 2

```{r}
postprob(hlme_m1)
```

# NFL & PTAU: Multivariate latent class mixed model

-   Significant coefficients, except genotype e3/e3 (collapse further down to e2 e3, e4, and e4/e4)

-   Converges in 29 iterations, posterior probabilities good

```{r}
mlcmm_m0 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  subject = "RID",
  data = jlcmm_data_test
)
```

```{r}
mlcmm_m1 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  mixture = ~ time_years_z,
  subject = "RID",
  classmb = ~ genotype + age_at_baseline + PTGENDER,
  ng = 2,
  data = jlcmm_data_test,
  B = random(mlcmm_m0)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

```{r}
plot(mlcmm_m1, var.time = "time_years_z", which = "fit")
```

# ..........................................................................................................

# JLCMM: NFL

## M0 (1 latent class)

-   Score test statistic for the assumption of conditional independence is significant: this means the assumption is violated. There is some residual dependence between the longitudinal outcome and the survival outcome. This is evidence for there being more than one latent class in the sample.

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

### Collapsing genotype down to 3 factors

-   e2/e3 containing genotypes

-   e4 heterozygotes

-   e4 homozygotes

-   Now all genotype factors significant

-   Now time_years variance still low, try with only random intercept

```{r}
# checking current factors
table(jlcmm_data_test$genotype)
```

```{r}
jlcmm_data_test <- jlcmm_data_test %>%
  mutate(
    genotype = case_when(
      genotype == "e2" ~ "e2 e3",
      genotype == "e3/e3" ~ "e2 e3",
      genotype == "e4" ~ "e4",
      genotype == "e4/e4" ~ "e4/e4"
    )
  )
```

```{r}
# checking changed factors
table(jlcmm_data_test$genotype)
```

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

# ..........................................................................................................

# GENOTYPE 3 LEVELS:

# ..........................................................................................................

# NFL & PTAU: mlcmm 3 genotype levels

### With baseline age, PTGENDER and genotype

-   $AIC: 6740.57 \to 6738.58$

-   $BIC: 6818.82 \to 6812.22$

-   $log-likelihood: -3353.29 \to -3353.29$

-   No big difference, but not worse (and would we really expect a change in these considering it is only a refactoring of a variable, we are not adding more information, but changing the way it is handled), and now all coefficients are significant

-   Bigger variance for the slope than in the univariate models

```{r}
mlcmm_m1 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  mixture = ~ time_years_z,
  subject = "RID",
  classmb = ~ genotype + age_at_baseline + PTGENDER,
  ng = 2,
  data = jlcmm_data_test,
  B = random(mlcmm_m0)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

```{r}
plot(mlcmm_m1)
```

```{r}
plot(mlcmm_m1, var.time = "time_years_z", which = "fit")

```

### Trying with no classmb covariates

-   Evidence again for disease trajectories being described by covariates as opposed to a true latent class

-   No significant covariates

-   Posterior probabilities show bad classification, random class structure

```{r}
mlcmm_m1 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  mixture = ~ time_years_z,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data_test,
  B = random(mlcmm_m0)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

### Checking results with ng = 1 and ng = 2 model comparison

-   $BIC$ lower in the 1 class model

-   Entropy extremely low in the 2 class model

```{r}
summarytable(mlcmm_m0, mlcmm_m1, which = c("G", "loglik", "BIC", "entropy"))
```

```{r}
plot(mlcmm_m1, var.time = "time_years_z", which = "fit", bty = "l")
```

### Intercept only and no classmb covariates

-   NO separation at all

```{r}
mlcmm_m0 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ 1,
  subject = "RID",
  data = jlcmm_data_test
)
```

```{r}
mlcmm_m1 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data_test,
  B = random(mlcmm_m0)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

### ..........................................................................................................

# JLCMM: NFL

### Try with only random intercept

-   Higher AIC and BIC

-   Actually figure out what is 'too small' slope variance...

-   Again, significant score test statistic is evidence for latent classes.

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

### Trying again with slope and intercept

-   Correlation should not be +- 1: no violation here

$$
Corr_{intercept, slope} = \frac{cov_{intercept, slope}}{\sigma_{intercept}\sigma_{slope}} =  \frac{0.00669}{\sqrt{0.19574}\sqrt{0.00255}} \approx 0.30 
$$

-   The variance of the intercept and slope should not be 0. The question is, what constitutes *too* small?

-   $AIC = 3236$

-   $BIC = 3282$

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
# calculating the correlation betweent the random intercept and slope, this should not be 1
0.00669 / (sqrt(0.19574) * sqrt(0.00255))
```

```{r}
plot(m0)
```

## M1 (2 latent classes)

-   Start values from m0

-   Intercept not significant at the 0.05 level

-   Score test statistic no longer significant, conditional independence assumption is met: is this evidence for 2 latent classes as opposed to 3? Is that why the joint latent class model with ng = 3 does not converge?

-   $AIC = 3145$

-   $BIC = 3214$

```{r}
m1 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  mixture = ~ time_years_z,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test,
  B = m0
)
```

```{r}
summary(m1)
```

```{r}
plot(m1)
```

### Posterior probabilities

-   Good classification

-   Differences between classification based on longitudinal only and longitudinal and time to event data

```{r}
postprob(m1)
```

# THIS WORKS!!!!! WITHOUT COVARIATE IN SURVIVAL MODEL

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
m1 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  mixture = ~ time_years_z,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data_test,
  B = m0
)
```

```{r}
summary(m1)
```

```{r}
postprob(m1)
```

### Random start values

-   Score test statistic no longer significant, conditional independence assumption is met

-   $AIC = 3145$

-   $BIC = 3214$

```{r}
m1 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  mixture = ~ time_years_z,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test,
  B = random(m0)
)
```

```{r}
summary(m1)
```

```{r}
plot(m1)
```

## M2 (3 latent classes)

-   Start values from m0

-   100 iterations without convergence

```{r}
m2 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  mixture = ~ time_years_z,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test,
  B = m0
)
```

```{r}
summary(m2)
```

```{r}
plot(m2)
```

### Try with max 500 iterations

-   Still does not converge

```{r}
# m2 <- jlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years_z,
#   random = ~ 1 + time_years_z,
#   mixture = ~ time_years_z,
#   classmb = ~ 1,
#   ng = 3,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ genotype,
#   data = jlcmm_data_test,
#   B = m0,
#   maxiter = 500
# )
```

```{r}
# summary(m2)
```

### Try with gridsearch()

```{r}
# not converging -> local maxima -> try to find global maximum with gridsearch()

# gridsearch(rep = 20, maxiter = 20,
#            minit = jlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years_z,
#   random = ~ 1 + time_years_z,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ genotype,
#   data = jlcmm_data_test
# ),
#            m = jlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years_z,
#   random = ~ 1 + time_years_z,
#   mixture = ~ time_years_z,
#   classmb = ~ 1,
#   ng = 3,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ genotype,
#   data = jlcmm_data_test,
#   B = m0
# ))
```

## 2 latent classes + same covariates as in hlme()

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
m1_genotype_age_gender <- jlcmm(
    fixed = ln_PLASMA_NFL ~ time_years_z,
    random = ~ 1 + time_years_z,
    mixture = ~ time_years_z,
    classmb = ~ genotype + age_at_baseline + PTGENDER,
    survival = Surv(time_to_event, event) ~ genotype,
    subject = "RID",
    ng = 2,
    data = jlcmm_data_test,
    B = random(m0)
  )
```

```{r}
summary(m1_genotype_age_gender)
```

```{r}
postprob(m1_genotype_age_gender)
```

# JLCMM: PTAU

### M0 (1 latent class)

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

### M1 (2 latent classes)

```{r}

m1 <- jlcmm(
  fixed = ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ 1 + time_years_z,
  mixture = ~ time_years_z,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data_test,
  B = m0
)
```

```{r}
summary(m1)
```

```{r}
postprob(m1)
```

# ..........................................................................................................

# Testing Weibull and Cox

### CoxPH and Weibull regressions

-   Surv object, removing time_to_event = 0

-   Both model outputs show that the e3 homozyogotic genotype does not have a hazard significantly different from the baseline level, which is genotypes containing the e2 allele (whether that be heterozygotic or homozygotic). This gives more support to collapsing the genotype into three levels instead of four.

```{r}
surv_data_test <- surv_data %>% filter(time_to_event > 0)
s_data <- Surv(surv_data_test$time_to_event, surv_data_test$event)
```

```{r}
# fitting a coxph model with surv_data_test 
fit_cox <- coxph(s_data ~ genotype, data = surv_data_test)

# fitting a weibull regression model with surv_data_test 
fit_weib <- survreg(s_data ~ genotype, dist="weibull", data = surv_data_test)
```

```{r}
summary(fit_cox)
```

```{r}
summary(fit_weib)
```

### Kaplan-Meier and log-log survival curves

```{r}
par(mfrow = c(1, 2))

surv_fit <- (survfit(s_data ~ genotype, data = surv_data_test))

ggsurvfit(surv_fit) +
          labs(x = "Time-to-diagnosis",
               y = "log-log survival",
               title = "Kaplan-Meier curves by genotype") +
  theme_classic()

ggsurvfit(surv_fit, type = "cloglog") +
          labs(x = "Time-to-diagnosis",
               y = "log-log survival",
               title = "log-log survival curve") +
  theme_classic()
```

```{r}
# rcorr.cens(predict(s_data), surv_data_test)["C Index"]
# rcorr.cens(predict(s_data), surv_data_test)["C Index"]
```

### Assumption of proportional hazards, residuals

-   Testing the proportional hazard assumption, test statistic is not significant, so we can assume proportionality between hazards of the different genotypes

-   Schoenfield residuals should be flat and centered around 0 to assume proportionality

-   <https://bookdown.org/sestelo/sa_financial/how-to-evaluate-the-ph-assumption.html>

```{r}
cox.zph(fit_cox)

plot(cox.zph(fit_cox))
```

### Log-rank and generalized Wilcoxon tests

-   Log-rank and generalised Wilcoxon both have significant test statistics, therefore we can conclude that there is a significant difference in baseline hazard for the different levels of genotype

```{r}
# computing the log-rank test statistic for genotype
survdiff(s_data ~ surv_data_test$genotype, rho = 0)

# computing the generalised Wilcoxon test statistic for genotype
survdiff(s_data ~ surv_data_test$genotype, rho = 1)
```

## Surv_data: collapsing genotype down to 3 factors

```{r}
surv_data_test <- surv_data_test %>%
  mutate(
    genotype = case_when(
      genotype == "e2" ~ "e2 e3",
      genotype == "e3/e3" ~ "e2 e3",
      genotype == "e4" ~ "e4",
      genotype == "e4/e4" ~ "e4/e4"
    )
  )
```

### CoxPH and Weibull regressions with 3 genotype levels

```{r}
s_data <- Surv(surv_data_test$time_to_event, surv_data_test$event)
```

```{r}
# fitting a coxph model with time_to_event > 0 and 3 levels to genotype
fit_cox <- coxph(s_data ~ genotype, data = surv_data_test)

# fitting a weibull regression model with time_to_event > 0 and 3 levels to genotype
fit_weib <- survreg(s_data ~ genotype, dist="weibull", data = surv_data_test)
```

```{r}
summary(fit_cox)
```

```{r}
summary(fit_weib)
```

### Kaplan-Meier and log-log survival curves

```{r}
par(mfrow = c(1, 2))

surv_fit <- (survfit(s_data ~ genotype, data = surv_data_test))

ggsurvfit(surv_fit) +
          labs(x = "Time-to-diagnosis",
               y = "log-log survival",
               title = "Kaplan-Meier curves by genotype") +
  theme_classic()

ggsurvfit(surv_fit, type = "cloglog") +
          labs(x = "Time-to-diagnosis",
               y = "log-log survival",
               title = "log-log survival curve") +
  theme_classic()
```

### Assumption of proportional hazards, residuals

```{r}
cox.zph(fit_cox)

plot(cox.zph(fit_cox))
```

### Log-rank and generalized Wilcoxon tests

-   Log-rank and generalised Wilcoxon both have significant test statistics, therefore we can conclude that there is a significant difference in baseline hazard for the different levels of genotype

```{r}
# computing the log-rank test statistic for genotype
survdiff(s_data ~ surv_data_test$genotype, rho = 0)

# computing the generalised Wilcoxon test statistic for genotype
survdiff(s_data ~ surv_data_test$genotype, rho = 1)
```

# Testing different hazards

### Trying with splines hazard

-   ng = 1

```{r}
m0_splines <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  hazard = "splines",
  data = jlcmm_data_test,
)
```

```{r}
summary(m0_splines)
```

```{r}
plot(m0_splines)
```

-   ng = 2

```{r}
m1_splines <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  mixture = ~ time_years_z,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  hazard = "splines",
  data = jlcmm_data_test,
  B = m0_splines
)
```

```{r}
summary(m1_splines)
```

```{r}
plot(m1_splines)
```

# ..........................................................................................................

# Multivariate JLCMM (mpjlcmm)

-   Joint latent class mixed models for multivariate longitudinal markers

-   Defines each longitudinal dimension as a latent process

-   Latent processes assumed to be independent given latent class

-   Survival handles right censoring and LEFT TRUNCATION (should we consider left truncation, how is this defined, does it apply to us?)

### Fitting multlcmm: M0 (1 latent class)

```{r}
mlcmm_m0 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  data = jlcmm_data_test
)
```

### Fitting multlcmm: M1 (2 latent class)

```{r}
mlcmm_m1 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  mixture = ~ time_years_z,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data_test,
  B = random(mlcmm_m0)
)
```

### M0 (1 latent class)

-   $AIC = 8769$

-   $BIC = 8829$

-   $log-likelihood = -4374$

```{r}
mult_jlcmm_m0 <- mpjlcmm(
  longitudinal = list(mlcmm_m0),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data_test
)
```

```{r}
summary(mult_jlcmm_m0)
```

### M1 (2 latent classes)

-   $AIC = 8670$

-   $BIC = 8743$

-   $log-likelihood = -4320$

-   Model with 2 latent classes, NO class membership covariates, NO survival covariates

-   All coefficients significant **(why is intercept not estimated for class1????)**

-   $AIC$ and $BIC$ both decrease, and $log-likelihood$ increases

```{r}
mult_jlcmm_m1 <- mpjlcmm(
  longitudinal = list(mlcmm_m1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 2,
  data = jlcmm_data_test,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m1)
```

### Posterior probabilities

-   Good classification

```{r}
postprob(mult_jlcmm_m1)
```

# ...................................................................................................

# MULTISTATE multivariate JLCMM (mpjlcmm)

## Data

-   Define current state, next state, transition (current state -\> next state),

-   In order to work with the diagnosis data and the biomarker data, the diagnosis data before the first biomarker data need to be removed

```{r}
# filtering the observations out of the diagnosis data set where EXAMDATE is much before baseline biomarker EXAMDATE
cognitive_states <- diagnosis_data %>%
  merge(
    nfl_ptau %>% select(RID, baseline_date), by = "RID") %>%
  filter(EXAMDATE - baseline_date >= -50) %>%
  distinct()
```

-   In order to correctly merge the biomarker and event data, dx data with a later VISCODE than biomarker data will be given the same VISCDE as the last biomarker VISCODE

-   So they can be compare with a logical expression, both are transformed to a numerical class

First, for the bimarker data

```{r}
# creating a VISCODE_last variable in order to merge later diagnosis data with the last biomaker measurements, stripping "m" from the start of VISCODE2 and changing to numeric class
nfl_ptau <- nfl_ptau %>% 
  group_by(RID) %>%
  mutate(
    VISCODE2 = as.numeric(gsub("\\D", "", VISCODE2)),
    VISCODE_last = last(VISCODE2)
  )
```

Then, for the diagnosis data

-   Additionally, a stat variable is created, the biomarker and dx data are merged and VISCODE2 for dx data is changed to VISCODE_last if a later VISCODE than the final biomarker data

```{r}
# creating a new data frame for multistate survial data
# state variable
# transforming VISCODE2 into a numeric variable, merging with VISCODE_last from nfl_ptau data, changing VISCODE2 to VISCODE_last if VISCODE2 > VISCODE_last
cognitive_states <- cognitive_states %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    state_current = case_when(
      DIAGNOSIS == "CN" ~ 1,
      DIAGNOSIS == "MCI" ~ 2,
      DIAGNOSIS == "Dementia" ~ 3
    ),
    VISCODE2 = as.numeric(gsub("\\D", "", VISCODE2))
  ) %>%
  merge(
    nfl_ptau %>% select(RID, VISCODE_last)
    ) %>%
  mutate(
    VISCODE2 = ifelse(VISCODE2 > VISCODE_last, VISCODE_last, VISCODE2)
  ) %>%
  distinct()
```

-   Time from current EXAMDATE and baseline EXAMDATE, time from EXAMDATE_next and baseline EXAMDATE are created

-   Transition variable created: state_current -\> state_next

```{r}
episodes <- cognitive_states %>%
  group_by(RID) %>%
  mutate(
    state_next = lead(state_current),
    EXAMDATE_next = lead(EXAMDATE)
  ) %>%
  filter(!is.na(state_next)) %>%
  mutate(
    time_start = as.numeric(EXAMDATE - first(EXAMDATE)) / 365.25,
    time_stop = as.numeric(EXAMDATE_next - first(EXAMDATE)) / 365.25,
    transition = paste0(state_current, "->", state_next)
  ) %>%
  ungroup()
```

# HERE

Defining the transitions we want to model, where:

-   1 -\> 2 : CN -\> MCI

-   1 -\> 3 : CN -\> Dementia

-   2 -\> 3: MCI -\> Demenetia

-   2 -\> 1 : MCI -\> CN

Filtering by the transitions of interest or a valid at risk state.

```{r}
# defining the transitions we want to account for
transitions_of_interest <- c("1->2", "1->3", "2->3", "2->1")  # CN->MCI, CN->AD, MCI->AD, MCI->CN

# all valid "at risk" states for each transition
at_risk_state <- c("1->2" = 1, "2->3" = 2, "1->3" = 1, "2->1" = 2)

episodes_expanded <- episodes %>%
  filter(transition %in% transitions_of_interest | 
         state_current %in% at_risk_state) %>%
  # For each interval, create one row per transition the person is at risk for
  rowwise() %>%
  mutate(
    at_risk_for = list(names(at_risk_state)[at_risk_state == state_current])
  ) %>%
  unnest(at_risk_for) %>%
  mutate(
    event = as.integer(transition == at_risk_for),
    trans = at_risk_for
  ) %>%
  select(RID, EXAMDATE, VISCODE2, EXAMDATE_next, baseline_date, time_start, time_stop,
         trans, event, state_current)
```

-   TRansforming the event variable to a multistate event variable

```{r}
# creating a multistate event variable
episodes_expanded <- episodes_expanded %>%
  mutate(
    event_multi = case_when(
      event == 1 & trans == "1->2" ~ 1,
      event == 1 & trans == "1->3" ~ 2,
      event == 1 & trans == "2->3" ~ 3,
      event == 1 & trans == "2->1" ~ 4,
      event == 0 ~ 0),
    event_multi = factor(event_multi, 
                         levels = c(0, 1, 2, 3, 4),
                         labels = c("censored", "CN->MCI", "CN->AD", "MCI->AD", "MCI->CN")))
```

```{r}
# filtering out obserations where there is no transition (last observation)
episodes_expanded <- episodes_expanded %>%
  filter(EXAMDATE_next > EXAMDATE)
```

```{r}
# selecting necessary variables
episodes_expanded <- episodes_expanded %>%
  select(RID, VISCODE2, time_start, time_stop, trans, event, event_multi, state_current)
```

```{r}
# merging biomarker, demographic, genotype data with event data
ms_joint_data <- nfl_ptau %>%
  select(
    RID, VISCODE2, PLASMAPTAU181, PLASMA_NFL, PTDOBYY, PTGENDER, age_at_baseline
    ) %>%
  merge(episodes_expanded, by = c("RID","VISCODE2")) %>%
  merge(apoe, by = "RID") %>%
  distinct()
```

```{r}
class(ms_joint_data$event_multi)
levels(ms_joint_data$event_multi)
```

```{r}
ms_joint_data %>%
  group_by(RID, trans) %>%
  arrange(time_start) %>%
  mutate(gap = time_start - lag(time_stop)) %>%
  filter(!is.na(gap) & gap < 0)
```

```{r}
ms_joint_data <- ms_joint_data %>% filter(RID != 4499)
```

```{r}
# Verify after merge
class(ms_joint_data$event_multi)
levels(ms_joint_data$event_multi)
table(ms_joint_data$event_multi)
```

```{r}
summary(coxph(
  Surv(time_start, time_stop, event_multi) ~
    age_at_baseline + genotype + strata(trans),
  data = ms_joint_data,
  id = interaction(RID, trans),
  cluster = RID))
```

```{r}
ms_joint_data %>%
  filter(trans == "2->3") %>%
  group_by(genotype, event) %>%
  summarise(n = n())
```

```{r}
ms_joint_data %>%
  group_by(RID, trans) %>%
  arrange(time_start) %>%
  mutate(overlap = time_start < lag(time_stop)) %>%
  filter(overlap == TRUE) %>%
  select(RID, trans, time_start, time_stop)
```

```{r}
ms_joint_data <- nfl_ptau %>%
  left_join(episodes_expanded, by = "RID","VISCODE2")
# %>%
#   filter(time_years >= time_start & time_years < time_stop)
```

```{r}
# defining the transitions we want to account for
# CN - > MCI    MCI -> AD   CN -> AD    MCI -> CN
valid_transitions <- c("1->2", "2->3", "1->3", "2->1")

# lead() to compare with NEXT values, lag() to compare with PREVIOUS values
# creating a new varible for the next dx in order to assign transition 
ms_surv_data <- ms_surv_data %>%
  group_by(RID) %>%
  mutate(
    state_next = lead(state_current),
    EXAMDATE_next = lead(EXAMDATE)
  ) %>%
  filter(!is.na(state_next)) %>%
  ungroup()

# creating a transition variable: current state -> next state
ms_surv_data <- ms_surv_data %>%
  mutate(
    transition = paste0(state_current, "->", state_next)
    ) %>%
  filter(transition %in% valid_transitions) %>%
  mutate(
    time_to_event = as.numeric(EXAMDATE_next - EXAMDATE) / 365.25)
```

```{r}
# checking number of individuals after removing not allowed states: n = 706
length(unique(ms_surv_data$RID))
```

```{r}
# checking number of transitions for each transition state: n = 11 for CN -> AD
table(ms_surv_data$transition)
```

### Wide survival data

-   Needed to merge longitudinal data with survival data

-   Gives unique rows for the longitudinal data with repeated rows for the survival data

-   Looking at the data sets in JMBAYES, this is fine

```{r}

wide_ms_surv_data <- ms_surv_data %>%
  group_by(RID) %>%
  summarise(
# Time to first MCI (from CN)
    time_CN_MCI = min(time_to_event[transition == "1->2"], na.rm = TRUE),
    event_CN_MCI = any(transition == "1->2"),
    
# Time to AD (from MCI)
    time_MCI_AD = min(time_to_event[transition == "2->3"], na.rm = TRUE),
    event_MCI_AD = any(transition == "2->3"),
    
# Time to AD direct (from CN)
    time_CN_AD = min(time_to_event[transition == "1->3"], na.rm = TRUE),
    event_CN_AD = any(transition == "1->3"),
    
# Backward transition MCI -> CN
    time_MCI_CN = min(time_to_event[transition == "2->1"], na.rm = TRUE),
    event_MCI_CN = any(transition == "2->1")
  ) %>%

# Replace Inf (from min of empty set) with NA
# Coerce event flags to 0/1
  mutate(
    across(starts_with("time_"), ~ ifelse(is.infinite(.x), NA, .x)),
    across(starts_with("event_"), ~ as.integer(.x)))

```

```{r}
ms_surv_data_test <- wide_ms_surv_data %>% merge(nfl_ptau, by = "RID")
```

```{r}
# creating a data frame with diagnosis times: CN -> MCI, MCI -> AD
dx_times <- diagnosis_data %>%
  arrange(RID, EXAMDATE) %>%
  group_by(RID) %>%
  mutate(
# creating a variable for the time from first MCI dx to every CN dx
    t_CN_MCI = ifelse(DIAGNOSIS == "CN",
                      as.numeric(first(EXAMDATE[DIAGNOSIS == "MCI"]) - EXAMDATE) / 365.25,
                      NA),
# creating a variable for the time from first AD dx to every MCI dx
    t_MCI_AD = ifelse(DIAGNOSIS == "MCI",
                      as.numeric(first(EXAMDATE[DIAGNOSIS == "Dementia"]) - EXAMDATE) / 365.25,
                      NA),

    t_CN_AD = ifelse(DIAGNOSIS == "CN",
                      as.numeric(first(EXAMDATE[DIAGNOSIS == "Dementia"]) - EXAMDATE) / 365.25,
                      NA),
# 
  ) %>% ungroup()

```

-   Creating a variable which tells us if the MCI dx reverts back to CN, and same from AD to MCI

```{r}
# adding a variable to the dx times data frame which tells us whether an individual goes "backwards" from MCI -> CN or from AD -> MCI
dx_times <- dx_times %>%
  arrange(RID, t_CN_MCI) %>%
  group_by(RID) %>% 
  mutate(backwards_MCI = case_when(
    (any(DIAGNOSIS == "CN") & any(DIAGNOSIS == "MCI")) == TRUE ~
      # (ifelse(is.na(t_CN_MCI) == TRUE,
      #         NA,
              first(t_CN_MCI) < 0),
    
    backwards_AD = ifelse(is.na(t_MCI_AD) == TRUE,
                          NA,
                          first(t_MCI_AD) < 0))

# number of individuals which move backwards from MCI -> CN
sum(na.omit((dx_times %>% select(backwards_MCI, RID) %>% distinct())$backwards_MCI))

# number of individuals which move backwards from MCI -> CN = 0
sum(na.omit((dx_times %>% select(backwards_AD, RID) %>% distinct())$backwards_AD))
```

```{r}
# removing the backwards_AD variable as there are 0 cases
dx_times <- dx_times %>% select(!backwards_AD)
```

```{r}
# creating a shorter version of dx_times where the t_MCI takes the first MCI dx which does not revert to CN dx
dx_times_short <- dx_times %>%
  arrange(RID, EXAMDATE) %>%
  group_by(RID) %>%
  summarise(
    t_CN = first(EXAMDATE[DIAGNOSIS == "CN"]),
    
    t_MCI = case_when(
      
# THE PROBLEM ARISES HERE, asssumes that there is only one MCI dx before reversion back to CN
      
# for cases where the individual converts back to CN, we remove the first EXAMDATE were dx = MCI and take the next earliest EXAMDATE
      unique(backwards_MCI) == TRUE ~ as.Date(0),
# otherwise, we take the first EXAMDATE where dx = MCI
      TRUE ~ first(EXAMDATE[DIAGNOSIS == "MCI"])
      ),

# there are no reversions from AD dx, so we just take the first AD dx date
    t_AD = first(EXAMDATE[DIAGNOSIS == "Dementia"]),
    censor_time = max(EXAMDATE)
  ) %>% ungroup()
```

```{r}
dx_times %>% filter(backwards_MCI == TRUE) %>% arrange(RID, EXAMDATE)
```

## Multistate data

Multistates

1.  CN

2.  MCI

3.  AD

```{r}
multistate_data <- dx_times_short %>%
  mutate(
# CN to MCI event variable 
    CN_MCI_event = ifelse(!is.na(t_CN) & !is.na(t_MCI), 1, 0),
# survival time for CN to MCI event
    t_CN_MCI = case_when(
      CN_MCI_event == 1 ~ as.numeric(t_MCI - t_CN) / 365.25,
      CN_MCI_event == 0 ~ as.numeric(censor_time - t_CN) / 365.25
    ),
# MCI to AD event variable
    MCI_AD_event = case_when(
      !is.na(t_MCI) & !is.na(t_AD) ~ 1,
      !is.na(t_MCI) &  is.na(t_AD) ~ 0,
      TRUE ~ NA
    ),
# survival time for MCI to AD event 
    t_MCI_AD = case_when(
      MCI_AD_event == 1 ~ as.numeric(t_AD - t_MCI) / 365.25,
      MCI_AD_event == 0 ~ as.numeric(censor_time - t_MCI) / 365.25,
      TRUE ~ NA
    )
  ) %>% relocate(MCI_AD_event, .after = CN_MCI_event)
```

```{r}
# recoding the event and time_to_event variables for multistate jlcmm
multistate_data <- multistate_data %>%
  mutate(
    ms_event = case_when(
      (CN_MCI_event == 0 & is.na(t_CN_MCI) == FALSE) ~ 1,
      (CN_MCI_event == 1 & MCI_AD_event == 0) ~ 2,
      (is.na(t_CN_MCI) == TRUE & is.na(t_MCI_AD) == FALSE) ~ 2,
      (CN_MCI_event == 1 & MCI_AD_event == 1) ~ 3,
      (is.na(t_CN_MCI) == TRUE & is.na(t_MCI_AD) == TRUE) ~ 3),
    ms_time_to_event = case_when(
      event == 1 ~ t_CN_MCI,
      event == 2 ~ ifelse(!is.na(t_CN_MCI), t_CN_MCI, t_MCI_AD),
      event == 3 ~ ifelse(!is.na(t_MCI_AD), t_CN_MCI, 0),
    ))
```

```{r}
# merging the multistate event and time to event data with the jlcmm data set
multistate_data_test <- multistate_data %>%
  select(RID, ms_event, ms_time_to_event) %>%
  merge(jlcmm_data_test, by = "RID")
```

-   There are negative time to event, the mpjlcmm does not converge

-   66 unique RID with negative time to event: check if these are MCI -\> CN reversions

-   **PROBLEM:** Code which checks for MCI -\> CN reversions assumes the next MCI event is AFTER the CN reversion

-   **SOLUTION:**

```{r}
# checking which individuals have negative time to event
CN_reversions <- unique((multistate_data_test %>% filter(ms_time_to_event < 0))$RID)

negative_event_times <- diagnosis_data %>%
  arrange(RID, EXAMDATE) %>%
  filter(RID %in% CN_reversions) %>%
  group_by(RID) %>%
  mutate(
    no_reversion =
      case_when(
        (DIAGNOSIS[EXAMDATE == min(EXAMDATE)] == "MCI" &
           DIAGNOSIS[EXAMDATE == max(EXAMDATE)] == "CN") ~ FALSE,
        TRUE ~ TRUE)) %>%
  merge(multistate_data %>% select(RID, ms_time_to_event, ms_event), by = "RID")
```

```{r}
# final version
multistate_data <- dx_times_short %>%
  mutate(
# CN to MCI event variable 
    CN_MCI_event = ifelse(!is.na(t_CN) & !is.na(t_MCI), 1, 0),
# survival time for CN to MCI event
    t_CN_MCI = case_when(
      CN_MCI_event == 1 ~ as.numeric(t_MCI - t_CN) / 365.25,
      CN_MCI_event == 0 ~ as.numeric(censor_time - t_CN) / 365.25
    ),
# MCI to AD event variable
    MCI_AD_event = case_when(
      !is.na(t_MCI) & !is.na(t_AD) ~ 1,
      !is.na(t_MCI) &  is.na(t_AD) ~ 0,
      TRUE ~ NA
    ),
# survival time for MCI to AD event 
    t_MCI_AD = case_when(
      MCI_AD_event == 1 ~ as.numeric(t_AD - t_MCI) / 365.25,
      MCI_AD_event == 0 ~ as.numeric(censor_time - t_MCI) / 365.25,
      TRUE ~ NA
    )
  ) %>% relocate(MCI_AD_event, .after = CN_MCI_event)
```

## JLCMM DATA

-   same number of rows - good

```{r}
jlcmm_data_test_multi <- jlcmm_data_test %>%
  merge(
    multistate_data %>%
      select(RID, censor_time, CN_MCI_event, MCI_AD_event, t_CN_MCI, t_MCI_AD))
```

```{r}
?mpjlcmm
```

# ...................................................................................................

# MULTISTATE model fitting

### Fitting multlcmm: M0 (1 latent class)

-   multivariate latent class mixed model with 1 class

```{r}
mlcmm_m0 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  data = df_test
)
```

```{r}
summary(mlcmm_m0)
```

### M0 (1 latent class) mpjlcmm

-   multistate multivariate joint latent class mixed model with 1 class

-   infinite parameters

```{r}
ms_mjlcmm_m0 <- mpjlcmm(
  longitudinal = list(ms_mlcmm_m0),
  subject = "RID",
  survival = time_event + state ~ 1,
  data = df_test,
  ng = 1
)
```

#### Checking time_to_event variable

```{r}
sum(df_test$time_event == 0)
range(df_test$time_event)
```

### Fitting multlcmm: M1 (2 latent classes)

-   multistate multivariate latent class mixed model

```{r}
mlcmm_m1 <- multlcmm(
  fixed = ln_PLASMA_NFL + ln_PLASMAPTAU181 ~ time_years_z,
  random = ~ time_years_z,
  mixture = ~ time_years_z,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = df_test,
  B = random(mlcmm_m0)
)
```

```{r}
summary(mlcmm_m1)
```

### M1 (2 latent classes)

-   multistate multivariate joint latent class mixed model

```{r}
ms_mjlcmm_m1 <- mpjlcmm(
  longitudinal = list(mlcmm_m1),
  subject = "RID",
  survival = time_event + state ~ 1,
  ng = 2,
  data = df_test,
  B = random(ms_mjlcmm_m0)
)
```

-   Remove NA values

```{r}
jlcmm_data_test_multi_short <- na.omit(jlcmm_data_test_multi)
```

```{r}
is.numeric(jlcmm_data_test_multi_short$t_CN_MCI)
is.numeric(jlcmm_data_test_multi_short$t_MCI_AD)
```

```{r}
# m0 <- jlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years_z,
#   random = ~ 1 + time_years_z,
#   subject = "RID",
#   survival = Surv(time = t_CN_MCI, event = CN_MCI_event) +
#     Surv(time = t_MCI_AD, event = MCI_AD_event) ~ genotype,
#   data = jlcmm_data_test_multi
# )
```

```{r}
# hlme_m1 <- hlme(
#     fixed = ln_PLASMA_NFL ~ time_years_z,
#     mixture = ~ time_years_z,
#     random = ~ time_years_z,
#     classmb = ~ 1,
#     subject = "RID",
#     ng = 2,
#     data = jlcmm_data_test,
#     B = random(hlme_m0)
#   )
```

```{r}
# model_multistate <- Jointlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years_z,
#   random = ~ time_years_z,
#   mixture = ~ time_years_z,
#   classmb = ~ 1,
#   ng = 2,
#   subject = "RID",
#   
#   survival = Surv(time1, event1) + Surv(time2, event2),
#   hazard = c("Weibull", "Weibull"),
#   
#   data = jlcmm_data_test_multi,
#   B = hlme_m1
# )
```
