---
title: "Joint Latent Class Mixed Models / Multistate Survival"
format: pdf
editor: visual
---

```{r}
# loading the packages
library(lcmm)
library(dplyr)
library(survival)
library(ADNIMERGE2)
```

```{r}
# loading data
# NfL
nfl_data <- BLENNOWPLASMANFLLONG

# pTAU
ptau_data <- UGOTPTAU181

# dx data
diagnosis_data <- DXSUM

# demographic data
demographics <- PTDEMOG

# genetic data
apoe <- APOERES
```

-   Check how we merge with diagnosis data: we have more data than is present here. For example RID 2 changes from CN to MCI to CN again, but with the full diagnosis data we see that they convert again to MCI. Let's think about how to include this data correctly. Should we make a different data set with conversions from the full diagnosis data set?

```{r}
# selecting relevant variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# adding a suffix to the dx data
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# merging the data sets
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))

# merging only nfl and ptau
nfl_ptau <- nfl_data %>%
  merge(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))
```

```{r}
# removing missing values
nfl_ptau <- na.omit(nfl_ptau)
diagnosis_data <- na.omit(diagnosis_data)
```

## Checking RID with more than one measurement recorded on the same exam date

-   There are 52 individuals with two measurements recorded on the same exam date.

-   How should we deal with these? Average the measurements? Remove the observations?

-   Some of the values are quite different, what do we assume by averaging them? What effect does it have to remove them?

-   For now, remove them:

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
# ptau_data %>%
#   filter(RID %in% ptau_double$RID) %>% 
#   filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

## Baseline date in nfl_ptau

```{r}
# checking which observations have the same examdate for nfl and ptau
common <- which(nfl_ptau$EXAMDATE_nfl %in% nfl_ptau$EXAMDATE_ptau)

# checking the observations with different exam dates
nfl_ptau[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
nfl_ptau <- nfl_ptau %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
  ungroup()
```

## Time since baseline in nfl ptau

```{r}
# creating a time since baseline visit variable (in years)
nfl_ptau <- nfl_ptau %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## Selected demographics in nfl_ptau

-   Merging DOB, gender and creating an age at baseline variable

-   Converting gender to factor variable

```{r}
# selecting relevant variables (age, gender) and keeping only one row per person
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# merging the demographic data with the biomarker data
nfl_ptau <- nfl_ptau %>% merge(demographics_small, by = "RID")
```

```{r}
# converting the gender variable to a factor variable 
nfl_ptau <- nfl_ptau %>% mutate(PTGENDER = as.factor(PTGENDER))

# creating an age at baseline variable
nfl_ptau <- nfl_ptau %>% mutate(
  age_at_baseline = as.numeric(format(baseline_date, "%Y")) - PTDOBYY)
```

## Factor variable for genotype in apoe

-   Trying with a factor for each specific genotype

-   If there is not good distinction between these, since we do not have many individuals with some genotypes, we can collpase down to fewer factors later

```{r}
# creating a factor variable for ApoE genotype and selecting relevant variables
apoe = apoe %>% mutate(
  genotype = as.factor(case_when(
    GENOTYPE == "4/4" ~ "e4/e4",
    GENOTYPE %in% c("3/4", "2/4") ~ "e4",
    GENOTYPE == "3/3" ~ "e3/e3",
    GENOTYPE %in% c("2/3", "2/2") ~ "e2"
  ))
) %>%
  select(RID, genotype)
```

## We filter participants with at least 3 measurements in nfl_ptau

```{r}
# Filtering individuals with at least 3 measurements
keep <- unique((nfl_ptau %>% group_by(RID) %>% filter(n() > 2))$RID)
nfl_ptau <- nfl_ptau %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

## Merging biomarker data with diagnosis data by RID but not by VISCODE

-   Keeps all diagnosis data

-   We can then use diagnoses which happen not on the same VISCODE as the biomarker data

-   Check VISCODE2 variable names -\> change to \_bio and \_dx

```{r}
merged_data <- nfl_ptau %>% merge(diagnosis_data_dx, by = "RID") 
```

```{r}
merged_data <- na.omit(merged_data)
```

```{r}
merged_data <- merged_data %>%
  mutate(
    Tevent = as.numeric(EXAMDATE_dx - baseline_date) / 365.25
  )
```

# Multi-state Survival Data

```{r}
# old code
# dx_times <- merged_data %>%
#   arrange(RID, EXAMDATE_dx) %>%
#   group_by(RID) %>%
#   summarise(
#     t_CN  = first(EXAMDATE_dx[DIAGNOSIS_dx == "CN"]),
#     t_MCI = first(EXAMDATE_dx[DIAGNOSIS_dx == "MCI"]),
#     t_AD  = first(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]),
#     censor_time = max(EXAMDATE_dx)
#   ) %>% ungroup()
```

```{r}
# creating a data frame with diagnosis times: CN -> MCI, MCI -> AD
dx_times <- diagnosis_data %>%
  arrange(RID, EXAMDATE) %>%
  group_by(RID) %>%
  mutate(
# creating a variable for the time from first MCI dx to every CN dx
    t_CN_MCI = ifelse(DIAGNOSIS == "CN",
                      as.numeric(first(EXAMDATE[DIAGNOSIS == "MCI"]) - EXAMDATE) / 365.25,
                      NA),
# creating a variable for the time from first AD dx to every MCI dx
    t_MCI_AD = ifelse(DIAGNOSIS == "MCI",
                      as.numeric(first(EXAMDATE[DIAGNOSIS == "Dementia"]) - EXAMDATE) / 365.25,
                      NA)
# 
  ) %>% ungroup()
```

```{r}
# adding a variable to the dx times data frame which tells us whether an individual goes "backwards" from MCI -> CN or from AD -> MCI
dx_times <- dx_times %>%
  arrange(RID, t_CN_MCI) %>%
  group_by(RID) %>% 
  mutate(backwards_MCI = case_when(
    (any(DIAGNOSIS == "CN") & any(DIAGNOSIS == "MCI")) == TRUE ~
      # (ifelse(is.na(t_CN_MCI) == TRUE,
      #         NA,
              first(t_CN_MCI) < 0),
    
    backwards_AD = ifelse(is.na(t_MCI_AD) == TRUE,
                          NA,
                          first(t_MCI_AD) < 0))

# number of individuals which move backwards from MCI -> CN
sum(na.omit((dx_times %>% select(backwards_MCI, RID) %>% distinct())$backwards_MCI))

# number of individuals which move backwards from MCI -> CN = 0
sum(na.omit((dx_times %>% select(backwards_AD, RID) %>% distinct())$backwards_AD))

# removing the backwards_AD variable as 
dx_times <- dx_times %>% select(!backwards_AD)
```

```{r}
# creating a shorter version of dx_times where the t_MCI takes the first MCI dx which does not revert
dx_times_short <- dx_times %>%
  arrange(RID, EXAMDATE) %>%
  group_by(RID) %>%
  summarise(
    t_CN  = first(EXAMDATE[DIAGNOSIS == "CN"]),
    
    t_MCI = case_when(
# for cases where the individual converts back to CN, we remove the first EXAMDATE were dx = MCI and take the next earliest EXAMDATE 
      unique(backwards_MCI) == TRUE ~ first(EXAMDATE[DIAGNOSIS == "MCI"][-1]),
# otherwise, we take the first EXAMDATE where dx = MCI
      TRUE ~ first(EXAMDATE[DIAGNOSIS == "MCI"])
      ),

# there are no reversions from AD dx, so we just take the first AD dx date
    t_AD  = first(EXAMDATE[DIAGNOSIS == "Dementia"]),
    censor_time = max(EXAMDATE)
  ) %>% ungroup()
```

### Transitions times - maybe not needed since we can't use multistate surv in jlcmm

```{r}
# times for transitions from CN -> MCI
# CN_MCI <- dx_times_short %>% filter(is.na(t_CN) == FALSE) %>%
#   mutate(
#     from = "CN",
#     to = "MCI",
#     Tstart = 0,
#     Tstop = as.numeric(case_when(
#       is.na(t_MCI) == TRUE ~ censor_time - t_CN,
#       TRUE ~ t_MCI - t_CN) / 365.25),
#     event = case_when(
#       is.na(t_MCI) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of MCI -> CN
#   filter(
#     Tstop > 0
#   ) %>%
#   select(RID, from, to, Tstart, Tstop, event)
```

```{r}
# times for transitions from MCI -> AD
# MCI_AD <- dx_times_short %>% filter(is.na(t_MCI) == FALSE) %>%
#   mutate(
#     from = "MCI",
#     to = "AD",
#     Tstart = case_when(
#       is.na(t_CN) == FALSE ~ as.numeric(t_MCI - t_CN) / 365.25,
#       TRUE ~ 0),
#     Tstop = as.numeric(case_when(
#       is.na(t_AD) == TRUE ~ censor_time - t_CN,
#       TRUE ~ t_AD - t_CN) / 365.25),
#     event = case_when(
#       is.na(t_AD) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of AD -> MCI
#   filter(
#     Tstart >= 0,
#     Tstop > 0
#   ) %>%
#   select(RID, from, to, Tstart, Tstop, event)
```

```{r}
# times for transitions from CN -> AD
# CN_AD <- dx_times_short %>% filter(is.na(t_CN) == FALSE, is.na(t_AD) == FALSE, is.na(t_MCI) == TRUE) %>%
#   mutate(
#     from = "CN",
#     to = "AD",
#     Tstart = 0,
#     Tstop = as.numeric(case_when(
#       is.na(t_AD) == TRUE ~ censor_time - t_CN,
#       TRUE ~ t_AD - t_CN) / 365.25),
#     event = case_when(
#       is.na(t_AD) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of AD -> CN
#   filter(
#     Tstart >= 0,
#     Tstop > 0
#   ) %>%
#   select(RID, from, to, Tstart, Tstop, event)
```

```{r}
# # times for transitions from CN -> AD or MCI -> AD
# AD_transitions <- dx_times_short %>%
#   mutate(
#     Tstart = 0,
#     Tstop = as.numeric(case_when(
#       is.na(t_AD) == TRUE ~ ifelse(is.na(t_CN) == TRUE,
#                                    censor_time - t_MCI,
#                                    censor_time - t_CN),
#       TRUE ~ ifelse(is.na(t_CN) == TRUE,
#                                    t_AD - t_MCI,
#                                    t_AD - t_CN)) / 365.25),
#     event = case_when(
#       is.na(t_AD) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of AD -> CN/MCI
#   filter(
#     Tstop > 0
#   ) %>%
#   select(RID, Tstart, Tstop, event)
```

```{r}
# multistate_data <- bind_rows(CN_MCI, MCI_AD, CN_AD) %>%
#   arrange(RID, Tstart) %>%
#   filter(Tstop >= Tstart)
```

```{r}
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarise(
    age_at_baseline = unique(age_at_baseline),
    PTGENDER = unique(PTGENDER),
    time_to_event = case_when(
      any(DIAGNOSIS_dx == "Dementia") ~
        as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),
      
      TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,
    
    event = as.integer(any(DIAGNOSIS_dx == "Dementia"))) %>%
  ungroup()

# surv_data = merged_data %>% group_by(RID) %>%
#   summarise(
#     age_at_baseline = unique(age_at_baseline),
#     PTGENDER = unique(PTGENDER),
# 
#     time_to_dementia = case_when(
# 
#     any(DIAGNOSIS_dx == "Dementia") ~
#       as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) -
#                    unique(baseline_date)),
# 
#     TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
#                     unique(baseline_date)) / 365),
# 
#     time_to_MCI = case_when(
# 
#     any(DIAGNOSIS_dx == "MCI") ~
#       as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "MCI"]) -
#                    unique(baseline_date)),
# 
#     TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
#                     unique(baseline_date)) / 365),
# 
#     last_dx = DIAGNOSIS_dx[which.max(EXAMDATE_dx)],
# 
#     event = case_when(
#       last_dx == "CN" ~ 0,
#       last_dx == "MCI" ~ 1,
#       last_dx == "Dementia" ~ 2,
#     ))
```

```{r}
# selecting the relevant variables for the survial model
# biomarkers <- nfl_ptau %>%
#   select(RID, time_years, PLASMA_NFL, PLASMAPTAU181)
# 
# # multistate data with the most recent biomarker value added
# multistate_data_biomarkers <- multistate_data %>%
#   left_join(biomarkers, by = "RID") %>%
#   group_by(RID, Tstart, Tstop, from, to, event) %>%
#   # keep only biomarker values measured before interval end
#   filter(time_years <= Tstop) %>%
#   # pick the most recent biomarker
#   slice_max(time_years, n = 1, with_ties = FALSE) %>%
#   ungroup()
```

# Multi-state Survival Model

```{r}
# fit <- coxph(
#   Surv(Tstart, Tstop, event) ~ log(PLASMA_NFL) + log(PLASMAPTAU181) + strata(from, to),
#   data = multistate_data_biomarkers
# )
# 
# summary(fit)
```

# Survival Model

-   merge event data to nfl_ptau

-   event

-   event time

```{r}
# jlcmm_data = nfl_ptau %>% merge(AD_transitions, by = "RID")
```

```{r}
nfl_ptau <- nfl_ptau %>% arrange(RID, time_years)
```

```{r}
# surv_data <- merged_data %>%
#   group_by(RID) %>%
#   summarise(
#     Tevent = ifelse(any(DIAGNOSIS_dx == "Dementia"),
#                     min(time_years[DIAGNOSIS_dx=="Dementia"]),
#                     max(time_years)),
#     event = ifelse(any(DIAGNOSIS_dx == "Dementia"), 1, 0)
#   )
```

```{r}
# Jointlcmm(
#   fixed = PLASMA_NFL ~ time_years,
#   random = ~ RID,
#   mixture = ~ time_years,
#   classmb = ~ 1,
#   subject = "RID",
#   survival = Surv(Tstop, event) ~ age_at_baseline,
#   data = jlcmm_data,
#   ng = 3
# )
```

## Creating a single EXAMDATE variable

```{r}
library(lubridate)
nfl_ptau <- nfl_ptau %>%
  mutate(EXAMDATE = EXAMDATE_nfl + (EXAMDATE_nfl - EXAMDATE_ptau)) %>%
  select(!c(EXAMDATE_nfl, EXAMDATE_ptau)) %>%
  relocate(c(baseline_date, EXAMDATE, time_years), .after = VISCODE2)
```

## Survival data (CN/MCI -\> AD only)

```{r}
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)

surv_data <- diagnosis_data %>%
  group_by(RID) %>%
  
  # We create baseline date per patient (whichever EXAMDATE comes first, nfl or ptau)
  mutate(baseline_date = min(EXAMDATE),
  
  # We calculate event and time_to_event based on all diagnosis data
    # 1 if patient ever developed dementia
  event = as.integer(any(DIAGNOSIS == "Dementia")),
  time_to_event = case_when(
    event == 1 ~
      as.numeric(min(EXAMDATE[DIAGNOSIS == "Dementia"]) - baseline_date)/365,
    TRUE ~
      as.numeric(max(EXAMDATE) - baseline_date)/365)) %>%
  
    # We calculate age at baseline
  # age_at_baseline = as.numeric(format(unique(baseline_date), "%Y")) - PTDOBYY) %>%
      ungroup() %>%
  select(RID, event, time_to_event) %>%
  distinct()
  
surv_data <- surv_data %>% filter(RID %in% keep)

  # # We keep only one baseline row per patient
  # slice_min(order_by = EXAMDATE_ptau, n = 1, with_ties = FALSE)
    
  # We remove grouping

# We check the number of events 
table(surv_data$event)
```

## Joint Latent Class Mixed Models DATA

```{r}
# reference
# lcmm package data - example that works with the jlcmm
d = data_lcmm
```

```{r}
# merging longitudinal and survival data
jlcmm_data <- nfl_ptau %>%
  merge(surv_data, by = "RID") %>%
  relocate(c(event, time_to_event), .after = time_years)
```

```{r}
# transforming the data
jlcmm_data <- jlcmm_data %>%
  mutate(ln_PLASMA_NFL = log(PLASMA_NFL))
```

```{r}
# merging genetic data
jlcmm_data <- jlcmm_data %>%
  merge(apoe, by = "RID")
```

# JLCMM fitting

## Specifying random components

-   random = \~ 1 (random intercept)

-   random = \~ time (random slope)

-   random = \~ 1 + time (random intercept + slope)

## Model summary output

-   Iteration process/convergence criteria

    -   

-   Goodness-of-fit statistics

    -   Information criteria are model selection criteria. Some information criteria are better than others for a specific purpose, depending on, for example the data generating process, suggested models and sample size (pg.2).

        Zhang, J., Yang, Y. and Ding, J. (2023). Information criteria for model selection. *Wiley interdisciplinary reviews. Computational statistics*, 15(5). doi:https://doi.org/10.1002/wics.1607.

    -   Akaike Information Criterion

    -   Bayesian Information Criterion

-   Estimates, Wald and p-value

-   Variance-covariance matrix for the random effects

```{r}
jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data,
  maxiter = 500
)

# to do on the 1 class model if not converging -> local maxima
# gridsearch(jlcmm(
#   fixed = ln_PLASMA_NFL ~ time_years,
#   random = ~ 1,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ age_,
#   data = jlcmm_data
#   ),
#   rep = 20, maxiter = 20, minit = 10)
```

# JLCMM PROBLEMS

-   Infinite parameters:

-   The program stopped abnormally:

-   

```{r}
# checking for slope variance = 0 or correlation = +-1
lme(ln_PLASMA_NFL ~ time_years, random = ~ time_years | RID, data = jlcmm_data)
```

```{r}
# checking for the event distribution by genotype
table(jlcmm_data$event, jlcmm_data$genotype)
```

```{r}
# checking the range for time_years and ln_PLASMA_NFL
summary(jlcmm_data$time_years)
summary(jlcmm_data$ln_PLASMA_NFL)
```

## Try again with time_years_z

```{r}
# adding a scaled time_years variable
jlcmm_data <- jlcmm_data %>%
  mutate(time_years_z = scale(time_years))
```

```{r}
# fitting a simple mixed model to check the random slope 
lme(ln_PLASMA_NFL ~ time_years_z, random = ~ time_years_z | RID, data = jlcmm_data)
```

```{r}
jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data,
  maxiter = 500
)
```

## Try again with time_years_z and ln_NFL_z

```{r}
jlcmm_data <- jlcmm_data %>%
  mutate(ln_PLASMA_NFL_z = scale(ln_PLASMA_NFL))
```

```{r}
# fitting a simple mixed model to check the random slope 
lme(ln_PLASMA_NFL_z ~ time_years_z, random = ~ time_years_z | RID, data = jlcmm_data)
```

```{r}
jlcmm(
  fixed = ln_PLASMA_NFL_z ~ time_years_z,
  random = ~ time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data,
  maxiter = 500
)
```

## Try again with only time_to_event \> 0

-   This works!

-   $AIC = 7419$

-   $BIC = 7470$

-   Non-significant intercept

-   Low slope variance

-   We see that genotype e3/e3 is not significantly different from genotypes containing e2 (maybe since we have lower number of individuals with e2 genotypes): we can try collapsing the genotype variable down to e2/e3 containing genotypes, e4 heterozygotes and e4 homozygotes (try later)

```{r}
# there are 83 individuals with a start dementia dx, so time_to_event is 0
# remove these and test the model again
jlcmm_data %>% filter(time_to_event <= 0)

# creating a test data frame with no time_to_event = 0
jlcmm_data_test <- jlcmm_data %>% filter(time_to_event > 0)
```

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL_z ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test,
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

## Try with normal time_years and ln_PLASMA_NFL 

-   $AIC = 3237.74$

-   $BIC = 3288.37$

-   Lower information criteria, and significant intercept and slope

-   Even lower slope variance

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years,
  random = ~ 1 + time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

## Try with time_years_z and ln_PLASMA_NFL 

-   $AIC = 3237$

-   $BIC = 3288$

-   Increased slope variance

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

## Collapsing genotype down to 3 factors

-   e2/e3 containing genotypes

-   e4 heterozygotes

-   e4 homozygotes

-   Now all genotype factors significant

-   Now time_years variance still low, try with only random intercept

```{r}
# checking current factors
table(jlcmm_data_test$genotype)
```

```{r}
jlcmm_data_test <- jlcmm_data_test %>%
  mutate(
    genotype = case_when(
      genotype == "e2" ~ "e2 e3",
      genotype == "e3/e3" ~ "e2 e3",
      genotype == "e4" ~ "e4",
      genotype == "e4/e4" ~ "e4/e4"
    )
  )
```

```{r}
# checking changed factors
table(jlcmm_data_test$genotype)
```

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

## Try with only random intercept

-   Higher AIC and BIC

-   Actually figure out what is 'too small' slope variance...

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
plot(m0)
```

## Trying again with slope and intercept

-   Correlation should not be +- 1: no violation here

$$
Corr_{intercept, slope} = \frac{cov_{intercept, slope}}{\sigma_{intercept}\sigma_{slope}} =  \frac{0.00669}{\sqrt{0.19574}\sqrt{0.00255}} \approx 0.30 
$$

-   The variance of the intercept and slope should not be 0. The question is, what constitutes *too* small?

-   

```{r}
m0 <- jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years_z,
  random = ~ 1 + time_years_z,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test
)
```

```{r}
summary(m0)
```

```{r}
# calculating the correlation betweent the random intercept and slope, this should not be 1
0.00669 / (sqrt(0.19574) * sqrt(0.00255))
```

```{r}
plot(m0)
```

## Trying with ng \> 1

```{r}
jlcmm(
  fixed = ln_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ genotype,
  data = jlcmm_data_test,
  ng = 3,
  B = m0,
  maxiter = 500
)
```

### Model fitting examples from the lcmm package

```{r}
jlcmm1 <- Jointlcmm(Ydep1 ~ X1 * Time,
                    random = ~ Time,
                    subject = "ID",
                    survival = Surv(Tevent, Event) ~ X1 + X2, hazard = "3-quant-splines",
                    data = d)
```

```{r}
jlcmm2 <- Jointlcmm(Ydep1 ~ Time * X1,
                    random = ~ Time,
                    subject = "ID",
                    mixture = ~ Time,
                    survival = Surv(Tevent, Event) ~ X1 + mixture(X2),
                    hazard = "3-quant-splines", hazardtype = "PH", ng = 2,
                    data = data_lcmm, B = jlcmm1)
```
