---
format:
  pdf:
    include-before-body:
      text: |
        \begin{titlepage}
        \raggedleft
        \vspace*{2cm}
        
        {\Huge\bfseries Kandidatuppsats\par}
        \vspace{1.5cm}
        
        {\Large Statistiska Instiutionen\par}
        \vspace{3cm}
        
        {\Huge\bfseries Nr 2026:x\par}
        \vspace{0.5cm}
        
        {\LARGE \bfseries Code \par}
        
        {\large \today\par}
        \vspace{3cm}
        
        {\Huge Charlotte Emily Iszatt & Milou Rossi\par}
        
        \vfill
        
        \end{titlepage}
        
    toc: false
    colorlinks: false
    number-sections: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, output = TRUE)
```

# Pre-processing

```{r, warning=FALSE}
# Loading libraries
library(dplyr)
library(tidyr)
library(kableExtra)
library(corrplot)
library(JMbayes2)
library(JM)
library(nlme) 
library(lme4)
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) 
library(lcmm)
library(timeROC)
library(pec)
```

```{r}
# Set seed for reproducibility
set.seed(123)
```

## Loading the data

The data was loaded from the ADNIMERGE2 package accessed via the Alzheimer's Disease Neuroimaging Initiative (ADNI). The data sets include longitudinal ptau-181 and NfL measurements, diagnosis, demographic and genetic data.

```{r}
# Loading data
# Milou
# NfL data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
# P-tau181 data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
# Diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
# Demographics data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
# Apoe data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

```{r}
# Charley
#library(ADNIMERGE2)
#nfl_data = BLENNOWPLASMANFLLONG
#ptau_data = UGOTPTAU181
#diagnosis_data = DXSUM
#demographics = PTDEMOG
#apoe = APOERES
```

- *There are 52 individuals with two measurements recorded on the same exam date.*

- *How should we deal with these? Average the measurements? Remove the observations?*

1.  *Describe the measurements, their differences and averages*

- *Some of the values are quite different, what do we assume by averaging them? Do we introduce bias but just just removing them?*

2.  *Show how different these values are from the nfl and ptau averages (report SD differences in log transformations)*

```{r}
# Checking which ADNI protocols the data are from
# 
table(nfl_data$ORIGPROT)
table(ptau_data$ORIGPROT)
table(diagnosis_data$ORIGPROT)
```

```{r}
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", diagnosis_data$PTID)
grep("381_S_10###", apoe$PTID)
```

## Selecting required variables

Variables for the unique identifier for each individual, visit code, exam date, biomarker measurements, diagnosis and genetic data were extracted.

```{r, warning=FALSE}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required genotype variables
apoe <- apoe %>%
  dplyr::select(RID, GENOTYPE)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  dplyr::select(ORIGPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))
```

## Checking RID with more than one measurement recorded on the same exam date

```{r}
## Checking which obs have double VISCODE2 codes
# for ptau data
ptau_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# for nfl data
nfl_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# changing VISCODE2 for RID 1097 to m60_1 and m60_2
ptau_data[ptau_data$RID == 1097, "VISCODE2"] <- c("m48", "m60_2", "m60_1")

nfl_data[nfl_data$RID == 1097, "VISCODE2"] <-  c("m48", "m60_1", "m60_2")
```

```{r}
# Replacing visits with double measurements with visits with mean-value of the measurements as well as saving both measurements for future sensitivity analysis
# Naming the mean as the original variable to do analysis first with this method and then sensitivity analysis with the first and last measurements

# Adding columns for each ptau measurement and their mean and then collapsing down to unique RID and EXAMDATE
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_ptau = first(PLASMAPTAU181),
    last_ptau = last(PLASMAPTAU181),
    PLASMAPTAU181 = mean(PLASMAPTAU181),
    ) %>%
  ungroup() %>%
  distinct()

# Adding columns for each nfl measurement and their mean and then collapsing down to unique RID and EXAMDATE
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_nfl = first(PLASMA_NFL),
    last_nfl = last(PLASMA_NFL),
    PLASMA_NFL = mean(PLASMA_NFL)
    ) %>%
  ungroup() %>%
  distinct()
```

```{r}
# checking the difference in method for those with 2 obs on the same examdate
# for ptau
head(ptau_data %>%
       filter(PLASMAPTAU181 != first_ptau) %>%
       dplyr::select(PLASMAPTAU181, first_ptau, last_ptau)
     )

# for nfl
head(nfl_data %>%
       filter(PLASMA_NFL != first_nfl) %>%
       dplyr::select(PLASMA_NFL, first_nfl, last_nfl)
     )
```

## Merging NfL and ptau-181 data for the longitudinal data set

```{r}
# We merge NFL and ptau data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))
```

## Creating a common EXAMDATE and EXAMDATE_baseline variable

This was performed before preparation of the other variables for the survival and longitudinal models in order to use EXAMDATE_baseline to remove any observations from the diagnosis data set which occur long before the first biomarker measurements for that individual.

```{r}
# checking how many observations where the EXAMDATE is different for nfl and ptau
diff <- long_data %>% mutate(diff = as.numeric(EXAMDATE_nfl != EXAMDATE_ptau))
table(diff$diff)
```

```{r}
# filtering out the observations with different EXAMDATE
diff <- diff %>% filter(diff == TRUE)

# checking how many of these are baseline visits
table(diff$VISCODE2)

# checking which observations have the same examdate for nfl and ptau
common <- which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# checking the observations with different exam dates
long_data[-common, ]
```

```{r}
# calculating the difference in time between EXAMDATE_nfl and EXAMDATE_ptau
diff <- diff %>% mutate(time_diff = EXAMDATE_nfl - EXAMDATE_ptau)

# fetching the time differences
diff$time_diff 

# checking how many observations have a time difference over 30 days (1 month)
diff %>% filter(abs(time_diff) > 30) %>% dplyr::select(time_diff)
```

### EXAMDATE is EXAMDATE_nfl or EXAMDATE_nfl +- time difference

```{r}
# creating a common EXAMDATE from two cases
# 1: EXAMDATE_nfl == EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl
# 2: EXAMDATE_nfl != EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl + the difference between the two dates
# then defining the baseline date from the earliest EXAMDATE 
long_data <- long_data %>%
  group_by(RID) %>%
  mutate(
    EXAMDATE = EXAMDATE_ptau + 0.5 * (EXAMDATE_nfl - EXAMDATE_ptau),
    EXAMDATE_baseline = min(EXAMDATE),
    time_years = as.numeric(EXAMDATE - EXAMDATE_baseline) / 365.25) %>%
  ungroup()
```

### Baseline date is the earliest of EXAMDATE_nfl and EXAMDATE_ptau

```{r}
# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
# long_data <- long_data %>%
#   group_by(RID) %>%
#   mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
#   ungroup()
```

## Removing early diagnosis observations

```{r}
# filtering the observations out of the diagnosis data set where EXAMDATE is much before baseline biomarker EXAMDATE
# creating a baseline_dx variable, so we can set time_to_event = 0 if baseline_dx == DEMENTIA
diagnosis_data_clean <- diagnosis_data %>%
  merge(
    long_data %>% dplyr::select(RID, EXAMDATE_baseline), by = "RID"
    ) %>%
  filter(EXAMDATE_dx - EXAMDATE_baseline >= -50) %>%
  distinct() %>%
  group_by(RID) %>%
  arrange(EXAMDATE_dx) %>%
  mutate(
    baseline_dx = first(DIAGNOSIS_dx),
    last_dx = last(DIAGNOSIS_dx)
  ) %>%
  dplyr::select(!EXAMDATE_baseline)
```

## Merging longitudinal and diagnosis data for survival data set

```{r}
# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- long_data %>% left_join(diagnosis_data_clean, by = c("RID"))
```

```{r}
# Checking how many dementia diagnoses are probably due to AD
#dx <- ADNIMERGE2::DXSUM
#dx <- dx %>% dplyr::select(RID, DXAPP)
#dx <- dx %>% merge(surv_data, by = "RID")
#table(dx %>% dplyr::select(RID, DXAPP) %>% distinct() %>% dplyr::select(DXAPP))
```

```{r}
# Checking with ADNI protocol the diagnosis data is from, ADNI 3 + uses dementia diagnosis an all-cause dementia diagnosis
table(surv_data$ORIGPROT_dx)
```

## Checking missing values

```{r}
# We count missing values per column in long_data: all these are variables we will use in the model, so we can remove with na.omit()
sapply(long_data, function(x) sum(is.na(x)))
```

```{r}
# We count missing values per column in surv_data: VISCODE2.y will not be used in the model, so we need to see how these observations look
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# There are no rows with missing VISCODE2.y with full data for the model variables, so we can remove all with na.omit()
surv_data %>% filter(is.na(VISCODE2.y))
```

```{r}
# Checking for ptau-181 and NfL values = -4, this is missing data code in ADNI documentation -> no -4 coded cells
nrow(long_data %>% filter(PLASMA_NFL == -4, PLASMAPTAU181 == -4))
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## Merging selected demographic variables (age and gender) and APOE4 status

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  dplyr::select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We define APOE4 status based on number of E4 alleles
apoe <- apoe %>%
  mutate(APOE4_status = as.factor(
    case_when(
      GENOTYPE == "4/4" ~ "2 alleles",
      GENOTYPE %in% c("2/4", "3/4") ~ "1 allele",
      TRUE ~ "0 alleles")),
  genotype = as.factor(GENOTYPE)
  )

apoe$APOE4_status <- factor(apoe$APOE4_status,
                            levels = c("0 alleles", "1 allele", "2 alleles"))
```

```{r}
# We merge gender with long_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We merge apoe with long_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status", "genotype")], by="RID")

# We make sure that gender is a factor
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

## Selecting participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements of ptau_data
 keep <- unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
 long_data <- long_data %>% filter(RID %in% keep)
 surv_data <- surv_data %>% filter(RID %in% keep)
 # We check number of individuals with at least 3 measurements
 length(keep)
```

## Selecting participants with at least 1 measurement

```{r}
# We filter individuals with at least 1 measurement of ptau_data
#keep <- unique((ptau_data %>% group_by(RID) %>% filter(n() > 0))$RID)
#long_data <- long_data %>% filter(RID %in% keep)
#surv_data <- surv_data %>% filter(RID %in% keep)
# We check number of individuals with at least 1 measurement
#length(keep)
```

## Preparing for Cox PH model

### Constructing global (common) baseline, age at baseline, event and time-to-event variables

```{r}
surv_data <- surv_data %>%
  group_by(RID) %>%
  summarise(
#     carrying these forward to check events and event times are coded correctly
    baseline_dx = unique(baseline_dx),
    last_dx = unique(last_dx),
    
#     baseline EXAMDATE defined as the earlier biomarker EXAMDATE (nfl or ptau)
    EXAMDATE_baseline = min(EXAMDATE_baseline),
    
#     age at baseline visit
    age_at_baseline = age_at_visit[
      EXAMDATE_ptau == min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)]) |
      EXAMDATE_nfl == min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])
      ][1],

#      last dx/nfl/ptau exam
      last_followup = max(c(EXAMDATE_ptau, EXAMDATE_nfl, EXAMDATE_dx), na.rm = TRUE),

#     1 = dementia, 0 = otherwise
      event = as.integer(any(DIAGNOSIS_dx == "Dementia")),

#     0 if dementia dx at baseline, otherwise time to dementia dx or date of censuring
      time_to_event = case_when(
        event == 1 & unique(baseline_dx) == "Dementia" ~ 0,
        event == 1 & unique(baseline_dx) != "Dementia" ~
          as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - EXAMDATE_baseline)/365,
        TRUE ~
          as.numeric(last_followup - EXAMDATE_baseline)/365),

#     Sensitivity: event time = midpoint between last clean exam and dementia dx exam
    time_to_event_midpoint = case_when(
      event == 1 & unique(baseline_dx) == "Dementia" ~ 0,
      event == 1 & unique(baseline_dx) != "Dementia" ~ {
        dx_date <- min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"])
        prev_date <- max(EXAMDATE_dx[DIAGNOSIS_dx != "Dementia" & 
                                     EXAMDATE_dx < dx_date], 
                         na.rm = TRUE)
        as.numeric((prev_date + (dx_date - prev_date) / 2) - EXAMDATE_baseline) / 365
      },
      TRUE ~
        as.numeric(last_followup - EXAMDATE_baseline) / 365
    )
)
```

```{r}
# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

## Removing individuals with dementia at start

```{r}
# Removing individuals who had a dementia dx at time = 0
surv_data <- surv_data %>% filter(time_to_event > 0)
long_data <- long_data %>% filter(RID %in% unique(surv_data$RID))

# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

### Constructing baseline values of each biomarker, to include in the Cox PH models as covariates

```{r}
# Creating a data frame with the first biomarker measurements
baseline <- long_data %>% 
  group_by(RID) %>%
  arrange(RID, EXAMDATE_nfl) %>% 
  mutate(baseline_value_nfl = first(PLASMA_NFL)) %>%
  arrange(RID, EXAMDATE_ptau) %>%
  mutate(baseline_value_ptau = first(PLASMAPTAU181)) %>%
  dplyr::select(RID, baseline_value_ptau, baseline_value_nfl) %>%
  ungroup() %>%
  distinct()
```

```{r}
# merging the survival data with demographics and the baseline biomarker values
surv_data <- surv_data %>%
  merge(demographics_small, by = "RID") %>% 
  merge(apoe %>% dplyr::select(RID, APOE4_status, genotype), by = "RID") %>%
  merge(baseline, by = "RID")
```

## Summaries and distributions of the model variables

```{r}
# Distribution of diagnosis assessments
dx_full_704 <- diagnosis_data %>%
  filter(RID %in% surv_data$RID)
table(dx_full_704$DIAGNOSIS_dx)

# Distribution of post-baseline diagnosis assessments (aligned to biomarker baseline)
dx_post <- dx_full_704 %>%
  left_join(
    long_data %>%
      dplyr::select(RID, EXAMDATE_baseline) %>%
      distinct(),
    by = "RID"
  ) %>%
  dplyr::filter(EXAMDATE_dx >= EXAMDATE_baseline)
table(dx_post$DIAGNOSIS_dx)

# Number of diagnosis assessments per participant
dx_full_704 %>%
  filter(!is.na(DIAGNOSIS_dx)) %>%
  count(RID) %>%
  summarise(
    min = min(n),
    median = median(n),
    max = max(n)
  )

# Summaries of p-tau181 and NfL 
summary(long_data[, c("PLASMAPTAU181", "PLASMA_NFL")])

# Examination date ranges for p-tau181, NfL and diagnosis
range(long_data$EXAMDATE_ptau, na.rm = TRUE)
range(long_data$EXAMDATE_nfl, na.rm = TRUE)
range(dx_full_704$EXAMDATE_dx, na.rm = TRUE)

# Number of biomarker measurements per participant (p-tau181)
ptau_table <- long_data %>%
  filter(!is.na(PLASMAPTAU181)) %>%
  count(RID) %>%
  count(n, name = "n_RID")
ptau_table

# Number of biomarker measurements per participant (NfL)
nfl_table <- long_data %>%
  filter(!is.na(PLASMA_NFL)) %>%
  count(RID) %>%
  count(n, name = "n_RID")
nfl_table

# Summary of age at baseline
summary(surv_data$age_at_baseline)

# Distributions of gender and APOE4 status
table(surv_data$PTGENDER)
table(surv_data$genotype)
```

## Checking outliers

```{r}
# We identify "extreme" values

# Values greater than 3 sd above ptau mean
long_data[scale(long_data$PLASMAPTAU181) > 3, ]
# Values smaller than 3 sd below ptau mean
long_data[scale(long_data$PLASMAPTAU181) < -3, ]

# Values greater than 3 sd above nfl mean
long_data[scale(long_data$PLASMA_NFL) > 3, ]
# Values smaller than 3 sd below nfl mean
long_data[scale(long_data$PLASMA_NFL) < -3, ]
```

## Transformation of the biomarkers

```{r}
# We check the distribution of each biomarker
hist(long_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(long_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(long_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(long_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

```{r}
# Transforming the baseline values in the survival data
surv_data$baseline_value_ptau <- log(surv_data$baseline_value_ptau)
surv_data$baseline_value_nfl <- log(surv_data$baseline_value_nfl)
```

## Rechecking longitudinal and survival data lengths

```{r}
# We check the number of individuals in each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

## Splitting data into in-sample and out-sample

```{r}
# Total number of individuals
RID <- unique(surv_data$RID)

# Number of individuals to test the model on
n_out_sample <- length(RID) * (1/4)

# Out-sample individuals
RID_out_sample <- sample(RID, n_out_sample)

# Longitudinal data with in-sample individuals
long_data_in <- long_data %>% filter(!RID %in% RID_out_sample)

# Longitudinal data with out-sample individuals
long_data_out <- long_data %>% filter(RID %in% RID_out_sample)

# Survival data with in-sample individuals
surv_data_in <- surv_data %>% filter(!RID %in% RID_out_sample)

# Survival data with out-sample individuals
surv_data_out <- surv_data %>% filter(RID %in% RID_out_sample)
```

```{r}
# Checking the number of individuals and event distribution for the in-sample observations
length(unique(surv_data_in$RID))
table(surv_data_in$event)
# Fraction of individuals who experience the event in in-sample
128/528

# Checking the number of individuals and event distribution for the out-sample observations
length(unique(surv_data_out$RID))
table(surv_data_out$event)
# Fraction of individuals who experience the event in out-sample
49/176
```

## Checking correlation between the biomarkers as well as the numerical variables going to be used in the models

```{r}
# Correlation between the biomarkers 
cor(long_data_in$PLASMAPTAU181, long_data_in$PLASMA_NFL, use = "complete.obs")
```

```{r}
# Saving the numerical variables used in the Cox PH models
cox_vars <- surv_data_in[, c("age_at_baseline",
                              "baseline_value_ptau",
                              "baseline_value_nfl")]

# Computing correlation matrix for Cox model variables
cox_cor <- cor(cox_vars, use = "complete.obs")

# Visualising correlations
corrplot(cox_cor,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)


# Saving the numerical variables used in the LME models
lme_vars <- long_data_in[, c("time_years",
                             "log_PLASMAPTAU181",
                             "log_PLASMA_NFL")]

# Computing correlation matrix for LME model variables
lme_cor <- cor(lme_vars, use = "complete.obs")

# Visualising correlations 
corrplot(lme_cor,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)
```

# Longitudinal models

## Comparing link functions for longitudinal models

Suggested models:

- Gamma: raw data could be modelled well with the gamma distribution since takes no negative values, variance increases with mean, and there is a right-skew to the raw distributions

- Linear mixed effects model with log-transformed data: approximately normal, yet still some skew

- Linear mixed effects model with cube root transformed data: again, approximately normal, still a bit of skew (why such a low AIC? seems wrong?)

- GLMM with log-normal distribution

- GLMM with Student-t distribution: longer tails

### PTAU

```{r}
# checking the mean and variance for ptau181
mean(long_data$PLASMAPTAU181)
var(long_data$PLASMAPTAU181)
```

#### Fitting the models

```{r}
# GLMM
fit_gamma_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID), 
                   data = long_data_in,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_ptau <- lmer(log(PLASMAPTAU181) ~ time_years + (1|RID), 
                data = long_data_in)

# LME approach with log transformed data
fit_lme_cubed_ptau <- lmer(sqrt(sqrt(PLASMAPTAU181)) ~ time_years + (1|RID), 
                           data = long_data_in)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                             data = long_data_in,
                             family = lognormal(link = "log"))

# Student-t
#fit_student_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                            #data = long_data_in,
                            #family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_ptau <- simulateResiduals(fit_gamma_ptau)
testResiduals(gamma_residuals_ptau)

lme_residuals_log_ptau <- simulateResiduals(fit_lme_log_ptau)
testResiduals(lme_residuals_log_ptau)

lme_residuals_cubed_ptau <- simulateResiduals(fit_lme_cubed_ptau)
testResiduals(lme_residuals_cubed_ptau)

lognormal_residuals_ptau <- simulateResiduals(fit_lognormal_ptau)
testResiduals(lognormal_residuals_ptau)

#student_residuals_ptau <- simulateResiduals(fit_student_ptau)
#testResiduals(student_residuals_ptau)
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(gamma_residuals_ptau$fittedResiduals)
shapiro.test(lme_residuals_log_ptau$fittedResiduals)
shapiro.test(lme_residuals_cubed_ptau$fittedResiduals)
shapiro.test(lognormal_residuals_ptau$fittedResiduals)
#shapiro.test(student_residuals_ptau$fittedResiduals)
```

#### AIC and log-likelihood

```{r}
# Comparing p-tau181 models with AIC and log likelihood
AIC(fit_lme_log_ptau, fit_lme_cubed_ptau, fit_gamma_ptau, fit_lognormal_ptau)
logLik(fit_lme_log_ptau)
logLik(fit_lme_cubed_ptau)
logLik(fit_gamma_ptau)
logLik(fit_lognormal_ptau)
#logLik(fit_student_ptau)
```

### NfL

```{r}
# checking the mean and variance for nfl
mean(long_data$PLASMA_NFL)
var(long_data$PLASMA_NFL)
```

#### Fitting the models

```{r}
# GLMM
fit_gamma_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID), 
                   data = long_data_in,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_nfl <- lmer(log(PLASMA_NFL) ~ time_years + (1|RID), 
                data = long_data_in)

# LME approach with log transformed data
fit_lme_cubed_nfl <- lmer(sqrt(sqrt(PLASMA_NFL)) ~ time_years + (1|RID), 
                          data = long_data_in)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                             data = long_data_in,
                             family = lognormal(link = "log"))

# Student-t
#fit_student_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                           #data = long_data_in,
                           #family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_nfl <- simulateResiduals(fit_gamma_nfl)
testResiduals(gamma_residuals_nfl)

lme_residuals_log_nfl <- simulateResiduals(fit_lme_log_nfl)
testResiduals(lme_residuals_log_nfl)

lme_residuals_cubed_nfl <- simulateResiduals(fit_lme_cubed_nfl)
testResiduals(lme_residuals_cubed_nfl)

lognormal_residuals_nfl <- simulateResiduals(fit_lognormal_nfl)
testResiduals(lognormal_residuals_nfl)

#student_residuals_nfl <- simulateResiduals(fit_student_nfl)
#testResiduals(student_residuals_nfl)
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(gamma_residuals_nfl$fittedResiduals)
shapiro.test(lme_residuals_log_nfl$fittedResiduals)
shapiro.test(lme_residuals_cubed_nfl$fittedResiduals)
shapiro.test(lognormal_residuals_nfl$fittedResiduals)
#shapiro.test(student_residuals_nfl$fittedResiduals)
```

#### AIC and log-likelihood

```{r}
# Comparing NfL-models with AIC and log likelihood
AIC(fit_lme_log_nfl, fit_lme_cubed_nfl, fit_gamma_nfl, fit_lognormal_nfl)
logLik(fit_lme_log_nfl)
logLik(fit_lme_cubed_nfl)
logLik(fit_gamma_nfl)
logLik(fit_lognormal_nfl)
#logLik(fit_student_nfl)
```

Osäker dock om man kan jämföra log-version och cube-version av lme med de andra modellerna (som använder råa versionen) eftersom de då är på olika skalor! Tror det är bäst att kolla på residualerna och QQ-plotsen

# Fitted LME models

## PTAU

#### With all genotype levels

- None of the coefficients are significant for any of the levels for genotype

```{r}
# Longitudinal model for log p-tau181 with normal distribution 
long_model_ptau_full <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + genotype,
  random = ~ time_years | RID,                    
  data = long_data_in
)
```

```{r}
summary(long_model_ptau_full)
```

#### APOE4 status

- All three levels of APOE4_status are significant

```{r}
# Longitudinal model for log p-tau181 with APOE4 status 
long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status,
  random = ~ time_years | RID,                    
  data = long_data_in
)
```

```{r}
summary(long_model_ptau)
```

#### Determining random effects

- Fitting the model with ML estimation in order to compare models using likelihood ratio test

- The results give support to fitting the simpler model with only a random intercept. However, $jm()$ does not currently support model fitting for univariate longitudinal data with only a random intercept.

```{r}
m1 <- lmer(log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status + (1 | RID),
           data = long_data_in, REML = FALSE)

m2 <- lmer(log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status + (time_years | RID),
           data = long_data_in, REML = FALSE)
```

```{r}
anova(m1, m2, type = "LRT")
```

## NFL

#### With all genotype levels

```{r}
# Longitudinal model for log NfL with normal distribution, all genotype levels
long_model_nfl_full <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER + genotype, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl_full)
```

#### APOE4 status

- None of the levels are significant for log NfL and neither is gender

```{r}
# Longitudinal model for log NfL with normal distribution, APOE4_status
long_model_nfl <- lme(
  fixed = log_PLASMA_NFL ~ time_years + PTGENDER + APOE4_status, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl)
```

#### Removing APOE4 status

```{r}
# Longitudinal model for log NfL without APOE4_status
long_model_nfl_reduced1 <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl_reduced1)
```

#### Removing PTGENDER

```{r}
# Longitudinal model for log NfL without covariates
long_model_nfl_reduced2 <- lme(
  log_PLASMA_NFL ~ time_years, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl_reduced2)
```

#### Determining the random effects

- The results of the likelihood-ratio test support including a random slope as well as a random intercept for the NfL model

```{r}
m1 <- lmer(log_PLASMA_NFL ~ time_years + (1 | RID),
           data = long_data_in, REML = FALSE)

m2 <- lmer(log_PLASMA_NFL ~ time_years + (time_years | RID),
           data = long_data_in, REML = FALSE)
```

```{r}
anova(m1, m2, type = "LRT")
```

```{r}
# Table of results from the likelihood ratio tests between models with only a ranodom intercept and those with random intercept and slope
long_re_lrt_table <- data.frame(
  "Model" = c("ptau-181", "NfL"),
  "Test statistic" = c(1.488, 10.506),
  "P-value" = c(0.475, 0.005),
  check.names = FALSE
  )
```

```{r}
# Mixed object for multivariate joint model
mixed <- list(long_model_ptau, long_model_nfl)
```

## Residual check

#### Residuals vs fitted values

```{r}
# residuals of the p-tau181 LME model
res_ptau <- residuals(long_model_ptau, type = "pearson")  

# fitted values of the p-tau181 LME model
fitted_vals_ptau <- fitted(long_model_ptau)  

# plot of residuals vs fitted values 
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for p-tau181")
abline(h = 0, col = "red", lty = 2)
```

```{r}
# residuals of the NfL LME model
res_nfl <- residuals(long_model_nfl, type = "pearson")  

# fitted values of the NfL LME model
fitted_vals_nfl <- fitted(long_model_nfl)  

# plot of residuals vs fitted values 
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NFL")
abline(h = 0, col = "red", lty = 2)
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(res_ptau)
```

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(res_nfl)
```

#### QQ plots

```{r}
# For P-tau
qqnorm(res_ptau, main = "QQ plot: p-tau181 residuals")
qqline(res_ptau, col = "red")

# For NFL 
qqnorm(res_nfl, main = "QQ plot: NfL residuals")
qqline(res_nfl, col = "red")
```

# Survival models

## For the different cox PH models and the separate joint models

## Fitted standard Cox PH model (for the joint models)

#### With all genotype levels

```{r}
# Cox survival-models, with covariates
# All genotype levels
surv_model_full <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + genotype,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_full)
```

#### APOE4 status

```{r}
# Cox survival-models, with covariates
# With APOE4_status
surv_model <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model)
```

## Fitted Cox PH models using baseline values of each biomarker

### PTAU

```{r}
# Cox survival-models, with covariates including baseline values of each biomarker 
# For p-tau181 with PTGENDER
surv_model_ptau_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_ptau_bl)
```

### NfL

```{r}
# For baseline NfL with covariates
surv_model_nfl_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_nfl_bl)
```

### PTAU and NfL

```{r}
# Inlcuding both p-tau181 and NfL
surv_model_both_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_both_bl) 
```

### Test of the assumption of proportional hazards

```{r}
# Extracting p-values from cox.zph models
extract_ph_p <- function(model, model_name) {
  test <- cox.zph(model)
  df <- as.data.frame(test$table)
  
  data.frame(
    variable = rownames(df),
    p_value = round(df$p, 3),
    model = model_name
  )
}

# Combining all Cox PH models into one dataset
ph_all <- rbind(
  extract_ph_p(surv_model, "Standard"),
  extract_ph_p(surv_model_ptau_bl, "p-tau181 baseline"),
  extract_ph_p(surv_model_nfl_bl, "NfL baseline"),
  extract_ph_p(surv_model_both_bl, "Both baseline")
)

# Forming a table where each model has its own column
table_ph <- ph_all %>%
  dplyr::select(variable, model, p_value) %>%
  tidyr::pivot_wider(names_from = model, values_from = p_value)

# Displaying final table
table_ph
```

#### Removing age-at-baseline due to evidence of non-proportional hazards

```{r}
# Cox survival-models with covariates but excluding age-at-baseline
# Standard cox model 
surv_model_final <- coxph(
  Surv(time_to_event, event) ~ PTGENDER + APOE4_status,
  data = surv_data_in,
  x = TRUE
)

# Including p-tau181
surv_model_ptau_bl_final <- coxph(
  Surv(time_to_event, event) ~
    PTGENDER + APOE4_status + baseline_value_ptau,
  data = surv_data_in,
  x = TRUE
)

# Including NfL
surv_model_nfl_bl_final <- coxph(
  Surv(time_to_event, event) ~
    PTGENDER + APOE4_status + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

# Inlcuding both p-tau181 and NfL
surv_model_both_bl_final <- coxph(
  Surv(time_to_event, event) ~
    PTGENDER + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)
```

#### Final test of the assumption of proportional hazards

```{r}
# Combining all final Cox PH models into one dataset
ph_all_final <- rbind(
  extract_ph_p(surv_model_final, "Standard"),
  extract_ph_p(surv_model_ptau_bl_final, "p-tau181 baseline"),
  extract_ph_p(surv_model_nfl_bl_final, "NfL baseline"),
  extract_ph_p(surv_model_both_bl_final, "Both baseline")
)

# Forming a table where each final model has its own column
table_ph_final <- ph_all_final %>%
  dplyr::select(variable, model, p_value) %>%
  tidyr::pivot_wider(names_from = model, values_from = p_value)

# Displaying final PH table 
table_ph_final
```

### Removing PTGENDER to run likelihood-ratio tests

```{r}
# Cox survival-models, with covariates including baseline values of each biomarker

# For p-tau181 without PTGENDER
surv_model_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status,
  data = surv_data_in,
  x = TRUE
  )

anova_surv <- anova(surv_model_final, surv_model_reduced, test = "LRT")
chi_surv <- anova_surv[2,2]

# For p-tau181 without PTGENDER
surv_model_ptau_bl_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status + baseline_value_ptau,
  data = surv_data_in,
  x = TRUE
  )

anova_ptau <- anova(surv_model_ptau_bl_final, surv_model_ptau_bl_reduced, test = "LRT")
chi_ptau <- anova_ptau[2,2]

# For NfL without PTGENDER
surv_model_nfl_bl_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

anova_nfl <- anova(surv_model_nfl_bl_final, surv_model_nfl_bl_reduced, test = "LRT")
chi_nfl <- anova_nfl[2,2]

# Without PTGENDER
surv_model_both_bl_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

anova_both <- anova(surv_model_both_bl_final, surv_model_both_bl_reduced, test = "LRT")
chi_both <- anova_both[2,2]
```

```{r}
# Viewing the LRT results
anova_ptau
anova_nfl
anova_both
```

```{r}
# Table of results from the likelihood ratio tests for cox models with and without PTGENDER
surv_lrt_table <- data.frame(
  "Model" = c("ptau-181 baseline", "NfL baseline", "ptau-181 and NfL baseline"),
  "Test statistic" = c(chi_ptau, chi_nfl, chi_both),
  "P-value" = c(0.98, 0.95, 0.73),
  check.names = FALSE
  )
```

### Cox model comparison table

```{r}
# Building Cox model comparison table
# Extracting model output
extract_model_cox <- function(model) {
  s <- summary(model)

  data.frame(
    Variable = rownames(s$coefficients),
    HR = exp(s$coefficients[, "coef"]),
    CI_low = s$conf.int[, "lower .95"],
    CI_high = s$conf.int[, "upper .95"],
    P = s$coefficients[, "Pr(>|z|)"],
    row.names = NULL
  )
}

# Formatting HR and CI
format_hr_ci <- function(hr, low, high) {
  paste0(round(hr, 3), " (", round(low, 3), "–", round(high, 3), ")")
}

# Formatting P-values
format_p <- function(x) {
  ifelse(is.na(x), NA,
         ifelse(x < 0.001, "<0.001", round(x, 3)))
}

# Extracting models
ptau_cox <- extract_model_cox(surv_model_ptau_bl_final)
nfl_cox <- extract_model_cox(surv_model_nfl_bl_final)
both_cox <- extract_model_cox(surv_model_both_bl_final)

# Building variable set
all_vars_cox <- unique(c(ptau_cox$Variable, nfl_cox$Variable, both_cox$Variable))

# Value extraction
get_value <- function(df, var, col) {
  out <- rep(NA, length(var))

  for (i in seq_along(var)) {
    idx <- which(df$Variable == var[i])
    if (length(idx) == 1) {
      out[i] <- df[idx, col]
    }
  }
  out
}

# Building final table
table_cox <- data.frame(
  Variable = all_vars_cox,

  "PTau HR (CI)" = format_hr_ci(
    get_value(ptau_cox, all_vars_cox, "HR"),
    get_value(ptau_cox, all_vars_cox, "CI_low"),
    get_value(ptau_cox, all_vars_cox, "CI_high")
  ),
  "PTau P" = format_p(get_value(ptau_cox, all_vars_cox, "P")),

  "NfL HR (CI)" = format_hr_ci(
    get_value(nfl_cox, all_vars_cox, "HR"),
    get_value(nfl_cox, all_vars_cox, "CI_low"),
    get_value(nfl_cox, all_vars_cox, "CI_high")
  ),
  "NfL P" = format_p(get_value(nfl_cox, all_vars_cox, "P")),

  "Both HR (CI)" = format_hr_ci(
    get_value(both_cox, all_vars_cox, "HR"),
    get_value(both_cox, all_vars_cox, "CI_low"),
    get_value(both_cox, all_vars_cox, "CI_high")
  ),
  "Both P" = format_p(get_value(both_cox, all_vars_cox, "P")),

  row.names = NULL, 
  check.names = FALSE
)

# Showing table
table_cox
```

### Concordance Index Table

```{r}
# C-index table
c_index_table <- data.frame(
  "Model" = c("Cox ptau-181", "Cox NfL", "Cox ptau-181 & NfL"),
  "C-Index" = c(0.679, 0.673, 0.681),
  check.names = FALSE)
```

# Two separate joint models

## Fitted joint model for p-tau181

```{r}
# Joint model for p-tau181
joint_model_ptau <- jm(
  surv_model_final,
  long_model_ptau,
  time_var = "time_years",
  n_chains = 3L,
  n_iter = 11000L,
  n_burnin = 1000L
  )
```

```{r}
# Checking convergence 
ggtraceplot(joint_model_ptau, "alphas")
```

## Fitted joint model for NfL

```{r}
# Joint model for NfL
joint_model_nfl <- jm(
  surv_model_final,
  long_model_nfl,
  time_var = "time_years",
  n_chains = 3L,
  n_iter = 11000L,
  n_burnin = 1000L
  )
```

```{r}
# Checking convergence 
ggtraceplot(joint_model_nfl, "alphas")
```

# Multivariate joint model

## To check if the models have stable random slopes

```{r}
VarCorr(long_model_ptau)
VarCorr(long_model_nfl)
```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

## fForms

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181),
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL)
  )
```

## Fitted multivariate joint model

```{r}
multi_joint_model <- jm(surv_model_final, mixed ,
                         time_var = "time_years",
                         functional_forms = fForms,
                         n_chains = 3L,
                         n_iter = 20000L,
                         n_burnin = 2000L
                         )
 
```

```{r}
# Checking convergence 
ggtraceplot(multi_joint_model, "alphas")
```

### Joint model comparison table

```{r}
# Building joint model comparison table
# Extracting models (JMbayes2 joint models use log-HR scale with HR = exp(Mean)), where Mean is the posterior mean of the log-hazard ratio
extract_model_jm <- function(model) {

  s <- summary(model)$Survival

  data.frame(
    Variable = rownames(s),

    HR = exp(s[, "Mean"]),
    CI_low = exp(s[, "2.5%"]),
    CI_high = exp(s[, "97.5%"]),

    P = s[, "P"],
    row.names = NULL
  )
}

# Extracting models
ptau_jm <- extract_model_jm(joint_model_ptau)
nfl_jm <- extract_model_jm(joint_model_nfl)
both_jm <- extract_model_jm(multi_joint_model)

# Building variable set
all_vars_jm <- unique(c(ptau_jm$Variable, nfl_jm$Variable, both_jm$Variable))

# Building final table
table_jm <- data.frame(
  Variable = all_vars_jm,

  "PTau HR (CI)" = format_hr_ci(
    get_value(ptau_jm, all_vars_jm, "HR"),
    get_value(ptau_jm, all_vars_jm, "CI_low"),
    get_value(ptau_jm, all_vars_jm, "CI_high")
  ),
  "PTau P" = format_p(get_value(ptau_jm, all_vars_jm, "P")),

  "NfL HR (CI)" = format_hr_ci(
    get_value(nfl_jm, all_vars_jm, "HR"),
    get_value(nfl_jm, all_vars_jm, "CI_low"),
    get_value(nfl_jm, all_vars_jm, "CI_high")
  ),
  "NfL P" = format_p(get_value(nfl_jm, all_vars_jm, "P")),

  "Both HR (CI)" = format_hr_ci(
    get_value(both_jm, all_vars_jm, "HR"),
    get_value(both_jm, all_vars_jm, "CI_low"),
    get_value(both_jm, all_vars_jm, "CI_high")
  ),
  "Both P" = format_p(get_value(both_jm, all_vars_jm, "P")),

  row.names = NULL,
  check.names = FALSE
)

# Showing table
table_jm
```

## Rhat value comparison table

```{r}
# Extracting Rhat for each model (survival component)
extract_rhat_surv <- function(model) {

  s <- summary(model)$Survival

  data.frame(
    Variable = rownames(s),
    Rhat = s[, "Rhat"],
    row.names = NULL
  )
}

# Applying to each model
ptau_surv <- extract_rhat_surv(joint_model_ptau)
nfl_surv <- extract_rhat_surv(joint_model_nfl)
multi_surv <- extract_rhat_surv(multi_joint_model)

# Combining with model labels to form a table
rhat_survival_table <- bind_rows(
  ptau_surv %>% mutate(Model = "p-tau181"),
  nfl_surv %>% mutate(Model = "NfL"),
  multi_surv %>% mutate(Model = "Multivariate")
) %>%
  pivot_wider(
    names_from = Model,
    values_from = Rhat
  ) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))

# Showing table
rhat_survival_table
```

```{r}
# Extracting Rhat for longitudinal outcome
extract_rhat_long <- function(model) {

  s <- summary(model)$Outcome1

  data.frame(
    Variable = rownames(s),
    Rhat = s[, "Rhat"],
    row.names = NULL
  )
}

# Applying to each model
ptau_long <- extract_rhat_long(joint_model_ptau)
nfl_long <- extract_rhat_long(joint_model_nfl)
multi_long <- extract_rhat_long(multi_joint_model)

# Combining with model labels to form a table
rhat_longitudinal_table <- bind_rows(
  ptau_long %>% mutate(Model = "p-tau181"),
  nfl_long %>% mutate(Model = "NfL"),
  multi_long %>% mutate(Model = "Multivariate")
) %>%
  pivot_wider(
    names_from = Model,
    values_from = Rhat
  ) %>%
  mutate(across(where(is.numeric), ~round(.x, 3)))

# Showing table
rhat_longitudinal_table
```

# Latent Class Mixed Models

Choice of p-value <https://pmc.ncbi.nlm.nih.gov/articles/PMC5059270/>

# LCMM: NFL

## M0 (1 latent class)

```{r}
# Fitting a one class latent class mixed model with a random intercept
hlme_m0_1 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(hlme_m0_1)
```

```{r}
# Fitting a one class latent class mixed model with a random intercept and slope
hlme_m0_2 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(hlme_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(hlme_m0_1, hlme_m0_2)
```

```{r}
# log likelihood for the full model
l = hlme_m0_2$loglik

# log likelihood for the reduced model
l_0 = hlme_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Fitting a two class latent class mixed model
hlme_m1 <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = long_data_in,
    B = random(hlme_m0_1)
  )

summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## Comparing M0, M1

```{r}
summarytable(hlme_m0_2, hlme_m1,
             which = c("G", "loglik", "BIC", "entropy"))
```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# JLCMM: NFL

```{r}
surv <- survfit(Surv(surv_data_in$time_to_event, surv_data_in$event) ~ 1)

plot(log(surv$time), log(-log(surv$surv)), main="Weibull Check")
```

```{r}
# merging the longitudinal and survival data for jlcmm
jlcmm_data <- surv_data_in %>%
  dplyr::select(RID, event, time_to_event) %>%
  merge(long_data_in, by = "RID")
```

## M0 (1 latent class)

```{r}
# Fitting a one class joint latent class mixed model with a random intercept
jlcmm_m0_1 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)

# Fitting a one class joint latent class mixed model with flexible hazard
jlcmm_m0_1_piecewise <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  hazard = "3-quant-piecewise",
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_1)
```

```{r}
summary(jlcmm_m0_1_piecewise)
```

```{r}
# Fitting a one class joint latent class mixed model with a random intercept and slope
jlcmm_m0_2 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1 + time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(jlcmm_m0_1, jlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = jlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = jlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Running gridsearch() for a two class model
# jlcmm_m1_nfl_search <- gridsearch(m = jlcmm(fixed = log_PLASMA_NFL ~ time_years,
#           random = ~ 1,
#           mixture = ~ 1,
#           classmb = ~ 1,
#           ng = 2,
#           subject = "RID",
#           survival = Surv(time_to_event, event) ~ 1,
#           data = jlcmm_data),
# minit = jlcmm_m0_1,
# rep = 50,
# maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m1_nfl <- jlcmm(
#   fixed = log_PLASMA_NFL ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 2,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m1_nfl_search$best
# )
```

```{r}
# Fitting a two class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m1_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m1_nfl)
```

```{r}
# Checking flexible hazard
jlcmm_m1_nfl_piecewise <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1_piecewise),
)
```

```{r}
# Does not converge
summary(jlcmm_m1_nfl_piecewise)
```

```{r}
# Defining a function to extract the posterior probability metrics as presenting in postprob() but in one table
get_pprob_table <- function(model) {
  G <- model$ng
  pprobs <- model$pprob
  
#  Mean posterior probabilities per class 
  mean_pprob <- do.call(rbind, lapply(1:G, function(g) {
    # Individuals assigned to class g
    members <- pprobs[pprobs$class == g, 3:(2 + G)]
    colMeans(members)
  }))
  
  rownames(mean_pprob) <- paste0("Class", 1:G)
  colnames(mean_pprob) <- paste0("Prob", 1:G)
  
#  Percentage of indiviudals above posterior probability thresholds per class
  thresholds <- c(0.7, 0.8, 0.9)
  
  threshold_table <- do.call(rbind, lapply(1:G, function(g) {
#    Extracting individuals whose highest posterior probability is for class g
    pprob_g <- pprobs[pprobs$class == g, 2 + g]
    sapply(thresholds, function(t)
      round(mean(pprob_g > t) * 100, 2))
  }))
  
  rownames(threshold_table) <- paste0("Class", 1:G)
  colnames(threshold_table) <- paste0("Prob > ", thresholds, " (%)")
  
#  Combine into one table
  out <- cbind(
    as.data.frame(round(mean_pprob, 4)),
    as.data.frame(threshold_table)
  )
  
  return(out)
}
```

```{r}
# defining a function to extract the model parameters
get_jlcmm_params <- function(model) {
  
  p_full <- length(model$best)
#   model$V is a vector of the upper triangle of a matrix, not the var-covar matrix itself 
  V_mat <- matrix(0, nrow = p_full, ncol = p_full)
  V_mat[upper.tri(V_mat, diag = TRUE)] <- model$V
#   Fill the matrix by adding its transpose, then removing the doubled values on the diagonal
  V_mat <- V_mat + t(V_mat) - diag(diag(V_mat))
  
#   Dropping unwanted values stored in model$best
  drop <- "varcov|stderr"
#   Creates a logical vector in orer to drop values from estimates and var-covar matrix 
  keep <- !grepl(drop, names(model$best))
  estimates <- model$best[keep]
  V_mat <- V_mat[keep, keep]
  parameters <- names(estimates)

  se <- sqrt(diag(V_mat))
  z <- estimates / se
  p_values <- 2 * (1 - pnorm(abs(z)))
  
  submodel <- c("Class Membership",
                "Survival", "Survival", "Survival", "Survival",
                "Longitudinal", "Longitudinal", "Longitudinal")
  
  out <- data.frame(
    "Submodel" = submodel,
    "Parameter" = parameters,
    "Estimate" = round(estimates, 3),
    "SE" = round(se, 3),
    "Z" = round(z, 3),
    "P value" = round(p_values, 3),
    check.names = FALSE
  )
  
  return(out)
}
```

## M2 (3 latent classes)

```{r}
# Running gridsearch() for a three class model
# jlcmm_m2_nfl_search <- gridsearch(m = jlcmm(fixed = log_PLASMA_NFL ~ time_years,
#                                             random = ~ 1,
#                                             mixture = ~ 1,
#                                             classmb = ~ 1,
#                                             ng = 3,
#                                             subject = "RID",
#                                             survival = Surv(time_to_event, event) ~ 1,
#                                             data = jlcmm_data),
#                                   minit = jlcmm_m0_1,
#                                   rep = 50,
#                                   maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m2_nfl <- jlcmm(
#   fixed = log_PLASMA_NFL ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 3,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m2_nfl_search$best
# )
```

```{r}
# Fitting a three class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m2_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m2_nfl)
```

```{r}
postprob(jlcmm_m2_nfl)
```

## Comparing M0, M1, M2

```{r}
#  Function for extracting the class proportions from each model
get_class_proportions <- function(model) {
#   Getting the number of classes
  G <- model$ng
#   Calculating the class proportions (number of individuals assigned to a class / number of inidividuals * 100)
  props <- table(model$pprob$class) / nrow(model$pprob) * 100
  
#   Creating a vector to store the proportions in, with NA for extra classes
  out <- rep(NA, G)
  out[1:G] <- as.numeric(props)
  names(out) <- paste0("Class", 1:G)
  
#   Returning results
  return(round(out, 2))
}

# Function for extracting entropy
get_entropy <- function(model) {
  G <- model$ng
  N <- nrow(model$pprob)
  
  if (G == 1) return(1.00)

  pprobs <- model$pprob[, 3:(2 + G)]
#   computing entropy
  entropy <- -sum(pprobs * log(pprobs + 0.0001))
  
#   scaling to between 0-1
  entropy_r2 <- 1 - (entropy / (N * log(G)))
  
  return(round(entropy_r2, 3))
}
```

```{r}
models <- list(jlcmm_m0_1, jlcmm_m1_nfl, jlcmm_m2_nfl)

# Find max number of classes across all models
G_max <- max(sapply(models, function(m) m$ng))

# Extract proportions for each model, with NA for cells where the model has fewer classes than the given row
props <- do.call(rbind, lapply(models, function(m) {
  p <- get_class_proportions(m)
  padded <- c(p, rep(NA, G_max - length(p)))
  return(padded)
}))

colnames(props) <- paste0("Class", 1:G_max, " (%)")

summary_jlcmm_nfl <- data.frame(
  "Candidate model" = paste0("Model", 1:length(models)),
  "G" = sapply(models, function(m) m$ng),
  "Log Likelihood" = sapply(models, function(m) m$loglik),
  "BIC" = sapply(models, function(m) m$BIC),
  "Entropy" = sapply(models, get_entropy),
  props,
  check.names = FALSE
)
```

```{r}
summary(jlcmm_m1_nfl)
```

```{r}
# params_all <- do.call(rbind, lapply(names(models), function(m)
#   get_jlcmm_params(models[[nm]], model_name = m)
# ))
# 
# params_all %>%
#   dplyr::select(Model, parameter, estimate, lower, upper, P_value) %>%
#   mutate(
#     "95% CI" = paste0("(", Lower95, ", ", Upper95, ")"),
#     "P-value" = case_when(
#       P_value < 0.001 ~ "<0.001",
#       TRUE ~ as.character(P_value)
#     )
#   ) %>%
#   dplyr::select(Model, Parameter, Estimate, `95% CI`, `P-value`) %>%
#   kable(
#     booktabs = TRUE,
#     caption = "Parameter estimates from joint latent class models",
#     escape = FALSE
#   ) %>%
#   kable_styling(latex_options = c("hold_position", "scale_down")) %>%
#   # Group rows by model
#   pack_rows(index = table(params_all$Model))
```

## Class Descriptions

```{r}
# # saving which class each individual was predicted to be in
class_preds <- jlcmm_m1_nfl$pprobY

# merging with genotype and demographic data
class_preds <- class_preds %>%
  merge((apoe %>% dplyr::select(RID, APOE4_status)), by = "RID") %>% 
  merge((surv_data %>% dplyr::select(RID, event, age_at_baseline, PTGENDER)), by = "RID")

# checking the distribution of genotype by class
table(class_preds$class, class_preds$APOE4_status)
# checking the distribution of gender by class
table(class_preds$class, class_preds$PTGENDER)
# checking the distribution of age by class
class_preds %>%
  group_by(class) %>%
  summarise(
    mean = mean(age_at_baseline, na.rm = TRUE),
    median = median(age_at_baseline, na.rm = TRUE),
    q1 = quantile(age_at_baseline, 0.25, na.rm = TRUE),
    q3 = quantile(age_at_baseline, 0.75, na.rm = TRUE)
  )
```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m1_nfl, which = "survival")
```

### Longitudinal trajectory by class

```{r}
# longitudinal trajectory by class
plot(jlcmm_m1_nfl, var.time = "time_years", which = "fit",
     ylim = c(2.5,4.2))
```

# JLCMM: PTAU

## M0 (1 latent class)

```{r}
# Fitting a one class joint latent class mixed model with random intercept
jlcmm_m0_1 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)

# Fitting a one class joint latent class mixed model with flexible baseline hazard to check performance against parametric hazard (Weibull)
jlcmm_m0_1_piecewise <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  hazard = "3-quant-piecewise",
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_1)
```

```{r}
summary(jlcmm_m0_1_piecewise)
```

```{r}
# Fitting a one class joint latent class mixed model with a random intercept and slope
jlcmm_m0_2 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(jlcmm_m0_1, jlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = jlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = jlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Running gridsearch() for a two class model
# jlcmm_m1_ptau_search <- gridsearch(m = jlcmm(fixed = log_PLASMAPTAU181 ~ time_years,
#                                               random = ~ 1,
#                                               mixture = ~ 1,
#                                               classmb = ~ 1,
#                                               ng = 2,
#                                               subject = "RID",
#                                               survival = Surv(time_to_event, event) ~ 1,
#                                               data = jlcmm_data),
#                                     minit = jlcmm_m0_1,
#                                     rep = 50,
#                                     maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m1_ptau <- jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 2,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m1_ptau_search$best
# )
```

```{r}
# Fitting a two class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m1_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m1_ptau)
```

```{r}
postprob(jlcmm_m1_ptau)
```

```{r}
# Posterior probability table for the 2 class jlcmm for ptau
get_pprob_table(jlcmm_m1_ptau)
```

```{r}
# The model estimates in a table
get_jlcmm_params(jlcmm_m1_ptau)
```

## M2 (3 latent classes)

```{r}
# Running gridsearch() for a three class model
# jlcmm_m2_ptau_search <- gridsearch(m = jlcmm(fixed = log_PLASMAPTAU181 ~ time_years,
#                                               random = ~ 1,
#                                               mixture = ~ 1,
#                                               classmb = ~ 1,
#                                               ng = 3,
#                                               subject = "RID",
#                                               survival = Surv(time_to_event, event) ~ 1,
#                                               data = jlcmm_data),
#                                     minit = jlcmm_m0_1,
#                                     rep = 50,
#                                     maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m2_ptau <- jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 3,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m2_ptau_search$best
#   )
```

```{r}
# Fitting a three class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m2_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m2_ptau)
```

```{r}
# Checking the posterior probabilities for the 3 class model
postprob(jlcmm_m2_ptau)
```

```{r}
# Posterior probability table for the 3 class jlcmm for ptau
get_pprob_table(jlcmm_m2_ptau)
```

```{r}
# Checking flexible baseline hazard (piecewise)
jlcmm_m2_ptau_piecewise <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  hazard = "3-quant-piecewise",
  data = jlcmm_data,
  B = random(jlcmm_m0_1_piecewise)
)
```

```{r}
# Flexible hazard with two parameters has higher BIC than Weibull hazard
summary(jlcmm_m2_ptau_piecewise)
```

## M3 (4 latent classes)

```{r}
# Running gridsearch() for a four class model
# jlcmm_m3_ptau_search <- gridsearch(m = jlcmm(fixed = log_PLASMAPTAU181 ~ time_years,
#                                               random = ~ 1,
#                                               mixture = ~ 1,
#                                               classmb = ~ 1,
#                                               ng = 4,
#                                               subject = "RID",
#                                               survival = Surv(time_to_event, event) ~ 1,
#                                               data = jlcmm_data),
#                                     minit = jlcmm_m0_1,
#                                     rep = 50,
#                                     maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m3_ptau <- jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 4,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m3_ptau_search$best
#   )
```

```{r}
# Fitting a four class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m3_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 4,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m3_ptau)
```

```{r}
postprob(jlcmm_m3_ptau)
```

## Comparing M0, M1, M2, M3

```{r}
models <- list(jlcmm_m0_1, jlcmm_m1_ptau, jlcmm_m2_ptau, jlcmm_m3_ptau)

# Find max number of classes across all models
G_max <- max(sapply(models, function(m) m$ng))

# Extract proportions for each model, with NA for cells where the model has fewer classes than the given row
props <- do.call(rbind, lapply(models, function(m) {
  p <- get_class_proportions(m)
  padded <- c(p, rep(NA, G_max - length(p)))
  return(padded)
}))

colnames(props) <- paste0("Class", 1:G_max, " (%)")

summary_jlcmm_ptau <- data.frame(
  "Candidate model" = paste0("Model", 1:length(models)),
  "G" = sapply(models, function(m) m$ng),
  "Log Likelihood" = sapply(models, function(m) m$loglik),
  "BIC" = sapply(models, function(m) m$BIC),
  "Entropy" = sapply(models, get_entropy),
  props,
  check.names = FALSE
)
```

## Class Descriptions

```{r}
# # saving which class each individual was predicted to be in
class_preds <- jlcmm_m1_ptau$pprobY

# merging with genotype and demographic data
class_preds <- class_preds %>%
  merge((apoe %>% dplyr::select(RID, APOE4_status)), by = "RID") %>% 
  merge((surv_data %>% dplyr::select(RID, event, age_at_baseline, PTGENDER)), by = "RID")

# checking the distribution of genotype by class
table(class_preds$class, class_preds$APOE4_status)
# checking the distribution of gender by class
table(class_preds$class, class_preds$PTGENDER)
# checking the distribution of age by class
class_preds %>%
  group_by(class) %>%
  summarise(
    mean = mean(age_at_baseline, na.rm = TRUE),
    median = median(age_at_baseline, na.rm = TRUE),
    q1 = quantile(age_at_baseline, 0.25, na.rm = TRUE),
    q3 = quantile(age_at_baseline, 0.75, na.rm = TRUE)
  )
```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m2_ptau, which = "survival")
```

### Longitudinal trajectories by class

```{r}
# longitudnal trajectories by class
plot(jlcmm_m1_ptau, var.time = "time_years", which = "fit")
```

# MJLCMM: NFL & PTAU

## M0 (1 latent class)

```{r}
# Modelling the longitudinal data only, these will be used as arguments in the joint models
# Fitting a one class multivariate latent class mixed model with random intercept
mlcmm_m0_1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = jlcmm_data
)

# Fitting a one class multivariate latent class model with random intercept and slope
mlcmm_m0_2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = jlcmm_data
)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a one class multivariate joint latent class model with random intercept
mult_jlcmm_m0_1 <- mpjlcmm(
  longitudinal = list(mlcmm_m0_1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data
)
```

```{r}
summary(mult_jlcmm_m0_1)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a one class multivariate joint latent class model with random intercept and slope
mult_jlcmm_m0_2 <- mpjlcmm(
  longitudinal = list(mlcmm_m0_2),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data
)
```

```{r}
summary(mult_jlcmm_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(mult_jlcmm_m0_1, mult_jlcmm_m0_2)
```

## M1 (2 latent classes)

```{r}
# Modelling longitudinal data, used as argument in the joint model
# Fitting a multivariate latent class model with two latent classes
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mlcmm_m0_1)
)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a multivariate joint latent class model with two latent classes
mult_jlcmm_m1 <- mpjlcmm(
  longitudinal = list(mlcmm_m1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0_1)
)
```

```{r}
summary(mult_jlcmm_m1)
```

```{r}
postprob(mult_jlcmm_m1)
```

```{r}
# Posterior probability table for the 2 class jlcmm for ptau and NfL
get_pprob_table(mult_jlcmm_m1)
```

```{r}
# # Saving the model estimates in a table
# jlcmm_m1_ptau_nfl_output_midpoint <- get_jlcmm_params(mult_jlcmm_m1)
# saveRDS(jlcmm_m1_ptau_nfl_output_midpoint, "tables/jlcmm_m1_ptau_nfl_output_midpoint.rds")
```

## M2 (3 latent classes)

```{r}
# Modelling longitudinal data, used as argument in the joint model
# Fitting a three class multivariate latent class model
mlcmm_m2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mlcmm_m0_1)
)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a three class multivariate joint latent class model
mult_jlcmm_m2 <- mpjlcmm(
  longitudinal = list(mlcmm_m2),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0_1)
)
```

```{r}
summary(mult_jlcmm_m2)
```

```{r}
postprob(mult_jlcmm_m2)
```

## Comparing M0, M1, M2

```{r}
models <- list(mult_jlcmm_m0_1, mult_jlcmm_m1, mult_jlcmm_m2)

# Find max number of classes across all models
G_max <- max(sapply(models, function(m) m$ng))

# Extract proportions for each model, with NA for cells where the model has fewer classes than the given row
props <- do.call(rbind, lapply(models, function(m) {
  p <- get_class_proportions(m)
  padded <- c(p, rep(NA, G_max - length(p)))
  return(padded)
}))

colnames(props) <- paste0("Class", 1:G_max, " (%)")

summary_jlcmm_ptau_nfl <- data.frame(
  "Candidate model" = paste0("Model", 1:length(models)),
  "G" = sapply(models, function(m) m$ng),
  "Log Likelihood" = sapply(models, function(m) m$loglik),
  "BIC" = sapply(models, function(m) m$BIC),
  "Entropy" = sapply(models, get_entropy),
  props,
  check.names = FALSE
)
```

### Survival curves by class

```{r}
#plot(mult_jlcmm_m1, which = "survival")
```

```{r}
# plot(mult_jlcmm_m1, var.time = "time_years", which = "fit")
```

### Survival curves by class

```{r}
# survival curves by class
#plot(mult_jlcmm_m2, which = "survival")
```

```{r}
# # longitudnal trajectories by class
# plot(mult_jlcmm_m2, outcome = 1, which = "fit")
```

## Defining time points for evaluation of predictive performance

```{r}
# We define evaluation time points for time-dependent AUC comparison
times <- c(2, 3, 4, 5, 6)
```

# Time-dependent AUC

## Time-dependent AUC for fitted Cox PH models

```{r}
# We define a function to compute time-dependent AUC for Cox models using timeROC
get_cox_auc <- function(model, times) {
  
  T <- surv_data_out$time_to_event
  delta <- surv_data_out$event
  
  lp <- predict(model, newdata = surv_data_out, type = "lp")
  
  timeROC::timeROC(
    T = T,
    delta = delta,
    marker = lp,
    cause = 1,
    times = times,
    iid = TRUE
  )$AUC
}
```

```{r}
# We compute time-dependent AUC for Cox model with p-tau181
auc_ptau_bl <- get_cox_auc(surv_model_ptau_bl_final, times)
 
# # We compute time-dependent AUC for Cox model with NfL
auc_nfl_bl <- get_cox_auc(surv_model_nfl_bl_final, times)

# We compute time-dependent AUC for Cox model with both p-tau181 and NfL 
auc_both_bl <- get_cox_auc(surv_model_both_bl_final, times)
```

## Time-dependent AUC for fitted joint models

```{r}
# We construct prediction datasets for the joint models by merging the longitudinal and survival data
newdata <- dplyr::left_join(
  long_data_out,
  dplyr::select(surv_data_out, RID, age_at_baseline, time_to_event, event),
  by = "RID"
)
```

```{r}
# We define a function to compute time-dependent AUC for joint models using tvAUC
get_jm_auc <- function(model, times) {
  
  aucs <- sapply(times, function(t) {
  
    tmp <- tvAUC(
      model,
      newdata = newdata,
      Tstart = 1,
      Thoriz = t
    )
    
    tmp$auc
  })
  
  return(as.numeric(aucs))
}
```

```{r}
# We compute time-dependent AUC for the joint model including p-tau181 
auc_ptau_jm <- get_jm_auc(joint_model_ptau, times)

# We compute time-dependent AUC for the joint model including NfL 
auc_nfl_jm <- get_jm_auc(joint_model_nfl, times)

# We compute time-dependent AUC for the multivariate joint model 
auc_multi_jm <- get_jm_auc(multi_joint_model, times)

```

## Time-dependent AUC for latent class joint models

#### Function for getting predictions from multivariate joint latent class model

```{r}
# Defining a function to predict cumulative hazard (1 - S(t)) for multivariate joint latent class models (these cannot be passed to dynpred in the lcmm package)
predict_mjlcmm <- function(model, data, ID, landmark, times) {
  
#   Extracting the unique ID's
  ids <- unique(data[[ID]])
#   Extracting number of individuals to make predictions for
  N <- length(unique(data[[ID]]))
#   Extracting the nuber of classes
  G <- model$ng
#   Number of horizons
  n_h <- length(times)
  print(n_h)
  
#   Extracting the posterior probabilities (given a two class model)
  post_prob <- model$pprob[, 3:4]
  
#   Extracting the Weibull hazard parameters from the model object 
  scale <- sapply(1:G, function(g) {
                   exp(model$best[paste0("event1 +/-sqrt(Weibull2) class ", as.character(g))])})
  shape <- sapply(1:G, function(g) {
                   exp(model$best[paste0("event1 +/-sqrt(Weibull1) class ", as.character(g))])})
  
#   Initiating a matrix to save the predictions for each individual at each horizon
  S_pred <- matrix(0, nrow = N, ncol = n_h)
  
#   Saving the class-specific survivals at the landmark time
  S_land <- sapply(1:G, function(g)
    exp(-(landmark / scale[g])^shape[g]))
  
#   Looping over all indiivudals
  for(i in 1:N) {
    
#     GEtting the posterior probabilities for each individual
    pp_i <- post_prob[i, ]
    
#     Looping over the horizons
    for(j in 1:n_h) {
      t <- times[j]
    
#      Getting the conditional class-specific survivals
      Sg <- sapply(1:G, function(g)
        exp(-((t) / scale[g])^shape[g]) / S_land[g])

#      Calculating the cumulative hazard (1 - survival) for inidivudal i based on their posterior probabilities and the class-specific survival functions
      S_pred[i, j] <- 1 - sum(pp_i * Sg)
    }
  }  
#   Convert to long format matching dynpred() output structure
  S_pred <- as.data.frame(S_pred)
  colnames(S_pred) <- as.character(times)
  S_pred[[ID]] <- ids

  S_pred_long <- S_pred %>%
    pivot_longer(
      cols = -all_of(ID),
      names_to  = "horizon",
      values_to = "pred"
    ) %>%
    mutate(horizon = as.numeric(horizon))
  
  names(S_pred_long) <- c("RID", "horizon", "pred")

  return(S_pred_long)
  }
```

#### Function for extracting time-dependent AUC

1.  Confidence intervals

2.  Overall model discrimination

3.  Class specific discrimination

```{r}
# Defining a function to compute time dependent AUC for joint latent class models
jlcmm_timeAUC <- function(model, newdata, ID, time_var, surv_time_var, event_var,
                          landmark, times, cause = 1) {
  
#   Initiating a data frame to save the results in at each horizon time
  output <- data.frame(
    times = times,
    cases = NA,
    AUC = NA
  )

#   Calculating the survival predictions previously defined function
  if(class(model) == "mpjlcmm") {
    dfpred <- predict_mjlcmm(
      model = model,
      data = newdata,
      ID = ID,
      landmark = landmark,
      times = times
      )
  }
  
#   Calculating the dynamic survival predictions from joint models using dynpred() in the lcmm package
  if(class(model) == "Jointlcmm") {
    dfpred <- dynpred(
      model = model,
      newdata = newdata,
      var.time = time_var,
      horizon = times,
      landmark = landmark
      )$pred
  }
  
#   Fetching the true events for each ID for each time point
  dftrue <- unique(newdata[c(ID, surv_time_var, event_var)])
  names(dftrue) <- c(ID, "time", "event")
  
#   Merging the true and predicted data frames
  merged <- merge(dfpred, dftrue, by = ID)
  
#   Looping through the horizon times
  for(i in 1:length(times)) {
    h = times[i]
    
#     Filtering for the predictions from horizon h
    data_h <- merged %>% filter(horizon == h)
    
    #     Identifiying individuals that convert (cases) in the time between landmark and landmark + horizon (l + h) or remain disease free past l + h (controls)
    cases <- data_h$event == 1 & data_h$time > landmark & data_h$time <= h
    controls <- data_h$time > h
    data_h <- data_h[cases | controls, ]
    
    roc <- timeROC(
      T = data_h$time,
      delta = data_h$event,
      marker = data_h$pred,
      cause = cause,
      times = h,
      iid = TRUE)
    
    output$AUC[i] <- roc$AUC[2]
#     Checking the events in the time window between l and l + h
    output$cases[i] <- sum(cases)
    
  }
  return(output)
}
```

```{r}
# Computing AUC for 2 class joint latent class mixed model for NfL
auc_jlcmm_m1_nfl <- jlcmm_timeAUC(jlcmm_m1_nfl, newdata,
                                  "RID", "time_years", "time_to_event", "event",
                                  1, times)

# Computing AUC for 2 class joint latent class mixed model for ptau
auc_jlcmm_m1_ptau <- jlcmm_timeAUC(jlcmm_m1_ptau, newdata,
                                   "RID", "time_years", "time_to_event", "event",
                                   1, times)

# # Computing AUC for 3 class joint latent class mixed model for ptau
# auc_jlcmm_m2_ptau <- jlcmm_timeAUC(jlcmm_m2_ptau, newdata,
#                                    "RID", "time_years", "time_to_event", "event",
#                                    1, times)
# 
# # ON IN-SAMPLE DATA
# # Computing AUC for 2 class multivariate joint latent class mixed model for ptau & NfL
# auc_mult_jlcmm_m1 <- jlcmm_timeAUC(mult_jlcmm_m1, jlcmm_data,
#                                    "RID", "time_years", "time_to_event", "event",
#                                    1, times)
```

# Time-depending Brier score

## Time-dependent Brier score for fitted Cox PH models

```{r}
# We specify Cox models for comparison
cox_models <- list(
   "Cox p-tau" = surv_model_ptau_bl_final,
   "Cox NfL" = surv_model_nfl_bl_final,
   "Cox both" = surv_model_both_bl_final
 )
 
# We compute Brier score using pec for Cox models
brier_cox <- pec(
   object = cox_models,
   formula = Surv(time_to_event, event) ~ 
     age_at_baseline + PTGENDER + APOE4_status,
   data = surv_data_out,
   times = times,
   cens.model = "cox"
 )
 
cox_brier <- lapply(brier_cox$AppErr, function(x)
   approx(brier_cox$time, x, xout = times)$y
 )
```

## Time-dependent Brier score for fitted joint models

```{r}
# We define a function to compute time-dependent Brier score for joint models using tvBrier
get_jm_brier <- function(model, times) {
  
  briers <- sapply(times, function(t) {
    
    tmp <- tvBrier(
      model,
      newdata = newdata,
      Tstart = 0,
      Thoriz = t
    )
    
    return(tmp$Brier)
  })
  
  return(as.numeric(briers))
}
```

```{r}
# We compute time-dependent Brier score for the joint model including p-tau181 
brier_ptau_jm <- get_jm_brier(joint_model_ptau, times)
brier_ptau_jm

# We compute time-dependent Brier score for the joint model including NfL 
brier_nfl_jm <- get_jm_brier(joint_model_nfl, times)
brier_nfl_jm

# We compute time-dependent Brier score for the multivariate joint model 
brier_multi_jm <- get_jm_brier(multi_joint_model, times)
brier_multi_jm
```

## Time-dependent Brier score for latent class joint models

```{r}
brier_jlcmm <- function(model, data, ID, time_var, surv_time_var, event_var, times, landmark) {
  
#   Initiating a data frame to save the results in at each horizon time
  output <- data.frame(
    times = times,
    Brier = NA,
    IBS = NA
  )
  
#   Calculating the survival predictions previously defined function
  if(class(model) == "mpjlcmm") {
    dfpred <- predict_mjlcmm(
      model = model,
      data = data,
      ID = ID,
      landmark = landmark,
      times = times
      )
  }
  
#   Calculating the dynamic survival predictions from joint models using dynpred() in the lcmm package
  if(class(model) == "Jointlcmm") {
    dfpred <- dynpred(
      model = model,
      newdata = newdata,
      var.time = time_var,
      horizon = times,
      landmark = landmark
      )$pred
  }
  
#   Fetching the true event times and events 
  dftrue <- unique(data[c(ID, surv_time_var, event_var)])
  names(dftrue) <- c(ID, "time", "event")
  
#   Merging prediction df with true df
  merged <- merge(dfpred, dftrue, by = ID)
  
#   Saving censoring indicator (1 - event indicator), times remain the same as event/censured times from the event indicator
  merged$cens_event <- 1 - merged$event
  
#  Estimting the censoring distribution with the Kaplan-Meier estimate
  km_cens <- survfit(Surv(merged$time, merged$cens_event) ~ 1)
  
#   Function to evaluate estimate of censuring probability
  Ghat <- function(t) {
    sapply(t, function(x) {
      idx <- max(which(km_cens$time <= x))
      if (length(idx) == 0) return(1)
      km_cens$surv[idx]
    })
  }
  
  brier_one_time <- function(t) {
    
#    Saving all rows belonging to horizon t
    sub <- merged[merged$horizon == t, ]
#    Saving the observed survival status
    status_t <- as.numeric(sub$time > t)
#    Saving the predicted survival probability: 1 - pred since dynpred outputs cumulative risk (event probability) instead of survival probabilities 
    pred_t <- 1 - sub$pred

#    Calculating the IPCW weights
    w <- ifelse(sub$time > t,
                1 / Ghat(t),
                ifelse(sub$event == 1 & sub$time <= t,
                       1 / Ghat(sub$time),
                       0))
    
#     Calculating the IPCW-adjusted Brier score
    mean(w * (status_t - pred_t)^2, na.rm = TRUE)
  }
  
#   Calculating the Brier score for each horizon
  brier_scores <- sapply(times, brier_one_time)
#   Saving to output
  output$Brier <- brier_scores
  
#   Integrated Brier Score (IBS)
  IBS <- sum(diff(times) *
               (brier_scores[-1] + brier_scores[-length(brier_scores)]) / 2) /
         (max(times) - landmark)
#   Saving to output
  output$IBS <- IBS
  
  return(output)
}
```

```{r}
# Computing Brier scores for 2 class jlcmm for NfL
brier_jlcmm_m1_nfl <- brier_jlcmm(
  model = jlcmm_m1_nfl,
  data = newdata,
  ID = "RID",
  time_var = "time_years",
  surv_time_var = "time_to_event",
  event_var = "event",
  times = times,
  landmark = 1
)

# Computing Brier scores for 2 class jlcmm for ptau
brier_jlcmm_m1_ptau <- brier_jlcmm(
  model = jlcmm_m1_ptau,
  data = newdata,
  ID = "RID",
  time_var = "time_years",
  surv_time_var = "time_to_event",
  event_var = "event",
  times = times,
  landmark = 1
)

# ON IN-SAMPLE DATA 
# Computing Brier scores for 2 class multivariate jlcmm for ptau & NfL
brier_mult_jlcmm_m1 <- brier_jlcmm(
  model = mult_jlcmm_m1,
  data = jlcmm_data,
  ID = "RID",
  time_var = "time_years",
  surv_time_var = "time_to_event",
  event_var = "event",
  times = times,
  landmark = 1
)
```

# Tables of the predictive performance results

## Time-depending AUC 

```{r}
# Time-dependent AUC for Cox and joint models summarized in a table
results_table_AUC <- data.frame(
  Model = c("Cox ptau-181", "Cox NfL", "Cox both",
            "JM ptau-181", "JM NfL", "JM both", "JLCM ptau-181", "JLCM NfL"),
  
  AUC_2y = c(auc_ptau_bl[1],
             auc_nfl_bl[1],
             auc_both_bl[1],
             auc_ptau_jm[1],
             auc_nfl_jm[1],
             auc_multi_jm[1],
             auc_jlcmm_m1_ptau[1,3],
             auc_jlcmm_m1_nfl[1,3]),
  
  AUC_3y = c(auc_ptau_bl[2],
             auc_nfl_bl[2],
             auc_both_bl[2],
             auc_ptau_jm[2],
             auc_nfl_jm[2],
             auc_multi_jm[2],
             auc_jlcmm_m1_ptau[2,3],
             auc_jlcmm_m1_nfl[2,3]),
  
  AUC_4y = c(auc_ptau_bl[3],
             auc_nfl_bl[3],
             auc_both_bl[3],
             auc_ptau_jm[3],
             auc_nfl_jm[3],
             auc_multi_jm[3],
             auc_jlcmm_m1_ptau[3,3],
             auc_jlcmm_m1_nfl[3,3]),
  
  AUC_5y = c(auc_ptau_bl[4],
             auc_nfl_bl[4],
             auc_both_bl[4],
             auc_ptau_jm[4],
             auc_nfl_jm[4],
             auc_multi_jm[4],
             auc_jlcmm_m1_ptau[4,3],
             auc_jlcmm_m1_nfl[4,3]),
  
  AUC_6y = c(auc_ptau_bl[5],
             auc_nfl_bl[5],
             auc_both_bl[5],
             auc_ptau_jm[5],
             auc_nfl_jm[5],
             auc_multi_jm[5],
             auc_jlcmm_m1_ptau[5,3],
             auc_jlcmm_m1_nfl[5,3])
)

results_table_AUC[, -1] <- round(results_table_AUC[, -1], 3)
results_table_AUC
```

## Time-depending Brier score 

```{r}
# Time-dependent Brier score for Cox and joint models summarized in a table
results_table_Brier <- data.frame(
  Model = c("Cox ptau-181", "Cox NfL", "Cox both",
            "JM ptau-181", "JM NfL", "JM both",
            "JLCMM ptau-181", "JLCMM NfL"),
  
  Brier_2y = c(cox_brier$`Cox p-tau`[1],
               cox_brier$`Cox NfL`[1],
               cox_brier$`Cox both`[1],
               brier_ptau_jm[1],
               brier_nfl_jm[1],
               brier_multi_jm[1],
               brier_jlcmm_m1_ptau$Brier[1],
               brier_jlcmm_m1_nfl$Brier[1]),
  
  Brier_3y = c(cox_brier$`Cox p-tau`[2],
               cox_brier$`Cox NfL`[2],
               cox_brier$`Cox both`[2],
               brier_ptau_jm[2],
               brier_nfl_jm[2],
               brier_multi_jm[2],
               brier_jlcmm_m1_ptau$Brier[2],
               brier_jlcmm_m1_nfl$Brier[2]),
  
  Brier_4y = c(cox_brier$`Cox p-tau`[3],
               cox_brier$`Cox NfL`[3],
               cox_brier$`Cox both`[3],
               brier_ptau_jm[3],
               brier_nfl_jm[3],
               brier_multi_jm[3],
               brier_jlcmm_m1_ptau$Brier[3],
               brier_jlcmm_m1_nfl$Brier[3]),
  
  Brier_5y = c(cox_brier$`Cox p-tau`[4],
               cox_brier$`Cox NfL`[4],
               cox_brier$`Cox both`[4],
               brier_ptau_jm[4],
               brier_nfl_jm[4],
               brier_multi_jm[4],
               brier_jlcmm_m1_ptau$Brier[4],
               brier_jlcmm_m1_nfl$Brier[4]),
  
  Brier_6y = c(cox_brier$`Cox p-tau`[5],
               cox_brier$`Cox NfL`[5],
               cox_brier$`Cox both`[5],
               brier_ptau_jm[5],
               brier_nfl_jm[5],
               brier_multi_jm[5],
               brier_jlcmm_m1_ptau$Brier[5],
               brier_jlcmm_m1_nfl$Brier[5])
)

results_table_Brier[, -1] <- round(results_table_Brier[, -1], 3)
results_table_Brier
```

## Saving the tables 

### Original tables 

```{r}
# Creating a directory to save tables into
dir.create("tables", showWarnings = FALSE)

# Saving the AUC results table
saveRDS(results_table_AUC, "tables/results_table_AUC.rds")

# Saving the table of the Brier score results
saveRDS(results_table_Brier, "tables/results_table_Brier.rds")
```

### Sensitivity analysis tables

#### Baseline is the first biomarker examdate 

```{r}
#saveRDS(results_table_AUC, "tables/results_table_AUC_first_biomarker_bl.rds")
#saveRDS(results_table_Brier, "tables/results_table_Brier_first_biomarker_bl.rds")B
```

#### Baseline is the first diagnosis examdate 

```{r}
#saveRDS(results_table_AUC, "tables/results_table_AUC_first_dx_bl.rds")
#saveRDS(results_table_Brier, "tables/results_table_Brier_first_dx_bl.rds")
```

#### The duplicate biomarker measurements are replaced by the first measurement

```{r}
#saveRDS(results_table_AUC, "tables/results_table_AUC_first_duplicate.rds")
#saveRDS(results_table_Brier, "tables/results_table_Brier_first_duplicate.rds")
```

### The duplicate biomarker measurements are replaced by the last measurement

```{r}
#saveRDS(results_table_AUC, "tables/results_table_AUC_last_duplicate.rds")
#saveRDS(results_table_Brier, "tables/results_table_Brier_last_duplicate.rds")
```

### Including participants with at least one biomarker measurement 

```{r}
#saveRDS(results_table_AUC, "tables/results_table_AUC_full.rds")
#saveRDS(results_table_Brier, "tables/results_table_Brier_full.rds")
```

# Plots of the predictive performance results

```{r}
# We reshape all AUC results into long format for plotting
plot_data <- data.frame(
  Time = rep(times, 8),
  Model = rep(c("Cox ptau-181", "Cox NfL", "Cox both",
                "JM ptau-181", "JM NfL", "JM multi", "JLCM ptau-181", "JLCM NfL"),
              each = length(times)),
  AUC = c(
    auc_ptau_bl,
    auc_nfl_bl,
    auc_both_bl,
    auc_ptau_jm,
    auc_nfl_jm,
    auc_multi_jm,
    auc_jlcmm_m1_ptau$AUC,
    auc_jlcmm_m1_nfl$AUC
  )
) 
```

```{r}
# We reshape all Brier score results into long format for plotting
plot_data_Brier <- data.frame(
  Time = rep(times, 8),
  Model = rep(c("Cox ptau-181", "Cox NfL", "Cox both",
                "JM ptau-181", "JM NfL", "JM multi",
                "JLCMM ptau-181", "JLCMM NfL"),
              each = length(times)),
  
  Brier = c(
    cox_brier$`Cox p-tau`,
    cox_brier$`Cox NfL`,
    cox_brier$`Cox both`,
    
    brier_ptau_jm,
    brier_nfl_jm,
    brier_multi_jm,
  
    brier_jlcmm_m1_ptau$Brier,
    brier_jlcmm_m1_nfl$Brier
  )
)
```

## For comparing modelling approaches

### Time-dependent AUC

```{r}
# Q1: Biomarker fixed (p-tau181) - comparing models (time-dependent AUC)
ggplot(data.frame(
  Time = times,
  Cox = auc_ptau_bl,
  JM = auc_ptau_jm,
  JLCMM = auc_jlcmm_m1_ptau$AUC
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Model",
    shape = "Model"
    ) +
  theme_minimal()
```

```{r}
# Creating a folder to save plots
dir.create("figures", showWarnings = FALSE)

# Saving the plot 
png("figures/q1_auc_ptau.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  Cox = auc_ptau_bl,
  JM = auc_ptau_jm,
  JLCMM = auc_jlcmm_m1_ptau$AUC
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Model",
    shape = "Model"
    ) +
  theme_minimal()
dev.off()
```

```{r}
# Q1: NfL - comparing models (time-dependent AUC)
ggplot(data.frame(
  Time = times,
  Cox = auc_nfl_bl,
  JM = auc_nfl_jm,
  JLCMM = auc_jlcmm_m1_nfl$AUC
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Model",
    shape = "Model"
    ) +
  theme_minimal()
```

```{r}
# Saving the plot 
png("figures/q1_auc_nfl.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  Cox = auc_nfl_bl,
  JM = auc_nfl_jm,
  JLCMM = auc_jlcmm_m1_nfl$AUC
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Model",
    shape = "Model"
    ) +
  theme_minimal()
dev.off()
```

```{r}
# Q1: Both biomarkers - comparing models (time-dependent AUC)
ggplot(data.frame(
  Time = times,
  Cox = auc_both_bl,
  JM = auc_multi_jm
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Model",
    shape = "Model"
    ) +
  theme_minimal()
```

```{r}
# Saving the plot
png("figures/q1_auc_multi.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  Cox = auc_both_bl,
  JM = auc_multi_jm
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Model",
    shape = "Model"
    ) +
  theme_minimal()
dev.off()
```

### Time-dependent Brier score

```{r}
# Q1: Biomarker fixed (p-tau181) - comparing models (time-dependent Brier score)
ggplot(data.frame(
  Time = times,
  Cox = cox_brier$`Cox p-tau`,
  JM = brier_ptau_jm,
  JLCMM = brier_jlcmm_m1_ptau$Brier
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
```

```{r}
# Saving the plot 
png("figures/q1_brier_ptau.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  Cox = cox_brier$`Cox p-tau`,
  JM = brier_ptau_jm,
  JLCMM = brier_jlcmm_m1_ptau$Brier
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
dev.off()
```

```{r}
# Q1: NfL - comparing models (time-dependent Brier score)
ggplot(data.frame(
  Time = times,
  Cox = cox_brier$`Cox NfL`,
  JM = brier_nfl_jm,
  JLCMM = brier_jlcmm_m1_nfl$Brier
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
```

```{r}
# Saving the plot 
png("figures/q1_brier_nfl.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  Cox = cox_brier$`Cox NfL`,
  JM = brier_nfl_jm,
  JLCMM = brier_jlcmm_m1_nfl$Brier
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  geom_line(aes(Time, JLCMM, color = "JLCMM"), linewidth = 0.3) +
  geom_point(aes(Time, JLCMM, color = "JLCMM", shape = "JLCMM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16, "JLCMM" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
dev.off()
```

```{r}
# Q1: Both biomarkers - comparing models (time-dependent Brier score)
ggplot(data.frame(
  Time = times,
  Cox = cox_brier$`Cox both`,
  JM = brier_multi_jm
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
```

```{r}
# Saving the plot 
png("figures/q1_brier_multi.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  Cox = cox_brier$`Cox both`,
  JM = brier_multi_jm
)) +
  geom_line(aes(Time, Cox, color = "Cox PH"), linewidth = 0.3) +
  geom_point(aes(Time, Cox, color = "Cox PH", shape = "Cox PH"), size = 1.5) +
  
  geom_line(aes(Time, JM, color = "JM"), linewidth = 0.3) +
  geom_point(aes(Time, JM, color = "JM", shape = "JM"), size = 1.5) +
  
  scale_shape_manual(values = c("Cox PH" = 15, "JM" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Model",
    shape = "Model"
  ) +
  theme_minimal()
dev.off()
```

## For comparing biomarker specifications

### Time-dependent AUC

```{r}
# Q2: Cox PH - comparing biomarkers (time-dependent AUC)
ggplot(data.frame(
  Time = times,
  NfL = auc_nfl_bl,
  pTau = auc_ptau_bl,
  Multi = auc_both_bl
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariable"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariable", shape = "Multivariable"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariable" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Biomarker",
    shape = "Biomarker"
    ) +
  theme_minimal()
```

```{r}
# Saving the plot
png("figures/q2_auc_cox.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  NfL = auc_nfl_bl,
  pTau = auc_ptau_bl,
  Multi = auc_both_bl
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariable"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariable", shape = "Multivariable"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariable" = 17)) +
  ylim(0,1) +
  labs(
    title = "Time-depending AUC curves (Cox PH)",
    x = "Time", 
    y = "AUC",
    color = "Biomarker",
    shape = "Biomarker"
    ) +
  theme_minimal()
dev.off()
```

```{r}
# Q2: JM - comparing biomarkers (time-dependent AUC)
ggplot(data.frame(
  Time = times,
  NfL = auc_nfl_jm,
  pTau = auc_ptau_jm,
  Multi = auc_multi_jm
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariate"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariate", shape = "Multivariate"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariate" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Biomarker",
    shape = "Biomarker"
    ) +
  theme_minimal()
```

```{r}
# Saving the plot
png("figures/q2_auc_jm.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  NfL = auc_nfl_jm,
  pTau = auc_ptau_jm,
  Multi = auc_multi_jm
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariate"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariate", shape = "Multivariate"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariate" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Biomarker",
    shape = "Biomarker"
    ) +
  theme_minimal()
dev.off()
```

```{r}
# Q2: LCJM - comparing biomarkers (time-dependent AUC)
ggplot(data.frame(
  Time = times,
  NfL = auc_jlcmm_m1_nfl$AUC,
  pTau = auc_jlcmm_m1_ptau$AUC
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Biomarker",
    shape = "Biomarker"
    ) +
  theme_minimal()
```

```{r}
# Saving the plot
png("figures/q2_auc_lcjm.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  NfL = auc_jlcmm_m1_nfl$AUC,
  pTau = auc_jlcmm_m1_ptau$AUC
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "AUC",
    color = "Biomarker",
    shape = "Biomarker"
    ) +
  theme_minimal()
dev.off()
```

### Time-dependent Brier score

```{r}
# Q2: Cox PH - comparing biomarkers (time-dependent Brier score)
ggplot(data.frame(
  Time = times,
  NfL = cox_brier$`Cox NfL`,
  pTau = cox_brier$`Cox p-tau`,
  Multi = cox_brier$`Cox both`
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariable"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariable", shape = "Multivariable"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariable" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Biomarker",
    shape = "Biomarker"
  ) +
  theme_minimal()
```

```{r}
# Saving the plot 
png("figures/q2_brier_cox.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  NfL = cox_brier$`Cox NfL`,
  pTau = cox_brier$`Cox p-tau`,
  Multi = cox_brier$`Cox both`
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariable"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariable", shape = "Multivariable"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariable" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Biomarker",
    shape = "Biomarker"
  ) +
  theme_minimal()
dev.off()
```

```{r}
# Q2: JM - comparing biomarkers (time-dependent AUC)
ggplot(data.frame(
  Time = times,
  NfL = brier_nfl_jm,
  pTau = brier_ptau_jm,
  Multi = brier_multi_jm
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariate"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariate", shape = "Multivariate"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariate" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Biomarker",
    shape = "Biomarker"
  ) +
  theme_minimal()
```

```{r}
# Saving the plot 
png("figures/q2_brier_jm.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  NfL = brier_nfl_jm,
  pTau = brier_ptau_jm,
  Multi = brier_multi_jm
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  geom_line(aes(Time, Multi, color = "Multivariate"), linewidth = 0.3) +
  geom_point(aes(Time, Multi, color = "Multivariate", shape = "Multivariate"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16, "Multivariate" = 17)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Biomarker",
    shape = "Biomarker"
  ) +
  theme_minimal()
dev.off()
```

```{r}
# Q2: LCJM - comparing biomarkers (time-dependent Brier score)
ggplot(data.frame(
  Time = times,
  NfL = brier_jlcmm_m1_nfl$Brier,
  pTau = brier_jlcmm_m1_ptau$Brier
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Biomarker",
    shape = "Biomarker"
  ) +
  theme_minimal()
```

```{r}
# Saving the plot 
png("figures/q2_brier_lcjm.png",
    width = 6, height = 4, units = "in", res = 300)
ggplot(data.frame(
  Time = times,
  NfL = brier_jlcmm_m1_nfl$Brier,
  pTau = brier_jlcmm_m1_ptau$Brier
)) +
  geom_line(aes(Time, NfL, color = "NfL"), linewidth = 0.3) +
  geom_point(aes(Time, NfL, color = "NfL", shape = "NfL"), size = 1.5) +
  
  geom_line(aes(Time, pTau, color = "p-tau181"), linewidth = 0.3) +
  geom_point(aes(Time, pTau, color = "p-tau181", shape = "p-tau181"), size = 1.5) +
  
  scale_shape_manual(values = c("NfL" = 15, "p-tau181" = 16)) +
  ylim(0,1) +
  labs(
    x = "Time", 
    y = "Brier score",
    color = "Biomarker",
    shape = "Biomarker"
  ) +
  theme_minimal()
dev.off()
```

# Saving other relevant tables and figures

```{r}
#  Saving log transformed ptau histogram
png("figures/hist_log_p_tau.png",
    width = 6, height = 4, units = "in", res = 300)
hist(long_data$log_PLASMAPTAU181,
     xlab = "Log-transformed p-tau181")
dev.off()
```

```{r}
#  Saving log transformed NfL histogram
png("figures/hist_log_nfl.png",
    width = 6, height = 4, units = "in", res = 300)
hist(long_data$log_PLASMA_NFL,
     xlab = "Log-transformed NfL" 
     )
dev.off()
```

```{r}
# Saving the correlation plots 
png("figures/corrplot_cox.png", width = 6, height = 4, units = "in", res = 300)
corrplot(cox_cor,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)
dev.off()

png("figures/corrplot_lme.png", width = 6, height = 4, units = "in", res = 300)
corrplot(lme_cor,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)
dev.off()
```

```{r}
# Saving the table of results from the likelihood ratio tests between models with only a ranodom intercept and those with random intercept and slope
saveRDS(long_re_lrt_table, "tables/long_re_lrt_table.rds")
```

```{r}
# Saving the plot of residuals vs fitted for ptau
png("figures/residuals_vs_fitted_ptau.png",
    width = 6, height = 4, units = "in", res = 300)
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for p-tau181")
abline(h = 0, col = "red", lty = 2)
dev.off()
```

```{r}
# Saving the residuals vs fitted plot for nfl
png("figures/residuals_vs_fitted_nfl.png",
    width = 6, height = 4, units = "in", res = 300)
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NfL")
abline(h = 0, col = "red", lty = 2)
dev.off()
```

```{r}
#  Saving the QQ-plots
png("figures/qq_ptau.png",
    width = 6, height = 4, units = "in", res = 300)
qqnorm(res_ptau, main = "QQ plot: p-tau181 residuals")
qqline(res_ptau, col = "red")
dev.off()

png("figures/qq_nfl.png",
    width = 6, height = 4, units = "in", res = 300)
qqnorm(res_nfl, main = "QQ plot: NfL residuals")
qqline(res_nfl, col = "red")
dev.off()
```

```{r}
# Saving table with PHT results
saveRDS(table_ph, "tables/table_ph.rds")
```

```{r}
# Saving the final PH table
saveRDS(table_ph_final, "tables/table_ph_final.rds")
```

```{r}
# Saving a table of results from the likelihood ratio tests for cox models with and without PTGENDER
saveRDS(surv_lrt_table, "tables/surv_lrt_table.rds")
```

```{r}
# Savig the cox models comparison table
saveRDS(table_cox, "tables/table_cox1.rds")
```

```{r}
# Saving the C-index table
saveRDS(c_index_table, "tables/c_index_table.rds")
```

```{r}
# Saving the convergence plot for the p-tau181 joint model
png("figures/convergence_plot_ptau.png",
    width = 6, height = 4, units = "in", res = 300)
ggtraceplot(joint_model_ptau, "alphas")
dev.off()
```

```{r}
# Saving the convergence plot for the NfL joint model
png("figures/convergence_plot_nfl.png",
    width = 6, height = 4, units = "in", res = 300)
ggtraceplot(joint_model_nfl, "alphas")
dev.off()
```

```{r}
# Saving the convergence plot for the multivariate joint model
png("figures/convergence_plot_multi.png",
    width = 6, height = 4, units = "in", res = 300)
 ggtraceplot(multi_joint_model, "alphas")
 dev.off()
 
 # Note: Since the function does not allow direct extraction of both plots, one is saved here, # while the other was captured via screenshot and added manually.
```

```{r}
# Saving the joint model comparison table
saveRDS(table_jm, "tables/table_jm1.rds")
```

```{r}
# Saving the rhat table (survival outcome)
saveRDS(rhat_survival_table, "tables/rhat_survival_table.rds")
```

```{r}
# Saving the rhat table (longitudinal outcome)
saveRDS(rhat_longitudinal_table, "tables/rhat_longitudinal_table.rds")
```

```{r}
# Saving the posterior probability table for the 2 class jlcmm for nfl
pprob_m1_nfl <- get_pprob_table(jlcmm_m1_nfl)
saveRDS(pprob_m1_nfl, "tables/pprob_m1_nfl.rds")
```

```{r}
# Saving the model estimates in a table
jlcmm_m1_nfl_output <- get_jlcmm_params(jlcmm_m1_nfl)
saveRDS(jlcmm_m1_nfl_output, "tables/jlcmm_m1_nfl_output.rds")
```

```{r}
# Saving comparison table of fit statistics (NfL) 
saveRDS(summary_jlcmm_nfl, "tables/summary_jlcmm_nfl.rds")
```

```{r}
# Survival curves by class (NfL) 
png("figures/class-specific-survival-jlcmm-nfl.png",
    width = 6, height = 4, units = "in", res = 300)
plot(jlcmm_m1_nfl, which = "survival",
     main = "",
     xlab = "Time (years)", ylab = "Survival Probability")
dev.off()
```

```{r}
# Longitudinal trajectory by class (NfL) 
png("figures/longitudinal-jlcmm-nfl.png",
    width = 6, height = 4, units = "in", res = 300)
plot(jlcmm_m1_nfl, var.time = "time_years", which = "fit",
     main = "",
     xlab = "Time (years)", ylab = "NfL")
dev.off()
```

```{r}
# Saving the posterior probability table for the 2 class jlcmm for p-tau181
pprob_m1_ptau <- get_pprob_table(jlcmm_m1_ptau)
saveRDS(pprob_m1_ptau, "tables/pprob_m1_ptau.rds")
```

```{r}
# Saving the model estimates in a table (p-tau181)
jlcmm_m1_ptau_output <- get_jlcmm_params(jlcmm_m1_ptau)
saveRDS(jlcmm_m1_ptau_output, "tables/jlcmm_m1_ptau_output.rds")
```

```{r}
# Saving the posterior probability table for the 3 class jlcmm for p-tau181
pprob_m2_ptau <- get_pprob_table(jlcmm_m2_ptau)
saveRDS(pprob_m2_ptau, "tables/pprob_m2_ptau.rds")
```

```{r}
# Comparing M0, M1, M2 (p-tau181)
saveRDS(summary_jlcmm_ptau, "tables/summary_jlcmm_ptau.rds")
```

```{r}
# Survival curves by class (p-tau181)
png("figures/class-specific-survival-jlcmm-ptau.png",
    width = 6, height = 4, units = "in", res = 300)
plot(jlcmm_m1_nfl, which = "survival",
     main = "",
     xlab = "Time (years)", ylab = "Survival Probability")
dev.off()
```

```{r}
# Longitudinal trajectories by class (p-tau181)
png("figures/longitudinal-jlcmm-ptau.png",
    width = 6, height = 4, units = "in", res = 300)
plot(jlcmm_m1_ptau, var.time = "time_years", which = "fit",
     main = "",
     xlab = "Time (years)", ylab = "ptau-181")
dev.off()
```

```{r}
# Saving the posterior probability table for the 2 class jlcmm for ptau and NfL
pprob_m1_ptau_nfl <- get_pprob_table(mult_jlcmm_m1)
saveRDS(pprob_m1_ptau_nfl, "tables/pprob_m1_ptau_nfl.rds")
```

```{r}
# Comparing M0, M1, M2 (NfL and p-tau181)
saveRDS(summary_jlcmm_ptau_nfl, "tables/summary_jlcmm_ptau_nfl.rds")
```
