---
format:
  pdf:
    include-before-body:
      text: |
        \begin{titlepage}
        \raggedleft
        \vspace*{2cm}
        
        {\Huge\bfseries Kandidatuppsats\par}
        \vspace{1.5cm}
        
        {\Large Statistiska Instiutionen\par}
        \vspace{3cm}
        
        {\Huge\bfseries Nr 2026:x\par}
        \vspace{0.5cm}
        
        {\LARGE \bfseries Code \par}
        
        {\large \today\par}
        \vspace{3cm}
        
        {\Huge Charlotte Emily Iszatt and Milou Rossi\par}
        
        \vfill
        
        \end{titlepage}
        
    toc: false
    colorlinks: false
    number-sections: true
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, output = TRUE)
```

# Pre-processing

```{r, warning=FALSE}
# Loading libraries
library(dplyr)
library(tidyr)
library(corrplot)
library(JMbayes2)
library(JM)
library(nlme) 
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) 
library(lcmm)
library(timeROC)
library(kableExtra)
```

```{r}
# Set seed for reproducibility
set.seed(123)
```

## Loading the data

The data was loaded from the ADNIMERGE2 package accessed via the Alzheimer's Disease Neuroimaging Initiative (ADNI). The data sets include longitudinal ptau-181 and NfL measurements, diagnosis, demographic and genetic data.

```{r}
# Loading data
# Milou
# NfL data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
  nfl_data <- BLENNOWPLASMANFLLONG
# # # P-tau181 data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
  ptau_data <- UGOTPTAU181
# # # Diagnosis data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
  diagnosis_data <- DXSUM
# # # Demographics data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
  demographics <- PTDEMOG
# # # Apoe data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
  apoe <- APOERES
```

```{r}
# Charley
#library(ADNIMERGE2)
#nfl_data = BLENNOWPLASMANFLLONG
#ptau_data = UGOTPTAU181
#diagnosis_data = DXSUM
#demographics = PTDEMOG
#apoe = APOERES
```

- *There are 52 individuals with two measurements recorded on the same exam date.*

- *How should we deal with these? Average the measurements? Remove the observations?*

1.  *Describe the measurements, their differences and averages*

- *Some of the values are quite different, what do we assume by averaging them? Do we introduce bias but just just removing them?*

2.  *Show how different these values are from the nfl and ptau averages (report SD differences in log transformations)*
3.  

```{r}
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", diagnosis_data$PTID)
grep("381_S_10###", apoe$PTID)
```

## Selecting required variables

Variables for the unique identifier for each individual, visit code, exam date, biomarker measurements, diagnosis and genetic data were extracted.

```{r, warning=FALSE}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required genotype variables
apoe <- apoe %>%
  dplyr::select(RID, GENOTYPE)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))
```

## Checking RID with more than one measurement recorded on the same exam date

For participants with duplicate measurements on the same examination date, differences were observed depending on how they were handled: using the first measurement, the mean, or the last measurement. The mean value was used as the primary approach, and sensitivity analyses were performed to evaluate the robustness of the findings.

```{r}
## Checking which obs have double VISCODE2 codes
# for ptau data
ptau_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# for nfl data
nfl_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# changing VISCODE2 for RID 1097 to m60_1 and m60_2
ptau_data[ptau_data$RID == 1097, "VISCODE2"] <- c("m48", "m60_2", "m60_1")

nfl_data[nfl_data$RID == 1097, "VISCODE2"] <-  c("m48", "m60_1", "m60_2")
```

```{r}
# Replacing visits with double measurements with visits with mean-value of the measurements as well as saving both measurements for future sensitivity analysis
# Naming the mean as the original variable to do analysis first with this method and then sensitivity analysis with the first and last measurements

# Adding columns for each ptau measurement and their mean and then collapsing down to unique RID and EXAMDATE
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_ptau = first(PLASMAPTAU181),
    last_ptau = last(PLASMAPTAU181),
    PLASMAPTAU181 = mean(PLASMAPTAU181),
    ) %>%
  ungroup() %>%
  distinct()

# Adding columns for each nfl measurement and their mean and then collapsing down to unique RID and EXAMDATE
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_nfl = first(PLASMA_NFL),
    last_nfl = last(PLASMA_NFL),
    PLASMA_NFL = mean(PLASMA_NFL)
    ) %>%
  ungroup() %>%
  distinct()
```

```{r}
# checking the difference in method for those with 2 obs on the same examdate
# for ptau
head(ptau_data %>%
       filter(PLASMAPTAU181 != first_ptau) %>%
       dplyr::select(PLASMAPTAU181, first_ptau, last_ptau)
     )

# for nfl
head(nfl_data %>%
       filter(PLASMA_NFL != first_nfl) %>%
       dplyr::select(PLASMA_NFL, first_nfl, last_nfl)
     )
```

## Merging NfL and ptau-181 data for the longitudinal data set

```{r}
# We merge NFL and ptau data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))
```

## Creating a common EXAMDATE and EXAMDATE_baseline variable

This was performed before preparation of the other variables for the survival and longitudinal models in order to use EXAMDATE_baseline to remove any observations from the diagnosis data set which occur long before the first biomarker measurements for that individual.

```{r}
# checking how many observations where the EXAMDATE is different for nfl and ptau
diff <- long_data %>% mutate(diff = as.numeric(EXAMDATE_nfl != EXAMDATE_ptau))
table(diff$diff)
```

```{r}
# filtering out the observations with different EXAMDATE
diff <- diff %>% filter(diff == TRUE)

# checking how many of these are baseline visits
table(diff$VISCODE2)

# checking which observations have the same examdate for nfl and ptau
common <- which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# checking the observations with different exam dates
long_data[-common, ]
```

- There are 5 observations with time differences greater than 30 days

- 231/115 days: clearly we can't say these are the same date... BUT with setting a new EXAMDATE as the midpoint between dates means the new EXAMDATE is 2 months out from EXAMDATE_nfl and EXAMDATE_ptau

- Less clear what the case is with a time difference of 31 days, even 49 days

```{r}
# calculating the difference in time between EXAMDATE_nfl and EXAMDATE_ptau
diff <- diff %>% mutate(time_diff = EXAMDATE_nfl - EXAMDATE_ptau)

# fetching the time differences
diff$time_diff 

# checking how many observations have a time difference over 30 days (1 month)
diff %>% filter(abs(time_diff) > 30) %>% dplyr::select(time_diff)
```

### EXAMDATE is EXAMDATE_nfl or EXAMDATE_nfl +- time difference

- Can create a new variable for a common EXAMDATE in order to define the baseline date from this

```{r}
# creating a common EXAMDATE from two cases
# 1: EXAMDATE_nfl == EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl
# 2: EXAMDATE_nfl != EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl + the difference between the two dates
# then defining the baseline date from the earliest EXAMDATE 
long_data <- long_data %>%
  group_by(RID) %>%
  mutate(
    EXAMDATE = EXAMDATE_ptau + 0.5 * (EXAMDATE_nfl - EXAMDATE_ptau),
    EXAMDATE_baseline = min(EXAMDATE),
    time_years = as.numeric(EXAMDATE - EXAMDATE_baseline) / 365.25) %>%
  ungroup()
```

### Baseline date is the earliest of EXAMDATE_nfl and EXAMDATE_ptau

```{r}
# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
# long_data <- long_data %>%
#   group_by(RID) %>%
#   mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
#   ungroup()
```

## Removing early diagnosis observations

Any diagnoses that were made more than 50 days before the first biomarker measurements were made were removed.

```{r}
# filtering the observations out of the diagnosis data set where EXAMDATE is much before baseline biomarker EXAMDATE
# creating a baseline_dx variable, so we can set time_to_event = 0 if baseline_dx == DEMENTIA
diagnosis_data <- diagnosis_data %>%
  merge(
    long_data %>% dplyr::select(RID, EXAMDATE_baseline), by = "RID"
    ) %>%
  filter(EXAMDATE_dx - EXAMDATE_baseline >= -50) %>%
  distinct() %>%
  group_by(RID) %>%
  arrange(EXAMDATE_dx) %>%
  mutate(
    baseline_dx = first(DIAGNOSIS_dx),
    last_dx = last(DIAGNOSIS_dx)
  ) %>%
  dplyr::select(!EXAMDATE_baseline)
```

## Merging longitudinal and diagnosis data for survival data set

```{r}
# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- long_data %>% left_join(diagnosis_data, by = c("RID"))
```

## Checking missing values

```{r}
# We count missing values per column in long_data: all these are variables we will use in the model, so we can remove with na.omit()
sapply(long_data, function(x) sum(is.na(x)))
```

```{r}
# We count missing values per column in surv_data: VISCODE2.y will not be used in the model, so we need to see how these observations look
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# There are no rows with missing VISCODE2.y with full data for the model variables, so we can remove all with na.omit()
surv_data %>% filter(is.na(VISCODE2.y))
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## Merging selected demographic variables (age and gender) and APOE4 status

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  dplyr::select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We define APOE4 status based on number of E4 alleles
apoe <- apoe %>%
  mutate(APOE4_status = as.factor(
    case_when(
      GENOTYPE == "4/4" ~ "2 alleles",
      GENOTYPE %in% c("2/4", "3/4") ~ "1 allele",
      TRUE ~ "0 alleles")),
  genotype = as.factor(GENOTYPE)
  )

apoe$APOE4_status <- factor(apoe$APOE4_status,
                            levels = c("0 alleles", "1 allele", "2 alleles"))
```

```{r}
# We merge gender with long_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We merge apoe with long_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status", "genotype")], by="RID")

# We make sure that gender is a factor
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

### Checking individual's final diagnoses for the full data set

```{r}
# final diagnoses for the full sample size
table(
  (surv_data %>%
     group_by(RID) %>%
     filter(EXAMDATE_dx == last(EXAMDATE_dx)) %>%
     dplyr::select(DIAGNOSIS_dx) %>%
     distinct()
   )$DIAGNOSIS_dx
  )
```

## Selecting participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements of ptau_data
# keep <- unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
# long_data <- long_data %>% filter(RID %in% keep)
# surv_data <- surv_data %>% filter(RID %in% keep)
# # We check number of individuals with at least 3 measurements
# length(keep)
```

## Selecting participants with at least 1 measurement

```{r}
# We filter individuals with at least 1 measurement of ptau_data
keep <- unique((ptau_data %>% group_by(RID) %>% filter(n() > 0))$RID)
long_data <- long_data %>% filter(RID %in% keep)
surv_data <- surv_data %>% filter(RID %in% keep)
# We check number of individuals with at least 1 measurement
length(keep)
```

### Checking individuals final diagnoses for the reduced data set

```{r}
# final diagnoses for the reduced sample size
table(
  (surv_data %>%
     group_by(RID) %>%
     filter(EXAMDATE_dx == last(EXAMDATE_dx)) %>%
     dplyr::select(DIAGNOSIS_dx) %>%
     distinct()
   )$DIAGNOSIS_dx
  )
```

## Preparing for Cox PH model

### Constructing global (common) baseline, age at baseline, event and time-to-event variables

```{r}
test <- surv_data %>%
  mutate(EXAMDATE_dx_lag = lag(EXAMDATE_dx))
```

```{r}
surv_data <- surv_data %>%
  group_by(RID) %>%
  summarise(
#     carrying these forward to check events and event times are coded correctly
    baseline_dx = unique(baseline_dx),
    last_dx = unique(last_dx),
    
#     baseline EXAMDATE defined as the earlier biomarker EXAMDATE (nfl or ptau)
    EXAMDATE_baseline = min(EXAMDATE_baseline),
    
#     age at baseline visit
    age_at_baseline = age_at_visit[
      EXAMDATE_ptau == min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)]) |
      EXAMDATE_nfl == min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])
      ][1],

#     last dx/nfl/ptau exam
      last_followup = max(c(EXAMDATE_ptau, EXAMDATE_nfl, EXAMDATE_dx), na.rm = TRUE),

#     1 = dementia, 0 = otherwise
      event = as.integer(any(DIAGNOSIS_dx == "Dementia")),

#     0 if dementia dx at baseline, otherwise time to dementia dx or date of censuring
      time_to_event = case_when(
        event == 1 & unique(baseline_dx) == "Dementia" ~ 0,
        event == 1 & unique(baseline_dx) != "Dementia" ~
          as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - EXAMDATE_baseline)/365,
        TRUE ~
          as.numeric(last_followup - EXAMDATE_baseline)/365),

#     Sensitivity: event time = midpoint between last clean exam and dementia dx exam
    time_to_event_midpoint = case_when(
      event == 1 & unique(baseline_dx) == "Dementia" ~ 0,
      event == 1 & unique(baseline_dx) != "Dementia" ~ {
        dx_date <- min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"])
        prev_date <- max(EXAMDATE_dx[DIAGNOSIS_dx != "Dementia" & 
                                     EXAMDATE_dx < dx_date], 
                         na.rm = TRUE)
        # Midpoint between last clean visit and dementia dx visit
        as.numeric((prev_date + (dx_date - prev_date) / 2) - EXAMDATE_baseline) / 365
      },
      TRUE ~
        as.numeric(last_followup - EXAMDATE_baseline) / 365
    )
)
```

```{r}
# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

## Removing individuals with dementia at start

```{r}
# Removing individuals who had a dementia dx at time = 0
surv_data <- surv_data %>% filter(time_to_event > 0)
long_data <- long_data %>% filter(RID %in% unique(surv_data$RID))

# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

### Constructing baseline values of each biomarker, to include in the Cox PH models as covariates

```{r}
# Creating a data frame with the first biomarker measurements
baseline <- long_data %>% 
  group_by(RID) %>%
  arrange(RID, EXAMDATE_nfl) %>% 
  mutate(baseline_value_nfl = first(PLASMA_NFL)) %>%
  arrange(RID, EXAMDATE_ptau) %>%
  mutate(baseline_value_ptau = first(PLASMAPTAU181)) %>%
  dplyr::select(RID, baseline_value_ptau, baseline_value_nfl) %>%
  ungroup() %>%
  distinct()
```

```{r}
# merging the survival data with demographics and the baseline biomarker values
surv_data <- surv_data %>%
  merge(demographics_small, by = "RID") %>% 
  merge(apoe %>% dplyr::select(RID, APOE4_status, genotype), by = "RID") %>%
  merge(baseline, by = "RID")
```

# Checking distribution of each biomarker

```{r}
# We check the distribution of each biomarker
hist(long_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(long_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# Creating a folder to save plots
dir.create("figures", showWarnings = FALSE)

# Saving ptau histogram
png("figures/hist_p_tau.png", width = 400, height = 300)
hist(long_data$PLASMAPTAU181,
     main = "PLASMAPTAU181",
     xlab = "PLASMAPTAU181"
     )
dev.off()
```

```{r}
# Saving NfL histogram
png("figures/hist_nfl.png", width = 400, height = 300)
hist(long_data$PLASMA_NFL,
     main = "PLASMA_NFL",
     xlab = "PLASMA_NFL" 
     )
dev.off()
```

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(long_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(long_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

```{r}
#  Saving log transformed ptau histogram
png("figures/hist_log_p_tau.png", width = 400, height = 300)
hist(long_data$log_PLASMAPTAU181,
     main = "log-transformed PLASMAPTAU181",
     xlab = "log_PLASMAPTAU181"
     )
dev.off()
```

```{r}
#  Saving log transformed NfL histogram
png("figures/hist_log_nfl.png", width = 400, height = 300)
hist(long_data$log_PLASMA_NFL,
     main = "log-transformed PLASMA_NFL",
     xlab = "log_PLASMA_NFL" 
     )
dev.off()
```

```{r}
# Transforming the baseline values in the survival data
surv_data$baseline_value_ptau <- log(surv_data$baseline_value_ptau)
surv_data$baseline_value_nfl <- log(surv_data$baseline_value_nfl)
```

## Checking outliers

Checking for extreme observations shows there are no values 3 standard deviations below the mean, only above. An assessment of these more extreme observations suggests that they are within a reasonable range and are assumed to be correct observations.

```{r}
# We identify "extreme" values

# Values greater than 3 sd above ptau mean
long_data[scale(long_data$PLASMAPTAU181) > 3, ]
# Values smaller than 3 sd below ptau mean
long_data[scale(long_data$PLASMAPTAU181) < -3, ]

# Values greater than 3 sd above nfl mean
long_data[scale(long_data$PLASMA_NFL) > 3, ]
# Values smaller than 3 sd below nfl mean
long_data[scale(long_data$PLASMA_NFL) < -3, ]
```

## Rechecking longitudinal and survival data lengths

```{r}
# We check the number of individuals in each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

## Splitting data into in-sample and out-sample

```{r}
# Total number of individuals
RID <- unique(surv_data$RID)

# Number of individuals to test the model on
n_out_sample <- length(RID) * (1/4)

# Out-sample individuals
RID_out_sample <- sample(RID, n_out_sample)

# Longitudinal data with in-sample individuals
long_data_in <- long_data %>% filter(!RID %in% RID_out_sample)

# Longitudinal data with out-sample individuals
long_data_out <- long_data %>% filter(RID %in% RID_out_sample)

# Survival data with in-sample individuals
surv_data_in <- surv_data %>% filter(!RID %in% RID_out_sample)

# Survival data with out-sample individuals
surv_data_out <- surv_data %>% filter(RID %in% RID_out_sample)
```

```{r}
# Checking the number of individuals and event distribution for the in-sample observations
length(unique(surv_data_in$RID))
table(surv_data_in$event)
# Fraction of individuals who experience the event in in-sample
128/528

# Checking the number of individuals and event distribution for the out-sample observations
length(unique(surv_data_out$RID))
table(surv_data_out$event)
# Fraction of individuals who experience the event in out-sample
49/176
```

## Checking correlation between the biomarkers as well as the numerical variables going to be used in the models

```{r}
# Correlation between the biomarkers 
cor(long_data_in$PLASMAPTAU181, long_data_in$PLASMA_NFL, use = "complete.obs")
```

```{r}
# Saving the numerical variables used in the Cox PH models
cox_vars <- surv_data_in[, c("age_at_baseline",
                              "baseline_value_ptau",
                              "baseline_value_nfl")]

# Computing correlation matrix for Cox model variables
cox_cor <- cor(cox_vars, use = "complete.obs")

# Visualising correlations
corrplot(cox_cor,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)


# Saving the numerical variables used in the LME models
lme_vars <- long_data_in[, c("time_years",
                             "log_PLASMAPTAU181",
                             "log_PLASMA_NFL")]

# Computing correlation matrix for LME model variables
lme_cor <- cor(lme_vars, use = "complete.obs")

# Visualising correlations 
corrplot(lme_cor,
         method = "color",
         type = "upper",
         addCoef.col = "black",
         tl.col = "black",
         tl.srt = 45)
```

# Longitudinal models

## Comparing link functions for longitudinal models

Suggested models:

- Gamma: raw data could be modelled well with the gamma distribution since takes no negative values, variance increases with mean, and there is a right-skew to the raw distributions

- Linear mixed effects model with log-transformed data: approximately normal, yet still some skew

- Linear mixed effects model with cube root transformed data: again, approximately normal, still a bit of skew (why such a low AIC? seems wrong?)

- GLMM with log-normal distribution

- GLMM with Student-t distribution: longer tails

### PTAU

```{r}
# checking the mean and variance for ptau181
mean(long_data$PLASMAPTAU181)
var(long_data$PLASMAPTAU181)
```

#### Fitting the models

```{r}
# GLMM
fit_gamma_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID), 
                   data = long_data_in,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_ptau <- lmer(log(PLASMAPTAU181) ~ time_years + (1|RID), 
                data = long_data_in)

# LME approach with log transformed data
fit_lme_cubed_ptau <- lmer(sqrt(sqrt(PLASMAPTAU181)) ~ time_years + (1|RID), 
                           data = long_data_in)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                             data = long_data_in,
                             family = lognormal(link = "log"))

# Student-t
#fit_student_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                            #data = long_data_in,
                            #family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_ptau <- simulateResiduals(fit_gamma_ptau)
testResiduals(gamma_residuals_ptau)

lme_residuals_log_ptau <- simulateResiduals(fit_lme_log_ptau)
testResiduals(lme_residuals_log_ptau)

lme_residuals_cubed_ptau <- simulateResiduals(fit_lme_cubed_ptau)
testResiduals(lme_residuals_cubed_ptau)

lognormal_residuals_ptau <- simulateResiduals(fit_lognormal_ptau)
testResiduals(lognormal_residuals_ptau)

#student_residuals_ptau <- simulateResiduals(fit_student_ptau)
#testResiduals(student_residuals_ptau)
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(gamma_residuals_ptau$fittedResiduals)
shapiro.test(lme_residuals_log_ptau$fittedResiduals)
shapiro.test(lme_residuals_cubed_ptau$fittedResiduals)
shapiro.test(lognormal_residuals_ptau$fittedResiduals)
#shapiro.test(student_residuals_ptau$fittedResiduals)
```

#### AIC and log-likelihood

```{r}
# Comparing p-tau181 models with AIC and log likelihood
AIC(fit_lme_log_ptau, fit_lme_cubed_ptau, fit_gamma_ptau, fit_lognormal_ptau)
logLik(fit_lme_log_ptau)
logLik(fit_lme_cubed_ptau)
logLik(fit_gamma_ptau)
logLik(fit_lognormal_ptau)
#logLik(fit_student_ptau)
```

### NfL

```{r}
# checking the mean and variance for nfl
mean(long_data$PLASMA_NFL)
var(long_data$PLASMA_NFL)
```

#### Fitting the models

```{r}
# GLMM
fit_gamma_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID), 
                   data = long_data_in,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_nfl <- lmer(log(PLASMA_NFL) ~ time_years + (1|RID), 
                data = long_data_in)

# LME approach with log transformed data
fit_lme_cubed_nfl <- lmer(sqrt(sqrt(PLASMA_NFL)) ~ time_years + (1|RID), 
                          data = long_data_in)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                             data = long_data_in,
                             family = lognormal(link = "log"))

# Student-t
#fit_student_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                           #data = long_data_in,
                           #family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_nfl <- simulateResiduals(fit_gamma_nfl)
testResiduals(gamma_residuals_nfl)

lme_residuals_log_nfl <- simulateResiduals(fit_lme_log_nfl)
testResiduals(lme_residuals_log_nfl)

lme_residuals_cubed_nfl <- simulateResiduals(fit_lme_cubed_nfl)
testResiduals(lme_residuals_cubed_nfl)

lognormal_residuals_nfl <- simulateResiduals(fit_lognormal_nfl)
testResiduals(lognormal_residuals_nfl)

#student_residuals_nfl <- simulateResiduals(fit_student_nfl)
#testResiduals(student_residuals_nfl)
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(gamma_residuals_nfl$fittedResiduals)
shapiro.test(lme_residuals_log_nfl$fittedResiduals)
shapiro.test(lme_residuals_cubed_nfl$fittedResiduals)
shapiro.test(lognormal_residuals_nfl$fittedResiduals)
#shapiro.test(student_residuals_nfl$fittedResiduals)
```

#### AIC and log-likelihood

```{r}
# Comparing NfL-models with AIC and log likelihood
AIC(fit_lme_log_nfl, fit_lme_cubed_nfl, fit_gamma_nfl, fit_lognormal_nfl)
logLik(fit_lme_log_nfl)
logLik(fit_lme_cubed_nfl)
logLik(fit_gamma_nfl)
logLik(fit_lognormal_nfl)
#logLik(fit_student_nfl)
```

Osäker dock om man kan jämföra log-version och cube-version av lme med de andra modellerna (som använder råa versionen) eftersom de då är på olika skalor! Tror det är bäst att kolla på residualerna och QQ-plotsen

# Fitted LME models

## PTAU

#### With all genotype levels

- None of the coefficients are significant for any of the levels for genotype

```{r}
# Longitudinal model for log p-tau181 with normal distribution 
long_model_ptau_full <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + genotype,
  random = ~ time_years | RID,                    
  data = long_data_in
)
```

```{r}
summary(long_model_ptau_full)
```

#### APOE4 status

- All three levels of APOE4_status are significant

```{r}
# Longitudinal model for log p-tau181 with APOE4 status 
long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status,
  random = ~ time_years | RID,                    
  data = long_data_in
)
```

```{r}
summary(long_model_ptau)
```

#### Determining random effects

- Fitting the model with ML estimation in order to compare models using likelihood ratio test

- The results give support to fitting the simpler model with only a random intercept. However, $jm()$ does not currently support model fitting for univariate longitudinal data with only a random intercept.

```{r}
m1 <- lmer(log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status + (1 | RID),
           data = long_data_in, REML = FALSE)

m2 <- lmer(log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status + (time_years | RID),
           data = long_data_in, REML = FALSE)
```

```{r}
anova(m1, m2, type = "LRT")
```

## NFL

#### With all genotype levels

```{r}
# Longitudinal model for log NfL with normal distribution, all genotype levels
long_model_nfl_full <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER + genotype, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl_full)
```

#### APOE4 status

- None of the levels are significant for log NfL and neither is gender

```{r}
# Longitudinal model for log NfL with normal distribution, APOE4_status
long_model_nfl <- lme(
  fixed = log_PLASMA_NFL ~ time_years + PTGENDER + APOE4_status, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl)
```

#### Removing APOE4 status

```{r}
# Longitudinal model for log NfL without APOE4_status
long_model_nfl_reduced1 <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl_reduced1)
```

#### Removing PTGENDER

```{r}
# Longitudinal model for log NfL without covariates
long_model_nfl_reduced2 <- lme(
  log_PLASMA_NFL ~ time_years, 
  random = ~ time_years | RID,
  data = long_data_in
  )
```

```{r}
summary(long_model_nfl_reduced2)
```

#### Determining the random effects

- The results of the likelihood-ratio test support including a random slope as well as a random intercept for the NfL model

```{r}
m1 <- lmer(log_PLASMA_NFL ~ time_years + (1 | RID),
           data = long_data_in, REML = FALSE)

m2 <- lmer(log_PLASMA_NFL ~ time_years + (time_years | RID),
           data = long_data_in, REML = FALSE)
```

```{r}
anova(m1, m2, type = "LRT")
```

```{r}
# Mixed object for multivariate joint model
mixed <- list(long_model_ptau, long_model_nfl)
```

## Residual check

#### Residuals vs fitted values

```{r}
# residuals of the p-tau181 LME model
res_ptau <- residuals(long_model_ptau, type = "pearson")  

# fitted values of the p-tau181 LME model
fitted_vals_ptau <- fitted(long_model_ptau)  

# plot of residuals vs fitted values 
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for p-tau181")
abline(h = 0, col = "red", lty = 2)
```

```{r}
# Saving the plot of residuals vs fitted for ptau
png("figures/residuals_vs_fitted_ptau.png", width = 400, height = 300)
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for p-tau181")
abline(h = 0, col = "red", lty = 2)
dev.off()
```

```{r}
# residuals of the NfL LME model
res_nfl <- residuals(long_model_nfl, type = "pearson")  

# fitted values of the NfL LME model
fitted_vals_nfl <- fitted(long_model_nfl)  

# plot of residuals vs fitted values 
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NFL")
abline(h = 0, col = "red", lty = 2)
```

```{r}
# Saving the residuals vs fitted plot for nfl
png("figures/residuals_vs_fitted_nfl.png", width = 400, height = 300)
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NfL")
abline(h = 0, col = "red", lty = 2)
dev.off()
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(res_ptau)
```

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(res_nfl)
```

#### QQ plots

```{r}
# For P-tau
qqnorm(res_ptau, main = "QQ plot: p-tau181 residuals")
qqline(res_ptau, col = "red")

# For NFL 
qqnorm(res_nfl, main = "QQ plot: NfL residuals")
qqline(res_nfl, col = "red")
```

```{r}
#  Saving the QQ-plots
png("figures/qq_ptau.png", width = 400, height = 300)
qqnorm(res_ptau, main = "QQ plot: p-tau181 residuals")
qqline(res_ptau, col = "red")
dev.off()

png("figures/qq_nfl.png", width = 400, height = 300)
qqnorm(res_nfl, main = "QQ plot: NfL residuals")
qqline(res_nfl, col = "red")
dev.off()
```

# Survival models

## For the different cox PH models and the separate joint models

## Fitted standard Cox PH model (for the joint models)

#### With all genotype levels

```{r}
# Cox survival-models, with covariates
# All genotype levels
surv_model_full <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + genotype,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_full)
```

#### APOE4 status

```{r}
# Cox survival-models, with covariates
# With APOE4_status
surv_model <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model)
```

#### Removing PTGENDER

```{r}
# Cox survival-models, with covariates
# With APOE4_status
surv_model_reduced <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + APOE4_status,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_reduced)
```

```{r}
anova(surv_model, surv_model_reduced)
```

## Fitted Cox PH models using baseline values of each biomarker

### PTAU

```{r}
# Cox survival-models, with covariates including baseline values of each biomarker 
# For p-tau181 with PTGENDER
surv_model_ptau_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_ptau_bl)
```

### NfL

```{r}
# For baseline NfL with covariates
surv_model_nfl_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_nfl_bl)
```

### PTAU and NfL

```{r}
# Inlcuding both p-tau181 and NfL
surv_model_both_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_both_bl) 
```

### Test of the assumption of proportional hazards

```{r}
# Extracting p-values from cox.zph models
extract_ph_p <- function(model, model_name) {
  test <- cox.zph(model)
  df <- as.data.frame(test$table)
  
  data.frame(
    variable = rownames(df),
    p_value = round(df$p, 3),
    model = model_name
  )
}

# Combining all Cox PH models into one dataset
ph_all <- rbind(
  extract_ph_p(surv_model, "Standard"),
  extract_ph_p(surv_model_ptau_bl, "p-tau181 baseline"),
  extract_ph_p(surv_model_nfl_bl, "NfL baseline"),
  extract_ph_p(surv_model_both_bl, "Both baseline")
)

# Forming a table where each model has its own column
table_ph <- ph_all %>%
  dplyr::select(variable, model, p_value) %>%
  tidyr::pivot_wider(names_from = model, values_from = p_value)

# Displaying final table
table_ph
```

```{r}
#  Creating a directory to save tables into
dir.create("tables", showWarnings = FALSE)

# Saving table with PHT results
saveRDS(table_ph, "tables/table_ph.rds")
```

#### Removing age-at-baseline due to evidence of non-proportional hazards

```{r}
# Cox survival-models with covariates but excluding age-at-baseline
# Standard cox model 
surv_model_final <- coxph(
  Surv(time_to_event, event) ~ PTGENDER + APOE4_status,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_final)

# Including p-tau181
surv_model_ptau_bl_final <- coxph(
  Surv(time_to_event, event) ~
    PTGENDER + APOE4_status + baseline_value_ptau,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_ptau_bl_final)

# Including NfL
surv_model_nfl_bl_final <- coxph(
  Surv(time_to_event, event) ~
    PTGENDER + APOE4_status + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_nfl_bl_final)

# Inlcuding both p-tau181 and NfL
surv_model_both_bl_final <- coxph(
  Surv(time_to_event, event) ~
    PTGENDER + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

summary(surv_model_both_bl_final) 
```

#### Final test of the assumption of proportional hazards

```{r}
# Combining all final Cox PH models into one dataset
ph_all_final <- rbind(
  extract_ph_p(surv_model_final, "Standard"),
  extract_ph_p(surv_model_ptau_bl_final, "p-tau181 baseline"),
  extract_ph_p(surv_model_nfl_bl_final, "NfL baseline"),
  extract_ph_p(surv_model_both_bl_final, "Both baseline")
)

# Forming a table where each final model has its own column
table_ph_final <- ph_all_final %>%
  dplyr::select(variable, model, p_value) %>%
  tidyr::pivot_wider(names_from = model, values_from = p_value)

# Displaying final PH table 
table_ph_final
```

```{r}
# Saving the final PH table
saveRDS(table_ph_final, "tables/table_ph_final.rds")
```

### Removing PTGENDER to run likelihood-ratio tests

```{r}
# Cox survival-models, with covariates including baseline values of each biomarker

# For p-tau181 without PTGENDER
surv_model_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status,
  data = surv_data_in,
  x = TRUE
  )

anova_surv <- anova(surv_model_final, surv_model_reduced, test = "LRT")
chi_surv <- anova_surv[2,2]

# For p-tau181 without PTGENDER
surv_model_ptau_bl_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status + baseline_value_ptau,
  data = surv_data_in,
  x = TRUE
  )

anova_ptau <- anova(surv_model_ptau_bl_final, surv_model_ptau_bl_reduced, test = "LRT")
chi_ptau <- anova_ptau[2,2]

# For NfL without PTGENDER
surv_model_nfl_bl_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

anova_nfl <- anova(surv_model_nfl_bl_final, surv_model_nfl_bl_reduced, test = "LRT")
chi_nfl <- anova_nfl[2,2]

# Without PTGENDER
surv_model_both_bl_reduced <- coxph(
  Surv(time_to_event, event) ~ APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data_in,
  x = TRUE
)

anova_both <- anova(surv_model_both_bl_final, surv_model_both_bl_reduced, test = "LRT")
chi_both <- anova_both[2,2]
```

```{r}
anova_ptau
anova_nfl
anova_both
```

```{r}
# Saving a table of results from the likelihood ratio tests
surv_lrt_table <- data.frame(
  "Model" = c("ptau-181 baseline", "NfL baseline", "ptau-181 and NfL baseline"),
  "Test statistic" = c(chi_ptau, chi_nfl, chi_both),
  "P-value" = c(0.98, 0.95, 0.73),
  check.names = FALSE
  )
```

### Cox comparison table

```{r}
# Building Cox comparison table
# Extracting model output
extract_model_cox <- function(model) {
  s <- summary(model)

  data.frame(
    Variable = rownames(s$coefficients),
    HR = exp(s$coefficients[, "coef"]),
    CI_low = s$conf.int[, "lower .95"],
    CI_high = s$conf.int[, "upper .95"],
    P = s$coefficients[, "Pr(>|z|)"],
    row.names = NULL
  )
}

# Formatting HR and CI
format_hr_ci <- function(hr, low, high) {
  paste0(round(hr, 3), " (", round(low, 3), "–", round(high, 3), ")")
}

# Formatting P-values
format_p <- function(x) {
  ifelse(is.na(x), NA,
         ifelse(x < 0.001, "<0.001", round(x, 3)))
}

# Extracting models
ptau_cox <- extract_model_cox(surv_model_ptau_bl_final)
nfl_cox  <- extract_model_cox(surv_model_nfl_bl_final)
both_cox <- extract_model_cox(surv_model_both_bl_final)

# Building variable set
all_vars_cox <- unique(c(ptau_cox$Variable, nfl_cox$Variable, both_cox$Variable))

# Value extraction
get_value <- function(df, var, col) {
  out <- rep(NA, length(var))

  for (i in seq_along(var)) {
    idx <- which(df$Variable == var[i])
    if (length(idx) == 1) {
      out[i] <- df[idx, col]
    }
  }
  out
}

# Building final table
table_cox <- data.frame(
  Variable = all_vars_cox,

  "PTau HR (CI)" = format_hr_ci(
    get_value(ptau_cox, all_vars_cox, "HR"),
    get_value(ptau_cox, all_vars_cox, "CI_low"),
    get_value(ptau_cox, all_vars_cox, "CI_high")
  ),
  "PTau P" = format_p(get_value(ptau_cox, all_vars_cox, "P")),

  "NfL HR (CI)" = format_hr_ci(
    get_value(nfl_cox, all_vars_cox, "HR"),
    get_value(nfl_cox, all_vars_cox, "CI_low"),
    get_value(nfl_cox, all_vars_cox, "CI_high")
  ),
  "NfL P" = format_p(get_value(nfl_cox, all_vars_cox, "P")),

  "Both HR (CI)" = format_hr_ci(
    get_value(both_cox, all_vars_cox, "HR"),
    get_value(both_cox, all_vars_cox, "CI_low"),
    get_value(both_cox, all_vars_cox, "CI_high")
  ),
  "Both P" = format_p(get_value(both_cox, all_vars_cox, "P")),

  row.names = NULL, 
  check.names = FALSE
)

# Showing table
table_cox
```

```{r}
# Savig the cox models comparison table
saveRDS(table_cox, "tables/table_cox.rds")
```

### Concordance Index Table

```{r}
# Saving the c-index table
c_index_table <- data.frame(
  "Model" = c("Cox ptau-181", "Cox NfL", "Cox ptau-181 & NfL"),
  "C-Index" = c(0.679, 0.673, 0.681),
  check.names = FALSE)
  
saveRDS(c_index_table, "tables/c_index_table.rds")

```

# Two separate joint models

## Fitted joint model for p-tau181

```{r}
# # Joint model for p-tau181 
# joint_model_ptau <- jm(
#   surv_model_final,
#   long_model_ptau,
#   time_var = "time_years",
#   n_chains = 3L,
#   n_iter = 11000L,
#   n_burnin = 1000L
#   )
```

```{r}
# summary(joint_model_ptau)
```

```{r}
# Checking convergence 
# ggtraceplot(joint_model_ptau, "alphas")
```

```{r}
# Saving the convergence plot
# png("figures/convergence_plot_ptau.png", width = 400, height = 300)
# ggtraceplot(joint_model_ptau, "alphas")
# dev.off()
```

## Fitted joint model for NfL

```{r}
# # Joint model for NfL 
# joint_model_nfl <- jm(
#   surv_model_final,
#   long_model_nfl,
#   time_var = "time_years",
#   n_chains = 3L,
#   n_iter = 11000L,
#   n_burnin = 1000L
#   )
```

```{r}
# summary(joint_model_nfl)
```

```{r}
# Checking convergence 
# ggtraceplot(joint_model_nfl, "alphas")
```

```{r}
# Saving the convergence plot
# png("figures/convergence_plot_nfl.png", width = 400, height = 300)
# ggtraceplot(joint_model_nfl, "alphas")
# dev.off()
```

### Traceplots

```{r}
#traceplot(joint_model_ptau)
#traceplot(joint_model_nfl)
```

```{r}
# For p-tau181
# traceplot(joint_model_ptau, parm = "betas")
# traceplot(joint_model_ptau, parm = "alphas")
# traceplot(joint_model_ptau, parm = "sigmas")
```

```{r}
# For NfL
# traceplot(joint_model_nfl, parm = "betas")
# traceplot(joint_model_nfl, parm = "alphas")
# traceplot(joint_model_nfl, parm = "sigmas")
```

### Density plots

```{r}
#densplot(joint_model_ptau)
#densplot(joint_model_nfl)
```

```{r}
# For p-tau181
# densplot(joint_model_ptau, parm = "betas")
# densplot(joint_model_ptau, parm = "alphas")
# densplot(joint_model_ptau, parm = "sigmas")
```

```{r}
# For NfL
# densplot(joint_model_nfl, parm = "betas")
# densplot(joint_model_nfl, parm = "alphas")
# densplot(joint_model_nfl, parm = "sigmas")
```

# Multivariate joint model

## To check if the models have stable random slopes

```{r}
VarCorr(long_model_ptau)
VarCorr(long_model_nfl)
```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

## fForms

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL)
  )      
```

## Fitted multivariate joint model

```{r}
# multi_joint_model <- jm(surv_model_final, mixed ,
#                         time_var = "time_years",
#                         functional_forms = fForms,
#                         n_chains = 3L,
#                         n_iter = 20000L,
#                         n_burnin = 2000L
#                         )
# 
# summary(multi_joint_model)
```

```{r}
# Checking convergence 
# ggtraceplot(multi_joint_model, "alphas")
```

```{r}
# Saving the convergence plot
# png("figures/convergence_plot_multi.png", width = 400, height = 300)
# ggtraceplot(multi_joint_model, "alphas")
# dev.off()
```

### Joint model comparison table

`{# {r} # # Building joint model comparison table # # Extracting models (JMbayes2 joint models use log-HR scale with HR = exp(Mean)), where Mean is the posterior mean of the log-hazard ratio # extract_model_jm <- function(model) { #  #   s <- summary(model)$Survival #  #   data.frame( #     Variable = rownames(s), #  #     HR = exp(s[, "Mean"]), #     CI_low = exp(s[, "2.5%"]), #     CI_high = exp(s[, "97.5%"]), #  #     P = s[, "P"], #     row.names = NULL #   ) # } #  # # Extracting models # ptau_jm <- extract_model_jm(joint_model_ptau) # nfl_jm <- extract_model_jm(joint_model_nfl) # both_jm <- extract_model_jm(multi_joint_model) #  # # Building variable set # all_vars_jm <- unique(c(ptau_jm$Variable, nfl_jm$Variable, both_jm$Variable)) #  # # Building final table # table_jm <- data.frame( #   Variable = all_vars_jm, #  #   "PTau HR (CI)" = format_hr_ci( #     get_value(ptau_jm, all_vars_jm, "HR"), #     get_value(ptau_jm, all_vars_jm, "CI_low"), #     get_value(ptau_jm, all_vars_jm, "CI_high") #   ), #   "PTau P" = format_p(get_value(ptau_jm, all_vars_jm, "P")), #  #   "NfL HR (CI)" = format_hr_ci( #     get_value(nfl_jm, all_vars_jm, "HR"), #     get_value(nfl_jm, all_vars_jm, "CI_low"), #     get_value(nfl_jm, all_vars_jm, "CI_high") #   ), #   "NfL P" = format_p(get_value(nfl_jm, all_vars_jm, "P")), #  #   "Both HR (CI)" = format_hr_ci( #     get_value(both_jm, all_vars_jm, "HR"), #     get_value(both_jm, all_vars_jm, "CI_low"), #     get_value(both_jm, all_vars_jm, "CI_high") #   ), #   "Both P" = format_p(get_value(both_jm, all_vars_jm, "P")), #  #   row.names = NULL, #   check.names = FALSE # ) #  # # Showing table # table_jm`

```{r}
# Saving the joint model comparison table
# saveRDS(table_jm, "tables/table_jm.rds")
```

## Rhat value comparison table

`{# {r} # # Extracting Rhat for each model (survival component) # extract_rhat_surv <- function(model) { #  #   s <- summary(model)$Survival #  #   data.frame( #     Variable = rownames(s), #     Rhat = s[, "Rhat"], #     row.names = NULL #   ) # } #  # # Applying to each model # ptau_surv <- extract_rhat_surv(joint_model_ptau) # nfl_surv <- extract_rhat_surv(joint_model_nfl) # multi_surv <- extract_rhat_surv(multi_joint_model) #  # # Combining with model labels to form a table  # rhat_survival_table <- bind_rows( #   ptau_surv %>% mutate(Model = "PTau"), #   nfl_surv %>% mutate(Model = "NfL"), #   multi_surv %>% mutate(Model = "Multi") # ) %>% #   pivot_wider( #     names_from = Model, #     values_from = Rhat #   ) %>% #   mutate(across(where(is.numeric), ~round(.x, 3))) #  # # Showing table # rhat_survival_table`

```{r}
# Saving the rhat table
# saveRDS(rhat_survival_table, "tables/rhat_survival_table.rds")
```

`{# {r} # # Extracting Rhat for longitudinal outcome  # extract_rhat_long <- function(model) { #  #   s <- summary(model)$Outcome1 #  #   data.frame( #     Variable = rownames(s), #     Rhat = s[, "Rhat"], #     row.names = NULL #   ) # } #  # # Applying to each model # ptau_long <- extract_rhat_long(joint_model_ptau) # nfl_long <- extract_rhat_long(joint_model_nfl) # multi_long <- extract_rhat_long(multi_joint_model) #  # # Combining with model labels to form a table # rhat_longitudinal_table <- bind_rows( #   ptau_long %>% mutate(Model = "Long_PTau"), #   nfl_long %>% mutate(Model = "Long_NfL"), #   multi_long %>% mutate(Model = "Long_Multi") # ) %>% #   pivot_wider( #     names_from = Model, #     values_from = Rhat #   ) %>% #   mutate(across(where(is.numeric), ~round(.x, 3))) #  # # Showing table # rhat_longitudinal_table`

`{# {r} # # Saving the rhat table # saveRDS(rhat_longitudinal_table, "tables/rhat_longitudinal_table.rds")`

# ...

# Latent Class Mixed Models

Choice of p-value <https://pmc.ncbi.nlm.nih.gov/articles/PMC5059270/>

# LCMM: NFL

## M0 (1 latent class)

```{r}
# Fitting a one class latent class mixed model with a random intercept
hlme_m0_1 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(hlme_m0_1)
```

```{r}
# Fitting a one class latent class mixed model with a random intercept and slope
hlme_m0_2 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(hlme_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(hlme_m0_1, hlme_m0_2)
```

```{r}
# log likelihood for the full model
l = hlme_m0_2$loglik

# log likelihood for the reduced model
l_0 = hlme_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Fitting a two class latent class mixed model
hlme_m1 <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = long_data_in,
    B = random(hlme_m0_1)
  )

summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## Comparing M0, M1

```{r}
summarytable(hlme_m0_2, hlme_m1,
             which = c("G", "loglik", "BIC", "entropy"))
```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# LCMM: PTAU

## M0 (1 latent class)

```{r}
# Fitting a one class latent class mixed model with a random intercept
hlme_m0_1 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(hlme_m0_1)
```

```{r}
# Fitting a one class latent class mixed model with a random intercept and slope
hlme_m0_2 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(hlme_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(hlme_m0_1, hlme_m0_2)
```

```{r}
# log likelihood for the full model
l = hlme_m0_2$loglik

# log likelihood for the reduced model
l_0 = hlme_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Fitting a two class latent class mixed model
hlme_m1 <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = long_data_in,
    B = random(hlme_m0_1)
  )
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## Comparing M0, M1

```{r}
summarytable(hlme_m0_2, hlme_m1,
             which = c("G", "loglik", "BIC", "entropy"))
```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
mean(pp$prob1)
mean(pp$prob2)
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# MLCMM: NFL & PTAU

Multivariate latent class mixed models with both NFL and PTAU

## M0 (1 latent class)

```{r}
# Fitting a one class multivariate latent class mixed model with a random intercept
mlcmm_m0_1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(mlcmm_m0_1)
```

```{r}
# Fitting a one class multivariate latent class mixed model with a random intercept and slope
mlcmm_m0_2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data_in
)
```

```{r}
summary(mlcmm_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(mlcmm_m0_1, mlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = mlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = mlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Fitting a two class multivariate latent class mixed model
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = long_data_in,
  B = random(mlcmm_m0_2)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

## Comparing M0, M1

```{r}
summarytable(mlcmm_m0_1, mlcmm_m1, which = c("G", "loglik", "BIC", "entropy"))
```

## M2 (2 latent classes) (including class covariates)

```{r}
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ APOE4_status + age_at_visit,
  ng = 2,
  data = long_data_in,
  B = random(mlcmm_m0_1)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- mlcmm_m1$pprob
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 2") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# JLCMM: NFL

```{r}
surv <- survfit(Surv(surv_data_in$time_to_event, surv_data_in$event) ~ 1)

plot(log(surv$time), log(-log(surv$surv)), main="Weibull Check")
```

```{r}
# merging the longitudinal and survival data for jlcmm
jlcmm_data <- surv_data_in %>%
  dplyr::select(RID, event, time_to_event) %>%
  merge(long_data_in, by = "RID")
```

## M0 (1 latent class)

```{r}
# Fitting a one class joint latent class mixed model with a random intercept
jlcmm_m0_1 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)

# Fitting a one class joint latent class mixed model with flexible hazard
jlcmm_m0_1_piecewise <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  hazard = "3-quant-piecewise",
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_1)
```

```{r}
summary(jlcmm_m0_1_piecewise)
```

```{r}
# Fitting a one class joint latent class mixed model with a random intercept and slope
jlcmm_m0_2 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1 + time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(jlcmm_m0_1, jlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = jlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = jlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Running gridsearch() for a two class model
# jlcmm_m1_nfl_search <- gridsearch(m = jlcmm(fixed = log_PLASMA_NFL ~ time_years,
#           random = ~ 1,
#           mixture = ~ 1,
#           classmb = ~ 1,
#           ng = 2,
#           subject = "RID",
#           survival = Surv(time_to_event, event) ~ 1,
#           data = jlcmm_data),
# minit = jlcmm_m0_1,
# rep = 50,
# maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m1_nfl <- jlcmm(
#   fixed = log_PLASMA_NFL ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 2,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m1_nfl_search$best
# )
```

```{r}
# Fitting a two class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m1_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m1_nfl)
```

```{r}
# Checking flexible hazard
jlcmm_m1_nfl_piecewise <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1_piecewise)
)
```

```{r}
# Does not converge
summary(jlcmm_m1_nfl_piecewise)
```

```{r}
postprob_m1_nfl <- data.frame(
  "Class" = c("Class1", "Class2"),
  "Prob1" = c(0.90, 0.03),
  "Prob2" = c(0.10, 0.97),
  "Prob > 0.7 (%)" = c(93.6, 96.0),
  "Prob > 0.8 (%)" = c(82.4, 95.2),
  "Prob > 0.9 (%)" = c(60.3, 92.8),
  check.names = FALSE
)
```

```{r}
postprob_m1_nfl <- postprob(jlcmm_m1_nfl)[1:2]
```

```{r}
saveRDS(postprob_m1_nfl, "tables/postprob_m1_nfl.rds")
```

## M2 (3 latent classes)

```{r}
# Running gridsearch() for a three class model
# jlcmm_m2_nfl_search <- gridsearch(m = jlcmm(fixed = log_PLASMA_NFL ~ time_years,
#                                             random = ~ 1,
#                                             mixture = ~ 1,
#                                             classmb = ~ 1,
#                                             ng = 3,
#                                             subject = "RID",
#                                             survival = Surv(time_to_event, event) ~ 1,
#                                             data = jlcmm_data),
#                                   minit = jlcmm_m0_1,
#                                   rep = 50,
#                                   maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m2_nfl <- jlcmm(
#   fixed = log_PLASMA_NFL ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 3,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m2_nfl_search$best
# )
```

```{r}
# Fitting a three class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m2_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m2_nfl)
```

```{r}
postprob(jlcmm_m2_nfl)
```

## Comparing M0, M1, M2

```{r}
#  Function for extracting the class proportions from each model
get_class_proportions <- function(model) {
#   Getting the number of classes
  G <- model$ng
#   Calculating the class proportions (number of individuals assigned to a class / number of inidividuals * 100)
  props <- table(model$pprob$class) / nrow(model$pprob) * 100
  
#   Creating a vector to store the proportions in, with NA for extra classes
  out <- rep(NA, G)
  out[1:G] <- as.numeric(props)
  names(out) <- paste0("Class", 1:G)
  
#   Returning results
  return(round(out, 2))
}

# Function for extracting entropy
get_entropy <- function(model) {
  G <- model$ng
  N <- nrow(model$pprob)
  
  if (G == 1) return(1.00)

  pprobs <- model$pprob[, 3:(2 + G)]
#   computing entropy
  entropy <- -sum(pprobs * log(pprobs + 0.0001))
  
#   scaling to between 0-1
  entropy_r2 <- 1 - (entropy / (N * log(G)))
  
  return(round(entropy_r2, 3))
}
```

```{r}
models <- list(jlcmm_m0_1, jlcmm_m1_nfl, jlcmm_m2_nfl)

# Find max number of classes across all models
G_max <- max(sapply(models, function(m) m$ng))

# Extract proportions for each model, with NA for cells where the model has fewer classes than the given row
props <- do.call(rbind, lapply(models, function(m) {
  p <- get_class_proportions(m)
  padded <- c(p, rep(NA, G_max - length(p)))
  return(padded)
}))

colnames(props) <- paste0("Class", 1:G_max, " (%)")

summary_jlcmm_nfl_full <- data.frame(
  "Candidate model" = paste0("Model", 1:length(models)),
  "G" = sapply(models, function(m) m$ng),
  "Log Likelihood" = sapply(models, function(m) m$loglik),
  "BIC" = sapply(models, function(m) m$BIC),
  "Entropy" = sapply(models, get_entropy),
  props,
  check.names = FALSE
)
```

```{r}
saveRDS(summary_jlcmm_nfl_full, "tables/summary_jlcmm_nfl_full.rds")
```

## Class Descriptions

```{r}
# # saving which class each individual was predicted to be in
class_preds <- jlcmm_m1_nfl$pprobY

# merging with genotype and demographic data
class_preds <- class_preds %>%
  merge((apoe %>% dplyr::select(RID, APOE4_status)), by = "RID") %>% 
  merge((surv_data %>% dplyr::select(RID, event, age_at_baseline, PTGENDER)), by = "RID")
```

```{r}
# checking the distribution of genotype by class
table(class_preds$class, class_preds$APOE4_status)
# checking the distribution of gender by class
table(class_preds$class, class_preds$PTGENDER)
```

## Visualisation

1.  Need to check probYT1 and probYT2

```{r}
# saving the posterior probabilities into a data frame
# pp <- jlcmm_m1$pprob
# pp1 <- pp %>% filter(class == 1)
# pp2 <- pp %>% filter(class == 2)
```

### Distribution of posterior probabilities (class 1)

1.  Check that YT1 is for the joint model, not the longitudinal model only

```{r}
# plotting posterior probabilities for class 1
# ggplot(pp1, aes(x = probYT1)) + 
#   geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
#   labs(title = "Posterior Probabilities for Class 1") +
#   theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
# ggplot(pp2, aes(x = probYT1)) + 
#   geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
#   labs(title = "Posterior Probabilities for Class 2") +
#   theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
# ggplot(pp, aes(prob1, prob2)) +
#   geom_point(alpha = .8) +
#   geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
#   labs(x = "P(Class 1)", y = "P(Class 2)") +
#   theme_classic()
```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m1_nfl, which = "survival")
```

```{r}
png("figures/class-specific-survival-jlcmm-nfl.png", width = 400, height = 300)
plot(jlcmm_m1_nfl, which = "survival",
     main = "",
     xlab = "Time (years)", ylab = "Survival Probability")
dev.off()
```

### Longitudinal trajectory by class

```{r}
# longitudinal trajectory by class
plot(jlcmm_m1_nfl, var.time = "time_years", which = "fit",
     ylim = c(2.5,4.2))
```

```{r}
png("figures/longitudinal-jlcmm-nfl.png", width = 400, height = 300)
plot(jlcmm_m1_nfl, var.time = "time_years", which = "fit",
     main = "",
     xlab = "Time (years)", ylab = "NfL")
dev.off()
```

# JLCMM: PTAU

## M0 (1 latent class)

```{r}
# Fitting a one class joint latent class mixed model with random intercept
jlcmm_m0_1 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)

# Fitting a one class joint latent class mixed model with flexible baseline hazard to check performance against parametric hazard (Weibull)
jlcmm_m0_1_piecewise <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  hazard = "3-quant-piecewise",
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_1)
```

```{r}
summary(jlcmm_m0_1_piecewise)
```

```{r}
# Fitting a one class joint latent class mixed model with a random intercept and slope
jlcmm_m0_2 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(jlcmm_m0_1, jlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = jlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = jlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
# Running gridsearch() for a two class model
# jlcmm_m1_ptau_search <- gridsearch(m = jlcmm(fixed = log_PLASMAPTAU181 ~ time_years,
#                                               random = ~ 1,
#                                               mixture = ~ 1,
#                                               classmb = ~ 1,
#                                               ng = 2,
#                                               subject = "RID",
#                                               survival = Surv(time_to_event, event) ~ 1,
#                                               data = jlcmm_data),
#                                     minit = jlcmm_m0_1,
#                                     rep = 50,
#                                     maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m1_ptau <- jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 2,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m1_ptau_search$best
# )
```

```{r}
# Fitting a two class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m1_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m1_ptau)
```

```{r}
postprob(jlcmm_m1_ptau)
```

```{r}
postprob_m1_ptau <- data.frame(
  "Class" = c("Class1", "Class2"),
  "Prob1" = c(0.97, 0.12),
  "Prob2" = c(0.03, 0.88),
  "Prob > 0.7 (%)" = c(93.8, 91.0),
  "Prob > 0.8 (%)" = c(91.47, 76.19),
  "Prob > 0.9 (%)" = c(91.47, 52.63),
  check.names = FALSE
)
```

```{r}
saveRDS(postprob_m1_ptau, "tables/postprob_m1_ptau.rds")
```

## M2 (3 latent classes)

*Things to assess:*

1.  *Fit 4 class model -\> does BIC improve any more or plateau?*

2.  *Score test now not significant compared with the two class model*

3.  *VERY small class 1*

4.  *CLass 2 and 3 are not cleanly separated (look at posterior probabilities)*

5.  Weibull class 3 parameters not significant at 0.05 level

6.  *Plot survival curves (how?)*

```{r}
# Running gridsearch() for a three class model
# jlcmm_m2_ptau_search <- gridsearch(m = jlcmm(fixed = log_PLASMAPTAU181 ~ time_years,
#                                               random = ~ 1,
#                                               mixture = ~ 1,
#                                               classmb = ~ 1,
#                                               ng = 3,
#                                               subject = "RID",
#                                               survival = Surv(time_to_event, event) ~ 1,
#                                               data = jlcmm_data),
#                                     minit = jlcmm_m0_1,
#                                     rep = 50,
#                                     maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m2_ptau <- jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 3,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m2_ptau_search$best
#   )
```

```{r}
# Fitting a three class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m2_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m2_ptau)
```

```{r}
postprob(jlcmm_m2_ptau)
```

```{r}
postprob_m2_ptau <- data.frame(
  "Class" = c("Class1", "Class2", "Class3"),
  "Prob1" = c(0.74, 0.05, 0.01),
  "Prob2" = c(0.24, 0.82, 0.04),
  "Prob3" = c(0.02, 0.12, 0.95),
  "Prob > 0.7 (%)" = c(65.0, 45.0, 15.0),
  "Prob > 0.8 (%)" = c(86.1, 62.8, 28.7),
  "Prob > 0.9 (%)" = c(93.2, 87.9, 84.1),
  check.names = FALSE
)
```

```{r}
saveRDS(postprob_m2_ptau, "tables/postprob_m2_ptau.rds")
```

```{r}
# Checking flexible baseline hazard (piecewise)
jlcmm_m2_ptau_piecewise <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  hazard = "3-quant-piecewise",
  data = jlcmm_data,
  B = random(jlcmm_m0_1_piecewise)
)
```

```{r}
# Flexible hazard with two parameters has higher BIC than Weibull hazard
summary(jlcmm_m2_ptau_piecewise)
```

## M3 (4 latent classes)

```{r}
# Running gridsearch() for a four class model
# jlcmm_m3_ptau_search <- gridsearch(m = jlcmm(fixed = log_PLASMAPTAU181 ~ time_years,
#                                               random = ~ 1,
#                                               mixture = ~ 1,
#                                               classmb = ~ 1,
#                                               ng = 4,
#                                               subject = "RID",
#                                               survival = Surv(time_to_event, event) ~ 1,
#                                               data = jlcmm_data),
#                                     minit = jlcmm_m0_1,
#                                     rep = 50,
#                                     maxiter = 50)
```

```{r}
# Estimating two class model using the best result from gridsearch() as a starting value
# jlcmm_m3_ptau <- jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 4,
#   subject = "RID",
#   survival = Surv(time_to_event, event) ~ 1,
#   data = jlcmm_data,
#   B = jlcmm_m3_ptau_search$best
#   )
```

```{r}
# Fitting a four class joint latent class mixed model
# Results from gridsearch() and B = random(jlcmm_m0_1) are very similar, global maximum found with random start from the 1 class model
jlcmm_m3_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 4,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m3_ptau)
```

```{r}
postprob(jlcmm_m3_ptau)
```

## Comparing M0, M1, M2, M3

```{r}
models <- list(jlcmm_m0_1, jlcmm_m1_ptau, jlcmm_m2_ptau, jlcmm_m3_ptau)

# Find max number of classes across all models
G_max <- max(sapply(models, function(m) m$ng))

# Extract proportions for each model, with NA for cells where the model has fewer classes than the given row
props <- do.call(rbind, lapply(models, function(m) {
  p <- get_class_proportions(m)
  padded <- c(p, rep(NA, G_max - length(p)))
  return(padded)
}))

colnames(props) <- paste0("Class", 1:G_max, " (%)")

summary_jlcmm_ptau_full <- data.frame(
  "Candidate model" = paste0("Model", 1:length(models)),
  "G" = sapply(models, function(m) m$ng),
  "Log Likelihood" = sapply(models, function(m) m$loglik),
  "BIC" = sapply(models, function(m) m$BIC),
  "Entropy" = sapply(models, get_entropy),
  props,
  check.names = FALSE
)
```

```{r}
saveRDS(summary_jlcmm_ptau_full, "tables/summary_jlcmm_ptau_full.rds")
```

## Class Descriptions

```{r}
# # saving which class each individual was predicted to be in
class_preds <- jlcmm_m1_ptau$pprobY

# merging with genotype data
class_preds <- class_preds %>%
  merge(apoe %>% dplyr::select(RID, APOE4_status), by = "RID")

# checking the distribution of genotype by class
table(class_preds$class, class_preds$APOE4_status)
```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}

```

### Distribution of posterior probabilities (class 2)

```{r}

```

### Class certainty scatterplot

```{r}

```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m2_ptau, which = "survival")
```

```{r}
png("figures/class-specific-survival-jlcmm-ptau.png", width = 400, height = 300)
plot(jlcmm_m1_nfl, which = "survival",
     main = "",
     xlab = "Time (years)", ylab = "Survival Probability")
dev.off()
```

### Longitudinal trajectories by class

```{r}
# longitudnal trajectories by class
plot(jlcmm_m1_ptau, var.time = "time_years", which = "fit")
```

```{r}
png("figures/longitudinal-jlcmm-ptau.png", width = 400, height = 300)
plot(jlcmm_m1_ptau, var.time = "time_years", which = "fit",
     main = "",
     xlab = "Time (years)", ylab = "ptau-181")
dev.off()
```

# MJLCMM: NFL & PTAU

## M0 (1 latent class)

```{r}
# Modelling the longitudinal data only, these will be used as arguments in the joint models
# Fitting a one class multivariate latent class mixed model with random intercept
mlcmm_m0_1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = jlcmm_data
)

# Fitting a one class multivariate latent class model with random intercept and slope
mlcmm_m0_2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = jlcmm_data
)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a one class multivariate joint latent class model with random intercept
mult_jlcmm_m0_1 <- mpjlcmm(
  longitudinal = list(mlcmm_m0_1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data
)
```

```{r}
summary(mult_jlcmm_m0_1)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a one class multivariate joint latent class model with random intercept and slope
mult_jlcmm_m0_2 <- mpjlcmm(
  longitudinal = list(mlcmm_m0_2),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data
)
```

```{r}
summary(mult_jlcmm_m0_2)
```

### Determining the random effects

```{r}
# Comparing BIC and log-likelihood for the two one class models with different random effects structures
summarytable(mult_jlcmm_m0_1, mult_jlcmm_m0_2)
```

## M1 (2 latent classes)

```{r}
# Modelling longitudinal data, used as argument in the joint model
# Fitting a multivariate latent class model with two latent classes
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mlcmm_m0_1)
)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a multivariate joint latent class model with two latent classes
mult_jlcmm_m1 <- mpjlcmm(
  longitudinal = list(mlcmm_m1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0_1)
)
```

```{r}
summary(mult_jlcmm_m1)
```

```{r}
postprob(mult_jlcmm_m1)
```

```{r}
postprob_ptau_nfl <- data.frame(
  "Class" = c("Class1", "Class2"),
  "Prob1" = c(0.96, 0.10),
  "Prob2" = c(0.04, 0.90),
  "Prob > 0.7 (%)" = c(93.8, 94.5),
  "Prob > 0.8 (%)" = c(93.0, 82.8),
  "Prob > 0.9 (%)" = c(91.4, 60.5),
  check.names = FALSE
)
```

```{r}
saveRDS(postprob_ptau_nfl, "tables/postprob_ptau_nfl.rds")
```

## M2 (3 latent classes)

```{r}
# Modelling longitudinal data, used as argument in the joint model
# Fitting a three class multivariate latent class model
mlcmm_m2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mlcmm_m0_1)
)
```

```{r}
# Modelling longitudinal and survival data
# Fitting a three class multivariate joint latent class model
mult_jlcmm_m2 <- mpjlcmm(
  longitudinal = list(mlcmm_m2),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0_1)
)
```

```{r}
summary(mult_jlcmm_m2)
```

```{r}
postprob(mult_jlcmm_m2)
```

## Comparing M0, M1, M2

```{r}
models <- list(mult_jlcmm_m0_1, mult_jlcmm_m1, mult_jlcmm_m2)

# Find max number of classes across all models
G_max <- max(sapply(models, function(m) m$ng))

# Extract proportions for each model, with NA for cells where the model has fewer classes than the given row
props <- do.call(rbind, lapply(models, function(m) {
  p <- get_class_proportions(m)
  padded <- c(p, rep(NA, G_max - length(p)))
  return(padded)
}))

colnames(props) <- paste0("Class", 1:G_max, " (%)")

summary_jlcmm_ptau_nfl_full <- data.frame(
  "Candidate model" = paste0("Model", 1:length(models)),
  "G" = sapply(models, function(m) m$ng),
  "Log Likelihood" = sapply(models, function(m) m$loglik),
  "BIC" = sapply(models, function(m) m$BIC),
  "Entropy" = sapply(models, get_entropy),
  props,
  check.names = FALSE
)
```

```{r}
saveRDS(summary_jlcmm_ptau_nfl_full, "tables/summary_jlcmm_ptau_nfl_full.rds")
```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}

```

### Distribution of posterior probabilities (class 2)

```{r}

```

### Class certainty scatterplot

```{r}

```

### Survival curves by class

```{r}
#plot(mult_jlcmm_m1, which = "survival")
```

```{r}
# plot(mult_jlcmm_m1, var.time = "time_years", which = "fit")
```

### Survival curves by class

```{r}
# survival curves by class
#plot(mult_jlcmm_m2, which = "survival")
```

```{r}
# # longitudnal trajectories by class
# plot(mult_jlcmm_m2, outcome = 1, which = "fit")
```

## TO DO

External validation: what do these classes look like? Do they just represent different genotypes? How do we check this and are they truly valid LATENT classes?

# Comparing models using time-dependent AUC

*For the Cox PH models = using timeROC, from the package timeROC*

*For the joint models = using tvAUC, from the package JMbayes2*

## Defining evaluation time points for time-dependent AUC comparison

```{r}
# We define evaluation time points for time-dependent AUC comparison
times <- c(2, 3, 4, 5, 6)
```

## Time-dependent AUC for fitted Cox PH models

```{r}
# We define a function to compute time-dependent AUC for Cox models using timeROC
get_cox_auc <- function(model, times) {
  
  T <- surv_data_out$time_to_event
  delta <- surv_data_out$event
  
  lp <- predict(model, newdata = surv_data_out, type = "lp")
  
  timeROC::timeROC(
    T = T,
    delta = delta,
    marker = lp,
    cause = 1,
    times = times,
    iid = TRUE
  )$AUC
}
```

```{r}
# We compute time-dependent AUC for Cox model with p-tau181
auc_ptau_bl <- get_cox_auc(surv_model_ptau_bl_final, times)
 
# # We compute time-dependent AUC for Cox model with NfL
auc_nfl_bl <- get_cox_auc(surv_model_nfl_bl_final, times)

# We compute time-dependent AUC for Cox model with both p-tau181 and NfL 
auc_both_bl <- get_cox_auc(surv_model_both_bl_final, times)
```

## Time-dependent AUC for fitted joint models

*From: <https://cloud.r-project.org/web/packages/JMbayes2/JMbayes2.pdf>*

*The function tvAUC requires the following:*

- *"newdata = a data.frame that contains the longitudinal and covariate information for the subjects for which prediction of survival probabilities is required. The names of the variables in this data.frame must be the same as in the data.frames that were used to fit the linear mixed effects and the event process model that were supplied as the two first argument of jm."*

```{r}
# We construct prediction datasets for the joint models by merging the  longitudinal and survival data
newdata <- dplyr::left_join(
  long_data_out,
  dplyr::select(surv_data_out, RID, age_at_baseline, time_to_event, event),
  by = "RID"
)
```

```{r}
# We define a function to compute time-dependent AUC for joint models using tvAUC
get_jm_auc <- function(model, times) {
  
  aucs <- sapply(times, function(t) {
  
    tmp <- tvAUC(
      model,
      newdata = newdata,
      Tstart = 1,
      Thoriz = t
    )
    
    tmp$auc
  })
  
  return(as.numeric(aucs))
}
```

```{r}
# We compute time-dependent AUC for the joint model including p-tau181 
auc_ptau_jm <- get_jm_auc(joint_model_ptau, times)

# We compute time-dependent AUC for the joint model including NfL 
auc_nfl_jm <- get_jm_auc(joint_model_nfl, times)

# We compute time-dependent AUC for the multivariate joint model 
auc_multi_jm <- get_jm_auc(multi_joint_model, times)

```

## Time-dependent AUC for latent class joint models

#### Function for getting predictions from multivariate joint latent class model

```{r}
# Defining a function to predict cumulative hazard (1 - S(t)) for multivariate joint latent class models (these cannot be passed to dynpred in the lcmm package)
predict_mjlcmm <- function(model, data, ID, landmark, times) {
  
#   Extracting the unique ID's
  ids <- unique(data[[ID]])
#   Extracting number of individuals to make predictions for
  N <- length(unique(data[[ID]]))
#   Extracting the nuber of classes
  G <- model$ng
#   Number of horizons
  n_h <- length(times)
  print(n_h)
  
#   Extracting the posterior probabilities (given a two class model)
  post_prob <- model$pprob[, 3:4]
  
#   Extracting the Weibull hazard parameters from the model object 
  scale <- sapply(1:G, function(g) {
                   exp(model$best[paste0("event1 +/-sqrt(Weibull2) class ", as.character(g))])})
  shape <- sapply(1:G, function(g) {
                   exp(model$best[paste0("event1 +/-sqrt(Weibull1) class ", as.character(g))])})
  
#   Initiating a matrix to save the predictions for each individual at each horizon
  S_pred <- matrix(0, nrow = N, ncol = n_h)
  
#   Saving the class-specific survivals at the landmark time
  S_land <- sapply(1:G, function(g)
    exp(-(landmark / scale[g])^shape[g]))
  
#   Looping over all indiivudals
  for(i in 1:N) {
    
#     GEtting the posterior probabilities for each individual
    pp_i <- post_prob[i, ]
    
#     Looping over the horizons
    for(j in 1:n_h) {
      t <- times[j]
    
#      Getting the conditional class-specific survivals
      Sg <- sapply(1:G, function(g)
        exp(-((t) / scale[g])^shape[g]) / S_land[g])

#      Calculating the cumulative hazard (1 - survival) for inidivudal i based on their posterior probabilities and the class-specific survival functions
      S_pred[i, j] <- 1 - sum(pp_i * Sg)
    }
  }  
#   Convert to long format matching dynpred() output structure
  S_pred <- as.data.frame(S_pred)
  colnames(S_pred) <- as.character(times)
  S_pred[[ID]] <- ids

  S_pred_long <- S_pred %>%
    pivot_longer(
      cols = -all_of(ID),
      names_to  = "horizon",
      values_to = "pred"
    ) %>%
    mutate(horizon = as.numeric(horizon))
  
  names(S_pred_long) <- c("RID", "horizon", "pred")

  return(S_pred_long)
  }
```

#### Function for extracting time-dependent AUC

1.  Confidence intervals

2.  Overall model discrimination

3.  Class specific discrimination

```{r}
# Defining a function to compute time dependent AUC for joint latent class models
jlcmm_timeAUC <- function(model, newdata, ID, time_var, surv_time_var, event_var,
                          landmark, times, cause = 1) {
  
#   Initiating a data frame to save the results in at each horizon time
  output <- data.frame(
    times = times,
    cases = NA,
    AUC = NA
  )

#   Calculating the survival predictions previously defined function
  if(class(model) == "mpjlcmm") {
    dfpred <- predict_mjlcmm(
      model = model,
      data = newdata,
      ID = ID,
      landmark = landmark,
      times = times
      )
  }
  
#   Calculating the dynamic survival predictions from joint models using dynpred() in the lcmm package
  if(class(model) == "Jointlcmm") {
    dfpred <- dynpred(
      model = model,
      newdata = newdata,
      var.time = time_var,
      horizon = times,
      landmark = landmark
      )$pred
  }
  
#   Fetching the true events for each ID for each time point
  dftrue <- unique(newdata[c(ID, surv_time_var, event_var)])
  names(dftrue) <- c(ID, "time", "event")
  
#   Merging the true and predicted data frames
  merged <- merge(dfpred, dftrue, by = ID)
  
#   Looping through the horizon times
  for(i in 1:length(times)) {
    h = times[i]
    
#     Filtering for the predictions from horizon h
    data_h <- merged %>% filter(horizon == h)
    
    #     Identifiying individuals that convert (cases) in the time between landmark and landmark + horizon (l + h) or remain disease free past l + h (controls)
    cases <- data_h$event == 1 & data_h$time > landmark & data_h$time <= h
    controls <- data_h$time > h
    data_h <- data_h[cases | controls, ]
    
    roc <- timeROC(
      T = data_h$time,
      delta = data_h$event,
      marker = data_h$pred,
      cause = cause,
      times = h,
      iid = TRUE)
    
    output$AUC[i] <- roc$AUC[2]
#     Checking the events in the time window between l and l + h
    output$cases[i] <- sum(cases)
    
  }
  return(output)
}
```

```{r}
# Computing AUC for 2 class joint latent class mixed model for NfL
auc_jlcmm_m1_nfl <- jlcmm_timeAUC(jlcmm_m1_nfl, newdata,
                                  "RID", "time_years", "time_to_event", "event",
                                  1, times)

# Computing AUC for 2 class joint latent class mixed model for ptau
auc_jlcmm_m1_ptau <- jlcmm_timeAUC(jlcmm_m1_ptau, newdata,
                                   "RID", "time_years", "time_to_event", "event",
                                   1, times)

# Computing AUC for 3 class joint latent class mixed model for ptau
auc_jlcmm_m2_ptau <- jlcmm_timeAUC(jlcmm_m2_ptau, newdata,
                                   "RID", "time_years", "time_to_event", "event",
                                   1, times)

# ON IN-SAMPLE DATA
# Computing AUC for 2 class multivariate joint latent class mixed model for ptau & NfL
auc_mult_jlcmm_m1 <- jlcmm_timeAUC(mult_jlcmm_m1, jlcmm_data,
                                   "RID", "time_years", "time_to_event", "event",
                                   1, times)
```

## Table of the results

```{r}
# Time-dependent AUC for Cox and joint models summarized in a table
results_table_AUC <- data.frame(
  Model = c("Cox ptau-181", "Cox NfL", "Cox both",
            "JM ptau-181", "JM NfL", "JM both", "JLCM ptau-181", "JLCM NfL"),
  
  AUC_2y = c(auc_ptau_bl[1],
             auc_nfl_bl[1],
             auc_both_bl[1],
             auc_ptau_jm[1],
             auc_nfl_jm[1],
             auc_multi_jm[1],
             auc_jlcmm_m2_ptau[1,3],
             auc_jlcmm_m1_nfl[1,3]),
  
  AUC_3y = c(auc_ptau_bl[2],
             auc_nfl_bl[2],
             auc_both_bl[2],
             auc_ptau_jm[2],
             auc_nfl_jm[2],
             auc_multi_jm[2],
             auc_jlcmm_m2_ptau[2,3],
             auc_jlcmm_m1_nfl[2,3]),
  
  AUC_4y = c(auc_ptau_bl[3],
             auc_nfl_bl[3],
             auc_both_bl[3],
             auc_ptau_jm[3],
             auc_nfl_jm[3],
             auc_multi_jm[3],
             auc_jlcmm_m2_ptau[3,3],
             auc_jlcmm_m1_nfl[3,3]),
  
  AUC_5y = c(auc_ptau_bl[4],
             auc_nfl_bl[4],
             auc_both_bl[4],
             auc_ptau_jm[4],
             auc_nfl_jm[4],
             auc_multi_jm[4],
             auc_jlcmm_m2_ptau[4,3],
             auc_jlcmm_m1_nfl[4,3]),
  
  AUC_6y = c(auc_ptau_bl[5],
             auc_nfl_bl[5],
             auc_both_bl[5],
             auc_ptau_jm[5],
             auc_nfl_jm[5],
             auc_multi_jm[5],
             auc_jlcmm_m2_ptau[5,3],
             auc_jlcmm_m1_nfl[5,3])
)

results_table_AUC[, -1] <- round(results_table_AUC[, -1], 3)
results_table_AUC
```

```{r}
# Saving the AUC results table
saveRDS(results_table_AUC, "tables/results_table_AUC.rds")
```

## Plot of the results

```{r}
# We reshape all AUC results into long format for plotting
plot_data <- data.frame(
  Time = rep(times, 8),
  Model = rep(c("Cox ptau-181", "Cox NfL", "Cox both",
                "JM ptau-181", "JM NfL", "JM multi", "JLCM ptau-181", "JLCM NfL"),
              each = length(times)),
  AUC = c(
    auc_ptau_bl,
    auc_nfl_bl,
    auc_both_bl,
    auc_ptau_jm,
    auc_nfl_jm,
    auc_multi_jm,
    auc_jlcmm_m2_ptau$AUC,
    auc_jlcmm_m1_nfl$AUC
  )
) 

# Plot of Cox models
ggplot(plot_data %>% filter(grepl("Cox", Model)),
       aes(x = Time, y = AUC, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  labs(
    title = "Time-dependent AUC comparison",
    x = "Time (years)",
    y = "AUC"
  ) +
  theme_minimal()

# Plot of joint models
ggplot(plot_data %>% filter(grepl("J", Model)),
       aes(x = Time, y = AUC, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  labs(
    title = "Time-dependent AUC comparison",
    x = "Time (years)",
    y = "AUC"
  ) +
  theme_minimal()
```

```{r}
png("figures/auc_plot_cox.png", width = 400, height = 300)
ggplot(plot_data %>% filter(grepl("Cox", Model)),
       aes(x = Time, y = AUC, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  labs(
    title = "Time-dependent AUC Comparison for Cox Models",
    x = "Time (years)",
    y = "AUC"
  ) +
  theme_minimal()
dev.off()

png("figures/auc_plot_jm.png", width = 400, height = 300)
ggplot(plot_data %>% filter(grepl("J", Model)),
       aes(x = Time, y = AUC, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  labs(
    title = "Time-dependent AUC Comparison for Joint Models",
    x = "Time (years)",
    y = "AUC"
  ) +
  theme_minimal()
dev.off()
```

# Comparing models using time-depending Brier score

## Time-dependent Brier score for fitted Cox PH models

```{r}
# We load required library
 library(pec)
 
# We specify Cox models for comparison
cox_models <- list(
   "Cox p-tau" = surv_model_ptau_bl_final,
   "Cox NfL" = surv_model_nfl_bl_final,
   "Cox both" = surv_model_both_bl_final
 )
 
# We compute Brier score using pec for Cox models
brier_cox <- pec(
   object = cox_models,
   formula = Surv(time_to_event, event) ~ 
     age_at_baseline + PTGENDER + APOE4_status,
   data = surv_data_out,
   times = times,
   cens.model = "cox"
 )
 
cox_brier <- lapply(brier_cox$AppErr, function(x)
   approx(brier_cox$time, x, xout = times)$y
 )
```

## Time-dependent Brier score for fitted joint models

```{r}
# We define a function to compute time-dependent Brier score for joint models using tvBrier
get_jm_brier <- function(model, times) {
  
  briers <- sapply(times, function(t) {
    
    tmp <- tvBrier(
      model,
      newdata = newdata,
      Tstart = 0,
      Thoriz = t
    )
    
    return(tmp$Brier)
  })
  
  return(as.numeric(briers))
}
```

```{r}
# We compute time-dependent Brier score for the joint model including p-tau181 
brier_ptau_jm <- get_jm_brier(joint_model_ptau, times)
brier_ptau_jm

# We compute time-dependent Brier score for the joint model including NfL 
brier_nfl_jm <- get_jm_brier(joint_model_nfl, times)
brier_nfl_jm

# We compute time-dependent Brier score for the multivariate joint model 
brier_multi_jm <- get_jm_brier(multi_joint_model, times)
brier_multi_jm
```

## Time-dependent Brier score for latent class joint models

```{r}
brier_jlcmm <- function(model, data, ID, time_var, surv_time_var, event_var, times, landmark) {
  
#   Initiating a data frame to save the results in at each horizon time
  output <- data.frame(
    times = times,
    Brier = NA,
    IBS = NA
  )
  
#   Calculating the survival predictions previously defined function
  if(class(model) == "mpjlcmm") {
    dfpred <- predict_mjlcmm
    (
      model = model,
      data = data,
      ID = ID,
      landmark = landmark,
      times = times
      )
  }
  
#   Calculating the dynamic survival predictions from joint models using dynpred() in the lcmm package
  if(class(model) == "Jointlcmm") {
    dfpred <- dynpred(
      model = model,
      newdata = newdata,
      var.time = time_var,
      horizon = times,
      landmark = landmark
      )$pred
  }
  
#   Fetching the true event times and events 
  dftrue <- unique(data[c(ID, surv_time_var, event_var)])
  names(dftrue) <- c(ID, "time", "event")
  
#   Merging prediction df with true df
  merged <- merge(dfpred, dftrue, by = ID)
  
#   Saving censoring indicator (1 - event indicator), times remain the same as event/censured times from the event indicator
  merged$cens_event <- 1 - merged$event
  
#  Estimting the censoring distribution with the Kaplan-Meier estimate
  km_cens <- survfit(Surv(merged$time, merged$cens_event) ~ 1)
  
#   Function to evaluate estimate of censuring probability
  Ghat <- function(t) {
    sapply(t, function(x) {
      idx <- max(which(km_cens$time <= x))
      if (length(idx) == 0) return(1)
      km_cens$surv[idx]
    })
  }
  
  brier_one_time <- function(t) {
    
#    Saving all rows belonging to horizon t
    sub <- merged[merged$horizon == t, ]
#    Saving the observed survival status
    status_t <- as.numeric(sub$time > t)
#    Saving the predicted survival probability: 1 - pred since dynpred outputs cumulative risk (event probability) instead of survival probabilities 
    pred_t <- 1 - sub$pred

#    Calculating the IPCW weights
    w <- ifelse(sub$time > t,
                1 / Ghat(t),
                ifelse(sub$event == 1 & sub$time <= t,
                       1 / Ghat(sub$time),
                       0))
    
#     Calculating the IPCW-adjusted Brier score
    mean(w * (status_t - pred_t)^2, na.rm = TRUE)
  }
  
#   Calculating the Brier score for each horizon
  brier_scores <- sapply(times, brier_one_time)
#   Saving to output
  output$Brier <- brier_scores
  
#   Integrated Brier Score (IBS)
  IBS <- sum(diff(times) *
               (brier_scores[-1] + brier_scores[-length(brier_scores)]) / 2) /
         (max(times) - landmark)
#   Saving to output
  output$IBS <- IBS
  
  return(output)
}
```

```{r}
# Computing Brier scores for 2 class jlcmm for NfL
brier_jlcmm_m1_nfl <- brier_jlcmm(
  model = jlcmm_m1_nfl,
  data = newdata,
  ID = "RID",
  time_var = "time_years",
  surv_time_var = "time_to_event",
  event_var = "event",
  times = times,
  landmark = 1
)

# Computing Brier scores for 2 class jlcmm for ptau
brier_jlcmm_m1_ptau <- brier_jlcmm(
  model = jlcmm_m1_ptau,
  data = newdata,
  ID = "RID",
  time_var = "time_years",
  surv_time_var = "time_to_event",
  event_var = "event",
  times = times,
  landmark = 1
)

# Computing Brier scores for 3 class jlcmm for ptau
brier_jlcmm_m2_ptau <- brier_jlcmm(
  model = jlcmm_m2_ptau,
  data = newdata,
  ID = "RID",
  time_var = "time_years",
  surv_time_var = "time_to_event",
  event_var = "event",
  times = times,
  landmark = 1
)

# ON IN-SAMPLE DATA 
# Computing Brier scores for 2 class multivariate jlcmm for ptau & NfL
brier_mult_jlcmm_m1 <- brier_jlcmm(
  model = mult_jlcmm_m1,
  data = jlcmm_data,
  ID = "RID",
  time_var = "time_years",
  surv_time_var = "time_to_event",
  event_var = "event",
  times = times,
  landmark = 1
)
```

## Table of the results

```{r}
# Time-dependent Brier score for Cox and joint models summarized in a table
results_table_Brier <- data.frame(
  Time = times,

  Cox_p_tau = cox_brier$`Cox p-tau`,
  Cox_NfL = cox_brier$`Cox NfL`,
  Cox_both = cox_brier$`Cox both`,

  JM_p_tau = brier_ptau_jm,
  JM_NfL = brier_nfl_jm,
  # JM_both = brier_multi_jm,
  
  JLCMM_2_NfL = brier_jlcmm_m1_nfl$Brier,
  JLCMM_2_ptau = brier_jlcmm_m1_ptau$Brier
  # JLCMM_3_ptau = brier_jlcmm_m2_ptau$Brier
)
round(results_table_Brier, 3)
```

```{r}
# Saving the table of the Brier score results
saveRDS(results_table_Brier, "tables/results_table_Brier.rds")
```

## Plot of the results

```{r}
plot_data <- results_table_Brier %>%
  pivot_longer(
    cols = -Time,
    names_to = "Model",
    values_to = "Brier"
  )

# plot for the Cox models
ggplot(plot_data %>% filter(grepl("Cox", Model)),
       aes(x = Time, y = Brier, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  theme_minimal() +
  labs(
    title = "Time-dependent Brier Score for Cox Models",
    x = "Time",
    y = "Brier score"
  )

# plot for the joint models
ggplot(plot_data %>% filter(grepl("J", Model)),
       aes(x = Time, y = Brier, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  theme_minimal() +
  labs(
    title = "Time-dependent Brier Score for Joint Models",
    x = "Time",
    y = "Brier score"
  )
```

```{r}
# Saving plots of the Brier score results
png("figures/Brier_plot_cox.png", width = 400, height = 300)
ggplot(plot_data %>% filter(grepl("Cox", Model)),
       aes(x = Time, y = Brier, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  theme_minimal() +
  labs(
    title = "Time-dependent Brier Score for Cox Models",
    x = "Time",
    y = "Brier score"
  ) +
  theme_minimal()

dev.off()

png("figures/Brier_plot_jm.png", width = 400, height = 300)
ggplot(plot_data %>% filter(grepl("J", Model)),
       aes(x = Time, y = Brier, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0, 1) +
  theme_minimal() +
  labs(
    title = "Time-dependent Brier Score for Joint Models",
    x = "Time",
    y = "Brier score"
  ) +
  theme_minimal()

dev.off()
```
