---
format:
  pdf:
    include-before-body:
      text: |
        \begin{titlepage}
        \raggedleft
        \vspace*{2cm}
        
        {\Huge\bfseries Kandidatuppsats\par}
        \vspace{1.5cm}
        
        {\Large Statistiska Instiutionen\par}
        \vspace{3cm}
        
        {\Huge\bfseries Nr 2026:x\par}
        \vspace{0.5cm}
        
        {\LARGE \bfseries Title in English\par}

        {\large Title in Swedish\par}
        \vspace{0.5cm}
        
        {\large \today\par}
        \vspace{3cm}
        
        {\Huge Charlotte Emily Iszatt and Milou Rossi\par}
        
        \vfill
        
        \end{titlepage}
        
        \begin{titlepage}
        \raggedleft
        
        {\normalsize Självständigt arbete 15 högskolepoäng inom Statistik III, VT2026\par}
        {\normalsize Handledare: Hans Nyqvist\par}
        \vspace{2cm}
        
        \raggedright
        {\huge\bfseries Abstract\par}
        \vspace{1cm}
        
        {\Large The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. Keywords: XXX, XXX, XXX\par}
        \vspace{3cm}
        
        \vfill
        
        \end{titlepage}
    toc: false
    colorlinks: false
    number-sections: true
editor: visual
---

```{=latex}
\tableofcontents

\newpage
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, output = FALSE)
```

```{r}
# loading libraries
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
```

```{r}
# load data sets: Charley
# loading adni library
library(ADNIMERGE2)

# biomarker data set
plasma_data = UPENN_PLASMA_FUJIREBIO_QUANTERIX

# PET scan data
amyloid_data = UCBERKELEY_AMY_6MM

# dx data
diagnosis_data = DXSUM

# APOE GENOTYPING
apoe_data = APOERES

# deprivation index
deprivation = ADI

# demographics
demog1 = PTDEMOG
demog2 = DM %>% rename(RID = SUBJID) %>% mutate(RID = as.numeric(RID))

demographics = demog1 %>%
  inner_join(demog2, by = c("RID"))



# # loading data sets: Milou# 
# # Plasma-biomarkörer
# load(file.path(outdir, "ADNIMERGE2/data/UPENN_PLASMA_FUJIREBIO_QUANTERIX.rda"))
# plasma_data <- UPENN_PLASMA_FUJIREBIO_QUANTERIX
# 
# # Amyloid-PET
# load(file.path(outdir, "ADNIMERGE2/data/UCBERKELEY_AMY_6MM.rda"))
# amyloid_data <- UCBERKELEY_AMY_6MM
# 
# # Diagnos
# load(file.path(outdir, "ADNIMERGE2/data/DXSUM.rda"))
# diagnosis_data <- DXSUM

# # Demographics
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
# demographics <- PTDEMOG

# # Genotype data
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
# apoe_data <- APOERES 
```

# Introduction

## Background

General background information.

Introducing some domain specific information about AD diagnosis and pathology, so that the reader is well prepared.

## Research Context

Assessing the current landscape in terms of the scientific literature.

What is known?

What has been done?

Where are there gaps??? (This leads on to the next section, why are we doing this research, how do we hope to answer some remaining question in the AD research, orient this towards the statistical analysis/approaches!)

## Research Aims

What are our research questions?

Why is this research relevant to the existing body of literature? What does it add? Gaps in previous literature which we outline in above \^\^.

How will we accomplish this?

# Method

Introduce the methods we will use.

Why did we choose this method to answer the research question?

## Linear Mixed Effect Model

**Theory about the model... (specifically why we should use it)**

-   Biomarker process

-   Describes the trajectory of biomarkers at the individual level 

**Linear mixed effects models?** What are these - why use them here?

$$
y_i(t) = m_i(t) + \epsilon_i(t) \qquad \epsilon_i(t) \sim N(0, \sigma^2)
$$

$$
m_i(t) = \mathbf{X_i^T}\beta + \mathbf{Z_i}^T\mathbf{b_i} \qquad \mathbf{b_i}(t) \sim N(0, \sigma^2)
$$

$y_i(t):$ the observed longitudinal biomarker

$m_i(t):$ the trajectory function which represents the **true unobserved** **value of the biomarker** at time $t$ for patient $i$ - we assume that we do not observe the true value due to measurement errors or other reasons.

$\mathbf{X_i}^T(t)\beta:$ the fixed effects

$\mathbf{Z_i}^T\mathbf{b_i}:$the random effects

### Assumptions!!

-   Random effects $\mathbf{b_i}$ are **time-invariant** - they are assumed to come from a normal distribution, $\mathbf{b_i} \sim N(0, \sigma^2)$, and represent stable, latent characteristics about the individuals disease, such as a generally fast declining disease trajectory but cannot represent other factors that effect disease trajectory over-time, e.g. medication. Any external time-varying process that effects the biomarker trajectory and the hazard cannot be represented by the random effects.

## Proportional Hazard Model 

**Theory about the model... (specifically why we should use it)**

-   Time-to-diagnosis process

-   Parametric baseline hazard: **flexible parametrics, splines??**

-   OR leave undefined -\> Cox PH model

Assuming a Cox PH survival sub-model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha m_i(t) \right\}
$$

$\mathbf{v_i}:$ vector of time-independent co-variates

$\phi:$ vector of log-hazard ratios

$\alpha:$ the association parameter

$M_i = \left\{m_i, \quad 0 \le s \le t\right\}:$ the **true unobserved biomarker profile** up to time $t$

### Assumptions!!

-   What do we assume about the censoring process?

-   Proportionality

## Joint Model

**Theory about the model... (specifically why we should use it)**

In the joint model, the hazard function at time $t$ is directly dependent on the estimate of the biomarker at time $t$.

The association structure of the joint model:

-   How is the longitudinal model linked to the survival model? 

-   Does this association structure change for different biomarkers?

-   What happens to time-to-diagnosis predictions if we assume the standard structure when another underlying association is a better fit?

If we include the current rate of change alongside the current value in the proportional-hazard model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha_1 m_i(t) + \alpha_2 m_i'(t) \right\} \qquad where \qquad m_i'(t) = \frac{d}{dt} m_i(t)
$$

$\alpha_1 :$ links the current value of the biomarker - $m_i(t)$

$\alpha_2 :$ links the current gradient/rate of change - $m_i'(t)$

Joint-model options:

-   Longitudinal sub-model: model of class lme()

### Assumptions!!

-   Full conditional independence

    -   Random effects $\mathbf{b_i}$ explain all statistical dependence between (1) the longitudinal outcome and the survival outcome and (2) repeated measurements within the longitudinal process.

    -   This assumption can fail when: there are time-varying confounders not captured by the random effects (see the longitudinal model for more info about $\mathbf{b_i}$ assumptions), if the measurement error, $e_i(t)$ in the longitudinal model $y_i(t) = m_i(t) + e_i(t)$ is informative, or there are sub-populations (APOE genotype??? Maybe we need to split into populations based on genotype)

-   So: the longitudinal outcome is independent of the time-to-event outcome and the repeated measurements in the longitudinal outcome are independent of each other

-   $\mathbf{b_i}$ drives the **true (unobserved)** biomarker trajectory and hazard of AD diagnosis: the observed $y_i(t)$ is the observed biomarker state, $m_i(t)$, plus noise. Where $m_i(t)$ is itself determined by $\mathbf{b_i}$, \*\*since the others are fixed ????? \*\*, once we know $\mathbf{b_i}$, the observed biomarker measurement and the survival times are no longer dependent. The observed $y_i$'s are useful in determining $\mathbf{b_i}$ which is the useful information in estimating survival times.

# Data

## ADNI

Introducing the study, it's aims and the type of information gathered.

**IMPORTANT: REMEMBER WE NEED TO REFERENCE ADNI PROPERLY.**

Explain the different protocols and what they mean for data analysis, if we can compare across protocols etc.

## Introducing variables... bio-markers/amyloid data...

Information about the variables we will use and why, what they can show, how are they recorded,

## Pre-processing

ADNI documentation: <https://adni.loni.usc.edu/quick-start-guide-asset/anatomy2.html>

Some bullet points to add to the final paper about things we thought about/did:

-   Merging the diagnosis data with biomarker/amyloid data by RID (unique identifier for patients)

-   Selecting relevant variables from diagnosis and biomarker/amyloid data set

-   What variables do we need for longitudinal sub-model? (Repeated measurement, time followed?)

-   What variables do we need for survival sub-model? (Event, time followed?)

-   Joint model: time-var: a character string indicating the time variable in the mixed-effects model

-   How many observations going from CN/MCI -\> AD? Do we need a multi-state survival model?

# Results

Present the results from our analyses.

Do not interpret results here! That should be done in the discussion.

## Joint Model

Results from the joint model

## Comparison of predictive accuracy...

Results from comparison of models... not sure what we will do but you get the idea haha

# Discussion

## Main findings

Interpret the results. What did we find?

## Relate to the research question(s)/previous literature

How do our results help to answer the research question?

Anything unexpected/not in accordance with previous studies?

## Further research

Suggest what kinds of questions this analysis leaves us with?

What more can be done/looked at?

# Conclusions

Summary

# References

<https://www.mybib.com/tools/harvard-referencing-generator>

```{r, warning=FALSE}
# MERGING DATA TEST
# merged_data <- plasma_data %>%
#   inner_join(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx")) %>%
#   mutate(time_diff = abs(EXAMDATE_plasma - EXAMDATE_dx)) %>%
#   group_by(RID, EXAMDATE_plasma) %>%
#   slice_min(time_diff, n = 1) %>%
#   ungroup() %>%
#   filter(time_diff <= 90)
```

# 2. Pre-processing

```{r}
# selecting the relevant variables -> why these ones???
plasma_data = plasma_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "pT217_F",
         "AB42_F", "AB40_F", "AB42_AB40_F", "pT217_AB42_F", "NfL_Q", "GFAP_Q")

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = plasma_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx")) %>%
  mutate(dx = case_when(DIAGNOSIS == "Dementia" ~ 1, TRUE ~ 0))
  
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", merged_data$PTID)

# checking for missing values in EXAMDATE_dx and EXAMDATE_plasma
sum(as.numeric(is.na(merged_data$EXAMDATE_plasma)))
sum(as.numeric(is.na(merged_data$EXAMDATE_dx)))

# removing the obs with no EXAMDATE_dx
merged = merged_data %>% filter(as.character(EXAMDATE_dx) != "NA")
```

```{r}
unique((plasma_data %>% group_by(RID) %>% filter(n() > 2))$RID)
```

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((plasma_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# 2) Skapa baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_plasma)) %>%
  ungroup()

# 3) Skapa event + time_to_event (Dementia räknas som event)
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarize(
    # Event = 1 om individen någonsin fått "Dementia", annars 0 (censurerad)
    event = ifelse(any(DIAGNOSIS == "Dementia"), 1, 0),
    
    # Tid till första "Dementia", annars tid till sista observation
    time_to_event = as.numeric(
      ifelse(
        any(DIAGNOSIS == "Dementia"),
        min(EXAMDATE_plasma[DIAGNOSIS == "Dementia"]) - min(EXAMDATE_plasma),  # dagar till event
        max(EXAMDATE_plasma) - min(EXAMDATE_plasma)                               # censurerad tid
      )
    ) / 365.25  # konvertera till år
  ) %>%
  ungroup()

# 4) Lägg survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_plasma - baseline_date) / 365.25)
```

```{r}
# We check NA's
colSums(is.na(merged_data[, c("pT217_F", "time_years", "DIAGNOSIS")]))
```

# 3. Longitudinal model

```{r}
# Longitudinal model

long_model <- lme(
  pT217_F ~ time_years, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data,
  na.action = na.omit
)

summary(long_model)
```

```{r}
# Residualer
residuals <- resid(long_model)

# Förväntade värden
fitted_values <- fitted(long_model)

plot(fitted_values, residuals,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted values")
abline(h = 0, col = "red", lty = 2)
```

```{r}
qqnorm(residuals)
qqline(residuals, col = "red")
```

# 4. Survival model

```{r}

# Cox survival-modell
surv_model <- coxph(
  Surv(time_to_event, event) ~ 1,   # ~1 = ingen covariat just nu
  data = surv_data)

# Kontrollera modellen
summary(surv_model)
```

# 5. Joint-model

```{r}
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```

```{r}

# joint_model <- jm(
#   Surv_object = surv_model,      # Cox-survivalmodell
#   Mixed_objects = long_model,       # Longitudinell modell
#   time_var = "time_years"        # tid i longitudinal data
# )
# 
# summary(joint_model)
```

# Kovariater som kan vara till användning?

```{r, warning=FALSE}
head(demographics[, c("RID", "PTDOBYY", "PTGENDER")])
```

```{r, warning=FALSE}
# Vi lägger till AGE och SEX i merged_data
merged_data <- merged_data %>%
  left_join(demographics %>% select(RID, PTDOBYY, PTGENDER), by = "RID") %>%
  mutate(
    AGE = as.numeric(format(EXAMDATE_plasma, "%Y")) - PTDOBYY,
    SEX = PTGENDER
  )
```

Vi kan kolla på andra variabler också såsom utbildning och etnicitet (variablerna PTEDUCAT och PTETHCAT/PTRACCAT) –\> lägga till de i cox-modellen (survival-modellen)

# Andra sätt vi kan utveckla vår uppsats

1\. Om vi vill lägga till kovariatet APOE4 = genetisk riskfaktor för AD (t.ex. antal alleler: 0,1,2). Kanske i variabeln "GENOTYPE" –\> lägga till de i cox-modellen (survival-modellen)

2\. Lägg till interaktioner i LME (t.ex. time_years \* AGE)

3\. Residualdiagnostik för LME

4\. Kolla antagandet om proportional hazards (att hazard ratio är konstant över tid)

5\. Sensitivitetsanalyser, t.ex. ändra definitionen av event, såsom olika kriterier för AD-diagnos.

**Användning av flera biomarkörer:**

6\. Analysera flera biomarkörer separat - skapa enskilda modeller för varje biomarkör (t.ex. pT217, AB42, tau) för att se deras individuella effekter.

7\. Modellera flera biomarkörer samtidigt - bygga en multivariat joint-modell där varje biomarkör är ett longitudinellt utfall i modellen.

8\. Kombinera biomarkörer till ratio eller index - t.ex. AB42/AB40 eller pTau/AB42, och använda dessa som en enda variabel i modellen.

9\. Reducera dimension med **PCA** - göra principal component analysis på flera biomarkörer och använda komponenter som utfall i joint-modellen.

10\. Jämföra modeller med olika biomarkörssammansättningar - utvärdera skillnader i prediktion och signifikans för olika kombinationer av biomarkörer.
