---
title: "test"
format: pdf
editor: visual
---

```{r, warning=FALSE}
# We load necessary packages 
library(dplyr)     # for data manipulation
```

## 1. Merging the datasets

```{r}
# We load plasma data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UPENN_PLASMA_FUJIREBIO_QUANTERIX.rda")
plasma_data <- UPENN_PLASMA_FUJIREBIO_QUANTERIX

# We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM

# We merge the two datasets using patient ID
merged_data <- merge(plasma_data, diagnosis_data, by = "RID")
```

# 2. Pre-processing

## 2.1 We select only the variables needed for longitudinal + survival + joint model

```{r}
merged_data <- merged_data[, c(

  # Patient ID
  "RID",

  # Visit indicators (used to detect baseline)
  "VISCODE2.x",   # plasma visit code
  "VISCODE2.y",   # diagnosis visit code

  # Visit dates
  "EXAMDATE.x",   # plasma visit date
  "EXAMDATE.y",   # diagnosis visit date

  # Longitudinal biomarker
  "pT217_F",

  # Diagnosis (used to define the event)
  "DIAGNOSIS"

  # Optional additional biomarkers (can be useful as covariates)
  #"AB42_F",
  #"AB40_F",
  #"AB42_AB40_F",
  #"NfL_Q",
  #"GFAP_Q"
)]
```

## 2.2 We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We load the demographics dataset
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG

# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# We merge age and gender into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
)

# We make sure that gender is categorical (factor)
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# We create age at each visit using the visit date (EXAMDATE.x)
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE.x, "%Y")) - merged_data$PTDOBYY
```

## 2.3 We create time variable for longitudinal model

We need this because the longitudinal model require a numeric measure of follow-up time to estimate how pT217_F changes over time.

```{r, warning=FALSE}
# Longitudinal time variable (from plasma visits)
merged_data$time_months <- as.numeric(sub("m", "", merged_data$VISCODE2.x))
merged_data$time_months[merged_data$VISCODE2.x == "bl"] <- 0

# We check that time variables were created correctly
table(merged_data$time_months)
```

## 2.4 We extract baseline visits (first plasma visit per patient)

```{r}
# First baseline plasma per person
baseline_plasma <- merged_data %>%
  filter(VISCODE2.x == "bl") %>%       # baseline plasma
  group_by(RID) %>%
  arrange(EXAMDATE.x) %>%
  slice(1) %>% # We choose the earliest baseline date 
  ungroup()
```

## 2.5 We keep only diagnosis visits after baseline

So that we don't get negative time-to-event values

Det fanns nämligen typ 300 observationer med diagnosdatum innan baseline plasmabesöken, om vi tar med de får vi t.ex. -300 i time-to-event

```{r}
# We keep only diagnosis visits after baseline 
baseline_plus_diag <- merged_data %>%
  filter(EXAMDATE.y >= baseline_plasma$EXAMDATE.x[match(RID, baseline_plasma$RID)])  # diagnosis after baseline
```

## 2.6 We create surv_data with an event variable (conversion to AD) and a time to event variable

```{r}
# We create surv_data: one row per person with event and time to event
surv_data <- baseline_plus_diag %>%
  group_by(RID) %>%
  summarise(
    baseline_date = min(EXAMDATE.x),
    event = ifelse(any(DIAGNOSIS == "Dementia"), 1, 0),
    time_to_event = ifelse(
      event == 1,
      min(EXAMDATE.y[DIAGNOSIS == "Dementia"] - baseline_date),  # time to first dementia (in days)
      max(EXAMDATE.y - baseline_date)                             # censored time
    ),
    .groups = "drop"
  )

table(surv_data$event)
```

## 2.7 We include demographics, to use in a coxph model

```{r}
# We merge age_at_baseline and PTGENDER into surv_data
surv_data <- surv_data %>%
  left_join(
    merged_data %>% select(RID, age_at_visit, PTGENDER) %>% 
      group_by(RID) %>% 
      slice(1) %>%  # keep first row per person (baseline)
      ungroup(),
    by = "RID"
  ) %>%
  rename(age_at_baseline = age_at_visit)
```

## We check the structure of surv_data

```{r}
head(surv_data)
```

## We check for negative event times or missing values in surv_data

```{r}
# We check for negative event times
negative_times <- surv_data %>%
  filter(time_to_event < 0)
nrow(negative_times)

# We check for NAs in event or time_to_event
nas <- surv_data %>%
  filter(is.na(event) | is.na(time_to_event))
nrow(nas)

# We check also for duplicates 
any(duplicated(surv_data$RID))
```

## To check if the event and time_to_event is correct (comparing surv_data with merged_data)

```{r}
# We extract RID = 42 from surv_data
surv_rid10 <- surv_data %>%
  filter(RID == 42)

surv_rid10  # shows baseline_date, event, time_to_event_days for RID 10

# We extract all rows with RID 42 from merged_data
merged_rid10 <- merged_data %>%
  filter(RID == 42)

merged_rid10  # shows all visits for RID 42, plasma and diagnosis
```

## We check missing values after pre-processing

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

## We remove all missing values

```{r}
merged_data <- merged_data %>%
  filter(!is.na(pT217_F) &
         !is.na(time_months) &
         !is.na(age_at_visit) &
         !is.na(PTGENDER) &
         !is.na(EXAMDATE.y) &
         !is.na(DIAGNOSIS))

sapply(merged_data, function(x) sum(is.na(x)))
```

# 3. Longitudinal model

## 3.1 We remove participants with only 1 measurement 

```{r}
# We check the number of measurements per subject
table(table(merged_data$RID))  # counts of measurements per individual

# We keep only RIDs with more than 1 measurement
counts <- table(merged_data$RID)
keep_rids <- names(counts[counts > 1])

# Subset merged_data
merged_data <- merged_data[merged_data$RID %in% keep_rids, ]
```

## 3.2 Fitted LME (linear mixed-effects) model

```{r}
library(nlme) 
# We fit longitudinal 
long_model <- lme(
  pT217_F ~ time_months + age_at_visit + PTGENDER,
  random = ~ time_months | RID,
  data = merged_data
)
summary(long_model)

# We confirm class
class(long_model)  # should return "lme" "nlme"
```

## 3.2 Residuals

```{r}
# Residuals
residuals <- resid(long_model)

# Fitted values
fitted_values <- fitted(long_model)

# Plot of residuals vs fitted values 
plot(fitted_values, residuals,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted values")
abline(h = 0, col = "red", lty = 2)
```

## 3.3 QQ plot

```{r}
qqnorm(residuals)
qqline(residuals, col = "red")
```

# 4. Survival model

## 4.1 We create a survival object

```{r}
library(survival)  
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

## 4.2 Kaplan-Meier curve

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline plasma", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

## 4.3 Fitted Coxph model

```{r}
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

# 5. Joint-model

```{r}
library(JMbayes2)
joint_model_bayes <- jm(
  long_model,
  surv_model,
  time_var = "time_months"
)
```

# Andra sätt vi kan utveckla vår uppsats

1\. Om vi vill lägga till kovariatet APOE4 = genetisk riskfaktor för AD (t.ex. antal alleler: 0,1,2). Kanske i variabeln "GENOTYPE" –\> lägga till de i cox-modellen (survival-modellen)

```{r}
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe_data <- APOERES  
```

1\. Lägga till fler kovariater, t.ex. utbildning och etnicitet (variablerna PTEDUCAT och PTETHCAT/PTRACCAT)

2\. Lägga till interaktioner i modellerna (t.ex. time_years \* AGE)

3\. Residualdiagnostik för LME

4\. Kolla antagandet om proportional hazards (att hazard ratio är konstant över tid)

5\. Sensitivitetsanalyser, t.ex. ändra definitionen av event, såsom olika kriterier för AD-diagnos.

**Användning av flera biomarkörer:**

6\. Analysera flera biomarkörer separat - skapa enskilda modeller för varje biomarkör (t.ex. pT217, AB42, tau) för att se deras individuella effekter.

7\. Modellera flera biomarkörer samtidigt - bygga en multivariat joint-modell där varje biomarkör är ett longitudinellt utfall i modellen.

8\. Kombinera biomarkörer till ratio eller index - t.ex. AB42/AB40 eller pTau/AB42, och använda dessa som en enda variabel i modellen.

9\. Reducera dimension med **PCA** - göra principal component analysis på flera biomarkörer och använda komponenter som utfall i joint-modellen.

10\. Jämföra modeller med olika biomarkörssammansättningar - utvärdera skillnader i prediktion och signifikans för olika kombinationer av biomarkörer.
