---
format:
  pdf:
    include-before-body:
      text: |
        \begin{titlepage}
        \raggedleft
        \vspace*{2cm}
        
        {\Huge\bfseries Kandidatuppsats\par}
        \vspace{1.5cm}
        
        {\Large Statistiska Instiutionen\par}
        \vspace{3cm}
        
        {\Huge\bfseries Nr 2026:x\par}
        \vspace{0.5cm}
        
        {\LARGE \bfseries Code \par}
        
        {\large \today\par}
        \vspace{3cm}
        
        {\Huge Charlotte Emily Iszatt and Milou Rossi\par}
        
        \vfill
        
        \end{titlepage}
        
    toc: false
    colorlinks: false
    number-sections: true
editor: visual
---

```{=latex}
\tableofcontents

\newpage
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, output = TRUE)
```

# Pre-processing

```{r, warning=FALSE}
# loading libraries
library(dplyr)
library(tidyr)
library(JMbayes2)
library(JM)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) # for running residual diagnostics on GLMM
library(lcmm)
library(timeROC)
library(kableExtra)
```

## Loading the data

The data was loaded from the ADNIMERGE2 package accessed via the Alzheimer's Disease Neuroimaging Initiative (ADNI). The data sets include longitudinal ptau-181 and NfL measurements, diagnosis, demographic and genetic data.

```{r}
# Loading data
#Milou
 # NfL data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
  nfl_data <- BLENNOWPLASMANFLLONG
 # # P-tau181 data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
  ptau_data <- UGOTPTAU181
 # # Diagnosis data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
  diagnosis_data <- DXSUM
 # # Demographics data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
  demographics <- PTDEMOG
 # # Apoe data
  load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
  apoe <- APOERES
```

```{r}
# Charley

library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

apoe = APOERES
```

- *There are 52 individuals with two measurements recorded on the same exam date.*

- *How should we deal with these? Average the measurements? Remove the observations?*

1.  *Describe the measurements, their differences and averages*

- *Some of the values are quite different, what do we assume by averaging them? Do we introduce bias but just just removing them?*

2.  *Show how different these values are from the nfl and ptau averages (report SD differences in log transformations)*
3.  

```{r}
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", diagnosis_data$PTID)
grep("381_S_10###", apoe$PTID)
```

## Selecting required variables

Variables for the unique identifier for each individual, visit code, exam date, biomarker measurements, diagnosis and genetic data were extracted.

```{r, warning=FALSE}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required genotype variables
apoe <- apoe %>%
  dplyr::select(RID, GENOTYPE)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))
```

## Checking RID with more than one measurement recorded on the same exam date

For participants with duplicate measurements on the same examination date, differences were observed depending on how they were handled: using the first measurement, the mean, or the last measurement. The mean value was used as the primary approach, and sensitivity analyses were performed to evaluate the robustness of the findings.

```{r}
## Checking which obs have double VISCODE2 codes
# for ptau data
ptau_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# for nfl data
nfl_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# changing VISCODE2 for RID 1097 to m60_1 and m60_2
ptau_data[ptau_data$RID == 1097, "VISCODE2"] <- c("m48", "m60_2", "m60_1")

nfl_data[nfl_data$RID == 1097, "VISCODE2"] <-  c("m48", "m60_1", "m60_2")
```

```{r}
# Replacing visits with double measurements with visits with mean-value of the measurements as well as saving both measurements for future sensitivity analysis
# Naming the mean as the original variable to do analysis first with this method and then sensitivity analysis with the first and last measurements

# Adding columns for each ptau measurement and their mean and then collapsing down to unique RID and EXAMDATE
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_ptau = first(PLASMAPTAU181),
    last_ptau = last(PLASMAPTAU181),
    PLASMAPTAU181 = mean(PLASMAPTAU181),
    ) %>%
  ungroup() %>%
  distinct()

# Adding columns for each nfl measurement and their mean and then collapsing down to unique RID and EXAMDATE
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_nfl = first(PLASMA_NFL),
    last_nfl = last(PLASMA_NFL),
    PLASMA_NFL = mean(PLASMA_NFL)
    ) %>%
  ungroup() %>%
  distinct()
```

```{r}
# checking the difference in method for those with 2 obs on the same examdate
# for ptau
head(ptau_data %>%
       filter(PLASMAPTAU181 != first_ptau) %>%
       dplyr::select(PLASMAPTAU181, first_ptau, last_ptau)
     )

# for nfl
head(nfl_data %>%
       filter(PLASMA_NFL != first_nfl) %>%
       dplyr::select(PLASMA_NFL, first_nfl, last_nfl)
     )
```

## Merging NfL and ptau-181 data for the longitudinal data set

```{r}
# We merge NFL and ptau data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))
```

## Creating a common EXAMDATE and EXAMDATE_baseline variable

This was performed before preparation of the other variables for the survival and longitudinal models in order to use EXAMDATE_baseline to remove any observations from the diagnosis data set which occur long before the first biomarker measurements for that individual.

```{r}
# checking how many observations where the EXAMDATE is different for nfl and ptau
diff <- long_data %>% mutate(diff = as.numeric(EXAMDATE_nfl != EXAMDATE_ptau))
table(diff$diff)
```

```{r}
# filtering out the observations with different EXAMDATE
diff <- diff %>% filter(diff == TRUE)

# checking how many of these are baseline visits
table(diff$VISCODE2)

# # checking which observations have the same examdate for nfl and ptau
# common <- which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# # checking the observations with different exam dates
# long_data[-common, ]
```

- There are 5 observations with time differences greater than 30 days

- 231/115 days: clearly we can't say these are the same date... BUT with setting a new EXAMDATE as the midpoint between dates means the new EXAMDATE is 2 months out from EXAMDATE_nfl and EXAMDATE_ptau

- Less clear what the case is with a time difference of 31 days, even 49 days

```{r}
# calculating the difference in time between EXAMDATE_nfl and EXAMDATE_ptau
diff <- diff %>% mutate(time_diff = EXAMDATE_nfl - EXAMDATE_ptau)

# fetching the time differences
diff$time_diff 

# checking how many observations have a time difference over 30 days (1 month)
diff %>% filter(abs(time_diff) > 30) %>% dplyr::select(time_diff)
```

### EXAMDATE is EXAMDATE_nfl or EXAMDATE_nfl +- time difference

- Can create a new variable for a common EXAMDATE in order to define the baseline date from this

```{r}
# testing the method
diff_examdate <- long_data %>%
  mutate(
    time_diff = EXAMDATE_nfl - EXAMDATE_ptau,
    EXAMDATE_new = case_when(
      time_diff == 0 ~ EXAMDATE_nfl,
      TRUE ~ EXAMDATE_ptau + 0.5 * time_diff
    )
  ) %>%
  filter(time_diff != 0) %>%
  relocate(where(~ inherits(.x, "Date")), time_diff, .after = EXAMDATE_new)
```

```{r}
# creating a common EXAMDATE from two cases
# 1: EXAMDATE_nfl == EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl
# 2: EXAMDATE_nfl != EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl + the difference between the two dates
# then defining the baseline date from the earliest EXAMDATE 
long_data <- long_data %>%
  group_by(RID) %>%
  mutate(
    EXAMDATE = EXAMDATE_ptau + 0.5 * (EXAMDATE_nfl - EXAMDATE_ptau),
    EXAMDATE_baseline = min(EXAMDATE),
    time_years = as.numeric(EXAMDATE - EXAMDATE_baseline) / 365.25) %>%
  ungroup()


```

### Baseline date is the earliest of EXAMDATE_nfl and EXAMDATE_ptau

```{r}
# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
# long_data <- long_data %>%
#   group_by(RID) %>%
#   mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
#   ungroup()
```

## Removing early diagnosis observations

Any diagnoses that were made more than 50 days before the first biomarker measurements were made were removed.

```{r}
# filtering the observations out of the diagnosis data set where EXAMDATE is much before baseline biomarker EXAMDATE
# creating a baseline_dx variable, so we can set time_to_event = 0 if baseline_dx == DEMENTIA
diagnosis_data <- diagnosis_data %>%
  merge(
    long_data %>% dplyr::select(RID, EXAMDATE_baseline), by = "RID"
    ) %>%
  filter(EXAMDATE_dx - EXAMDATE_baseline >= -50) %>%
  distinct() %>%
  group_by(RID) %>%
  arrange(EXAMDATE_dx) %>%
  mutate(
    baseline_dx = first(DIAGNOSIS_dx),
    last_dx = last(DIAGNOSIS_dx)
  ) %>%
  dplyr::select(!EXAMDATE_baseline)
```

## Merging longitudinal and diagnosis data for survival data set

```{r}
# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- long_data %>% left_join(diagnosis_data, by = c("RID"))
```

## Checking missing values

```{r}
# We count missing values per column in long_data: all these are variables we will use in the model, so we can remove with na.omit()
sapply(long_data, function(x) sum(is.na(x)))
```

```{r}
# We count missing values per column in surv_data: VISCODE2.y will not be used in the model, so we need to see how these observations look
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# There are no rows with missing VISCODE2.y with full data for the model variables, so we can remove all with na.omit()
surv_data %>% filter(is.na(VISCODE2.y))
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## Checking correlation between biomarkers

```{r}
# Correlation between the biomarkers 
cor(long_data$PLASMAPTAU181, long_data$PLASMA_NFL, use = "complete.obs")
```

## Merging selected demographic variables (age and gender) and APOE4 status

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  dplyr::select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We define APOE4 status based on number of E4 alleles
apoe <- apoe %>%
  mutate(APOE4_status = as.factor(
    case_when(
      GENOTYPE == "4/4" ~ "2 alleles",
      GENOTYPE %in% c("2/4", "3/4") ~ "1 allele",
      TRUE ~ "0 alleles")),
  genotype = as.factor(GENOTYPE)
  )

apoe$APOE4_status <- factor(apoe$APOE4_status,
                            levels = c("0 alleles", "1 allele", "2 alleles"))

```

```{r}
# We merge gender with long_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We merge apoe with long_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status", "genotype")], by="RID")

# We make sure that gender is a factor
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

### Checking individual's final diagnoses for the full data set

```{r}
# final diagnoses for the full sample size
table(
  (surv_data %>%
     group_by(RID) %>%
     filter(EXAMDATE_dx == last(EXAMDATE_dx)) %>%
     dplyr::select(DIAGNOSIS_dx) %>%
     distinct()
   )$DIAGNOSIS_dx
  )
```

## Selecting participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements of ptau_data
keep <- unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
long_data <- long_data %>% filter(RID %in% keep)
surv_data <- surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

```{r}
# We filter individuals with at least 1 measurement of ptau_data
#keep <- unique((ptau_data %>% group_by(RID) %>% filter(n() > 0))$RID)
#long_data <- long_data %>% filter(RID %in% keep)
#surv_data <- surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
#length(keep)
```

### Checking individuals final diagnoses for the reduced data set

```{r}
# final diagnoses for the reduced sample size
table(
  (surv_data %>%
     group_by(RID) %>%
     filter(EXAMDATE_dx == last(EXAMDATE_dx)) %>%
     dplyr::select(DIAGNOSIS_dx) %>%
     distinct()
   )$DIAGNOSIS_dx
  )
```

## Preparing for Cox PH model

### Constructing global (common) baseline, age at baseline, event and time-to-event variables

```{r}
surv_data <- surv_data %>%
  group_by(RID) %>%
  summarise(
#     carrying these forward to check events and event times are coded correctly
    baseline_dx = unique(baseline_dx),
    last_dx = unique(last_dx),
    
#     baseline EXAMDATE defined as the earlier biomarker EXAMDATE (nfl or ptau)
    EXAMDATE_baseline = min(EXAMDATE_baseline),
    
#     age at baseline visit
    age_at_baseline = age_at_visit[
      EXAMDATE_ptau == min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)]) |
      EXAMDATE_nfl == min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])
      ][1],

#     last dx/nfl/ptau exam
      last_followup = max(c(EXAMDATE_ptau, EXAMDATE_nfl, EXAMDATE_dx), na.rm = TRUE),

#     1 = dementia, 0 = otherwise
      event = as.integer(any(DIAGNOSIS_dx == "Dementia")),

#     0 if dementia dx at baseline, otherwise time to dementia dx or date of censuring
      time_to_event = case_when(
        event == 1 & unique(baseline_dx) == "Dementia" ~ 0,
        event == 1 & unique(baseline_dx) != "Dementia" ~
          as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - EXAMDATE_baseline)/365,
        TRUE ~
          as.numeric(last_followup - EXAMDATE_baseline)/365))
```

```{r}
# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

## Removing individuals with dementia at start

```{r}
# Removing individuals who had a dementia dx at time = 0
surv_data <- surv_data %>% filter(time_to_event > 0)
long_data <- long_data %>% filter(RID %in% unique(surv_data$RID))

# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

### Constructing baseline values of each biomarker, to include in the Cox PH models as covariates

```{r}
# Creating a data frame with the first biomarker measurements
baseline <- long_data %>% 
  group_by(RID) %>%
  arrange(RID, EXAMDATE_nfl) %>% 
  mutate(baseline_value_nfl = first(PLASMA_NFL)) %>%
  arrange(RID, EXAMDATE_ptau) %>%
  mutate(baseline_value_ptau = first(PLASMAPTAU181)) %>%
  dplyr::select(RID, baseline_value_ptau, baseline_value_nfl) %>%
  ungroup() %>%
  distinct()
```

```{r}
# merging the survival data with demographics and the baseline biomarker values
surv_data <- surv_data %>%
  merge(demographics_small, by = "RID") %>% 
  merge(apoe %>% dplyr::select(RID, APOE4_status, genotype), by = "RID") %>%
  merge(baseline, by = "RID")
```

# Checking distribution of each biomarker

```{r}
# We check the distribution of each biomarker
hist(long_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(long_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(long_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(long_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

```{r}
# cube-root transformed
ggplot(long_data, aes(sqrt(sqrt(PLASMA_NFL)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()

# cube-root transformed
ggplot(long_data, aes(sqrt(sqrt(PLASMAPTAU181)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()
```

```{r}
# Transforming the baseline values in the survival data
surv_data$baseline_value_ptau <- log(surv_data$baseline_value_ptau)
surv_data$baseline_value_nfl <- log(surv_data$baseline_value_nfl)
```

## Checking outliers

Checking for extreme observations shows there are no values 3 standard deviations below the mean, only above. An assessment of these more extreme observations suggests that they are within a reasonable range and are assumed to be correct observations.

```{r}
# We identify "extreme" values

# Values greater than 3 sd above ptau mean
long_data[scale(long_data$PLASMAPTAU181) > 3, ]
# Values smaller than 3 sd below ptau mean
long_data[scale(long_data$PLASMAPTAU181) < -3, ]

# Values greater than 3 sd above nfl mean
long_data[scale(long_data$PLASMA_NFL) > 3, ]
# Values smaller than 3 sd below nfl mean
long_data[scale(long_data$PLASMA_NFL) < -3, ]
```

## Rechecking longitudinal and survival data lengths

```{r}
# We check the number of individuals in each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

# ...

# Longitudinal models

## Comparing longitudinal models

Suggested models:

- Gamma: raw data could be modelled well with the gamma distribution since takes no negative values, variance increases with mean, and there is a right-skew to the raw distributions

- Linear mixed effects model with log-transformed data: approximately normal, yet still some skew

- Linear mixed effects model with cube root transformed data: again, approximately normal, still a bit of skew (why such a low AIC? seems wrong?)

- GLMM with log-normal distribution

- GLMM with Student-t distribution: longer tails

### PTAU

```{r}
# checking the mean and variance for ptau181
mean(long_data$PLASMAPTAU181)
var(long_data$PLASMAPTAU181)
```

#### Fitting the models

```{r}
# GLMM
fit_gamma_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID), 
                   data = long_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_ptau <- lmer(log(PLASMAPTAU181) ~ time_years + (1|RID), 
                data = long_data)

# LME approach with log transformed data
fit_lme_cubed_ptau <- lmer(sqrt(sqrt(PLASMAPTAU181)) ~ time_years + (1|RID), 
                           data = long_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                             data = long_data,
                             family = lognormal(link = "log"))

# Student-t
#fit_student_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                            #data = long_data,
                            #family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_ptau <- simulateResiduals(fit_gamma_ptau)
testResiduals(gamma_residuals_ptau)

lme_residuals_log_ptau <- simulateResiduals(fit_lme_log_ptau)
testResiduals(lme_residuals_log_ptau)

lme_residuals_cubed_ptau <- simulateResiduals(fit_lme_cubed_ptau)
testResiduals(lme_residuals_cubed_ptau)

lognormal_residuals_ptau <- simulateResiduals(fit_lognormal_ptau)
testResiduals(lognormal_residuals_ptau)

#student_residuals_ptau <- simulateResiduals(fit_student_ptau)
#testResiduals(student_residuals_ptau)
```

```{r}
# LME model for p-tau181 with splines 
library(splines)

fit_spline_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ ns(time_years, df = 3),
  random = ~ time_years | RID,
  data = long_data,
  method = "REML"
)

res_ptau <- residuals(fit_spline_ptau, type = "pearson")
fit_ptau <- fitted(fit_spline_ptau)

plot(fit_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

qqnorm(res_ptau, main = "QQ plot: p-tau181 residuals")
qqline(res_ptau, col = "red")
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(gamma_residuals_ptau$fittedResiduals)
shapiro.test(lme_residuals_log_ptau$fittedResiduals)
shapiro.test(lme_residuals_cubed_ptau$fittedResiduals)
shapiro.test(lognormal_residuals_ptau$fittedResiduals)
#shapiro.test(student_residuals_ptau$fittedResiduals)
shapiro.test(res_ptau)
```

#### AIC and log-likelihood

```{r}
# Comparing p-tau181 models with AIC and log likelihood
AIC(fit_lme_log_ptau, fit_lme_cubed_ptau, fit_gamma_ptau, fit_lognormal_ptau, fit_spline_ptau)
logLik(fit_lme_log_ptau)
logLik(fit_lme_cubed_ptau)
logLik(fit_gamma_ptau)
logLik(fit_lognormal_ptau)
#logLik(fit_student_ptau)
logLik(fit_spline_ptau)
```

### NfL

- Negative AIC for cube root transformed -\> take the absolute value

```{r}
# checking the mean and variance for nfl
mean(long_data$PLASMA_NFL)
var(long_data$PLASMA_NFL)
```

#### Fitting the models

```{r}
# GLMM
fit_gamma_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID), 
                   data = long_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_nfl <- lmer(log(PLASMA_NFL) ~ time_years + (1|RID), 
                data = long_data)

# LME approach with log transformed data
fit_lme_cubed_nfl <- lmer(sqrt(sqrt(PLASMA_NFL)) ~ time_years + (1|RID), 
                          data = long_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                             data = long_data,
                             family = lognormal(link = "log"))

# Student-t
#fit_student_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                           #data = long_data,
                           #family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_nfl <- simulateResiduals(fit_gamma_nfl)
testResiduals(gamma_residuals_nfl)

lme_residuals_log_nfl <- simulateResiduals(fit_lme_log_nfl)
testResiduals(lme_residuals_log_nfl)

lme_residuals_cubed_nfl <- simulateResiduals(fit_lme_cubed_nfl)
testResiduals(lme_residuals_cubed_nfl)

lognormal_residuals_nfl <- simulateResiduals(fit_lognormal_nfl)
testResiduals(lognormal_residuals_nfl)

#student_residuals_nfl <- simulateResiduals(fit_student_nfl)
#testResiduals(student_residuals_nfl)
```

```{r}
# LME model for NfL with splines 
library(splines)

fit_spline_nfl <- lme(
  fixed = log_PLASMA_NFL ~ ns(time_years, df = 3),
  random = ~ time_years | RID,
  data = long_data,
  method = "REML"
)

res_nfl <- residuals(fit_spline_nfl, type = "pearson")
fit_nfl <- fitted(fit_spline_nfl)

plot(fit_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

qqnorm(res_nfl, main = "QQ plot: NfL residuals")
qqline(res_nfl, col = "red")
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(gamma_residuals_nfl$fittedResiduals)
shapiro.test(lme_residuals_log_nfl$fittedResiduals)
shapiro.test(lme_residuals_cubed_nfl$fittedResiduals)
shapiro.test(lognormal_residuals_nfl$fittedResiduals)
#shapiro.test(student_residuals_nfl$fittedResiduals)
shapiro.test(res_nfl)
```

#### AIC and log-likelihood

```{r}
# Comparing NfL-models with AIC and log likelihood
AIC(fit_lme_log_nfl, fit_lme_cubed_nfl, fit_gamma_nfl, fit_lognormal_nfl, fit_spline_nfl)
logLik(fit_lme_log_nfl)
logLik(fit_lme_cubed_nfl)
logLik(fit_gamma_nfl)
logLik(fit_lognormal_nfl)
#logLik(fit_student_nfl)
logLik(fit_spline_nfl)

```

Osäker dock om man kan jämföra log-version och cube-version av lme med de andra modellerna (som använder råa versionen) eftersom de då är på olika skalor! Tror det är bäst att kolla på residualerna och QQ-plotsen

# Fitted LME models

## PTAU

#### With all genotype levels

- None of the coefficients are significant for any of the levels for genotype

```{r}
# Longitudinal model for log p-tau181 with normal distribution 
long_model_ptau_full <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + genotype,
  random = ~ time_years | RID,                    
  data = long_data
)
```

```{r}
summary(long_model_ptau_full)
```

#### APOE4 status

- All three levels of APOE4_status are significant

```{r}
# Longitudinal model for log p-tau181 with APOE4 status 
long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status,
  random = ~ time_years | RID,                    
  data = long_data
)
```

```{r}
summary(long_model_ptau)
```

#### Determining random effects

- Fitting the model with ML estimation in order to compare models using likelihood ratio test

- The results give support to fitting the simpler model with only a random intercept. However, $jm()$ does not currently support model fitting for univariate longitudinal data with only a random intercept.

```{r}
m1 <- lmer(log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status + (1 | RID),
           data = long_data, REML = FALSE)

m2 <- lmer(log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status + (time_years | RID),
           data = long_data, REML = FALSE)
```

```{r}
anova(m1, m2, type = "LRT")
```

## NFL

#### With all genotype levels

```{r}
# Longitudinal model for log NfL with normal distribution, all genotype levels
long_model_nfl_full <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER + genotype, 
  random = ~ time_years | RID,
  data = long_data
  )
```

```{r}
summary(long_model_nfl_full)
```

#### APOE4 status

- None of the levels are significant for log NfL and neither is gender

```{r}
# Longitudinal model for log NfL with normal distribution, APOE4_status
long_model_nfl <- lme(
  fixed = log_PLASMA_NFL ~ time_years + PTGENDER + APOE4_status, 
  random = ~ time_years | RID,
  data = long_data
  )
```

```{r}
summary(long_model_nfl)
```

#### Removing APOE4 status

```{r}
# Longitudinal model for log NfL without APOE4_status
long_model_nfl_reduced1 <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER, 
  random = ~ time_years | RID,
  data = long_data
  )
```

```{r}
summary(long_model_nfl_reduced1)
```

#### Removing PTGENDER

```{r}
# Longitudinal model for log NfL without covariates
long_model_nfl_reduced2 <- lme(
  log_PLASMA_NFL ~ time_years, 
  random = ~ time_years | RID,
  data = long_data
  )
```

```{r}
summary(long_model_nfl_reduced2)
```

#### Determining the random effects

- The results of the likelihood-ratio test support including a random slope as well as a random intercept for the NfL model

```{r}
m1 <- lmer(log_PLASMA_NFL ~ time_years + (1 | RID),
           data = long_data, REML = FALSE)

m2 <- lmer(log_PLASMA_NFL ~ time_years + (time_years | RID),
           data = long_data, REML = FALSE)
```

```{r}
anova(m1, m2, type = "LRT")
```

```{r}
# Mixed object for multivariate joint model
mixed <- list(long_model_ptau, long_model_nfl)
```

## Residual check

#### Residuals vs fitted values

```{r}
# residuals of the p-tau181 LME model
res_ptau <- residuals(long_model_ptau, type = "pearson")  

# fitted values of the p-tau181 LME model
fitted_vals_ptau <- fitted(long_model_ptau)  

# plot of residuals vs fitted values 
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for p-tau181")
abline(h = 0, col = "red", lty = 2)
```

```{r}
# residuals of the NfL LME model
res_nfl <- residuals(long_model_nfl, type = "pearson")  

# fitted values of the NfL LME model
fitted_vals_nfl <- fitted(long_model_nfl)  

# plot of residuals vs fitted values 
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NFL")
abline(h = 0, col = "red", lty = 2)
```

#### Shapiro-Wilk test

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(res_ptau)
```

```{r}
# Shapiro-Wilk test of normality of the residuals
shapiro.test(res_nfl)
```

#### QQ plots

```{r}
# For P-tau
qqnorm(res_ptau, main = "QQ plot: p-tau181 residuals")
qqline(res_ptau, col = "red")

# For NFL 
qqnorm(res_nfl, main = "QQ plot: NfL residuals")
qqline(res_nfl, col = "red")
```

# Survival models

## For the different cox PH models and the separate joint models

## Fitted standard Cox PH models (for the joint models)

#### With all genotype levels

```{r}
# Cox survival-models, with covariates
# All genotype levels
surv_model_full <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + genotype,
  data = surv_data,
  x = TRUE
)

summary(surv_model_full)
```

#### APOE4 status

```{r}
# Cox survival-models, with covariates
# With APOE4_status
surv_model <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status,
  data = surv_data,
  x = TRUE
)

summary(surv_model)
```

#### Removing PTGENDER

```{r}
# Cox survival-models, with covariates
# With APOE4_status
surv_model_reduced <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + APOE4_status,
  data = surv_data,
  x = TRUE
)

summary(surv_model_reduced)
```

```{r}
anova(surv_model, surv_model_reduced)
```

## Fitted Cox PH models using baseline values of each biomarker

### PTAU

```{r}
# Cox survival-models, with covariates including baseline values of each biomarker 
# For p-tau181 with PTGENDER
surv_model_ptau_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau,
  data = surv_data,
  x = TRUE
)

summary(surv_model_ptau_bl)
```

#### Removing PTGENDER

```{r}
# Cox survival-models, with covariates including baseline values of each biomarker 
# For p-tau181 without PTGENDER
surv_model_ptau_bl_reduced <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + APOE4_status + baseline_value_ptau,
  data = surv_data,
  x = TRUE
  )

summary(surv_model_ptau_bl_reduced)
```

#### Comparing models

Testing the significance of gender in the Cox proportional hazard model shows that there is no evidence of an improved fit compared with a reduced model with only age at baseline and APOE- $\epsilon 4$ status and baseline ptau-181 value as covariates.

```{r}
anova(surv_model_ptau_bl, surv_model_ptau_bl_reduced, test = "LRT")
```

### NfL

```{r}
# For NfL with PTGENDER
surv_model_nfl_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_nfl,
  data = surv_data,
  x = TRUE
)

summary(surv_model_nfl_bl)
```

#### Removing PTGENDER

```{r}
# For NfL without PTGENDER
surv_model_nfl_bl_reduced1 <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + APOE4_status + baseline_value_nfl,
  data = surv_data,
  x = TRUE
)

summary(surv_model_nfl_bl_reduced1)
```

#### Removing age at baseline

```{r}
# For NfL without PTGENDER and age at baseline
surv_model_nfl_bl_reduced2 <- coxph(
  Surv(time_to_event, event) ~
    APOE4_status + baseline_value_nfl,
  data = surv_data,
  x = TRUE
)

summary(surv_model_nfl_bl_reduced2)
```

#### Comparing models

Testing the significance of gender in the Cox proportional hazard model for NfL again shows that there is no evidence to include gender as a covariate. Additionally, there is no support for including age at baseline in the model with only baseline NfL values.

```{r}
anova(surv_model_nfl_bl, surv_model_nfl_bl_reduced1, surv_model_nfl_bl_reduced2, test = "LRT")
```

### PTAU and NfL

```{r}
# Inlcuding both p-tau181 and NfL
surv_model_both_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data,
  x = TRUE
)

summary(surv_model_both_bl) 
```

#### Removing PTGENDER

```{r}
# Without PTGENDER
surv_model_both_bl_reduced1 <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data,
  x = TRUE
)

summary(surv_model_both_bl_reduced1)
```

#### Removing age at baseline

```{r}
# Without PTGENDER and age at baseline
surv_model_both_bl_reduced2 <- coxph(
  Surv(time_to_event, event) ~
    APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data,
  x = TRUE
)

summary(surv_model_both_bl_reduced2)
```

#### Comparing models

```{r}
anova(surv_model_both_bl, surv_model_both_bl_reduced1, surv_model_both_bl_reduced2,
      test = "LRT")
```

### Test of the assumption of proportional hazards

```{r}
##### For the standard Cox PH model
cox.zph(surv_model)

##### For the Cox PH models with baseline values of each biomarker 
# With p-tau181 baseline values
cox.zph(surv_model_ptau_bl)
# With NfL baseline values
cox.zph(surv_model_nfl_bl)
# With both p-tau181 and NfL baseline values
cox.zph(surv_model_both_bl)
```

To confirm that age at baseline has a time-dependent effect that violates the proportional hazards assumption in the Cox models, a Cox model with a time-varying effect of age at baseline is fitted.

### Test of time-varying coefficient Cox models for age at baseline due to evidence of non-proportional hazards

```{r}
# We fit a time-varying coefficient Cox model where age at baseline is modeled as time-varying
surv_model_tv <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status + tt(age_at_baseline),
  data = surv_data,
  tt = function(x, t, ...) x * log(t)
)

summary(surv_model_tv)
```

```{r}
AIC(surv_model, surv_model_tv)

logLik(surv_model)
logLik(surv_model_tv)
```

# Two separate joint models

## Fitted joint model for p-tau181

- High Rhat for time_years

```{r}
# Joint model for p-tau181 using package JMbayes2 (bayesian approach)
joint_model_ptau <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years",
  n_chains = 3L,
  n_iter = 11000L,
  n_burnin = 1000L
  )
```

```{r}
summary(joint_model_ptau)
```

#### 20000 iterations

- $\widehat R$ for $time\_years$ with $20,000$ iterations is acceptable

```{r}
joint_model_ptau <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years",
  n_chains = 3L,
  n_iter = 20000L,
  n_burnin = 1000L
  )
```

```{r}
summary(joint_model_ptau)
```

```{r}
# Checking convergence 
library(ggplot2)
ggtraceplot(joint_model_ptau, "alphas")
```

## Fitted joint model for NfL

- Rhat are fine with newly specified long_model_nfl

```{r}
# Joint model for NfL using package JMbayes2 (bayesian approach)
joint_model_nfl <- jm(
  surv_model,
  long_model_nfl,
  time_var = "time_years",
  n_chains = 3L,
  n_iter = 11000L,
  n_burnin = 1000L
  )
```

```{r}
summary(joint_model_nfl)
```

```{r}
# Checking convergence 
library(ggplot2)
ggtraceplot(joint_model_nfl, "alphas")
```

### Traceplots

```{r}
#traceplot(joint_model_ptau)
#traceplot(joint_model_nfl)
```

```{r}
# For p-tau181
traceplot(joint_model_ptau, parm = "betas")
traceplot(joint_model_ptau, parm = "alphas")
traceplot(joint_model_ptau, parm = "sigmas")
```

```{r}
# For NfL
traceplot(joint_model_nfl, parm = "betas")
traceplot(joint_model_nfl, parm = "alphas")
traceplot(joint_model_nfl, parm = "sigmas")
```

### Density plots

```{r}
#densplot(joint_model_ptau)
#densplot(joint_model_nfl)
```

```{r}
# For p-tau181
densplot(joint_model_ptau, parm = "betas")
densplot(joint_model_ptau, parm = "alphas")
densplot(joint_model_ptau, parm = "sigmas")
```

```{r}
# For NfL
densplot(joint_model_nfl, parm = "betas")
densplot(joint_model_nfl, parm = "alphas")
densplot(joint_model_nfl, parm = "sigmas")
```

### Sensitivity analysis: frequentist joint models

As a sensitivity analysis, joint models using a maximum likelihood approach are also fitted. Given the known sensitivity of this estimation method to model specification and scaling, biomarker values are standardised, all covariates are excluded, and only random intercepts are included in the longitudinal submodels.

```{r}
simple_surv_model <- coxph(
  Surv(time_to_event, event) ~ 1,
  data = surv_data,
  x = TRUE
)
```

```{r}
long_data$log_PLASMAPTAU181_s <- scale(long_data$log_PLASMAPTAU181)
long_data$log_PLASMA_NFL_s <- scale(long_data$log_PLASMA_NFL)

simple_long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181_s ~ time_years,
  random = ~ 1 | RID,                    
  data = long_data
)

simple_long_model_nfl <- lme(
  fixed = log_PLASMA_NFL_s ~ time_years, 
  random = ~ 1 | RID,
  data = long_data
)
```

```{r}
# Joint model for p-tau181 using package JM (frequentist approach)
joint_model_ptau_freq <- jointModel(
  simple_long_model_ptau,
  simple_surv_model,
  timeVar = "time_years"
)

summary(joint_model_ptau_freq)
```

```{r}
# Joint model for NfL using package JM (frequentist approach)
joint_model_nfl_freq <- jointModel(
  simple_long_model_nfl,
  simple_surv_model,
  timeVar = "time_years"
)

summary(joint_model_nfl_freq)
```

Neither of the joint models fitted using maximum likelihood converged successfully, even after scaling the biomarkers, removing all covariates, and simplifying the longitudinal submodels to include only random intercepts.

# Multivariate joint model

## To check if the models have stable random slopes

```{r}
VarCorr(long_model_ptau)
VarCorr(long_model_nfl)
```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

## fForms

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL)
  )      
```

## Fitted multivariate joint model

- High $\widehat R$ for $time_years$

```{r}
multi_joint_model <- jm(surv_model, mixed ,
                        time_var = "time_years",
                        functional_forms = fForms,
                        n_chains = 3L,
                        n_iter = 20000L,
                        n_burnin = 2000L
                        )

summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival
```

#### With random slope for NfL

- $\widehat R$ for $time\_years$ now fine, but many others now high

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) + slope(log_PLASMA_NFL)
  )  
```

```{r}
multi_joint_model <- jm(surv_model, mixed ,
                        time_var = "time_years",
                        functional_forms = fForms,
                        n_chains = 3L,
                        n_iter = 20000L,
                        n_burnin = 2000L
                        )

summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival
```

# ...

# Latent Class Mixed Models

Choice of p-value <https://pmc.ncbi.nlm.nih.gov/articles/PMC5059270/>

# LCMM: NFL

## M0 (1 latent class)

```{r}
hlme_m0_1 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_1)
```

```{r}
hlme_m0_2 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_2)
```

### Determining the random effects

The $BIC$ is reduced and the log-likelihood increases when adding a random slope to the linear mixed effects model. Additionally, testing whether the improvement in the model with the random slope is statistically significant with a likelihood ratio tests gave $\lambda_{LR} = 16.09$ on $2$ degrees of freedom ($p < 0.05$).

```{r}
summarytable(hlme_m0_1, hlme_m0_2)
```

$$\lambda_{LR} = - 2 \left\{ \ell(\theta_0) - \ell(\widehat\theta) \right\} \quad \text{where, under } H_0 \quad \lambda_{LR} \sim \chi^2$$

```{r}
# log likelihood for the full model
l = hlme_m0_2$loglik

# log likelihood for the reduced model
l_0 = hlme_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
hlme_m1 <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = long_data,
    B = random(hlme_m0_1)
  )

summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## M2 (3 latent classes)

```{r}
hlme_m2 <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 3,
    data = long_data,
    B = random(hlme_m0_1)
  )

summary(hlme_m2)
```

### Posterior probabilities

```{r}
postprob(hlme_m2)
```

## M3 (4 latent classes)

```{r}
hlme_m3 <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 4,
    data = long_data,
    B = random(hlme_m0_1)
  )

summary(hlme_m3)
```

### Posterior probabilities

```{r}
postprob(hlme_m3)
```

## Comparing M0, M1, M2, M3

```{r}
summarytable(hlme_m0_2, hlme_m1, hlme_m2, hlme_m3,
             which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# LCMM: PTAU

Latent class mixed models for PTAU

## M0 (1 latent class)

```{r}
hlme_m0_1 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_1)
```

```{r}
hlme_m0_2 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_2)
```

### Determining the random effects

Comparing the $BIC$ from a model with random intercept versus a model with both random intercept and slope and the results of a log-likelihood test give support to the reduced model with only a random intercept in the longitudinal model. Therefore, only a random intercept will be used when modelling the longitudinal ptau-181 data.

```{r}
summarytable(hlme_m0_1, hlme_m0_2)
```

```{r}
# log likelihood for the full model
l = hlme_m0_2$loglik

# log likelihood for the reduced model
l_0 = hlme_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

A latent mixed effects model with 2 classes was estimated on the longitudinal ptau181 data which produced significant coefficients for the longitudinal model ($p < 0.01$) but not for the class-membership model ($p = 1.00$). The $BIC$ increased when compared to the 1 class model and the posterior probabilities show no clear differentiation between class 1 and class 2, with the average posterior probability being $0.50$ for both classes.

```{r}
hlme_m1 <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = long_data,
    B = random(hlme_m0_1)
  )
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## M2 (3 latent classes)

Similarly, the three class model had significant coefficients for the lonitudinal model ($p < 0.01$), whereas the coefficients for the intercepts of the class membership model were insignificant ($p = 0.98, p = 0.97$). Further, the model assaigned no-one to class three, $97\%$ to class one and $3\%$ to class two, showing poor discriminatory power. Considering the results of the models with two and three classes, it was decided not to run the four class model.

```{r}
hlme_m2 <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 3,
    data = long_data,
    B = random(hlme_m0_1)
  )
```

```{r}
summary(hlme_m2)
```

### Posterior probabilities

```{r}
postprob(hlme_m2)
```

## Comparing M0, M1, M2

```{r}
summarytable(hlme_m0_2, hlme_m1, hlme_m2,
             which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
mean(pp$prob1)
mean(pp$prob2)
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# LCMM: NFL (including classmb covariates)

Latent class mixed models for NFL including covariates in the class model

## M0 (1 latent class)

```{r}
# updating the 1 class model
hlme_m0 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0)
```

## M1 (2 latent classes)

The 2 class model with covariates $APOE4\_status$ and $age\_at\_visit$ is a significantly better fit than the 1 class model according to a likelihood ratio test and comparison of $BIC$ for the two models.

### Including PTGENDER

```{r}
hlme_m1_PTGENDER <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ time_years,
    random = ~ time_years,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit + PTGENDER,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1_PTGENDER)
```

```{r}
postprob(hlme_m1_PTGENDER)
```

### Excluding PTGENDER

```{r}
hlme_m1_nfl_classmb <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ time_years,
    random = ~ time_years,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1_nfl_classmb)
```

```{r}
postprob(hlme_m1_nfl_classmb)
```

## M2 (3 latent classes)

```{r}
hlme_m2_nfl_classmb <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ time_years,
    random = ~ time_years,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit + PTGENDER,
    ng = 3,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m2_nfl_classmb)
```

```{r}
postprob(hlme_m2_nfl_classmb)
```

## M3 (4 latent classes)

```{r}
hlme_m3_nfl_classmb <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ time_years,
    random = ~ time_years,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit + PTGENDER,
    ng = 4,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m3_nfl_classmb)
```

```{r}
postprob(hlme_m3_nfl_classmb)
```

## Comparing M0, M1, M2, M3

```{r}
summarytable(hlme_m0, hlme_m1_nfl_classmb, hlme_m2_nfl_classmb, hlme_m3_nfl_classmb,
             which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1_nfl_classmb$pprob
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# LCMM: PTAU (including classmb covariates)

Latent class mixed models for PTAU including covariates in the class model

## M0 (1 latent class)

```{r}
# updating the 1 class model
hlme_m0 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0)
```

## M1 (2 latent classes)

TO CHECK: difference between genotype in latent class file

```{r}
levels(long_data$APOE4_status)
```

### Including PTGENDER

```{r}
hlme_m1_PTGENDER <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit + PTGENDER,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1_PTGENDER)
```

### Excluding PTGENDER

```{r}
hlme_m1_ptau_classmb <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1_ptau_classmb)
```

```{r}
postprob(hlme_m1_ptau_classmb)
```

## M2 (3 latent classes)

```{r}
hlme_m2_ptau_classmb <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit,
    ng = 3,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m2_ptau_classmb)
```

```{r}
postprob(hlme_m2_ptau_classmb)
```

## M3 (4 latent classes)

```{r}
hlme_m3_ptau_classmb <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit,
    ng = 4,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m3_ptau_classmb)
```

```{r}
postprob(hlme_m3_ptau_classmb)
```

## Comparing M0, M1, M2, M3

```{r}
summarytable(hlme_m0, hlme_m1_ptau_classmb, hlme_m2_ptau_classmb, hlme_m3_ptau_classmb,
             which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 2") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# MLCMM: NFL & PTAU

Multivariate latent class mixed models with both NFL and PTAU

## M0 (1 latent class)

```{r}
mlcmm_m0_1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(mlcmm_m0_1)
```

```{r}
mlcmm_m0_2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(mlcmm_m0_2)
```

### Determining the random effects

```{r}
summarytable(mlcmm_m0_1, mlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = mlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = mlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = long_data,
  B = random(mlcmm_m0_2)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

```{r}
plot(mlcmm_m1, which="fit", var.time="time_years")
```

## M2 (3 latent classes)

```{r}
mlcmm_m2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 3,
  data = long_data,
  B = random(mlcmm_m0_1),
  nproc = 10
)
```

```{r}
summary(mlcmm_m2)
```

```{r}
postprob(mlcmm_m2)
```

## M3 (4 latent classes)

```{r}
mlcmm_m3 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 4,
  data = long_data,
  B = random(mlcmm_m0_1),
  nproc = 10
)
```

```{r}
summary(mlcmm_m3)
```

```{r}
postprob(mlcmm_m3)
```

## Comparing M0, M1

```{r}
summarytable(mlcmm_m0_1, mlcmm_m1, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

## M2 (2 latent classes) (including class covariates)

```{r}
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ APOE4_status + age_at_visit,
  ng = 2,
  data = long_data,
  B = random(mlcmm_m0_1)
)
```

```{r}
summary(mlcmm_m1)
```

```{r}
postprob(mlcmm_m1)
```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- mlcmm_m1$pprob
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 2") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

# JLCMM: NFL

1.  Posterior probabilities show good classification

2.  Score test statistic for conditional independence is significant with 1 class and not significant with 2 classes: evidence for 2 classes

3.  No covariates in the survival model OR the class membership model, evidence for latent classes?

```{r}
# merging the longitudinal and survival data for jlcmm
jlcmm_data <- surv_data %>%
  dplyr::select(RID, event, time_to_event) %>%
  merge(long_data, by = "RID")
```

## M0 (1 latent class)

```         
             G    loglik      BIC   entropy
jlcmm_m0_1   1 -1479.537 2998.415 1.0000000
jlcmm_m1_nfl 2 -1433.037 2931.642 0.6833857
jlcmm_m2_nfl 3 -1432.401 2956.598 0.5160664
```

```{r}
jlcmm_m0_1 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_1)
```

```{r}
jlcmm_m0_2 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1 + time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_2)
```

### Determining the random effects

```{r}
summarytable(jlcmm_m0_1, jlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = jlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = jlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
jlcmm_m1_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m1_nfl)
```

```{r}
postprob(jlcmm_m1_nfl)
```

## M2 (3 latent classes)

```{r}
jlcmm_m2_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m2_nfl)
```

```{r}
postprob(jlcmm_m2_nfl)
```

## M3 (4 latent classes)

```{r}
jlcmm_m3_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 4,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_1)
)
```

```{r}
summary(jlcmm_m3_nfl)
```

```{r}
postprob(jlcmm_m3_nfl)
```

## Comparing M0, M1, M2, M3

```{r}
summarytable(jlcmm_m0_1, jlcmm_m1_nfl, jlcmm_m2_nfl,
                                                which = c("G", "loglik", "BIC", "entropy"))
```

```{r}

# saving the summary metrics as a data frame
jlcmm_nfl_summary <- as.data.frame(summarytable(jlcmm_m0_1, jlcmm_m1_nfl, jlcmm_m2_nfl,
                                                which = c("G", "loglik", "BIC", "entropy")))

# renaming the models for the table output
row.names(jlcmm_nfl_summary) <- c("1 class model", "2 class model", "3 class model")

```

```{r}
# # saving which class each individual was predicted to be in
class_preds <- jlcmm_m2_nfl$pprobY

# merging with genotype data
class_preds <- class_preds %>%
  merge(apoe %>% dplyr::select(RID, APOE4_status), by = "RID")

# checking the distribution of genotype by class
table(class_preds$class, class_preds$APOE4_status)
```

### Adjusted LRT

```{r}

```

## Visualisation

1.  Need to check probYT1 and probYT2

```{r}
# saving the posterior probabilities into a data frame
# pp <- jlcmm_m1$pprob
# pp1 <- pp %>% filter(class == 1)
# pp2 <- pp %>% filter(class == 2)
```

### Distribution of posterior probabilities (class 1)

1.  Check that YT1 is for the joint model, not the longitudinal model only

```{r}
# plotting posterior probabilities for class 1
# ggplot(pp1, aes(x = probYT1)) + 
#   geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
#   labs(title = "Posterior Probabilities for Class 1") +
#   theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
# ggplot(pp2, aes(x = probYT1)) + 
#   geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
#   labs(title = "Posterior Probabilities for Class 2") +
#   theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
# ggplot(pp, aes(prob1, prob2)) +
#   geom_point(alpha = .8) +
#   geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
#   labs(x = "P(Class 1)", y = "P(Class 2)") +
#   theme_classic()
```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m1_nfl, which = "survival")
```

### Survival curves by class

```{r}
# longitudinal trajectory by class
plot(jlcmm_m1_nfl, var.time = "time_years", which = "fit")
```

# JLCMM: PTAU

## M0 (1 latent class)

The one class joint latent class model was fit with a random intercept. The one class model has statistically significant coefficients ($p < 0.01$

```{r}
jlcmm_m0 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0)
```

## M1 (2 latent classes)

```{r}
jlcmm_m1_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0)
)
```

```{r}
summary(jlcmm_m1_ptau)
```

```{r}
postprob(jlcmm_m1_ptau)
```

## M2 (3 latent classes)

*Things to assess:*

1.  *Fit 4 class model -\> does BIC improve any more or plateau?*

2.  *Score test now not significant compared with the two class model*

3.  *VERY small class 1*

4.  *CLass 2 and 3 are not cleanly separated (look at posterior probabilities)*

5.  Weibull class 3 parameters not significant at 0.05 level

6.  *Plot survival curves (how?)*

### gridsearch()

```{r}
# gridsearch(m = jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years_ptau,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 2,
#   subject = "RID",
#   survival = Surv(time_to_event_ptau, event_ptau) ~ 1,
#   data = jlcmm_data_ptau,
#   B = random(jlcmm_m0)),
#   minit = jlcmm_m1,
#   rep = 50,
#   maxiter = 30)

```

```{r}
jlcmm_m2_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0)
  )
```

```{r}
summary(jlcmm_m2_ptau)
```

```{r}
postprob(jlcmm_m2_ptau)
```

## M3 (4 latent classes)

```{r}
jlcmm_m3_ptau <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 4,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0)
  )
```

```{r}
summary(jlcmm_m3_ptau)
```

```{r}
postprob(jlcmm_m3_ptau)
```

## Comparing M0, M1, M2, M3

```{r}
summarytable(jlcmm_m0, jlcmm_m1_ptau, jlcmm_m2_ptau, jlcmm_m3_ptau,
             which = c("G", "loglik", "BIC", "entropy"))
```

```{r}
# # saving which class each individual was predicted to be in
class_preds <- jlcmm_m2_ptau$pprobY

# merging with genotype data
class_preds <- class_preds %>%
  merge(apoe %>% dplyr::select(RID, APOE4_status), by = "RID")

# checking the distribution of genotype by class
table(class_preds$class, class_preds$APOE4_status)
```

### Adjusted LRT

```{r}

```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}

```

### Distribution of posterior probabilities (class 2)

```{r}

```

### Class certainty scatterplot

```{r}

```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m1_ptau, which = "survival")
```

### Longitudinal trajectories by class

```{r}
# longitudnal trajectories by class
plot(jlcmm_m1_ptau, var.time = "time_years", which = "fit")
```

## Visualisation

### Survival curves by class

```{r}
# survival curves by class
#plot(jlcmm_m2_ptau, which = "survival")
```

### Longitudinal trajectories by class

```{r}
# longitudnal trajectories by class
#plot(jlcmm_m2_ptau, var.time = "time_years", which = "fit")
```

# MJLCMM: NFL & PTAU

## 1: Random intercept and slope

TWO CLASS MULTIVARIATE LATENT CLASS MIXED MODEL -\> significant

```{r}
# # merging the longitudinal and survival data for NfL
# jlcmm_data_multi <- surv_test_1 %>%
#   dplyr::select(RID, event, time_to_event) %>%
#   merge(long_data, by = "RID") %>%
#   filter(time_to_event != 0)
```

```{r}
# fitting multivariate latent class model with 1, 2, 3 and 4 classes with random intercept and random slope

# 1 class model
mlcmm_m0 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = jlcmm_data
)

# 2 class model
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)

# 3 class model
mlcmm_m2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)

# 4 class model
mlcmm_m3 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 4,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)

```

## M0 (1 latent class)

```{r}
mult_jlcmm_m0 <- mpjlcmm(
  longitudinal = list(mlcmm_m0),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data
)
```

```{r}
summary(mult_jlcmm_m0)
```

## M1 (2 latent classes)

```{r}
mult_jlcmm_m1 <- mpjlcmm(
  longitudinal = list(mlcmm_m1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m1)
```

```{r}
postprob(mult_jlcmm_m1)
```

## M2 (3 latent classes)

```{r}
mult_jlcmm_m2 <- mpjlcmm(
  longitudinal = list(mlcmm_m2),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m2)
```

```{r}
postprob(mult_jlcmm_m2)
```

## M3 (4 latent classes)

```{r}
mult_jlcmm_m3 <- mpjlcmm(
  longitudinal = list(mlcmm_m3),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 4,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m3)
```

```{r}
postprob(mult_jlcmm_m3)
```

## Comparing M0, M1, M2, M3

```{r}
summarytable(mult_jlcmm_m0, mult_jlcmm_m1, mult_jlcmm_m2, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

## 1: Random intercept

```{r}
# fitting multivariate latent class model with 1, 2, 3 and 4 classes with random intercept and random slope

# 1 class model
mlcmm_m0 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = jlcmm_data
)

# 2 class model
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)

# 3 class model
mlcmm_m2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)

# 4 class model
mlcmm_m3 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 4,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)

```

## M0 (1 latent class)

```{r}
mult_jlcmm_m0 <- mpjlcmm(
  longitudinal = list(mlcmm_m0),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data
)
```

```{r}
summary(mult_jlcmm_m0)
```

## M1 (2 latent classes)

```{r}
mult_jlcmm_m1 <- mpjlcmm(
  longitudinal = list(mlcmm_m1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m1)
```

```{r}
postprob(mult_jlcmm_m1)
```

## M2 (3 latent classes)

```{r}
mult_jlcmm_m2 <- mpjlcmm(
  longitudinal = list(mlcmm_m2),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m2)
```

```{r}
postprob(mult_jlcmm_m2)
```

## M3 (4 latent classes)

```{r}
mult_jlcmm_m3 <- mpjlcmm(
  longitudinal = list(mlcmm_m3),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 4,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m3)
```

```{r}
postprob(mult_jlcmm_m3)
```

## Comparing M0, M1, M2, M3

```{r}
summarytable(mult_jlcmm_m0, mult_jlcmm_m1, which = c("G", "loglik", "BIC", "entropy"))

```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}

```

### Distribution of posterior probabilities (class 2)

```{r}

```

### Class certainty scatterplot

```{r}

```

### Survival curves by class

```{r}
#plot(mult_jlcmm_m1, which = "survival")
```

```{r}
# plot(mult_jlcmm_m1, var.time = "time_years", which = "fit")
```

## Visualisation

### Survival curves by class

```{r}
# survival curves by class
#plot(mult_jlcmm_m2, which = "survival")
```

```{r}
# # longitudnal trajectories by class
# plot(mult_jlcmm_m2, outcome = 1, which = "fit")
```

## TO DO

Check sensitivity to starting values by using gridsearch()

External validation: what do these classes look like? Do they just represent different genotypes? How do we check this and are they truly valid LATENT classes?

Adjusted LRT (Lo-Mendell-Rubin test)

Bootstrap LRT???

# Comparing models using time-dependent AUC

*For the Cox PH models = using timeROC, from the package timeROC*

*For the joint models = using tvAUC, from the package JMbayes2*

## Defining evaluation time points for time-dependent AUC comparison

```{r}
# We define evaluation time points for time-dependent AUC comparison
times <- c(3, 4, 5, 6, 7, 8)
```

## Time-dependent AUC for fitted Cox PH models

```{r}
# We define a function to compute time-dependent AUC for Cox models using timeROC
get_cox_auc <- function(model, times) {
  
  T <- surv_data$time_to_event
  delta <- surv_data$event
  
  lp <- predict(model, newdata = surv_data, type = "lp")
  
  timeROC::timeROC(
    T = T,
    delta = delta,
    marker = lp,
    cause = 1,
    times = times,
    iid = TRUE
  )$AUC
}
```

```{r}
# We compute time-dependent AUC for Cox model with p-tau181
auc_ptau_bl <- get_cox_auc(surv_model_ptau_bl, times)
 
# # We compute time-dependent AUC for Cox model with NfL
auc_nfl_bl <- get_cox_auc(surv_model_nfl_bl, times)

# We compute time-dependent AUC for Cox model with both p-tau181 and NfL 
auc_both_bl <- get_cox_auc(surv_model_both_bl, times)
```

## Time-dependent AUC for fitted joint models

*From: <https://cloud.r-project.org/web/packages/JMbayes2/JMbayes2.pdf>*

*The function tvAUC requires the following:*

- *"newdata = a data.frame that contains the longitudinal and covariate information for the subjects for which prediction of survival probabilities is required. The names of the variables in this data.frame must be the same as in the data.frames that were used to fit the linear mixed effects and the event process model that were supplied as the two first argument of jm."*

```{r}
# We construct prediction datasets for the joint models by merging the  longitudinal and survival data
newdata <- dplyr::left_join(
  long_data,
  dplyr::select(surv_data, RID, age_at_baseline, time_to_event, event),
  by = "RID"
)
```

```{r}
library(JMbayes2)
# We define a function to compute time-dependent AUC for joint models using tvAUC
get_jm_auc <- function(model, times) {
  
  aucs <- sapply(times, function(t) {
    
    tmp <- tvAUC(
      model,
      newdata = newdata,
      Tstart = 2,
      Thoriz = t
    )
    
    tmp$auc
  })
  
  return(as.numeric(aucs))
}
```

```{r}
# We compute time-dependent AUC for the joint model including p-tau181 
auc_ptau_jm <- get_jm_auc(joint_model_ptau, times)

# We compute time-dependent AUC for the joint model including NfL 
auc_nfl_jm <- get_jm_auc(joint_model_nfl, times)

# We compute time-dependent AUC for the multivariate joint model 
auc_multi_jm <- get_jm_auc(multi_joint_model, times)

```

## Time-dependent AUC for latent class joint models

Landmark: the time at which predictions are made

Horizon: the predict event within $h$ years

Threshold: e.g if risk \> 0.5 -\> conversion

ROC evaluates the model (sensitivity and specificity) at all possible thresholds, plotting sensitivity (y axis) vs 1 - specificity (x axis)

We can fix the landmark, or the horizon

Fixing the landmark, varying the horizon: how well does the model predict over different horizons? Does prediction accuracy change with horizon, e.g model predicts well over short horizons but not long.

Fiing the horizon, varying the landmark: how well does the model predict conversion within $h$ years as more longitudinal data is gathered?

#### Function for extracting time-dependent AUC

1.  Confidence intervals

2.  Overall model discrimination

3.  Class specific discrimination

```{r}
# Defining a function to compute time dependent AUC for joint latent class models
jlcmm_timeAUC <- function(model, newdata, ID, time_var, surv_time_var, event_var,
                          landmark, times, cause = 1) {
  
#   Initiating a data frame to save the results in at each horizon time
  output <- data.frame(
    times = times,
    cases = NA,
    AUC = NA
  )

#   Calculating the dynamic survival predictions from joint models using dynpred() in the lcmm package
  dynpreds <- dynpred(
    model = model,
    newdata = newdata,
    var.time = time_var,
    horizon = times,
    landmark = landmark
  )
  
#   Extracting the predicted survival probabilities 
  preds <- dynpreds$pred["pred"]
#   Extracting the unique identifiers (RID) relating to the predicitons
  ids <- dynpreds$pred[ID]
#   Extracting the horizon relating to the predicitons
  horizons <- dynpreds$pred["horizon"]
  
#   Creatng a data frame of predictions, horizons and IDs
  dfpred <- data.frame(id = ids, horizon = horizons, pred = preds)
  
#   Fetching the true events for each ID for each time point
  dftrue <- unique(newdata[c(ID, surv_time_var, event_var)])
  names(dftrue) <- c(ID, "time", "event")
  
#   Merging the true and predicted data frames
  merged <- merge(dfpred, dftrue, by = ID)
  
#   Looping through the horizon times
  for(i in 1:length(times)) {
    h = times[i]
    
#     Filtering for the predictions from horizon h
    data_h <- merged %>% filter(horizon == h)
    
    #     Identifiying individuals that convert (cases) in the time between landmark and landmark + horizon (l + h) or remain disease free past l + h (controls)
    cases <- data_h$event == 1 & data_h$time > landmark & data_h$time <= h
    controls <- data_h$time > h
    data_h <- data_h[cases | controls, ]
    
    roc <- timeROC(
      T = data_h$time,
      delta = data_h$event,
      marker = data_h$pred,
      cause = cause,
      times = h,
      iid = TRUE)
    
    output$AUC[i] <- roc$AUC[2]
#     Checking the events in the time window between l and l + h
    output$cases[i] <- sum(cases)
    
  }
  return(output)
}
```

```{r}
auc_jlcmm_m2_nfl <- jlcmm_timeAUC(jlcmm_m2_nfl, jlcmm_data,
                                  "RID", "time_years", "time_to_event", "event",
                                  2, c(3, 4, 5, 6, 7, 8))
```

```{r}
auc_jlcmm_m2_ptau <- jlcmm_timeAUC(jlcmm_m2_ptau, jlcmm_data,
                                   "RID", "time_years", "time_to_event", "event",
                                   2, c(3, 4, 5, 6, 7, 8))
```

```{r}
#  model must be jlcmm for dynpred()
# AUC_mult_jlcmm_m1 <- jlcmm_timeAUC(mult_jlcmm_m1, jlcmm_data, "RID", "time_to_event", "event", 0, times)
```

## Table of the results

```{r}
# Time-dependent AUC for Cox and joint models summarized in a table
results_table_AUC <- data.frame(
  Model = c("Cox ptau-181", "Cox NfL", "Cox both",
            "JM ptau-181", "JM NfL", "JM both", "JLCM ptau-181", "JLCM NfL"),
  
  AUC_3y = c(auc_ptau_bl[1],
             auc_nfl_bl[1],
             auc_both_bl[1],
             auc_ptau_jm[1],
             auc_nfl_jm[1],
             auc_multi_jm[1],
             auc_jlcmm_m2_ptau[1,3],
             auc_jlcmm_m2_nfl[1,3]),
  
  AUC_4y = c(auc_ptau_bl[2],
             auc_nfl_bl[2],
             auc_both_bl[2],
             auc_ptau_jm[2],
             auc_nfl_jm[2],
             auc_multi_jm[2],
             auc_jlcmm_m2_ptau[2,3],
             auc_jlcmm_m2_nfl[2,3]),
  
  AUC_5y = c(auc_ptau_bl[3],
             auc_nfl_bl[3],
             auc_both_bl[3],
             auc_ptau_jm[3],
             auc_nfl_jm[3],
             auc_multi_jm[3],
             auc_jlcmm_m2_ptau[3,3],
             auc_jlcmm_m2_nfl[3,3]),
  
  AUC_6y = c(auc_ptau_bl[4],
             auc_nfl_bl[4],
             auc_both_bl[4],
             auc_ptau_jm[4],
             auc_nfl_jm[4],
             auc_multi_jm[4],
             auc_jlcmm_m2_ptau[4,3],
             auc_jlcmm_m2_nfl[4,3]),
  
  AUC_7y = c(auc_ptau_bl[5],
             auc_nfl_bl[5],
             auc_both_bl[5],
             auc_ptau_jm[5],
             auc_nfl_jm[5],
             auc_multi_jm[5],
             auc_jlcmm_m2_ptau[5,3],
             auc_jlcmm_m2_nfl[5,3])
)

results_table_AUC[, -1] <- round(results_table_AUC[, -1], 3)
results_table_AUC
```

## Plot of the results

```{r}
# We reshape all AUC results into long format for plotting
plot_data <- data.frame(
  Time = rep(times, 8),
  Model = rep(c("Cox ptau-181", "Cox NfL", "Cox both",
                "JM ptau-181", "JM NfL", "JM multi", "JLCM ptau-181", "JLCM NfL"),
              each = length(times)),
  AUC = c(
    auc_ptau_bl,
    auc_nfl_bl,
    auc_both_bl,
    auc_ptau_jm,
    auc_nfl_jm,
    auc_multi_jm,
    auc_jlcmm_m2_ptau$AUC,
    auc_jlcmm_m2_nfl$AUC
  )
) 

# Plot of all models
ggplot(plot_data, aes(x = Time, y = AUC, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  scale_shape_manual(values = c(0,1,2,3,4,5,6,7,8)) +
  ylim(0.5, 1) +
  labs(
    title = "Time-dependent AUC comparison",
    x = "Time (years)",
    y = "AUC"
  )

# Plot of Cox models
ggplot(plot_data[1:18, ], aes(x = Time, y = AUC, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0.5, 1) +
  labs(
    title = "Time-dependent AUC comparison",
    x = "Time (years)",
    y = "AUC"
  )

# Plot of joint models
ggplot(plot_data[18:48, ], aes(x = Time, y = AUC, color = Model)) +
  geom_line(linewidth = 0.5) +
  geom_point(aes(shape = Model), size = 2) +
  ylim(0.5, 1) +
  labs(
    title = "Time-dependent AUC comparison",
    x = "Time (years)",
    y = "AUC"
  )
```

# Comparing models using time-depending Brier score

## Time-dependent Brier score for fitted Cox PH models

```{r}
# We load required library
 library(pec)
 
# We specify Cox models for comparison
cox_models <- list(
   "Cox p-tau" = surv_model_ptau_bl,
   "Cox NfL" = surv_model_nfl_bl,
   "Cox both" = surv_model_both_bl
 )
 
# We compute Brier score using pec for Cox models
brier_cox <- pec(
   object = cox_models,
   formula = Surv(time_to_event, event) ~ 
     age_at_baseline + PTGENDER + APOE4_status,
   data = surv_data,
   times = times,
   cens.model = "cox"
 )
 
 cox_brier <- lapply(brier_cox$AppErr, function(x)
   approx(brier_cox$time, x, xout = times)$y
 )
cox_brier
```

## Time-dependent Brier score for fitted joint models

```{r}
# We define a function to compute time-dependent Brier score for joint models using tvBrier
get_jm_brier <- function(model, times) {
  
  briers <- sapply(times, function(t) {
    
    tmp <- tvBrier(
      model,
      newdata = newdata,
      Tstart = 0,
      Thoriz = t
    )
    
    return(tmp$Brier)
  })
  
  return(as.numeric(briers))
}
```

```{r}
# We compute time-dependent Brier score for the joint model including p-tau181 
brier_ptau_jm <- get_jm_brier(joint_model_ptau, times)
brier_ptau_jm

# We compute time-dependent Brier score for the joint model including NfL 
brier_nfl_jm <- get_jm_brier(joint_model_nfl, times)
brier_nfl_jm

# We compute time-dependent Brier score for the multivariate joint model 
brier_multi_jm <- get_jm_brier(multi_joint_model, times)
brier_multi_jm
```

## Time-dependent Brier score for latent class joint models

...

## Table of the results

```{r}
# Time-dependent Brier score for Cox and joint models summarized in a table
results_table_Brier <- data.frame(
  Time = times,

  Cox_p_tau = cox_brier$`Cox p-tau`,
  Cox_NfL = cox_brier$`Cox NfL`,
  Cox_both = cox_brier$`Cox both`,

  JM_p_tau = brier_ptau_jm,
  JM_NfL = brier_nfl_jm,
  JM_both = brier_multi_jm
)
round(results_table_Brier, 3)
```

## Plot of the results

```{r}
# We reshape all AUC results into long format for plotting
all_table <- data.frame(
  Time = times,
  Cox_p_tau = cox_brier$`Cox p-tau`,
  Cox_NfL = cox_brier$`Cox NfL`,
  Cox_both = cox_brier$`Cox both`,
  JM_p_tau = brier_ptau_jm,
  JM_NfL = brier_nfl_jm
)

plot_data <- all_table %>%
  pivot_longer(
    cols = -Time,
    names_to = "Model",
    values_to = "Brier"
  )

# plot
ggplot(plot_data, aes(x = Time, y = Brier, color = Model)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(
    title = "Time-dependent Brier Score: Cox vs Joint Models",
    x = "Time",
    y = "Brier score"
  )
```
