---
title: "Checking"
format: pdf
editor: visual
---

```{r}
library(dplyr)
library(ggplot2)
library(ADNIMERGE2)

library(lme4)
library(glmmTMB)
library(DHARMa) # for running residual diagnostics on GLMM
```

## Loading data

```{r}
# loading data
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

APOE = APOERES

# selecting relevant variables
nfl_data = nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

ptau_data = ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# 1. Lägg _dx på diagnos-kolumner
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# 2. Slå ihop NFL, ptau och diagnos
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))
```

## Merging nfl and ptau181 data

```{r}
# merging ptai and nfl data
nfl_ptau = nfl_data %>% merge(ptau_data, by = c("RID", "VISCODE2"))

# checking how many EXAMDATE match out of all observations
sum(as.integer(nfl_ptau$EXAMDATE.x %in% nfl_ptau$EXAMDATE.y)) / 3855

# checking for missing values: none
table(is.na(nfl_ptau))
```

## Plotting distributions

Seems like the raw variables are Gamma distributed:

-   Only positive values

-   Right-skewed

-   Gamma accomodates increasing variance with increased mean. This is present in the data, as is shown by the computations below.

-   <https://observablehq.com/>@mattiasvillani/gamma-distribution?collection=@mattiasvillani/distributions

So we have a choice: model using a LME on the log-transformed data, or model using a GLMM and assuming Gamma distribution for the response variables (both nfl and ptau181)

```{r}
# checking the mean and variance for ptau181
mean(nfl_ptau$PLASMAPTAU181)
var(nfl_ptau$PLASMAPTAU181)
```

```{r}
# checking the mean and variance for nfl
mean(nfl_ptau$PLASMA_NFL)
var(nfl_ptau$PLASMA_NFL)
```

## nfl, raw and transformed

```{r}
# original scale
ggplot(nfl_ptau, aes(PLASMA_NFL)) +
  geom_histogram(binwidth = 5, colour = "white") +
  theme_classic()

# log-transformed
ggplot(nfl_ptau, aes(log(PLASMA_NFL))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()

# cube-root transformed
ggplot(nfl_ptau, aes(sqrt(sqrt(PLASMA_NFL)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()
```

## ptau181, raw and transformed

```{r}
# orignal scale
ggplot(nfl_ptau, aes(PLASMAPTAU181)) +
  geom_histogram(binwidth = 5, colour = "white") +
  theme_classic()

# log transformed
ggplot(nfl_ptau, aes(log(PLASMAPTAU181))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()

# cube-root transformed
ggplot(nfl_ptau, aes(sqrt(sqrt(PLASMAPTAU181)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()

```

## We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# 1. Håll bara demografi-variabler vi behöver och en rad per person
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# 2. Slå ihop med merged_data
merged_data <- merged_data %>%
  left_join(demographics_small, by = "RID")

# 3. Säkerställ att kön är faktor
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# 4. Konvertera EXAMDATE_ptau till Date
merged_data$EXAMDATE_ptau <- as.Date(as.character(merged_data$EXAMDATE_ptau), format="%Y-%m-%d")

# 5. Skapa age_at_visit baserat på ptau datum
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_ptau, "%Y")) - merged_data$PTDOBYY
```

## We filter participants with at least 4 measurements

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 3))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

## Baseline date, survival data

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event) + lägger till age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_ptau)],
    
    time_to_event = case_when(

    any(DIAGNOSIS_dx == "Dementia") ~
      as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,

  event = as.integer(any(DIAGNOSIS_dx == "Dementia"))
) %>% ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)

# adding PTGENDER to the survival data
ptgender = merged_data %>% select(RID, PTGENDER)
surv_data = surv_data %>% merge(ptgender, by = "RID") 
surv_data = surv_data[!duplicated(surv_data$RID), ]
```

## We create a time variable (time from baseline)

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## Removing missing values

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))

# getting rid of missing values in both merged and surv
merged_data = na.omit(merged_data)
surv_data = na.omit(surv_data)
```

```{r}
summary(merged_data$PLASMAPTAU181)
any(is.infinite(merged_data$PLASMAPTAU181)) # should be FALSE
# Inf = extreme values 

summary(merged_data$PLASMA_NFL)
any(is.infinite(merged_data$PLASMA_NFL)) # should be FALSE
# Inf = extreme values 
```

```{r}
length(unique(merged_data$RID)) # longitudinala data
length(unique(surv_data$RID)) # survival data
```

## Comparing GLMM and LME

### pTAU-181

```{r}
# GLMM
fit_gamma_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID), 
                   data = merged_data,
                   family = Gamma(link = "log"))

# LME approach
fit_lme_ptau <- lmer(log(PLASMAPTAU181) ~ time_years + (1|RID), 
                data = merged_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                         data = merged_data,
                         family = lognormal(link = "log"))

# residual tests for each model
gamma_residuals_ptau <- simulateResiduals(fit_gamma_ptau)
testResiduals(gamma_residuals_ptau)

lme_residuals_ptau <- simulateResiduals(fit_lme_ptau)
testResiduals(lme_residuals_ptau)

# lognormal_residuals_ptau <- simulateResiduals(fit_lognormal_ptau)
# testResiduals(lognormal_residuals_ptau)
```

```{r}
# comparing the fits between models
AIC(fit_lme_ptau, fit_gamma_ptau, fit_lognormal_ptau)
```

```{r}
plotResiduals(gamma_residuals_ptau, form = merged_data$time_years)
```

### NfL

```{r}
# GLMM
fit_gamma_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID), 
                   data = merged_data,
                   family = Gamma(link = "log"))

# LME approach
fit_lme_nfl <- lmer(log(PLASMA_NFL) ~ time_years + (1|RID), 
                data = merged_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                         data = merged_data,
                         family = lognormal(link = "log"))


# residual tests for each model
gamma_residuals_nfl <- simulateResiduals(fit_gamma_nfl)
testResiduals(gamma_residuals_nfl)

lme_residuals_nfl <- simulateResiduals(fit_lme_nfl)
testResiduals(lme_residuals_nfl)

# lognormal_residuals_nfl <- simulateResiduals(fit_lognormal_nfl)
# testResiduals(lognormal_residuals_nfl)
```

```{r}
# comparing the fits between models
AIC(fit_lme_nfl, fit_gamma_nfl, fit_lognormal_nfl)
```

```{r}
# troubleshooting fit_gamma
# residuals over time
plotResiduals(gamma_residuals_nfl, form = merged_data$time_years)
```
