---
title: "Test"
format: pdf
editor: visual
---

# Pre-processing först

(Baseline här = första biomarkörs-datumet (oavsett ptau eller nfl), INTE första diagnosstatus-datumet, som vi har i officiella dokumentet)

```{r}
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) # for running residual diagnostics on GLMM
```

```{r}
# Loading data
#Milou
# NfL data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
# P-tau181 data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
# Diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
# Demographics data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
# Apoe data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
# ptau_data %>%
#    filter(RID %in% ptau_double$RID) %>% 
#    filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

```{r}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# We merge NFL, ptau and diagnosis data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))

# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID"))
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We define APOE4 status based on number of E4 alleles
apoe <- apoe %>%
  mutate(APOE4_status = case_when(
    GENOTYPE == "4/4" ~ "2 alleles",
    GENOTYPE %in% c("2/4", "3/4") ~ "1 allele",
    TRUE ~ "0 alleles"
  ))

# We merge gender with merged_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We merge apoe4 status with merged_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status")], by="RID")
surv_data <- merge(surv_data, apoe[, c("RID", "APOE4_status")], by="RID")

# We make sure that gender and apoe4 status are factors
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)
long_data$APOE4_status <- factor(long_data$APOE4_status,
                                 levels = c("0 alleles", "1 allele", "2 alleles"))
surv_data$APOE4_status <- factor(surv_data$APOE4_status,
                                 levels = c("0 alleles", "1 allele", "2 alleles"))

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

```{r}
# We filter individuals with at least 3 measurements of ptau_data
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
long_data <- long_data %>% filter(RID %in% keep)
surv_data <- surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

```{r}
# checking which observations have the same examdate for nfl and ptau
common = which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# checking the observations with different exam dates
long_data[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
merged_data <- long_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
  ungroup()

# checking which observations have the same examdate for nfl and ptau
common = which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# checking the observations with different exam dates (9 obs)
long_data[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
long_data <- long_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
ungroup()
```

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)

surv_data$log_PLASMAPTAU181 <- log(surv_data$PLASMAPTAU181)
surv_data$log_PLASMA_NFL <- log(surv_data$PLASMA_NFL)
```

```{r}
core <- surv_data %>%
  group_by(RID) %>%
  summarise(
    
    # Baseline = first observed contact across all sources
    baseline = min(c(EXAMDATE_ptau, EXAMDATE_nfl), na.rm = TRUE),
    
    # Age at baseline 
    age_at_baseline = age_at_visit[
      EXAMDATE_ptau == baseline |
      EXAMDATE_nfl  == baseline 
    ][1],
    
    # First dementia diagnosis after baseline
    event_date = {
      dx_dates <- EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]
      dx_dates <- dx_dates[!is.na(dx_dates)]
      dx_dates <- dx_dates[dx_dates > baseline]
      
      if (length(dx_dates) == 0) as.Date(NA) else min(dx_dates)
    },
    
    # Last follow-up across all sources
    last_followup = max(c(EXAMDATE_ptau, EXAMDATE_nfl, EXAMDATE_dx), na.rm = TRUE),
    
    .groups = "drop"
  ) %>%
  
  mutate(
    
    # Event indicator
    event = as.integer(!is.na(event_date)),
    
    # Time-to-event
    time_to_event = if_else(
      event == 1,
      as.numeric(event_date - baseline) / 365.25,
      as.numeric(last_followup - baseline) / 365.25
    )
  )

# We keep one row per participant and attach all original variables
surv_data <- surv_data %>%
  distinct(RID, .keep_all = TRUE) %>%
  left_join(core, by = "RID")

# We tabulate the ptau event indicator (1 = dementia, 0 = censored)
table(surv_data$event)
```

```{r}
long_data <- long_data %>%
  left_join(core %>% select(RID, baseline), by = "RID") %>%
  
  mutate(
    time_years_ptau = as.numeric(EXAMDATE_ptau - baseline) / 365.25,
    time_years_nfl  = as.numeric(EXAMDATE_nfl  - baseline) / 365.25
  )
```

# Models

```{r}
# P-tau181 baseline 
long_data_ptau <- long_data[order(long_data$RID, long_data$EXAMDATE_ptau), ]

baseline_ptau <- long_data[!duplicated(long_data$RID),
                                c("RID",
                                  "log_PLASMAPTAU181")]

names(baseline_ptau)[2] <- "baseline_value_ptau"

# NfL baseline
long_data <- long_data[order(long_data$RID, long_data$EXAMDATE_nfl), ]

baseline_nfl <- long_data[!duplicated(long_data$RID),
                              c("RID",
                                "log_PLASMA_NFL")]

names(baseline_nfl)[2] <- "baseline_value_nfl"

# We merge the two variables into a baseline table 
baseline <- merge(baseline_ptau, baseline_nfl, by="RID")

# We merge the baseline table with the p-tau181 survival dataset 
surv_data <- merge(surv_data,
                        baseline,
                        by="RID"
                        )
```

```{r}
# Cox survival-models, with covariates
surv_model <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status,
  data = surv_data,
  x = TRUE
)

```

```{r}
# For p-tau181
surv_model_ptau_bl <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau,
  data = surv_data
)

summary(surv_model_ptau_bl)

# For NfL
surv_model_nfl_bl <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status + baseline_value_nfl,
  data = surv_data
)

summary(surv_model_nfl_bl)

# For p-tau181 including NfL 
surv_model_both_bl <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data
)

summary(surv_model_both_bl)

```

```{r}
# Longitudinal model for p-tau181 with normal distribution 
long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years_ptau + PTGENDER + APOE4_status,
  random = ~ time_years_ptau | RID,                    
  data = long_data
)

#summary(long_model_ptau)

# Longitudinal model for NfL with normal distribution 
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years_nfl + PTGENDER + APOE4_status, 
  random = ~ time_years_nfl | RID, # random intercept + slope
  data = long_data
  )

#summary(long_model_nfl)
```

```{r}
library(JMbayes2)
# Joint model for p-tau181 using package JMbayes2 (bayesian approach)
joint_model_ptau <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years_ptau",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )
 
summary(joint_model_ptau)
```

```{r}
# Joint model for NfL using package JMbayes2 (bayesian approach)
joint_model_nfl <- jm(
  surv_model,
  long_model_nfl,
  time_var = "time_years_nfl",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

# summary(joint_model_nfl)
```

```{r}
mixed <- list(long_model_ptau, long_model_nfl)

fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )    
```

```{r}
multi_joint_model <- jm(surv_model, mixed ,
                         time_var = "time_years_ptau",
                         functional_forms = fForms,
                         n_chains = 3L,    
                         n_iter = 20000L,
                         n_burnin = 2000L
                         )
 
# summary(multi_joint_model)
```

# M3 - Time-dependent Cox PH model

Construction of start–stop time intervals and event indicators

```{r}
td_data <- long_data %>%
  
  # We add survival information to each longitudinal record
  left_join(
    surv_data %>%
      select(RID, event, time_to_event),
    by = "RID"
  ) %>%
  
  # We sort observations within each individual by time
  arrange(RID, time_years_ptau) %>%
  group_by(RID) %>%
  
  # We define start–stop intervals based on visit times
  mutate(
    time_start = lag(time_years_ptau, default = 0),
    time_end   = time_years_ptau
  ) %>%
  
  # We remove invalid (zero-length) intervals
  filter(time_end > time_start) %>%
  
  # We define the event as occurring in the interval that contains the event time
  mutate(
    event_interval = as.integer(
      event == 1 &
      time_to_event > time_start &
      time_to_event <= time_end
    )
  ) %>%
  
  ungroup()

# Checking event interval distribution
table(td_data$event_interval)
```

Fitted time-dependent Cox PH model

```{r}
# Fitted Cox PH model with time-dependent covariates using p-tau181
surv_model_ptau_td <- coxph(
  Surv(time_start, time_end, event_interval) ~ 
    log_PLASMAPTAU181 + age_at_visit + PTGENDER + APOE4_status,
  data = td_data
)

summary(surv_model_ptau_td)
```

```{r}
# Fitted Cox PH model with time-dependent covariates using NfL 
surv_model_nfl_td <- coxph(
  Surv(time_start, time_end, event_interval) ~ 
    log_PLASMA_NFL +
    age_at_visit +
    PTGENDER +
    APOE4_status,
  data = td_data
)

summary(surv_model_nfl_td)
```

```{r}
# Fitted Cox PH model with time-dependent covariates using both p-tau181 and NfL 
surv_model_both_td <- coxph(
  Surv(time_start, time_end, event_interval) ~ 
    log_PLASMAPTAU181 +
    log_PLASMA_NFL +
    age_at_visit +
    PTGENDER +
    APOE4_status,
  data = td_data
)

summary(surv_model_both_td)
```

```{r}
##### 3. For the Cox PH models with time-dependent covariates
# For model with p-tau181
cox.zph(surv_model_ptau_td)

# For model with NfL
cox.zph(surv_model_nfl_td)

# For model with both
cox.zph(surv_model_both_td)
```

# Test of AUC comparing

We evaluated predictive performance using time-dependent AUC based on incident/dynamic ROC curves at 3-, 5-, and 10-year horizons.

For the Cox PH models = using timeROC, from the package timeROC

For the joint models = using tvAUC, from the package JMbayes2

```{r}
library(timeROC)
library(JMbayes2)
```

```{r}
# We define evaluation time points for time-dependent AUC comparison
times <- c(3, 5, 10)
```

## For the Cox PH models

```{r}
# We define a function to compute time-dependent AUC for Cox models using timeROC
# We use the linear predictor as the risk score and survival outcome from the dataset
get_cox_auc <- function(model, data, times) {
  
  # We extract survival outcome variables
  T <- data$time_to_event
  delta <- data$event
  
  # We compute the linear predictor from the Cox model
  lp <- predict(model, newdata = data, type = "lp")
  
  # We compute time-dependent AUC using timeROC
  auc <- timeROC::timeROC(
    T = T,
    delta = delta,
    marker = lp,
    cause = 1,
    times = times,
    iid = TRUE
  )$AUC
  
  return(auc)
}
```

```{r}
# We compute time-dependent AUC for Cox models (p-tau and NfL)
auc_ptau_bl <- get_cox_auc(surv_model_ptau_bl, surv_data, times)
auc_nfl_bl  <- get_cox_auc(surv_model_nfl_bl, surv_data, times)
auc_both_bl  <- get_cox_auc(surv_model_both_bl, surv_data, times)
```

## For the joint models

Från: <https://cloud.r-project.org/web/packages/JMbayes2/JMbayes2.pdf>

The funcition tvAUC requires "newdata = a data.frame that contains the longitudinal and covariate information for the subjects for which prediction of survival probabilities is required. The names of the variables in this data.frame must be the same as in the data.frames that were used to fit the linear mixed effects and the event process model that were supplied as the two first argument of jm."

```{r}
# We create a unified longitudinal + survival dataset for prediction
newdata <- long_data %>%
  left_join(
    surv_data %>% select(RID, age_at_baseline, time_to_event, event),
    by = "RID"
  )

# We create the p-tau181 specific prediction dataset 
newdata_ptau <- newdata %>%
  select(
    RID,
    time_years_ptau,
    log_PLASMAPTAU181,
    PTGENDER,
    APOE4_status,
    age_at_baseline,
    time_to_event,
    event
  )

# We create the NfL specific prediction dataset 
newdata_nfl <- newdata %>%
  select(
    RID,
    time_years_nfl,
    log_PLASMA_NFL,
    PTGENDER,
    APOE4_status,
    age_at_baseline,
    time_to_event,
    event
  )

# We create a combined prediction dataset for a multivariate joint model
newdata_multi <- newdata %>%
  select(
    RID,
    time_years_ptau,
    log_PLASMAPTAU181,
    time_years_nfl,
    log_PLASMA_NFL,
    PTGENDER,
    APOE4_status,
    age_at_baseline,
    time_to_event,
    event
  )
```

```{r}
# We define a function to compute time-dependent AUC for joint models using tvAUC
# We extract the AUC value from the tvAUC output for each horizon time point
get_jm_auc <- function(model, newdata, times) {
  
  aucs <- sapply(times, function(t) {
    
    tmp <- tvAUC(
      model,
      newdata = newdata,
      Tstart = 0,
      Thoriz = t
    )
    
    tmp$auc   
  })
  
  return(as.numeric(aucs))
}
```

```{r}
# We compute joint model time-dependent AUC for p-tau181 using tvAUC
auc_ptau_jm <- get_jm_auc(
  joint_model_ptau,
  newdata_ptau,
  times
)
```

```{r}
# We compute joint model time-dependent AUC for NfL using tvAUC
auc_nfl_jm <- get_jm_auc(
  joint_model_nfl,
  newdata_nfl,
  times
)
```

```{r}
auc_multi_jm <- get_jm_auc(
  multi_joint_model,
  newdata_multi,
  times
)
```

# Results

```{r}
results <- data.frame(
  Model = c("Cox p-tau181", "Cox NfL", "Cox both", "JM p-tau181", "JM NfL", "JM both"),
  
  AUC_3y  = c(auc_ptau_bl[1], auc_nfl_bl[1], auc_both_bl[1], auc_ptau_jm[1], auc_nfl_jm[1], auc_multi_jm[1]),
  AUC_5y  = c(auc_ptau_bl[2], auc_nfl_bl[2], auc_both_bl[2], auc_ptau_jm[2], auc_nfl_jm[2], auc_multi_jm[2]),
  AUC_10y = c(auc_ptau_bl[3], auc_nfl_bl[3], auc_both_bl[3], auc_ptau_jm[3], auc_nfl_jm[3], auc_multi_jm[3])
)

results[, -1] <- round(results[, -1], 3)

results
```
