---
title: "test"
format: pdf
editor: visual
---

```{r, warning=FALSE}
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) # for running residual diagnostics on GLMM
```

```{r}
# Loading data
#Milou
# NfL data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
# P-tau181 data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
# Diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
# Demographics data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
# Apoe data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

# Preprocessing

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
# ptau_data %>%
#    filter(RID %in% ptau_double$RID) %>% 
#    filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

```{r}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# We merge NFL and ptau data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))

# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID"))
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# We check the length of each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We create a variable for apoe4 status
apoe$APOE4_status <- ifelse(apoe$GENOTYPE %in% c("2/4","3/4","4/4"), 1, 0)

# We merge gender with merged_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We make sure that gender is a factor 
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)

# We merge apoe with merged_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status")], by="RID")
surv_data <- merge(surv_data, apoe[, c("RID", "APOE4_status")], by="RID")

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

```{r}
# We filter individuals with at least 3 measurements of ptau_data
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
long_data <- long_data %>% filter(RID %in% keep)
surv_data <- surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)
```

# Two separate joint models

## Two separate baseline, age_at_baseline and time_to_event (one for ptau, one for nfl) 

```{r}
# We calculate baseline for each biomarker per participant
surv_data_simple <- surv_data %>%
  group_by(RID) %>%
  mutate(
    # We calculate baseline for pTau using the first non-missing measurement
    baseline_pTau = {
      if(all(is.na(PLASMAPTAU181))) NA else min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)])
    },    
    
    # We calculate baseline for NFL using the first non-missing measurement
    baseline_NFL  = {
      if(all(is.na(PLASMA_NFL))) NA else min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])
    },
    
    # We record age at baseline for each biomarker
    age_at_baseline_pTau = age_at_visit[EXAMDATE_ptau == baseline_pTau][1],
    age_at_baseline_NFL  = age_at_visit[EXAMDATE_nfl == baseline_NFL][1],
    
    # We indicate whether the participant ever developed dementia
    event = as.integer(any(DIAGNOSIS_dx == "Dementia")),
    
    # We calculate time-to-event from each biomarker baseline
    time_to_event_pTau = case_when(
  event == 1 ~ as.numeric(
    min(EXAMDATE_dx[
      DIAGNOSIS_dx == "Dementia" & EXAMDATE_dx >= baseline_pTau
    ]) - baseline_pTau
  ) / 365,
  
  TRUE ~ as.numeric(max(EXAMDATE_dx) - baseline_pTau) / 365
),
    time_to_event_NFL = case_when(
  event == 1 ~ as.numeric(
    min(EXAMDATE_dx[
      DIAGNOSIS_dx == "Dementia" & EXAMDATE_dx >= baseline_NFL
    ]) - baseline_NFL
  ) / 365,
  
  TRUE ~ as.numeric(max(EXAMDATE_dx) - baseline_NFL) / 365
)
    
  ) %>%
  ungroup()

# We only keep one baseline row per participant for pTau or NFL
surv_data_simple <- surv_data_simple %>%
  group_by(RID) %>%
  filter(EXAMDATE_ptau == baseline_pTau | EXAMDATE_nfl == baseline_NFL) %>%
  slice(1) %>%  # We take the first row if multiple rows match
  ungroup()

# We count how many participants experienced the event
table(surv_data_simple$event)

# Example: show participant with RID 61
surv_data_simple %>%
  filter(RID == 61)
```

## Two separate time_years (one for ptau, one for nfl) 

```{r}
long_data_simple <- long_data %>%
  group_by(RID) %>%
  # We create baseline for PTau
  mutate(
    baseline_date_ptau = min(EXAMDATE_ptau, na.rm = TRUE),
    time_years_ptau = as.numeric(EXAMDATE_ptau - baseline_date_ptau) / 365.25
  ) %>%
  # We create baseline for NFL
  mutate(
    baseline_date_nfl = min(EXAMDATE_nfl, na.rm = TRUE),
    time_years_nfl = as.numeric(EXAMDATE_nfl - baseline_date_nfl) / 365.25
  ) %>%
  ungroup()
```

## Longitudinal models

```{r}
long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years_ptau + PTGENDER + APOE4_status,
  random = ~ time_years_ptau | RID,                    
  data = long_data_simple
)

summary(long_model_ptau)
```

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years_nfl + APOE4_status, 
  random = ~ time_years_nfl | RID, # random intercept + slope
  data = long_data_simple
  )

summary(long_model_nfl)
```

## Survival model

```{r}
# We create a Surv object: time and event
surv_object_ptau <- Surv(
  time = surv_data_simple$time_to_event_pTau, 
  event = surv_data_simple$event
)

surv_object_nfl <- Surv(
  time = surv_data_simple$time_to_event_NFL, 
  event = surv_data_simple$event
)
```

```{r}
# Cox survival-model, with covariates
surv_model_ptau <- coxph(
  surv_object_ptau ~ age_at_baseline_pTau + APOE4_status,
  data = surv_data_simple,
   x = TRUE
)

summary(surv_model_ptau)

surv_model_nfl <- coxph(
  surv_object_nfl ~ age_at_baseline_NFL + APOE4_status,
  data = surv_data_simple, 
  x = TRUE
)

summary(surv_model_nfl)
```

## Two separate joint models (one for ptau, one for nfl) 

```{r}
joint_model_ptau <- jm(
  surv_model_ptau,
  long_model_ptau,
  time_var = "time_years_ptau",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

summary(joint_model_ptau)
```

```{r}
# Joint model 
joint_model_nfl <- jm(
  surv_model_nfl,
  long_model_nfl,
  time_var = "time_years_nfl",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

summary(joint_model_nfl)
```

# One multivariate joint model

## Global/common version of baseline, age_at_baseline and time_to_event

```{r}
# We calculate global baseline for each participant
surv_data_multi <- surv_data %>%
  group_by(RID) %>%
  mutate(
    
    # Global baseline = first available biomarker visit (pTau or NFL)
    global_baseline = {
      dates <- c(EXAMDATE_ptau, EXAMDATE_nfl)
      if(all(is.na(dates))) NA else min(dates, na.rm = TRUE)
    },
    
    # Age at global baseline
    age_at_global_baseline = age_at_visit[
      EXAMDATE_ptau == global_baseline | EXAMDATE_nfl == global_baseline
    ][1],
    
    # Event indicator (ever dementia)
    event = as.integer(any(DIAGNOSIS_dx == "Dementia")),
    
    # Time-to-event from global baseline
    time_to_event_global = case_when(
      event == 1 ~ as.numeric(
        min(EXAMDATE_dx[
          DIAGNOSIS_dx == "Dementia" & EXAMDATE_dx >= global_baseline
        ]) - global_baseline
      ) / 365,
      
      TRUE ~ as.numeric(max(EXAMDATE_dx) - global_baseline) / 365
    )
    
  ) %>%
  ungroup()

# We keep only 1 row per participant (global baseline row)
surv_data_multi <- surv_data_multi %>%
  group_by(RID) %>%
  filter(
    EXAMDATE_ptau == global_baseline |
    EXAMDATE_nfl == global_baseline
  ) %>%
  slice(1) %>%
  ungroup()

# We check event distribution
table(surv_data_multi$event)

# Example participant
surv_data_multi %>%
  filter(RID == 61)
```

## Global version of time_years

```{r}
long_data_multi <- long_data %>%
  group_by(RID) %>%
  
  mutate(
    global_baseline = min(c(EXAMDATE_ptau, EXAMDATE_nfl), na.rm = TRUE)
  ) %>%
  
  mutate(
    time_years = as.numeric(
      coalesce(EXAMDATE_ptau, EXAMDATE_nfl) - global_baseline
    ) / 365.25
  ) %>%
  
  ungroup()
```

## Survival model with global time_to_event and global age_at_baseline

```{r}
surv_object_multi <- Surv(
  time = surv_data_multi$time_to_event_global, 
  event = surv_data_multi$event
  )
```

```{r}
coxph(
  surv_object_multi ~ age_at_global_baseline + PTGENDER + APOE4_status,
  data = surv_data_multi
)
```

```{r}
surv_model_multi <- coxph(
  surv_object_multi ~ age_at_global_baseline + APOE4_status,
  data = surv_data_multi
)
```

## Longitudinal models with global time_years

```{r}
# Nya lme modeller
long_model_ptau_multi <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status,
  random = ~ time_years | RID,                    
  data = long_data_multi
)

summary(long_model_ptau_multi)

long_model_nfl_multi <- lme(
  log_PLASMA_NFL ~ time_years + APOE4_status, 
  random = ~ time_years | RID, # random intercept + slope
  data = long_data_multi
  )

summary(long_model_nfl_multi)
```

## We create a mixed object

```{r}
mixed <- list(long_model_ptau_multi, long_model_nfl_multi)
```

## We check if the models have stable random slopes

```{r}
VarCorr(long_model_ptau_multi)
VarCorr(long_model_nfl_multi)
```

## fForms

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )      
```

## One multivariate joint model 

```{r}
multi_joint_model <- jm(surv_model_multi, mixed,
                        time_var = "time_years",
                        functional_forms = fForms,
                        n_chains = 3L,    
                        n_iter = 20000L,
                        n_burnin = 2000L
                        )

summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival
```
