---
title: "Multi-state survival model"
format: pdf
editor: visual
---

```{r, warning=FALSE}
# loading the packages
library(lcmm)
library(dplyr)
library(survival)
library(nlme)
library(lcmm)
library(JMbayes2)
library(msm)
library(mstate)
```

```{r}
# loading data
# Milou
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
# nfl_data <- BLENNOWPLASMANFLLONG
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
# ptau_data <- UGOTPTAU181
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
# diagnosis_data <- DXSUM
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
# demographics <- PTDEMOG
```

```{r}
library(ADNIMERGE2)
# loading data
# NfL
nfl_data <- BLENNOWPLASMANFLLONG

# pTAU
ptau_data <- UGOTPTAU181

# dx data
diagnosis_data <- DXSUM

# demographic data
demographics <- PTDEMOG

# genetic data
apoe <- APOERES
```

-   Check how we merge with diagnosis data: we have more data than is present here. For example RID 2 changes from CN to MCI to CN again, but with the full diagnosis data we see that they convert again to MCI. Let's think about how to include this data correctly. Should we make a different data set with conversions from the full diagnosis data set?

```{r}
# selecting relevant variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# adding a suffix to the dx data
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# merging the data sets
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))

# merging only nfl and ptau
nfl_ptau <- nfl_data %>%
  merge(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))
```

```{r}
# checking for missing values with NA
# NfL
sum(is.na(nfl_data))
nfl_data[which(is.na(nfl_data)), ]

# pTAU
sum(is.na(ptau_data))

# checking with ADNI code -4
# NfL
sum(nfl_data$PLASMA_NFL < 0, na.rm = TRUE)

# pTAU
sum(ptau_data$PLASMAPTAU181 < 0, na.rm = TRUE)
```

```{r}
# removing missing values
nfl_ptau <- na.omit(nfl_ptau)
diagnosis_data <- na.omit(diagnosis_data)
```

## Checking RID with more than one measurement recorded on the same exam date

-   There are 52 individuals with two measurements recorded on the same exam date.

-   How should we deal with these? Average the measurements? Remove the observations?

-   Some of the values are quite different, what do we assume by averaging them? What effect does it have to remove them?

-   For now, remove them:

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
# ptau_data %>%
#   filter(RID %in% ptau_double$RID) %>% 
#   filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

## Baseline date in nfl_ptau

```{r}
# checking which observations have the same examdate for nfl and ptau
common <- which(nfl_ptau$EXAMDATE_nfl %in% nfl_ptau$EXAMDATE_ptau)

# checking the observations with different exam dates
nfl_ptau[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
nfl_ptau <- nfl_ptau %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
  ungroup()
```

## Time since baseline in nfl ptau

```{r}
# creating a time since baseline visit variable (in years)
nfl_ptau <- nfl_ptau %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## Selected demographics in nfl_ptau

-   Merging DOB, gender and creating an age at baseline variable

-   Converting gender to factor variable

```{r}
# selecting relevant variables (age, gender) and keeping only one row per person
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# merging the demographic data with the biomarker data
nfl_ptau <- nfl_ptau %>% merge(demographics_small, by = "RID")
```

```{r}
# converting the gender variable to a factor variable 
nfl_ptau <- nfl_ptau %>% mutate(PTGENDER = as.factor(PTGENDER))

# creating an age at baseline variable
nfl_ptau <- nfl_ptau %>% mutate(
  age_at_baseline = as.numeric(format(baseline_date, "%Y")) - PTDOBYY)
```

## Factor variable for genotype in apoe

-   Trying with a factor for each specific genotype

-   If there is not good distinction between these, since we do not have many individuals with some genotypes, we can collpase down to fewer factors later

```{r}
# creating a factor variable for ApoE genotype and selecting relevant variables
apoe = apoe %>% mutate(
  genotype = as.factor(case_when(
    GENOTYPE == "4/4" ~ "e4/e4",
    GENOTYPE %in% c("3/4", "2/4") ~ "e4",
    GENOTYPE == "3/3" ~ "e3/e3",
    GENOTYPE %in% c("2/3", "2/2") ~ "e2"
  ))
) %>%
  select(RID, genotype)
```

## We filter participants with at least 3 measurements in nfl_ptau

```{r}
# Filtering individuals with at least 3 measurements
keep <- unique((nfl_ptau %>% group_by(RID) %>% filter(n() > 2))$RID)
nfl_ptau <- nfl_ptau %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

## Merging biomarker data with diagnosis data by RID but not by VISCODE

-   Keeps all diagnosis data

-   We can then use diagnoses which happen not on the same VISCODE as the biomarker data

-   Check VISCODE2 variable names -\> change to \_bio and \_dx

```{r}
merged_data <- nfl_ptau %>% merge(diagnosis_data_dx, by = "RID") 
```

```{r}
merged_data <- na.omit(merged_data)
```

```{r}
merged_data <- merged_data %>%
  mutate(
    Tevent = as.numeric(EXAMDATE_dx - baseline_date) / 365.25
  )
```

# Multi-State Survival Data

-   For a multistate cox PH model

```{r}
# old code
# dx_times <- merged_data %>%
#   arrange(RID, EXAMDATE_dx) %>%
#   group_by(RID) %>%
#   summarise(
#     t_CN  = first(EXAMDATE_dx[DIAGNOSIS_dx == "CN"]),
#     t_MCI = first(EXAMDATE_dx[DIAGNOSIS_dx == "MCI"]),
#     t_AD  = first(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]),
#     censor_time = max(EXAMDATE_dx)
#   ) %>% ungroup()
```

```{r}
# creating a data frame with diagnosis times: CN -> MCI, MCI -> AD
dx_times <- diagnosis_data %>%
  arrange(RID, EXAMDATE) %>%
  group_by(RID) %>%
  mutate(
# creating a variable for the time from first MCI dx to every CN dx
    t_CN_MCI = ifelse(DIAGNOSIS == "CN",
                      as.numeric(first(EXAMDATE[DIAGNOSIS == "MCI"]) - EXAMDATE) / 365.25,
                      NA),
# creating a variable for the time from first AD dx to every MCI dx
    t_MCI_AD = ifelse(DIAGNOSIS == "MCI",
                      as.numeric(first(EXAMDATE[DIAGNOSIS == "Dementia"]) - EXAMDATE) / 365.25,
                      NA)
# 
  ) %>% ungroup()
```

```{r}
# adding a variable to the dx times data frame which tells us whether an individual goes "backwards" from MCI -> CN or from AD -> MCI
dx_times <- dx_times %>%
  arrange(RID, t_CN_MCI) %>%
  group_by(RID) %>% 
  mutate(backwards_MCI = case_when(
    (any(DIAGNOSIS == "CN") & any(DIAGNOSIS == "MCI")) == TRUE ~
      # (ifelse(is.na(t_CN_MCI) == TRUE,
      #         NA,
              first(t_CN_MCI) < 0),
    
    backwards_AD = ifelse(is.na(t_MCI_AD) == TRUE,
                          NA,
                          first(t_MCI_AD) < 0))

# number of individuals which move backwards from MCI -> CN
sum(na.omit((dx_times %>% select(backwards_MCI, RID) %>% distinct())$backwards_MCI))

# number of individuals which move backwards from MCI -> CN = 0
sum(na.omit((dx_times %>% select(backwards_AD, RID) %>% distinct())$backwards_AD))

# removing the backwards_AD variable as 
dx_times <- dx_times %>% select(!backwards_AD)
```

```{r}
# creating a shorter version of dx_times where the t_MCI takes the first MCI dx which does not revert to CN dx
dx_times_short <- dx_times %>%
  arrange(RID, EXAMDATE) %>%
  group_by(RID) %>%
  summarise(
    t_CN  = first(EXAMDATE[DIAGNOSIS == "CN"]),
    
    t_MCI = case_when(
# for cases where the individual converts back to CN, we remove the first EXAMDATE were dx = MCI and take the next earliest EXAMDATE 
      unique(backwards_MCI) == TRUE ~ first(EXAMDATE[DIAGNOSIS == "MCI"][-1]),
# otherwise, we take the first EXAMDATE where dx = MCI
      TRUE ~ first(EXAMDATE[DIAGNOSIS == "MCI"])
      ),

# there are no reversions from AD dx, so we just take the first AD dx date
    t_AD  = first(EXAMDATE[DIAGNOSIS == "Dementia"]),
    censor_time = max(EXAMDATE)
  ) %>% ungroup()
```

### Transition times

```{r}
# times for transitions from CN -> MCI
# CN_MCI <- dx_times_short %>% filter(is.na(t_CN) == FALSE) %>%
#   mutate(
#     from = "CN",
#     to = "MCI",
#     Tstart = 0,
#     Tstop = as.numeric(case_when(
#       is.na(t_MCI) == TRUE ~ censor_time - t_CN,
#       TRUE ~ t_MCI - t_CN) / 365.25),
#     event = case_when(
#       is.na(t_MCI) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of MCI -> CN
#   filter(
#     Tstop > 0
#   ) %>%
#   select(RID, from, to, Tstart, Tstop, event)
```

```{r}
# times for transitions from MCI -> AD
# MCI_AD <- dx_times_short %>% filter(is.na(t_MCI) == FALSE) %>%
#   mutate(
#     from = "MCI",
#     to = "AD",
#     Tstart = case_when(
#       is.na(t_CN) == FALSE ~ as.numeric(t_MCI - t_CN) / 365.25,
#       TRUE ~ 0),
#     Tstop = as.numeric(case_when(
#       is.na(t_AD) == TRUE ~ censor_time - t_CN,
#       TRUE ~ t_AD - t_CN) / 365.25),
#     event = case_when(
#       is.na(t_AD) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of AD -> MCI
#   filter(
#     Tstart >= 0,
#     Tstop > 0
#   ) %>%
#   select(RID, from, to, Tstart, Tstop, event)
```

```{r}
# times for transitions from CN -> AD
# CN_AD <- dx_times_short %>% filter(is.na(t_CN) == FALSE, is.na(t_AD) == FALSE, is.na(t_MCI) == TRUE) %>%
#   mutate(
#     from = "CN",
#     to = "AD",
#     Tstart = 0,
#     Tstop = as.numeric(case_when(
#       is.na(t_AD) == TRUE ~ censor_time - t_CN,
#       TRUE ~ t_AD - t_CN) / 365.25),
#     event = case_when(
#       is.na(t_AD) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of AD -> CN
#   filter(
#     Tstart >= 0,
#     Tstop > 0
#   ) %>%
#   select(RID, from, to, Tstart, Tstop, event)
```

```{r}
# # times for transitions from CN -> AD or MCI -> AD
# AD_transitions <- dx_times_short %>%
#   mutate(
#     Tstart = 0,
#     Tstop = as.numeric(case_when(
#       is.na(t_AD) == TRUE ~ ifelse(is.na(t_CN) == TRUE,
#                                    censor_time - t_MCI,
#                                    censor_time - t_CN),
#       TRUE ~ ifelse(is.na(t_CN) == TRUE,
#                                    t_AD - t_MCI,
#                                    t_AD - t_CN)) / 365.25),
#     event = case_when(
#       is.na(t_AD) == TRUE ~ 0,
#       TRUE ~ 1
#     )) %>%
# #   filtering out any instances of AD -> CN/MCI
#   filter(
#     Tstop > 0
#   ) %>%
#   select(RID, Tstart, Tstop, event)
```

```{r}
# multistate_data <- bind_rows(CN_MCI, MCI_AD, CN_AD) %>%
#   arrange(RID, Tstart) %>%
#   filter(Tstop >= Tstart)
```

```{r}
surv_data <- merged_data %>%
  group_by(RID) %>%
  summarise(
    age_at_baseline = unique(age_at_baseline),
    PTGENDER = unique(PTGENDER),
    time_to_event = case_when(
      any(DIAGNOSIS_dx == "Dementia") ~
        as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),
      
      TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,
    
    event = as.integer(any(DIAGNOSIS_dx == "Dementia"))) %>%
  ungroup()

# surv_data = merged_data %>% group_by(RID) %>%
#   summarise(
#     age_at_baseline = unique(age_at_baseline),
#     PTGENDER = unique(PTGENDER),
# 
#     time_to_dementia = case_when(
# 
#     any(DIAGNOSIS_dx == "Dementia") ~
#       as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) -
#                    unique(baseline_date)),
# 
#     TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
#                     unique(baseline_date)) / 365),
# 
#     time_to_MCI = case_when(
# 
#     any(DIAGNOSIS_dx == "MCI") ~
#       as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "MCI"]) -
#                    unique(baseline_date)),
# 
#     TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
#                     unique(baseline_date)) / 365),
# 
#     last_dx = DIAGNOSIS_dx[which.max(EXAMDATE_dx)],
# 
#     event = case_when(
#       last_dx == "CN" ~ 0,
#       last_dx == "MCI" ~ 1,
#       last_dx == "Dementia" ~ 2,
#     ))
```

```{r}
# selecting the relevant variables for the survial model
# biomarkers <- nfl_ptau %>%
#   select(RID, time_years, PLASMA_NFL, PLASMAPTAU181)
# 
# # multistate data with the most recent biomarker value added
# multistate_data_biomarkers <- multistate_data %>%
#   left_join(biomarkers, by = "RID") %>%
#   group_by(RID, Tstart, Tstop, from, to, event) %>%
#   # keep only biomarker values measured before interval end
#   filter(time_years <= Tstop) %>%
#   # pick the most recent biomarker
#   slice_max(time_years, n = 1, with_ties = FALSE) %>%
#   ungroup()
```

# Multistate model with coxph()

```{r}
# fit <- coxph(
#   Surv(Tstart, Tstop, event) ~ log(PLASMA_NFL) + log(PLASMAPTAU181) + strata(from, to),
#   data = multistate_data_biomarkers
# )
# 
# summary(fit)
```

# Survival Model

-   merge event data to nfl_ptau

-   event

-   event time

```{r}
# jlcmm_data = nfl_ptau %>% merge(AD_transitions, by = "RID")
```

```{r}
nfl_ptau <- nfl_ptau %>% arrange(RID, time_years)
```

```{r}
# surv_data <- merged_data %>%
#   group_by(RID) %>%
#   summarise(
#     Tevent = ifelse(any(DIAGNOSIS_dx == "Dementia"),
#                     min(time_years[DIAGNOSIS_dx=="Dementia"]),
#                     max(time_years)),
#     event = ifelse(any(DIAGNOSIS_dx == "Dementia"), 1, 0)
#   )
```

```{r}
# Jointlcmm(
#   fixed = PLASMA_NFL ~ time_years,
#   random = ~ RID,
#   mixture = ~ time_years,
#   classmb = ~ 1,
#   subject = "RID",
#   survival = Surv(Tstop, event) ~ age_at_baseline,
#   data = jlcmm_data,
#   ng = 3
# )
```

## Creating a single EXAMDATE variable

```{r}
library(lubridate)
nfl_ptau <- nfl_ptau %>%
  mutate(EXAMDATE = EXAMDATE_nfl + (EXAMDATE_nfl - EXAMDATE_ptau)) %>%
  select(!c(EXAMDATE_nfl, EXAMDATE_ptau)) %>%
  relocate(c(baseline_date, EXAMDATE, time_years), .after = VISCODE2)
```

## Survival data (CN/MCI -\> AD only)

```{r}
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)

surv_data <- diagnosis_data %>%
  group_by(RID) %>%
  
  # We create baseline date per patient (whichever EXAMDATE comes first, nfl or ptau)
  mutate(baseline_date = min(EXAMDATE),
  
  # We calculate event and time_to_event based on all diagnosis data
    # 1 if patient ever developed dementia
  event = as.integer(any(DIAGNOSIS == "Dementia")),
  time_to_event = case_when(
    event == 1 ~
      as.numeric(min(EXAMDATE[DIAGNOSIS == "Dementia"]) - baseline_date)/365,
    TRUE ~
      as.numeric(max(EXAMDATE) - baseline_date)/365)) %>%
  
    # We calculate age at baseline
  # age_at_baseline = as.numeric(format(unique(baseline_date), "%Y")) - PTDOBYY) %>%
      ungroup() %>%
  select(RID, event, time_to_event) %>%
  distinct()
  
surv_data <- surv_data %>% filter(RID %in% keep)

  # # We keep only one baseline row per patient
  # slice_min(order_by = EXAMDATE_ptau, n = 1, with_ties = FALSE)
    
  # We remove grouping

# We check the number of events 
table(surv_data$event)
```

## Joint Latent Class Mixed Models DATA

```{r}
# reference
# lcmm package data - example that works with the jlcmm
d = data_lcmm
```

```{r}
# merging longitudinal and survival data
jlcmm_data <- nfl_ptau %>%
  merge(surv_data, by = "RID") %>%
  relocate(c(event, time_to_event), .after = time_years)
```

```{r}
# transforming the data
jlcmm_data <- jlcmm_data %>%
  mutate(ln_PLASMA_NFL = log(PLASMA_NFL))
```

```{r}
# merging genetic data
jlcmm_data <- jlcmm_data %>%
  merge(apoe, by = "RID")
```

# 

## Collapsing genotype down to 3 factors

-   e2/e3 containing genotypes

-   e4 heterozygotes

-   e4 homozygotes

-   Now all genotype factors significant

-   Now time_years variance still low, try with only random intercept

```{r}
# checking current factors
table(jlcmm_data_test$genotype)
```

```{r}
jlcmm_data_test <- jlcmm_data_test %>%
  mutate(
    genotype = case_when(
      genotype == "e2" ~ "e2 e3",
      genotype == "e3/e3" ~ "e2 e3",
      genotype == "e4" ~ "e4",
      genotype == "e4/e4" ~ "e4/e4"
    )
  )
```

```{r}
# checking changed factors
table(jlcmm_data_test$genotype)
```

# OLD

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event) + lägger till age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_ptau)],
    PTGENDER = PTGENDER[which.min(EXAMDATE_ptau)],
    
    time_to_event = case_when(

    any(DIAGNOSIS_dx == "Dementia") ~
      as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,

  event = case_when(
    DIAGNOSIS_dx == "CN" ~ 1,
    DIAGNOSIS_dx == "MCI" ~ 2,
    DIAGNOSIS_dx == "Dementia" ~ 3
  )
) %>% ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

```{r}
length(unique(merged_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

# Multistate model with msm()

```{r}
# multi-state data
ms_data <- merge(nfl_data, ptau_data,
                by = c("RID", "VISCODE2"),
                suffix = c("_nfl", "_ptau")) %>%
  merge(diagnosis_data %>% select(RID, VISCODE2, EXAMDATE, DIAGNOSIS),
        by = c("RID", "VISCODE2")) %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_nfl)) %>%
  mutate(time_years = as.numeric(as.character((EXAMDATE - baseline_date) / 365.25))) %>%
  mutate(event = case_when(
    DIAGNOSIS == "CN" ~ 1,
    DIAGNOSIS == "MCI" ~ 2,
    DIAGNOSIS == "Dementia" ~ 3,
  ))

# removing any observations with time_years < - 0.1, and recoding any time_years from -0.1 - 0 as 0
ms_data <- ms_data %>%
  filter(time_years > - 0.09) %>% mutate(
    time_years = case_when(
      time_years < 0 ~ 0,
      TRUE ~ time_years
    )
  )
```

### Specifying which transitions are possible

-   CN -\> MCI

-   CN -\> AD

-   MCI -\> CN

-   MCI -\> AD

-   AD -\> MCI

```{r}
# Specifying the transitions
Q <- rbind(
# CN: can convert to MCI or AD
  c(0, 0.1, 0.01),
# MCI: can covert to CN or AD
  c(0.1, 0, 0.2),
# AD -> MCI only
  c(0, 0.1, 0)
)

# removing obs with 1 measurement
ms_data <- ms_data %>%
  group_by(RID) %>%
  filter(n() > 1) %>%
  ungroup()

# grouping the data by RID and arranging by time, the msm() function requires this form the data set
ms_data <- ms_data %>%
  group_by(RID) %>%
  arrange(time_years, .by_group = TRUE)

# Use crudeinits.msm to get better starting values
Q_crude <- crudeinits.msm(event ~ time_years, 
                           subject = RID, 
                           data = ms_data, 
                           qmatrix = Q)
```

### Multi-state model with ptau and nfl as covariates

```{r}
# basic msm, no covariates
fit_msm <- msm(event ~ time_years,
               subject = RID,
               data = ms_data,
               covariates = ~ log(PLASMA_NFL) + log(PLASMAPTAU181),
               qmatrix = Q_crude)

```

### Outputs

```{r}
# Transition intensity matrix
qmatrix.msm(fit_msm)

# Hazard ratios for covariates
hazard.msm(fit_msm)

# Transition probability matrix at a given time
# probabilities at 1 year
pmatrix.msm(fit_msm, t = 1)

# mean time in each state
sojourn.msm(fit_msm)

# # Plot transition probabilities over time
# range = range(ms_data$time_years)
# plot(fit_msm, range = c(0, 8.01))
```
