---
title: "Multi-state survival model"
format: pdf
editor: visual
---

```{r, warning=FALSE}
library(dplyr)
library(nlme)
library(survival)
library(JMbayes2)
library(msm)
library(mstate)
```

```{r}
# loading data
# Milou
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
# nfl_data <- BLENNOWPLASMANFLLONG
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
# ptau_data <- UGOTPTAU181
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
# diagnosis_data <- DXSUM
# 
# load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
# demographics <- PTDEMOG


# Charley
library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

APOE = APOERES
```

## 1. We select only the required variables and merge them

```{r}
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# 1. Lägg _dx på diagnos-kolumner
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# 2. Slå ihop NFL, ptau och diagnos
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))
```

Checking missing values

```{r}
# checking for missing values with NA

# NfL
sum(is.na(nfl_data))
nfl_data[which(is.na(nfl_data)), ]

# pTAU
sum(is.na(ptau_data))

# checking with ADNI code -4
# NfL
sum(nfl_data$PLASMA_NFL < 0, na.rm = TRUE)

# pTAU
sum(ptau_data$PLASMAPTAU181 < 0, na.rm = TRUE)
```

## 2. We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# 1. Håll bara demografi-variabler vi behöver och en rad per person
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# 2. Slå ihop med merged_data
merged_data <- merged_data %>%
  left_join(demographics_small, by = "RID")

# 3. Säkerställ att kön är faktor
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# 4. Konvertera EXAMDATE_ptau till Date
merged_data$EXAMDATE_ptau <- as.Date(as.character(merged_data$EXAMDATE_ptau), format="%Y-%m-%d")

# 5. Skapa age_at_visit baserat på ptau datum
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_ptau, "%Y")) - merged_data$PTDOBYY
```

## 3. We filter participants with at least 3 measurements

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event) + lägger till age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_ptau)],
    PTGENDER = PTGENDER[which.min(EXAMDATE_ptau)],
    
    time_to_event = case_when(

    any(DIAGNOSIS_dx == "Dementia") ~
      as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,

  event = case_when(
    DIAGNOSIS_dx == "CN" ~ 1,
    DIAGNOSIS_dx == "MCI" ~ 2,
    DIAGNOSIS_dx == "Dementia" ~ 3
  )
) %>% ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## 7. We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(merged_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## 8. We make sure we don't have any extreme values (for the longitudinal model)

```{r}
summary(merged_data$log_PLASMAPTAU181)
any(is.infinite(merged_data$log_PLASMAPTAU181)) # should be FALSE
# Inf = extreme values 

summary(merged_data$log_PLASMA_NFL)
any(is.infinite(merged_data$log_PLASMA_NFL)) # should be FALSE
# Inf = extreme values 
```

## 9. We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
length(unique(merged_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

# Longitudinal model

```{r}
# Longitudinal model
long_model <- lme(
  log_PLASMAPTAU181 ~ time_years + PTGENDER.y,
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model

```{r}
# multi-state data
ms_data <- merge(nfl_data, ptau_data,
                by = c("RID", "VISCODE2"),
                suffix = c("_nfl", "_ptau")) %>%
  merge(diagnosis_data %>% select(RID, VISCODE2, EXAMDATE, DIAGNOSIS),
        by = c("RID", "VISCODE2")) %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_nfl)) %>%
  mutate(time_years = as.numeric(as.character((EXAMDATE - baseline_date) / 365.25))) %>%
  mutate(event = case_when(
    DIAGNOSIS == "CN" ~ 1,
    DIAGNOSIS == "MCI" ~ 2,
    DIAGNOSIS == "Dementia" ~ 3,
  ))

# removing any observations with time_years < - 0.1, and recoding any time_years from -0.1 - 0 as 0
ms_data <- ms_data %>%
  filter(time_years > - 0.09) %>% mutate(
    time_years = case_when(
      time_years < 0 ~ 0,
      TRUE ~ time_years
    )
  )
```

###  Specifying which transitions are possible

-   CN -\> MCI

-   CN -\> AD

-   MCI -\> CN

-   MCI -\> AD

-   AD -\> MCI

```{r}
# Specifying the transitions
Q <- rbind(
# CN: can convert to MCI or AD
  c(0, 0.1, 0.01),
# MCI: can covert to CN or AD
  c(0.1, 0, 0.2),
# AD -> MCI only
  c(0, 0.1, 0)
)

# removing obs with 1 measurement
ms_data <- ms_data %>%
  group_by(RID) %>%
  filter(n() > 1) %>%
  ungroup()

# grouping the data by RID and arranging by time, the msm() function requires this form the data set
ms_data <- ms_data %>%
  group_by(RID) %>%
  arrange(time_years, .by_group = TRUE)

# Use crudeinits.msm to get better starting values
Q_crude <- crudeinits.msm(event ~ time_years, 
                           subject = RID, 
                           data = ms_data, 
                           qmatrix = Q)
```

### Multi-state model with ptau and nfl as covariates

```{r}
# basic msm, no covariates
fit_msm <- msm(event ~ time_years,
               subject = RID,
               data = ms_data,
               covariates = ~ log(PLASMA_NFL) + log(PLASMAPTAU181),
               qmatrix = Q_crude)

```

### Outputs

```{r}
# Transition intensity matrix
qmatrix.msm(fit_msm)

# Hazard ratios for covariates
hazard.msm(fit_msm)

# Transition probability matrix at a given time
# probabilities at 1 year
pmatrix.msm(fit_msm, t = 1)

# mean time in each state
sojourn.msm(fit_msm)

# # Plot transition probabilities over time
# range = range(ms_data$time_years)
# plot(fit_msm, range = c(0, 8.01))
```

### Need to fit a coxPH model 

```{r}

```

# Joint model

How to use a msm object in jm() function? -\> pass coxPH multistate model instead of msm()!!

```{r}
# Joint model 
joint_model <- jm(
  fit_msm,
  long_model,
  time_var = "time_years"
  )
```
