---
format:
  pdf:
    include-before-body:
      text: |
        \begin{titlepage}
        \raggedleft
        \vspace*{2cm}
        
        {\Huge\bfseries Kandidatuppsats\par}
        \vspace{1.5cm}
        
        {\Large Statistiska Instiutionen\par}
        \vspace{3cm}
        
        {\Huge\bfseries Nr 2026:x\par}
        \vspace{0.5cm}
        
        {\LARGE \bfseries Title in English\par}

        {\large Title in Swedish\par}
        \vspace{0.5cm}
        
        {\large \today\par}
        \vspace{3cm}
        
        {\Huge Charlotte Emily Iszatt and Milou Rossi\par}
        
        \vfill
        
        \end{titlepage}
        
        \begin{titlepage}
        \raggedleft
        
        {\normalsize Självständigt arbete 15 högskolepoäng inom Statistik III, VT2026\par}
        {\normalsize Handledare: Hans Nyqvist\par}
        \vspace{2cm}
        
        \raggedright
        {\huge\bfseries Abstract\par}
        \vspace{1cm}
        
        {\Large The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. Keywords: XXX, XXX, XXX\par}
        \vspace{3cm}
        
        \vfill
        
        \end{titlepage}
    toc: false
    colorlinks: false
    number-sections: true
editor: visual
---

```{=latex}
\tableofcontents

\newpage
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, output = TRUE)
```

```{r, warning=FALSE}
# loading libraries
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) # for running residual diagnostics on GLMM
```

```{r}
# Loading data
#Milou
# NfL data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
# P-tau181 data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
# Diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
# Demographics data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
# Apoe data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

```{r}
# Charley
library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

apoe = APOERES
```

# Introduction

## Background

*General background information. Introducing some domain specific information about AD diagnosis and pathology, so that the reader is well prepared.*

According to World Health Organization (2025), Alzheimer's disease (AD) is the most common neurodegenerative disorder and a leading cause of dementia around the world. First described by Dr. Alois Alzheimer in 1901, the disease continues to present significant challenges for the medical field, and no definitive cure exists to date (WHO, 2025).

The 2011 National Institute on Aging-Alzheimer’s Association (NIA-AA) diagnostic guidelines describe the disease as a continuum with three stages: a symptom-free stage (preclinical AD), a predementia stage (mild cognitive impairment (MCI) due to AD) and a dementia stage (dementia due to AD) (Carrillo et al. 2013: p. 596). Pathophysiological changes in AD can begin up to 20 years before clinical symptoms appear, highlighting the crucial importance of early diagnosis and timely intervention to potentially slow the progression of the disease (Carrillo et al. 2013: p. 595).

In the context of Alzheimer's disease, as well as diseases in general, the term "biomarker" is commonly used. According to the *Biomarkers Definitions Working Group*, a biomarker can be defined as "a characteristic that is objectively measured and evaluated as an indicator of normal biological processes, pathogenic processes, or pharmacologic responses to a therapeutic intervention." (2001: p. 91). Similarly, the *International Programme on Chemical Safety*, which is a a joint venture on chemical safety led by the *World Health Organization* (WHO) and in coordination with the *United Nations* and the *International Labor Organization*, defines it as "any substance, structure, or process that can be measured in the body or its products and influence or predict the incidence of outcome or disease" (2001: p. 1).

When it comes to Alzheimer’s disease, the detection of AD pathophysiology was historically done by positron emission topography (PET) to detect the presence of amyloid plaques in the brain, or by analysis of cerbrospinal fluid (CSF) to quantify the concentration of biological molecules implicated in AD pathology. Recent advances in research have made it possible to replace these invasive procedures with blood tests.

WHO published in 2024 a report highlighting the promising progress made in recent years regarding the use of blood-based biomarkers to support AD diagnosis. Among the ones currently being explored and used in research settings, are mentioned p-tau181, a plasma phosphorylated tau, and NfL, a plasma neurofilament light (2024: p. 7).

<https://iris.who.int/server/api/core/bitstreams/43c83ef7-7a34-4b56-ae17-b584f2fc0798/content>

Från: <https://journals.sagepub.com/doi/10.1177/25424823261432112> –\> kanske hitta bättre, mer konkreta källor för infon, men bra info?

Tau proteins, which are microtubule-binding proteins, play an essential role in maintaining neuronal stability and morphology under normal conditions. The regional progression of brain atrophy in AD is closely associated with the accumulation of tau proteins. Hyperphosphorylated tau (p-tau), in its full-length or truncated forms, represents a key pathological hallmark of AD, exhibiting over 70 post-translational modifications, including more than 40 phosphorylation sites. Various forms of p-tau can be measured in cerebrospinal fluid (CSF) and plasma. According to the NIA-AA, there are several core biomarkers to be used in the diagnosis and staging of AD (Ghahremani et al., 2022). One such core biomarker is tau protein phosphorylated at the threonine-181 position (p-tau181), where concentrations are expected to be higher in individuals with AD pathology (Wong et al., 2012). **MORE ON p-tau181 here...**

-   Among these, p-tau181 is significantly elevated in the CSF of AD patients and is considered a critical biomarker for the disease. The ability of p-tau181 to differentiate AD from other neurodegenerative conditions, such as frontotemporal lobar degeneration and chronic traumatic encephalopathy, is crucial for the accurate diagnosis of AD. Furthermore, p-tau181 levels correlate with AD severity and predict disease progression, providing valuable insights for evaluating therapeutic efficacy and guiding treatment strategies. Recent findings indicate that p-tau181 have excellent discriminative performance in plasma, paving the way for non-invasive monitoring of AD.

Neurofilament light chain (NfL) is instead a marker of neuronal injury not specific to AD (Ghahremani et al.).

The abnormal concentrations of these biomarkers are the result of the neuropathological changes symptomatic of AD.

These two biomarkers, along with various other biomarkers and clinical measures related to Alzheimer’s disease, are the main focus of the *Alzheimer’s Disease Neuroimaging Initiative* (ADNI). Since 2004, researchers across more than 60 clinical sites in the USA and Canada have continuously collected and analyzed data throughout the different phases of the study. The current phase, ADNI4, began in 2022 and is expected to be completed in 2027. Its main goals are to provide open-access data to researchers worldwide and to improve the diagnosis of Alzheimer’s disease (ADNI, n.d.)

## Research Context

*Assessing the current landscape in terms of the scientific literature. What is known? What has been done Where are there gaps??? (This leads on to the next section, why are we doing this research, how do we hope to answer some remaining question in the AD research, orient this towards the statistical analysis/approaches!)*

Recent research has applied a range of prognostic models, from traditional to advanced, to estimate disease progression and related adverse outcomes in Alzheimer’s disease. For instance, one article from 2024 investigated the predictive value of plasma p-tau181 for conversion from mild cognitive impairment (MCI) to Alzheimer’s disease dementia using Cox proportional hazards models (Aguilera, 2024: p. 1): the most widely used procedure for modeling survival outcomes and other censored data in relation to covariates (Grambsch & Therneau, 2000: p. 39). The study found that higher baseline plasma p-tau181 levels were associated with a significantly increased risk of progression, with a hazard ratio of around 1.84, compared with lower levels, even after adjusting for age, sex, and APOE genotype (Aguilera, 2024: p. 1).

Another article from the same year further extended this approach by using a joint modeling framework to analyze multiple longitudinal biomarkers, including plasma p-tau181 and NfL (Fang et al., 2024: p. 1). The joint model distinguishes itself from other models focusing on survival outcome by accounting for the effect of an endogenous time-dependent covariate measured with error, such as a biomarker (Rizopoulos, 2012: p. vii). More precisely, it couples the survival model, which is of primary interest, with an appropriate model for the repeated measurements of the endogenous covariate, thereby accounting for its specific characteristics (Rizopoulos, 2012: p. 51-52). In fact, several studies have examined these biomarkers using only a longitudinal model (Andreasson et al., 2019; Chen et al., 2021). The study using a joint model found that higher longitudinal p-tau181 levels were associated with a significantly increased risk of progression, with the joint model estimating a hazard ratio of approximately 5.83 for p-tau181 and 4.18 for NfL (Fang et al., 2024: p. 1). Additionally, the two biomarkers were combined into a multivariate joint model, leading to improved performance (Fang et al., 2024: p. 1, 3).

Thus, the question arises as to which modeling approach is best suited for predicting progression to Alzheimer’s disease dementia, particularly when considering the added value of longitudinal biomarker information.

Additionally, an important consideration is whether to include all participants with repeated biomarker measurements. Although the two previously mentioned studies included all participants regardless of the number of measurements per individual, this is not always the case. For example, a study examining different cognitive biomarkers to predict progression to Alzheimer’s disease explicitly states: "Participants were included if they had a baseline diagnosis of MCI and at least three follow-up visits with complete longitudinal cognitive assessments" (Cui et al., 2026). Similarly, in a study on type 1 diabetes and primary biliary cirrhosis (PBC), the joint model analysis was restricted to participants with recorded height and weight and at least three biomarker measurements (Xu, 2021: p. 9). Another study, focusing on cardiovascular events, restricted their joint model analysis to the cohort of 3784 patients who had at least 10 measurements of the selected biomarker. This was done to ensure that the approximation method did not lead to "non-ignorable biases" (Charalambous et al., 2024: p. 591-592).

In other words, some studies include all available participants, whereas others choose to include only participants with a minimum number of measurements, meaning that those with only a single measurement are left excluded from the modeling process.

Inför en källa som säger att individerna borde ha 3 eller fler mätningar

**Behövs källa** –\> On one hand, the longitudinal effects are only estimated from participants with more than one measurement. On the other hand, the excluded ones still contribute in estimating the cross-sectional effects of covariates, such as time effects. Additionally, there is the aspect of missing data, which means that a likelihood-based repeated measurements analysis will be valid under the missing at random assumption if they are in. It is thus recommended to still include the participants with only one measurement.

## Research Aims

What are our research questions?

Förslag 1.

This study evaluates the prediction of Alzheimer’s disease/dementia progression using longitudinal biomarker data, comparing conventional and advanced statistical approaches. More precisely, the research questions are:

1.  Which modeling approach provides the best prediction of progression to Alzheimer’s disease/dementia when using longitudinal biomarker data (Including baseline levels, average levels over time, single-biomarker joint models, and multivariate joint models)?
2.  How do these associations differ between p-tau181 and NfL? (Comparison across biomarkers)
3.  How does the precision of the estimated longitudinal trajectories and their associations with progression (survival outcomes) change when including individuals with different numbers of repeated measurements? (Precision analysis)

Förslag 2.

1.  How do longitudinal changes in plasma biomarker levels over time relate to the risk of progression to dementia? (Joint modeling)

2.  How do these associations differ between p-tau181 and NfL? (Comparison across biomarkers)

3.  How does the precision change for the estimated longitudinal trajectories and survival associations to the inclusion of individuals with different numbers of repeated measurements? (Precision analysis)

Sensitivity → how do the estimates of the the joint model parameters change when excluding sparce individuals with only one biomarker measurement?

Why is this research relevant to the existing body of literature? What does it add? Gaps in previous literature which we outline in above \^\^.

# Methodology

How will we accomplish this? –\> Introduce the methods we will use.

Why did we choose this method to answer the research question? Why not other methods, such as cox models with time-dependent covariates - just den anses inte som lämplig för vår data, bara lämpligt för viss typ av data. Dessutom har vi begränsad tid och utrymme, så om vi hade valt fler modeller hade vi inte kunnat utforska respektive en lika mycket på djupet, vi valde de vi ansåg som viktigast och som skiljer sig mest från varandra, för att verkligen kunna se bredden/variationen av modeller som finns.

Why did we only choose to look at p-tau181 and NfL and not other biomarkers? - För att vi har begränsad tid och begränsat utrymme att skriva på, och två är tillräckligt för att vi ska kunna jämföra mellan biomarkörer. Om vi hade valt fler hade vi inte kunnat utforska vardera biomarkör på djupet lika mycket. Dessa två dataset hade även väldigt många individer med samma RID, så det var väldigt lämpligt. Många individer med många mätningar i båda dataseten.

Varför inte mer aktuell data? För det inte finns tillräckligt många individer med 3 eller fler mätningar i aktuella än, skulle ha gett för litet stickprovsstorlek om vi valde det och därmed stor osäkerhet i skattningarna.

Varför just ADNI data? Inte lika relevant för oss som är i Sverige. - För det är den enda data som vi fick tillgång tid på så kort tid, vi hade tidsbegränsning, och i Sverige är det svårt generellt att få tillgång på individ-hälsodata på grund av regler.

# Statistical theory

Describe the statistical theories behind the chosen methods

## Longitudinal Data

Longitudinal data in this analysis is defined as $n$ repeated measurements of p-tau181 and NfL on $N$ individuals in the ADNI1, ADNIGO and ADNI2 cohorts. $T_{i, j}$ and $N_{i, j}$ denote the measured concentrations of p-tau181 and NfL respectively for the $i^{th}$ individual at the $j^{th}$ occasion. Assuming that the expected concentration of the biomarkers varies with both time and individual level covariates, the mean responses have the notation $\mu_{T:\text{ } i, j} = E(T_{i, j})$ and $\mu_{N:\text{ } i, j} = E(N_{i, j})$. Considering the data are made up of repeated responses on the same individuals, the assumption of independence between observations is illogical. Clearly, the baseline levels of p-tau181 and NfL for an individual are correlated with the measurements on subsequent occasions.

## Mixed Effects Models

Why???

We have grouped data, specifically longitudinal data where groups of observations are correlated, since we have repeated measurements on the same individuals. We expect data *within* the groups to be more correlated than those *between* the groups. Mixed effects models are not only for longitudinal data, but for any data where there is non-independence between observations.

### Linear Mixed Effect Model

**Theory about the model... (specifically why we should use it)**

-   Biomarker process

-   Describes the trajectory of biomarkers at the individual level 

**Linear mixed effects models?** What are these - why use them here?

$$
y_i(t) = m_i(t) + \epsilon_i(t) \qquad \epsilon_i(t) \sim N(0, \sigma^2)
$$

$$
m_i(t) = \mathbf{X_i^T}\beta + \mathbf{Z_i}^T\mathbf{b_i} \qquad \mathbf{b_i}(t) \sim N(0, \sigma^2)
$$

$y_i(t):$ the observed longitudinal biomarker

$m_i(t):$ the trajectory function which represents the **true unobserved** **value of the biomarker** at time $t$ for patient $i$ - we assume that we do not observe the true value due to measurement errors or other reasons.

$\mathbf{X_i}^T(t)\beta:$ the fixed effects

$\mathbf{Z_i}^T\mathbf{b_i}:$the random effects

The full model becomes:

$$
y_i(t) = \mathbf{X_i^T}\beta + \mathbf{Z_i}^T\mathbf{b_i} + \epsilon_i(t) \qquad \mathbf{b_i}(t) \sim N(0, \sigma^2), \quad \epsilon_i(t) \sim N(0, \sigma^2)
$$

### Generalised Linear Mixed Effects Model

Ref: <https://stats.oarc.ucla.edu/other/mult-pkg/introduction-to-generalized-linear-mixed-models/>

An extension of GLMs, where the linear predictor contains random effects alongside the fixed effects. This is used to extend the linear mixed model to non-normal data.

Given random effects, $\mathbf{u}$ the dependent variable, $\mathbf{y}$, is distributed according to the exponential family, such that $g(E[\mathbf{y}|\mathbf{u}]) = \mathbf{X𝛽} + \mathbf{Zu}$. Some link function, $g(x)$, links the expected value of $\mathbf{y}$ given the random effects to the linear predictor which included both fixed effects and random effects.

Where:

$\mathbf{X}$ = the fixed effects design matrix

$\mathbf{\beta}$ = the fixed effects

$\mathbf{Z}$ = the random effects design matrix

$\mathbf{u}$ = the random effects

What do we model as random?

We can model random intercept and keep all other effects fixed. In that case, for $i$ biomarker measurements on $j$ individuals, the random effects design matrix - $\mathbf{Z}$ - is a $i \cdot j$ matrix containing 1's and 0's, where $z_{i,j} = 1$ if the $i^{th}$ biomarker measurements were taken from the $j^{th}$ individuals, and $z_{i,j} = 0$ if the $i^{th}$ biomarker measurements were not taken from the $j^{th}$ individual.

We do not tend to estimate $\mathbf{u}$, but instead assume that $u_i \sim N(0, \mathbf{G})$, where $\mathbf{G}$ is the variance-covariance matrix of the random effects. The fixed effects are estimated, and the random effects are seen as deviations from the fixed effects and have $E(u) = 0$.

For a model with only a random intercept, $\mathbf{G}$ becomes $G = \sigma^2_{intercept}$. If we extend the model to have random slope, we get:

$$
G =
\begin{bmatrix}
\sigma^2_{intercept} & \sigma_{intercept, slope} \\
\sigma_{intercept, slope} & \sigma^2_{slope}
\end{bmatrix}
$$

Like all variance-covariance matrices, some elements are redundant since they are repeated. Estimating instead some function of $\mathbf{G}$, where $\mathbf{G} = \sigma(\theta)$, means estimating $\frac{p(p + 1)}{2}$ elements instead. Assuming (**?????**) the random effects are independent, the variance-covariance matrix for the random effects becomes diagonal:

$$
G =
\begin{bmatrix}
\sigma^2_{intercept} & 0 \\
0 & \sigma^2_{slope}
\end{bmatrix}
$$

### Assumptions!!

-   Random effects $\mathbf{b_i}$ are **time-invariant** - they are assumed to come from a normal distribution, $\mathbf{b_i} \sim N(0, \sigma^2)$, and represent stable, latent characteristics about the individuals disease, such as a generally fast declining disease trajectory but cannot represent other factors that effect disease trajectory over-time, e.g. medication. Any external time-varying process that effects the biomarker trajectory and the hazard cannot be represented by the random effects.

## Proportional Hazard Model

**Theory about the model... (specifically why we should use it)**

-   Time-to-diagnosis process

-   Parametric baseline hazard: **flexible parametrics, splines??**

-   OR leave undefined -\> Cox PH model

Assuming a Cox PH survival sub-model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha m_i(t) \right\}
$$

$\mathbf{v_i}:$ vector of time-independent co-variates

$\phi:$ vector of log-hazard ratios

$\alpha:$ the association parameter

$M_i = \left\{m_i, \quad 0 \le s \le t\right\}:$ the **true unobserved biomarker profile** up to time $t$

### Assumptions!!

-   What do we assume about the censoring process?

-   Proportionality

### Cox PH model using baseline values of each biomarker

### Cox PH model using time-averaged values of each biomarker

Additionally, the cumulative average observed up through each time could be used as a time-dependent covariate, particularly if that covariate is under study as a cause of mortality (Strawderman & Wolfe,1996: p.128). This average provides an unbiased estimate of the true mean, assuming no underlying trend, bias or heteroskedasticity (Barrett et al., 2017: p.4516). The resulting time-varying predictor, called the cumulative average (CA), is defined as:

$$
H^{(CA)}_t(X_i) = \frac{\sum_{j=0}^{m_i} I(t_{ij} < t) X_i(t_{ij})}{\sum_{j=0}^{m_i} I(t_{ij} < t)}
$$

where $I(\cdot)$ is an indicator function equal to 1 if the measurement occurred before time $t$ and 0 otherwise. It can replace the original predictor in the model (Barrett et al., 2017, p.4516).

## Joint Models

### Simple Joint Model

**Theory about the model... (specifically why we should use it)**

In the joint model, the hazard function at time $t$ is directly dependent on the estimate of the biomarker at time $t$.

The association structure of the joint model:

-   How is the longitudinal model linked to the survival model? 

-   Does this association structure change for different biomarkers?

-   What happens to time-to-diagnosis predictions if we assume the standard structure when another underlying association is a better fit?

If we include the current rate of change alongside the current value in the proportional-hazard model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha_1 m_i(t) + \alpha_2 m_i'(t) \right\} \qquad where \qquad m_i'(t) = \frac{d}{dt} m_i(t)
$$

$\alpha_1 :$ links the current value of the biomarker - $m_i(t)$

$\alpha_2 :$ links the current gradient/rate of change - $m_i'(t)$

Joint-model options:

-   Longitudinal sub-model: model of class lme()

### Assumptions!!

-   Full conditional independence

    -   Random effects $\mathbf{b_i}$ explain all statistical dependence between (1) the longitudinal outcome and the survival outcome and (2) repeated measurements within the longitudinal process.

    -   This assumption can fail when: there are time-varying confounders not captured by the random effects (see the longitudinal model for more info about $\mathbf{b_i}$ assumptions), if the measurement error, $e_i(t)$ in the longitudinal model $y_i(t) = m_i(t) + e_i(t)$ is informative, or there are sub-populations (APOE genotype??? Maybe we need to split into populations based on genotype)

-   So: the longitudinal outcome is independent of the time-to-event outcome and the repeated measurements in the longitudinal outcome are independent of each other

-   $\mathbf{b_i}$ drives the **true (unobserved)** biomarker trajectory and hazard of AD diagnosis: the observed $y_i(t)$ is the observed biomarker state, $m_i(t)$, plus noise. Where $m_i(t)$ is itself determined by $\mathbf{b_i}$, \*\*since the others are fixed ????? \*\*, once we know $\mathbf{b_i}$, the observed biomarker measurement and the survival times are no longer dependent. The observed $y_i$'s are useful in determining $\mathbf{b_i}$ which is the useful information in estimating survival times.

### Frequentist vs Bayesian approach

However, as the joint model includes a linear mixed-effects model with the normal assumption for a single longitudinal outcome, joint model-based analyses may not provide robust inference when longitudinal measurements exhibit skewness and/or heavy tails (Huang et al, 2022: p.?). Consequently, joint models, besides with the traditional approach of maximum likelihood estimation, can also be estimated within a Bayesian framework using MCMC techniques (Rizopoulos, 2012, p. 62). This approach allows for more flexible and complex association structures (Andrinopoulou & Rizopoulos, 2016: p.?).

### Multivariate Joint Model

The data collected for the joint model-based analyses are also often featured by multivariate longitudinal outcomes which are significantly correlated, and ignoring their correlation may lead to biased estimation (Huang et al, 2022: p.?). The bayesian approach allows for multivariate joint modelling (Andrinopoulou & Rizopoulos, 2016: p.?).

### Joint latent process model

#### The Gelman-Rubin statistic

According to Peng (2022), convergence can be evaluated by running multiple chains from different starting values. All chains are expected to eventually reach the stationary distribution and become indistinguishable. Convergence can be assessed by comparing variation between chains to variation within chains, with small between-chain variation indicating that the chains are similar.

Peng (2022) further explains that the Gelman-Rubin statistic R formalizes this idea. Consider $J$ Markov chains, each with samples $x_1^{(j)}, x_2^{(j)}, \dots$ from the $j$-th chain. For each chain, remove the first $D$ samples as “burn-in” and retain the following $L$ samples: $x_D^{(j)}, x_{D+1}^{(j)}, \dots, x_{D+L-1}^{(j)}$. One common choice is to set $D = L$.

Then calculate the chain mean for each chain,

$$
\bar{x}_j = \frac{1}{L} \sum_{t=1}^{L} x_t^{(j)}
$$

the overall mean across all chains (the grand mean),

$$
\bar{x}_\cdot = \frac{1}{J} \sum_{j=1}^{J} \bar{x}_j
$$

the between-chain variance,

$$
B = \frac{L}{J-1} \sum_{j=1}^{J} \left(\bar{x}_j - \bar{x}_\cdot\right)^2
$$

the within-chain variance for each chain,

$$
s_j^2 = \frac{1}{L-1} \sum_{t=1}^{L} \left(x_t^{(j)} - \bar{x}_j\right)^2
$$

and finally, the average within-chain variance,

$$
W = \frac{1}{J} \sum_{j=1}^{J} s_j^2
$$

Subsequently, the Gelman-Rubin statistic is:

$$
R = \frac{\frac{L-1}{L} W + \frac{1}{L} B}{W}
$$

As the number of iterations L increases and the between-chain variation B decreases, R approaches 1. Thus, chains are typically run until R is near 1, often around 1.1 or 1.2. Being a unit-free ratio, R provides a convenient summary of convergence without requiring specification of a particular parameter, making it a practical tool for monitoring MCMC chains before any inference is made (Peng, 2022).

### Latent Class Joint Models

-   Useful for modelling the relationship between longitudinal data and survival data when the population is thought to be heterogeneous

-   Disease trajectories assumed to differ between these different sub-groups

-   Not often implemented in research, as it can be difficult

-   Composed of 3 models: multinomial logistic submodel (latent class model), the linear mixed submodel (biomarker trajectory), and the survival submodel

-   Assume that the study population, $N$, is made up of $G$ homogenous, unobservable classes

-   $c_i$ is the class that individual $i, i = (1, ... , N)$ belongs to, where the multinomial logistic submodel models the probabilities for each individual of belonging to latent class $g$, given the predefined variables $X_{1,i}$. HOWEVER: no prior assumptions about the latent classes, no covariates included in the model

$$
\pi_{i, g} = P(c_i = g | X_{1,i})
$$

-   The linear submodel now becomes unique to each latent class

-   The survival submodel is defined for all latent classes

-   Assumption: conditional independence, correlation between biomarker trajectories and event occurrence are fully described by the latent class (Jacqmin-Gadda: score test for residual dependence between longitudinal data and event risk )

Latent class joint models offer a way to model the expected longitudinal biomarker measurements alongside time-to-diagnosis given that the population is made up of latent, unobserved classes that are not described by the observed covariates (Rizopoulos, 2012, pg 122).

-   Correlations between observations (biomarker measurements) are modelled by the random effects, $b_i$, in the longitudinal model

-   The relationship between the biomarker trajectory and the time-to-diagnosis is described by the latent class indicator, $c_i$

-   THESE ARE THE CONDITIONAL INDEPENDENCE ASSUMPTIONS

According to Rizopoulos, the multinomial logistic regression models the latent classes, such that:

$$
Pr(c_i = g) = \frac{e^{\mathbf{X}_i \boldsymbol{\beta}_g}}{\sum^{G}_{l = 1}e^{\mathbf{X}_i \boldsymbol{\beta}_l}}
\qquad where \qquad \overset{\mathbf{X}_i \text{ is the set of covariates for individual }i}{\boldsymbol{\beta}_l \text{ is the set of coefficients for latent group } g}
$$

If there are no prior assumptions about the latent classes, then there will be no covariates in the multinomial logistic regression model (Kyheng, 2025). This then simplifies the model to:

$$
Pr(c_i = g) = \frac{e^{\boldsymbol{\beta}_{0,g}}}{\sum^{G}_{l = 1}e^{\boldsymbol{\beta}_{0,l}}}
$$

An advantage of this formulation of the latent class joint model is that the log-likelihood is much more tractable than for the regular joint model (Rizopoulos, pg. 123) ... **MORE, WHY, SHOW**

# Data

The data used in this analysis comes from the ADNI cohort under the ADNI1 and ADNI2 protocols. Measurements of the concentration of p-tau181 and NfL in blood plasma of $n = xxx$ individuals were analysed. In accordance with the ADNI leadership guidelines, which states that there were serious data quality concerns for individuals with a PTID in the format "381_S_10###" , checks were run to assure that no observations with these PTIDs were in the data set (ADNI, 2026).

Inclusion/exclusion criteria

## ADNI

*Explain the different protocols and what they mean for data analysis, if we can compare across protocols etc.*

All protocols allow enrollment for individuals aged 55-90.

-   ADNI1

800 subjects (200 controls, 400 MCI, 200 mild AD) recruited at around 50 sites in the United States and Canada

-   ADNI2

Cognitively normal, significant memory concern, early MCI, late MCI, mild AD

Inclusion criteria for rollover participants: originally diagnosed as either MCI (early or late) or CN

-   ADNI3

700 - 800 rollover participants

370 - 1200 newly enrolled

Three cohorts: CN, MCI, mild AD

-   ADNI4

Introducing the study, it's aims and the type of information gathered.

**IMPORTANT: REMEMBER WE NEED TO REFERENCE ADNI PROPERLY.**

## Introducing variables... biomarkers

Information about the variables we will use and why, what they can show, how are they recorded

# Pre-processing

ADNI documentation: <https://adni.loni.usc.edu/quick-start-guide-asset/anatomy2.html>

Some bullet points to add to the final paper about things we thought about/did:

-   Merging the diagnosis data with biomarker/amyloid data by RID (unique identifier for patients)

-   Selecting relevant variables from diagnosis and biomarker/amyloid data set

-   What variables do we need for longitudinal sub-model? (Repeated measurement, time followed?)

-   What variables do we need for survival sub-model? (Event, time followed?)

-   Joint model: time-var: a character string indicating the time variable in the mixed-effects model

-   How many observations going from CN/MCI -\> AD? Do we need a multi-state survival model?

# Results

Present the results from our analyses. Do not interpret results here! That should be done in the discussion.

# Discussion

## Main findings/summary

Interpret the results. What did we find?

## Discussion - Relating to the research question(s)/previous literature

How do our results help to answer the research question? Anything unexpected/not in accordance with previous studies?

## Further research

Suggest what kinds of questions this analysis leaves us with? What more can be done/looked at?

# Conclusions

What we conclude based on our findings.

# Bibliography

Aguilera, N., Alcolea, D., Alegret, M., Arranz, J., Boada, M., Cano, A., Capdevila, M., de Rojas, I., Fernández, V., Fortea, J., García-González, P., García-Gutiérrez, F., Lafuente, A., Lleó, A., Marquié, M., Montrreal, L., Muñoz, N., Olivé, C., … Ruiz, A. (2024). "Clinical value of plasma pTau181 to predict Alzheimer's disease pathology in a large real-world cohort of a memory clinic". *EBioMedicine*, 108(1), 105345. DOI: <https://doi.org/10.1016/j.ebiom.2024.105345>.

Alzheimer’s Disease Neuroimaging Initiative (ADNI). (n.d.). "About ADNI". URL: <https://adni.loni.usc.edu/about/>.

Andreasson, U., Blennow, K., Cullen, N. C., Mattsson, N., & Zetterberg, H. (2019). "Association between longitudinal plasma neurofilament light and neurodegeneration in patients with Alzheimer disease". *JAMA Neurology*, 76(7), 791-799. DOI: <https://doi.org/10.1001/jamaneurol.2019.0765>.

Andrinopoulou, E. R., & Rizopoulos, D. (2016). "Bayesian shrinkage approach for a joint model of longitudinal and survival outcomes assuming different association structures". *Statistics in medicine*, 35(26), 4813-4823. DOI: <https://doi.org/10.1002/sim.7027>.

Barrett, J. K., Sweeting, M. J., Thompson, S. G. & Wood, A. M. (2017). "The use of repeated blood pressure measures for cardiovascular risk prediction: a comparison of statistical models in the ARIC study". *Statistics in medicine*, 36(28), 4514-4528. DOI: <https://doi.org/10.1002/sim.7144>.

Biomarkers Definitions Working Group. (2001). "Biomarkers and surrogate endpoints: preferred definitions and conceptual framework". *Clinical pharmacology and therapeutics*, 69(3), 89-95. DOI: <https://doi.org/10.1067/mcp.2001.113989>.

Carrillo, M. C., Dean, R. A., Nicolas, F., Miller, D. S., Berman, R., Khachaturian, Z., Bain, L. J., Schindler, R., Knopman, D., & Alzheimer's Association Research Roundtable (2013). "Revisiting the framework of the National Institute on Aging-Alzheimer's Association diagnostic criteria". *Alzheimer's & dementia : the journal of the Alzheimer's Association*, 9(5), 594-601. DOI: <https://doi.org/10.1016/j.jalz.2013.05.1762>.

Charalambous, C., Pan, J., Shen, J., & Wang, C. (2024). "Modeling biomarker variability in joint analysis of longitudinal and time-to-event data". *Biostatistics*, 25(2), 577-596. DOI: <https://doi.org/10.1093/biostatistics/kxad009>.

Chen, S. D., Dong, Q., Guo, Y., Huang, Y. Y., Shen, X. N., Tan, L. & Yu, J. T. (2021). "Longitudinal plasma phosphorylated tau 181 tracks disease progression in Alzheimer’s disease". *Translational Psychiatry*, 11(1), 356. DOI: <https://doi.org/10.1038/s41398-021-01476-7>.

Ghahremani, M., Wang, M., Chen, H.-Y., Zetterberg, H., Smith, E. and Ismail, Z. (2022). Plasma P-Tau181 and Neuropsychiatric Symptoms in Preclinical and Prodromal Alzheimer Disease. *Neurology*, p.10.1212. DOI: [10.1212/WNL.0000000000201517](https://doi.org/10.1212/WNL.0000000000201517)

Cui, Q., Guo, G., Han, H., Li, Z., Ma, Y., Meng, W., Shi, F., Song, W., Wang, A., Wang, S., Wang, Y., Yang, X. & Zhang, Z. (2026). "Sensitivity comparison of longitudinal cognitive function indicators of Alzheimer’s disease after mild cognitive impairment: a prospective cohort study". *Sci Rep*. DOI: <https://doi.org/10.1038/s41598-026-44192-2>.

Fang, Y., Li, X., Lian, S., Long, X. & Yuan, M. (2024). "Blood biomarkers in dynamic prediction of conversion to Alzheimer's disease: An application of joint modeling". *International journal of geriatric psychiatry*, 39(3), e6079. DOI: <https://doi.org/10.1002/gps.6079>.

Grambsch, P.M & Therneau, T.M. (2000). *Modeling Survival Data: Extending the Cox Model*. Springer, Berlin. DOI: <https://doi.org/10.1007/978-1-4757-3294-8>.

Huang, Y., Chen, J., Tang, N. S. & Xu, L. (2022). "Bayesian Joint Modeling of Multivariate Longitudinal and Survival Data With an Application to Diabetes Study". *Frontiers in big data*, 5, 812725. DOI: <https://doi.org/10.3389/fdata.2022.812725>.

Kyheng M, Babykina G, Duhamel A. Joint Latent Class Models: A Tutorial on Practical Applications in Clinical Research. Stat Med. 2025 Apr;44(8-9):e70047. doi: [10.1002/sim.70047](https://doi.org/10.1002/sim.70047). PMID: 40277369; PMCID: PMC12023844.

Peng. R.E. (2022). "Advanced Statistical Computing". Chapter 7.4.2: Gelman-Rubin Statistic. URL: <https://bookdown.org/rdpeng/advstatcomp/monitoring-convergence.html#gelman-rubin-statistic> \[Last updated 27 may 2022\].

Rizopoulos, D. (2012). *Joint Models for Longitudinal and Time-to-Event Data: With Applications in R* (1st ed.). Chapman and Hall/CRC. DOI: <https://doi.org/10.1201/b12208>.

Strawderman, R. L. & Wolfe, R. A. (1996). "Logical and statistical fallacies in the use of Cox regression models". *American journal of kidney diseases : the official journal of the National Kidney Foundation*, 27(1), 124-129. DOI: <https://doi.org/10.1016/s0272-6386(96)90039-6>.

**Wong, P.C., Savonenko, A., Li, T., Price, DL. (2012). 'Neurobiology of Alzheimer’s Disease' in Brady, S.T., Siegel, G.J., Albers, RW., Price, D.L (8 ed.) *Basic neurochemistry : principles of molecular, cellular, and medical neurobiology*. Waltham, Massachusetts; Oxford: Academic Press / Elsevier.**

World Health Organization (2025). "Dementia". URL: [https://www.who.int/news-room/fact-sheets/detal/dementia](https://www.who.int/news-room/fact-sheets/detail/dementia) \[Last updated 31 march 2025\].

-   (2024). "Preferred product characteristics of blood-based biomarker diagnostics for Alzheimer disease". URL: <https://iris.who.int/server/api/core/bitstreams/43c83ef7-7a34-4b56-ae17-b584f2fc0798/content>.

Xu, Lan, "Bayesian Multivariate Joint Modeling for Skewed-longitudinal and Time-to-event Data" (2021). *USF Tampa Graduate Theses and Dissertations*. URL: <https://digitalcommons.usf.edu/etd/9268>.

# Appendix

# Pre-processing

## Checking RID with more than one measurement recorded on the same exam date

-   There are 52 individuals with two measurements recorded on the same exam date.

-   How should we deal with these? Average the measurements? Remove the observations?

-   Some of the values are quite different, what do we assume by averaging them? Do we introduce bias but just just removing them?

-   For now, remove them:

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
# ptau_data %>%
#    filter(RID %in% ptau_double$RID) %>% 
#    filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

```{r}
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", diagnosis_data$PTID)
```

## We select only the required variables and merge them

```{r, warning=FALSE}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# We merge NFL and ptau data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))

# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID"))
```

## We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in long_data: all these are variables we will use in the model, so we can remove with na.omit()
sapply(long_data, function(x) sum(is.na(x)))
```

```{r}
# We count missing values per column in surv_data: VISCODE2.y will not be used in the model, so we need to see how these observations look
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# no rows with missing VISCODE2.y with full data for the model variables, so we can remove all with na.omit()
surv_data %>% filter(is.na(VISCODE2.y))
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# We check the length of each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

## We check the correlation between the biomarkers

```{r}
# Correlation between the biomarkers 
cor(long_data$PLASMAPTAU181, long_data$PLASMA_NFL, use = "complete.obs")
```

## We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We create a variable for apoe4 status
apoe$APOE4_status <- ifelse(apoe$GENOTYPE %in% c("2/4","3/4","4/4"), 1, 0)

# We merge gender with merged_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We make sure that gender is a factor 
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)

# We merge apoe with merged_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status")], by="RID")
surv_data <- merge(surv_data, apoe[, c("RID", "APOE4_status")], by="RID")

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

## We filter participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements of ptau_data
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
long_data <- long_data %>% filter(RID %in% keep)
surv_data <- surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

## We extract baseline visits, and create an event variable and a time-to-event variable

```{r}
# checking which observations have the same examdate for nfl and ptau
common = which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# checking the observations with different exam dates
long_data[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
merged_data <- long_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
  ungroup()

# checking which observations have the same examdate for nfl and ptau
common = which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# checking the observations with different exam dates (9 obs)
long_data[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
long_data <- long_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
ungroup()
```

```{r}
# We calculate baseline for each biomarker per participant
surv_data <- surv_data %>%
  group_by(RID) %>%
  mutate(
    # We calculate baseline for pTau using the first non-missing measurement
    baseline_pTau = ifelse(all(is.na(PLASMAPTAU181)), NA, min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)])),
    
    # We calculate baseline for NFL using the first non-missing measurement
    baseline_NFL  = ifelse(all(is.na(PLASMA_NFL)), NA, min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])),
    
    # We define a common baseline date as the earliest biomarker measurement
    baseline_date = pmin(baseline_pTau, baseline_NFL, na.rm = TRUE),
    
    # Age at baseline (age at the participant's baseline visit)
    age_at_baseline = age_at_visit[EXAMDATE_ptau == baseline_date | EXAMDATE_nfl == baseline_date][1],
    
    # We indicate whether the participant ever developed dementia
    event = as.integer(any(DIAGNOSIS_dx == "Dementia")),
    
    # We calculate time-to-event from the baseline date
    time_to_event = case_when(
      event == 1 ~ as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - baseline_date)/365,
      TRUE      ~ as.numeric(max(EXAMDATE_dx) - baseline_date)/365
    )
    
  ) %>%
  ungroup()

# We only keep one baseline row per participant
surv_data <- surv_data %>%
  group_by(RID) %>%
  # We keep rows where either pTau or NFL measurement matches the baseline
  filter(EXAMDATE_ptau == baseline_date | EXAMDATE_nfl == baseline_date) %>%
  slice(1) %>%  # We take the first row if multiple rows match
  ungroup()

# We count how many participants experienced the event
table(surv_data$event)

# Example
surv_data %>%
    filter(RID == 61)
```

## We create a time variable (time from baseline)

```{r}
long_data <- long_data %>%
  group_by(RID) %>%
  # We create baseline for PTau
  mutate(
    baseline_date_ptau = min(EXAMDATE_ptau, na.rm = TRUE),
    time_years_ptau = as.numeric(EXAMDATE_ptau - baseline_date_ptau) / 365.25
  ) %>%
  # We create baseline for NFL
  mutate(
    baseline_date_nfl = min(EXAMDATE_nfl, na.rm = TRUE),
    time_years_nfl = as.numeric(EXAMDATE_nfl - baseline_date_nfl) / 365.25
  ) %>%
  ungroup()
```

## We check the distribution of each relevant variable for multivariate joint model and transform

```{r}
# We check the distribution of each biomarker
hist(long_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(long_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(long_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(long_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

```{r}
# cube-root transformed
ggplot(long_data, aes(sqrt(sqrt(PLASMA_NFL)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()

# cube-root transformed
ggplot(long_data, aes(sqrt(sqrt(PLASMAPTAU181)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()
```

## We check for outliers

We still keep them, as we assume they are correct observations.

-   Checking for observations either side of the mean of the standardised data shows there are no values 3 sds below the mean, only above.

```{r}
# Scaling before 
long_data$ptau_z <- scale(long_data$PLASMAPTAU181)
long_data$nfl_z <- scale(long_data$PLASMA_NFL)

# We identify observations that are 3 standard deviations above or below the mean (z > 3 or < -3)
large_outliers_ptau <- long_data[long_data$ptau_z > 3, ]
small_outliers_ptau <- long_data[long_data$ptau_z < -3, ]
large_outliers_ptau
small_outliers_ptau

large_outliers_nfl <- long_data[long_data$nfl_z > 3, ]
small_outliers_nfl <- long_data[long_data$nfl_z < -3, ]
large_outliers_nfl
small_outliers_nfl
```

## We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
# We re-count missing values after pre-processing
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# We check the length of each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

# Longitudinal model

## Comparing longitudinal models

Suggested models:

-   Gamma: raw data could be modelled well with the gamma distribution since takes no negative values, variance increases with mean, and there is a right-skew to the raw distributions

-   Linear mixed effects model with log-transformed data: approximately normal, yet still some skew

-   Linear mixed effects model with cube root transformed data: again, approximately normal, still a bit of skew (why such a low AIC? seems wrong?)

-   GLMM with log-normal distribution

-   GLMM with Student-t distribution: longer tails

### p-tau181

```{r}
# checking the mean and variance for ptau181
mean(long_data$PLASMAPTAU181)
var(long_data$PLASMAPTAU181)
```

```{r}
# GLMM
fit_gamma_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years_ptau + (1|RID), 
                   data = long_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_ptau <- lmer(log(PLASMAPTAU181) ~ time_years_ptau + (1|RID), 
                data = long_data)

# LME approach with log transformed data
fit_lme_cubed_ptau <- lmer(sqrt(sqrt(PLASMAPTAU181)) ~ time_years_ptau + (1|RID), 
                           data = long_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years_ptau + (1|RID),
                              data = long_data,
                              family = lognormal(link = "log"))

# Student-t
fit_student_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years_ptau + (1|RID),
                            data = long_data,
                            family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_ptau <- simulateResiduals(fit_gamma_ptau)
testResiduals(gamma_residuals_ptau)

lme_residuals_log_ptau <- simulateResiduals(fit_lme_log_ptau)
testResiduals(lme_residuals_log_ptau)

lme_residuals_cubed_ptau <- simulateResiduals(fit_lme_cubed_ptau)
testResiduals(lme_residuals_cubed_ptau)

lognormal_residuals_ptau <- simulateResiduals(fit_lognormal_ptau)
testResiduals(lognormal_residuals_ptau)

student_residuals_ptau <- simulateResiduals(fit_student_ptau)
testResiduals(student_residuals_ptau)

AIC(fit_lme_log_ptau, fit_lme_cubed_ptau, fit_gamma_ptau, fit_lognormal_ptau, fit_student_ptau)
```

### NfL

-   Negative AIC for cube root transformed -\> take the absolute value

```{r}
# checking the mean and variance for nfl
mean(merged_data$PLASMA_NFL)
var(merged_data$PLASMA_NFL)
```

```{r}
# GLMM
fit_gamma_nfl <- glmmTMB(PLASMA_NFL ~ time_years_nfl + (1|RID), 
                   data = long_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_nfl <- lmer(log(PLASMA_NFL) ~ time_years_nfl + (1|RID), 
                data = long_data)

# LME approach with log transformed data
fit_lme_cubed_nfl <- lmer(sqrt(sqrt(PLASMA_NFL)) ~ time_years_nfl + (1|RID), 
                          data = long_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_nfl <- glmmTMB(PLASMA_NFL ~ time_years_nfl + (1|RID),
                             data = long_data,
                             family = lognormal(link = "log"))

# Student-t
fit_student_nfl <- glmmTMB(PLASMA_NFL ~ time_years_nfl + (1|RID),
                           data = long_data,
                           family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_nfl <- simulateResiduals(fit_gamma_nfl)
testResiduals(gamma_residuals_nfl)

lme_residuals_log_nfl <- simulateResiduals(fit_lme_log_nfl)
testResiduals(lme_residuals_log_nfl)

lme_residuals_cubed_nfl <- simulateResiduals(fit_lme_cubed_nfl)
testResiduals(lme_residuals_cubed_nfl)

lognormal_residuals_nfl <- simulateResiduals(fit_lognormal_nfl)
testResiduals(lognormal_residuals_nfl)

student_residuals_nfl <- simulateResiduals(fit_student_nfl)
testResiduals(student_residuals_nfl)

AIC(fit_lme_log_nfl, fit_lme_cubed_nfl, fit_gamma_nfl, fit_lognormal_nfl, fit_student_nfl)
```

## Fitted LME model for P-TAU

```{r}
long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years_ptau + PTGENDER + APOE4_status,
  random = ~ time_years_ptau | RID,                    
  data = long_data
)

summary(long_model_ptau)
```

```{r}
# Longitudinal model with t-distribution (identity link) 
library(GLMMadaptive)
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years_ptau + PTGENDER + APOE4_status,
  random = ~ time_years_ptau | RID,                    
  data = long_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

## Fitted LME model for NFL

```{r}
# Longitudinal model, with chosen covariates 
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years_nfl + PTGENDER + APOE4_status, 
  random = ~ time_years_nfl | RID,   # random intercept + slope
  data = long_data
  )

summary(long_model_nfl)
```

## Removing gender from NFL LME model

We see that gender is not significant in the LME model for NFL, so we remove it:

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years_nfl + APOE4_status, 
  random = ~ time_years_nfl | RID, # random intercept + slope
  data = long_data
  )

summary(long_model_nfl)
```

## Removing APOE4 status from ptau LME model

Even though APOE4 status was significant in the LME model for ptau, including it in both the LME model and the survival model led to poor MCMC convergence (high Rhat) in the joint model, probably due to difficulties in separating its direct effect on survival from its indirect effect through the biomarker. Therefore, to ensure stable estimation, we keep the APOE4 status only in the survival model, removing it from the LME modell:

```{r}
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years_ptau + PTGENDER,
  random = ~ time_years_ptau | RID,                    
  data = long_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

## Residuals

```{r}
# residuals 
res_ptau <- residuals(long_model_ptau, type = "subject_specific")

# fitted values
fitted_vals_ptau <- fitted(long_model_ptau, type = "subject_specific")

# plot
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for P-TAU")
abline(h = 0, col = "red", lty = 2)
```

```{r}
# residuals
res_nfl <- residuals(long_model_nfl, type = "pearson")  

# fitted values
fitted_vals_nfl <- fitted(long_model_nfl)  

# plot
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NFL")
abline(h = 0, col = "red", lty = 2)
```

## QQ plot

```{r}
# For P-tau
qqnorm(res_ptau)
qqline(res_ptau, col = "red")

# For NFL 
qqnorm(res_nfl)
qqline(res_nfl, col = "red")
```

# Survival model

## We create a survival object

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

## Kaplan-Meier curve

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

## Fitted Coxph model

```{r}
# Cox survival-model, with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

## Removing gender

Gender is not significant in the survival model, so we remove it:

```{r}
# Cox survival-model, without gender 
surv_model <- coxph(
  surv_object ~ age_at_baseline + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

## Testing for the assumption of proportional hazards

```{r}
cox.zph(surv_model)
```

# Joint-model

## Fitted joint model for P-tau

```{r}
# Joint model 
# joint_model <- jm(
#   surv_model,
#   long_model_ptau,
#   time_var = "time_years",
#   n_chains = 3L, 
#   n_iter = 11000L, 
#   n_burnin = 1000L
#   )
# summary(joint_model)
```

## Fitted joint model for NFL

```{r}
# Joint model 
# joint_model_nfl <- jm(
#   surv_model,
#   long_model_nfl,
#   time_var = "time_years",
#   n_chains = 3L, 
#   n_iter = 11000L, 
#   n_burnin = 1000L
#   )
# 
# summary(joint_model_nfl)
```

# Multivariate joint model

## We create a mixed object

```{r}
Mixed <- list(long_model_ptau, long_model_nfl)
```

## We check wether the models have stable random slopes

```{r}
summary(long_model_ptau)
VarCorr(long_model_nfl)

```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

## fForms

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )      
```

## Fitted multivariate joint model

```{r}
# multi_joint_model <- jm(surv_model, Mixed,
#                   time_var = "time_years",
#                   functional_forms = fForms,
#                   n_chains = 3L,    
#                   n_iter = 20000L,
#                   n_burnin = 2000L
#                   )
# summary(multi_joint_model)
```

```{r}
# summary <- summary(multi_joint_model)
# summary$Survival
```
