---
format:
  pdf:
    include-before-body:
      text: |
        \begin{titlepage}
        \raggedleft
        \vspace*{2cm}
        
        {\Huge\bfseries Kandidatuppsats\par}
        \vspace{1.5cm}
        
        {\Large Statistiska Instiutionen\par}
        \vspace{3cm}
        
        {\Huge\bfseries Nr 2026:x\par}
        \vspace{0.5cm}
        
        {\LARGE \bfseries Title in English\par}

        {\large Title in Swedish\par}
        \vspace{0.5cm}
        
        {\large \today\par}
        \vspace{3cm}
        
        {\Huge Charlotte Emily Iszatt and Milou Rossi\par}
        
        \vfill
        
        \end{titlepage}
        
        \begin{titlepage}
        \raggedleft
        
        {\normalsize Självständigt arbete 15 högskolepoäng inom Statistik III, VT2026\par}
        {\normalsize Handledare: Hans Nyqvist\par}
        \vspace{2cm}
        
        \raggedright
        {\huge\bfseries Abstract\par}
        \vspace{1cm}
        
        {\Large The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. Keywords: XXX, XXX, XXX\par}
        \vspace{3cm}
        
        \vfill
        
        \end{titlepage}
    toc: false
    colorlinks: false
    number-sections: true
editor: visual
---

```{=latex}
\tableofcontents

\newpage
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, output = TRUE)
```

```{r, warning=FALSE}
# loading libraries
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggplot2)
```

```{r}
# Loading data
# Milou
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
#
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
#
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
#
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
#
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

```{r}
# Charley
library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

apoe = APOERES
```

# Introduction

## Background

According to World Health Organization (2025), Alzheimer's disease (AD) is the most common neurodegenerative disorder and a leading cause of dementia around the world. First described by Dr. Alois Alzheimer in 1901, AD continues to present significant challenges for the medical field, and no definitive cure exists to date (WHO, 2025).

The 2011 National Institute on Aging-Alzheimer’s Association (NIA-AA) diagnostic guidelines describe the disease as a continuum with three stages: a symptom-free stage (preclinical AD), a predementia stage (mild cognitive impairment (MCI) due to AD) and a dementia stage (dementia due to AD) (Carrillo et al. 2013: p. 596). Pathophysiological changes in AD can begin up to 20 years before clinical symptoms appear, highlighting the crucial importance of early diagnosis and timely intervention to potentially slow the progression of the disease (Carrillo et al. 2013: p. 595).

Historically, the detection of AD pathophysiology was done by positron emission topography (PET) to detect the presence of amyloid plaques in the brain, or by analysis of cerbrospinal fluid (CSF) to quantify the concentration of biological molecules implicated in AD pathology. Recent advances in research have made it possible to replace these invasive procedures with blood tests.

Från: <https://journals.sagepub.com/doi/10.1177/25424823261432112> –\> kanske hitta bättre, mer konkreta källor för infon, men bra info?

Tau proteins, which are microtubule-binding proteins, play an essential role in maintaining neuronal stability and morphology under normal conditions. The regional progression of brain atrophy in AD is closely associated with the accumulation of tau proteins. Hyperphosphorylated tau (p-tau), in its full-length or truncated forms, represents a key pathological hallmark of AD, exhibiting over 70 post-translational modifications, including more than 40 phosphorylation sites. Various forms of p-tau can be measured in cerebrospinal fluid (CSF) and plasma. According to the NIA-AA, there are several core biomarkers to be used in the diagnosis and staging of AD (Ghahremani et al., 2022). One such core biomarker is tau protein phosphorylated at the threonine-181 position (p-tau181), where concentrations are expected to be higher in individuals with AD pathology (Wong et al., 2012). **MORE ON p-tau181 here...**

-   Among these, p-tau181 is significantly elevated in the CSF of AD patients and is considered a critical biomarker for the disease. The ability of p-tau181 to differentiate AD from other neurodegenerative conditions, such as frontotemporal lobar degeneration and chronic traumatic encephalopathy, is crucial for the accurate diagnosis of AD. Furthermore, p-tau181 levels correlate with AD severity and predict disease progression, providing valuable insights for evaluating therapeutic efficacy and guiding treatment strategies. Recent findings indicate that p-tau181 have excellent discriminative performance in plasma, paving the way for non-invasive monitoring of AD.

Neurofilament light chain (NfL) is instead a marker of neuronal injury not specific to AD (Ghahremani et al.).

The abnormal concentrations of these biomarkers are the result of the neuropathological changes symptomatic of AD.

*General background information. Introducing some domain specific information about AD diagnosis and pathology, so that the reader is well prepared.*

## Research Context

*Assessing the current landscape in terms of the scientific literature. What is known? What has been done Where are there gaps??? (This leads on to the next section, why are we doing this research, how do we hope to answer some remaining question in the AD research, orient this towards the statistical analysis/approaches!)*

## Research Aims

Förslag –\>

This study evaluates the prediction of Alzheimer’s disease/dementia progression using longitudinal biomarker data, comparing conventional and advanced statistical approaches. More precisely, the research questions are:

1.  Can the initial (baseline) levels of p-tau181 and NfL predict the progression to Alzheimer’s disease/dementia?
2.  Do the average biomarker levels over time improve prediction compared to baseline alone?
3.  Does using the biomarkers’ changing levels over time give a better prediction of disease progression?
4.  Using a joint model, can modeling the full trajectory of each biomarker improve prediction accuracy?
5.  Can a multivariate joint model combining both biomarkers provide even better prediction of progression?

What are our research questions?

1.  How do longitudinal changes in plasma biomarker levels over time relate to the risk of progression to dementia? (Joint modeling)

2.  How do these associations differ between p-tau181 and NfL? (Comparison across biomarkers)

3.  How does the precision change for the estimated longitudinal trajectories and survival associations to the inclusion of individuals with different numbers of repeated measurements? (Precision analysis)

Sensitivity → how do the estimates of the the joint model parameters change when excluding sparce individuals with only one biomarker measurement?

Why is this research relevant to the existing body of literature? What does it add? Gaps in previous literature which we outline in above \^\^.

How will we accomplish this?

# Statistical Theory and Methods

Introduce the methods we will use.

Why did we choose this method to answer the research question?

## Longitudinal Data

Longitudinal data in this analysis is defined as $n$ repeated measurements of p-tau181 and NfL on $N$ individuals in the ADNI1, ADNIGO and ADNI2 cohorts. $T_{i, j}$ and $N_{i, j}$ denote the measured concentrations of p-tau181 and NfL respectively for the $i^{th}$ individual at the $j^{th}$ occasion. Assuming that the expected concentration of the biomarkers varies with both time and individual level covariates, the mean responses have the notation $\mu_{T:\text{ } i, j} = E(T_{i, j})$ and $\mu_{N:\text{ } i, j} = E(N_{i, j})$. Considering the data are made up of repeated responses on the same individuals, the assumption of independence between observations is illogical. Clearly, the baseline levels of p-tau181 and NfL for an individual are correlated with the measurements on subsequent occasions.

## Mixed Effects Models

Why???

We have grouped data, specifically longitudinal data where groups of observations are correlated, since we have repeated measurements on the same individuals. We expect data *within* the groups to be more correlated than those *between* the groups. Mixed effects models are not only for longitudinal data, but for any data where there is non-independence between observations.

### Linear Mixed Effect Model

**Theory about the model... (specifically why we should use it)**

-   Biomarker process

-   Describes the trajectory of biomarkers at the individual level 

**Linear mixed effects models?** What are these - why use them here?

$$
y_i(t) = m_i(t) + \epsilon_i(t) \qquad \epsilon_i(t) \sim N(0, \sigma^2)
$$

$$
m_i(t) = \mathbf{X_i^T}\beta + \mathbf{Z_i}^T\mathbf{b_i} \qquad \mathbf{b_i}(t) \sim N(0, \sigma^2)
$$

$y_i(t):$ the observed longitudinal biomarker

$m_i(t):$ the trajectory function which represents the **true unobserved** **value of the biomarker** at time $t$ for patient $i$ - we assume that we do not observe the true value due to measurement errors or other reasons.

$\mathbf{X_i}^T(t)\beta:$ the fixed effects

$\mathbf{Z_i}^T\mathbf{b_i}:$the random effects

The full model becomes:

$$
y_i(t) = \mathbf{X_i^T}\beta + \mathbf{Z_i}^T\mathbf{b_i} + \epsilon_i(t) \qquad \mathbf{b_i}(t) \sim N(0, \sigma^2), \quad \epsilon_i(t) \sim N(0, \sigma^2)
$$

### Generalised Linear Mixed Effects Model

Ref: <https://stats.oarc.ucla.edu/other/mult-pkg/introduction-to-generalized-linear-mixed-models/>

An extension of GLMs, where the linear predictor contains random effects alongside the fixed effects. This is used to extend the linear mixed model to non-normal data.

Given random effects, $\mathbf{u}$ the dependent variable, $\mathbf{y}$, is distributed according to the exponential family, such that $g(E[\mathbf{y}|\mathbf{u}]) = \mathbf{X𝛽} + \mathbf{Zu}$. Some link function, $g(x)$, links the expected value of $\mathbf{y}$ given the random effects to the linear predictor which included both fixed effects and random effects.

Where:

$\mathbf{X}$ = the fixed effects design matrix

$\mathbf{\beta}$ = the fixed effects

$\mathbf{Z}$ = the random effects design matrix

$\mathbf{u}$ = the random effects

What do we model as random?

We can model random intercept and keep all other effects fixed. In that case, for $i$ biomarker measurements on $j$ individuals, the random effects design matrix - $\mathbf{Z}$ - is a $i \cdot j$ matrix containing 1's and 0's, where $z_{i,j} = 1$ if the $i^{th}$ biomarker measurements were taken from the $j^{th}$ individuals, and $z_{i,j} = 0$ if the $i^{th}$ biomarker measurements were not taken from the $j^{th}$ individual.

We do not tend to estimate $\mathbf{u}$, but instead assume that $u_i \sim N(0, \mathbf{G})$, where $\mathbf{G}$ is the variance-covariance matrix of the random effects. The fixed effects are estimated, and the random effects are seen as deviations from the fixed effects and have $E(u) = 0$.

For a model with only a random intercept, $\mathbf{G}$ becomes $G = \sigma^2_{intercept}$. If we extend the model to have random slope, we get:

$$
G =
\begin{bmatrix}
\sigma^2_{intercept} & \sigma_{intercept, slope} \\
\sigma_{intercept, slope} & \sigma^2_{slope}
\end{bmatrix}
$$

Like all variance-covariance matrices, some elements are redundant since they are repeated. Estimating instead some function of $\mathbf{G}$, where $\mathbf{G} = \sigma(\theta)$, means estimating $\frac{p(p + 1)}{2}$ elements instead. Assuming (**?????**) the random effects are independent, the variance-covariance matrix for the random effects becomes diagonal:

$$
G =
\begin{bmatrix}
\sigma^2_{intercept} & 0 \\
0 & \sigma^2_{slope}
\end{bmatrix}
$$

### Assumptions!!

-   Random effects $\mathbf{b_i}$ are **time-invariant** - they are assumed to come from a normal distribution, $\mathbf{b_i} \sim N(0, \sigma^2)$, and represent stable, latent characteristics about the individuals disease, such as a generally fast declining disease trajectory but cannot represent other factors that effect disease trajectory over-time, e.g. medication. Any external time-varying process that effects the biomarker trajectory and the hazard cannot be represented by the random effects.

## Proportional Hazard Model

**Theory about the model... (specifically why we should use it)**

-   Time-to-diagnosis process

-   Parametric baseline hazard: **flexible parametrics, splines??**

-   OR leave undefined -\> Cox PH model

Assuming a Cox PH survival sub-model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha m_i(t) \right\}
$$

$\mathbf{v_i}:$ vector of time-independent co-variates

$\phi:$ vector of log-hazard ratios

$\alpha:$ the association parameter

$M_i = \left\{m_i, \quad 0 \le s \le t\right\}:$ the **true unobserved biomarker profile** up to time $t$

### Assumptions!!

-   What do we assume about the censoring process?

-   Proportionality

### Cox PH model using baseline (initial) values of each biomarker

### Cox PH model using time-averaged values of each biomarker?

### Extended Cox PH model using time-depending covariates?

In the proportional hazards model, the hazard is assumed to depend solely on covariates with values that remain constant throughout the follow-up period, such as baseline age, sex, and randomized treatment. However, in longitudinal research, it is also of interest to assess whether time-dependent covariates influence the risk of an event. Examples are environmental factors, biochemical or clinical parameters measured during follow-up, and changes in treatment dosage (Rizopoulos, 2012: p. 44).

Time-dependent covariates can be classified into two categories: exogenous (external) and endogenous (internal). Exogenous covariates are those whose future values are not influenced by the occurrence of an event and are predictable at any time. Examples include time of day or season, or treatment strategies adjusted according to pre-specified rules. Endogenous covariates, in contrast, are affected by or associated with the event process and are generally unpredictable. These typically are time-varying measurements taken on the subject, such as biomarkers or clinical parameters (Rizopoulos, 2012: p.45)

The Cox model can be extended to handle exogenous time-dependent covariates using the so called counting process framework. In this formulation, event occurrences are treated as realizations of a slow Poisson process. For subject $i$, the event process is denoted by ${N_i(t), R_i(t)}$, where $N_i(t)$ is the number of events by time $t$ and $R_i(t)$ is a left-continuous at-risk indicator ($R_i(t) = 1$ if the subject is at risk at time $t$, $0$ otherwise). The extended Cox model is written as:

$$h_i(t \mid Y_i(t), w_i) = h_0(t) \, R_i(t) \, \exp\{\gamma^\top w_i + \alpha \, y_i(t)\}$$

where $w_i$ is a vector of baseline covariates (e.g., sex or randomized treatment) and $y_i(t)$ is a vector of time-dependent covariates. The coefficients $\alpha$ have the same interpretation as $\gamma$: for a single time-dependent covariate, $\exp(\alpha)$ represents the relative increase in the hazard at time $t$ associated with a one-unit increase in $y_i(t)$ (Rizopoulos, 2012: p. 47).

Han skriver:

-   "The only difference compared to the syntax used to fit the standard Cox model in Section 3.3, is in the specification of the Surv() function in which the first two arguments denote the limits of the interval, and the last argument is the corresponding event indicator. Moreover, in the data argument the data should be supplied in the long format (s.48)"

MEN sen skriver han också:

-   "However, despite this flexibility the extended Cox model is not appropriate when the time-dependent covariates are of endogenous nature (t.ex. biomarkörer). This is because the extended Cox model assumes that time-dependent covariates are predictable processes, measured without error, and have their complete path fully specified (s. 48). "

-   **Men den artikel jag visade gjorde det ändå, fastän de också kollade på biomarkörer. Om man gör det ändå visar man hur det inte är anpassat för vår typ av data, och får resultaten av de andra modeller man jämför med att se ännu bättre ut.**

## Joint Models

### Simple Joint Model

**Theory about the model... (specifically why we should use it)**

In the joint model, the hazard function at time $t$ is directly dependent on the estimate of the biomarker at time $t$.

The association structure of the joint model:

-   How is the longitudinal model linked to the survival model? 

-   Does this association structure change for different biomarkers?

-   What happens to time-to-diagnosis predictions if we assume the standard structure when another underlying association is a better fit?

If we include the current rate of change alongside the current value in the proportional-hazard model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha_1 m_i(t) + \alpha_2 m_i'(t) \right\} \qquad where \qquad m_i'(t) = \frac{d}{dt} m_i(t)
$$

$\alpha_1 :$ links the current value of the biomarker - $m_i(t)$

$\alpha_2 :$ links the current gradient/rate of change - $m_i'(t)$

Joint-model options:

-   Longitudinal sub-model: model of class lme()

### Assumptions!!

-   Full conditional independence

    -   Random effects $\mathbf{b_i}$ explain all statistical dependence between (1) the longitudinal outcome and the survival outcome and (2) repeated measurements within the longitudinal process.

    -   This assumption can fail when: there are time-varying confounders not captured by the random effects (see the longitudinal model for more info about $\mathbf{b_i}$ assumptions), if the measurement error, $e_i(t)$ in the longitudinal model $y_i(t) = m_i(t) + e_i(t)$ is informative, or there are sub-populations (APOE genotype??? Maybe we need to split into populations based on genotype)

-   So: the longitudinal outcome is independent of the time-to-event outcome and the repeated measurements in the longitudinal outcome are independent of each other

-   $\mathbf{b_i}$ drives the **true (unobserved)** biomarker trajectory and hazard of AD diagnosis: the observed $y_i(t)$ is the observed biomarker state, $m_i(t)$, plus noise. Where $m_i(t)$ is itself determined by $\mathbf{b_i}$, \*\*since the others are fixed ????? \*\*, once we know $\mathbf{b_i}$, the observed biomarker measurement and the survival times are no longer dependent. The observed $y_i$'s are useful in determining $\mathbf{b_i}$ which is the useful information in estimating survival times.

### Frequentist vs Bayesian approach

However, as the joint model includes a linear mixed-effects model with the normal assumption for a single longitudinal outcome, joint model-based analyses may not provide robust inference when longitudinal measurements exhibit skewness and/or heavy tails (Huang et al, 2022: p.?). Consequently, joint models, besides with the traditional approach of maximum likelihood estimation, can also be estimated within a Bayesian framework using MCMC techniques (Rizopoulos, 2012, p. 62). This approach allows for more flexible and complex association structures (Andrinopoulou & Rizopoulos, 2016: p.?).

### Multivariate Joint Model

The data collected for the joint model-based analyses are also often featured by multivariate longitudinal outcomes which are significantly correlated, and ignoring their correlation may lead to biased estimation (Huang et al, 2022: p.?). The bayesian approach allows for multivariate joint modelling (Andrinopoulou & Rizopoulos, 2016: p.?).

#### The Gelman-Rubin statistic

According to Peng (2022), convergence can be evaluated by running multiple chains from different starting values. All chains are expected to eventually reach the stationary distribution and become indistinguishable. Convergence can be assessed by comparing variation between chains to variation within chains, with small between-chain variation indicating that the chains are similar.

Peng (2022) further explains that the Gelman-Rubin statistic R formalizes this idea. As the number of iterations L increases and the between-chain variation B decreases, R approaches 1. Chains are typically run until R is near 1, often around 1.1 or 1.2. Being a unit-free ratio, R provides a convenient summary of convergence without requiring specification of a particular parameter, making it a practical tool for monitoring MCMC chains before any inference is made (Peng, 2022).

# Data

The data used in this analysis comes from the ADNI cohort under the ADNI1 and ADNI2 protocols. Measurements of the concentration of p-tau181 and NfL in blood plasma of $n = xxx$ individuals were analysed. In accordance with the ADNI leadership guidelines, which states that there were serious data quality concerns for individuals with a PTID in the format "381_S_10###" , checks were run to assure that no observations with these PTIDs were in the data set (ADNI, 2026).

## ADNI

*Explain the different protocols and what they mean for data analysis, if we can compare across protocols etc.*

All protocols allow enrollment for individuals aged 55-90.

-   ADNI1

800 subjects (200 controls, 400 MCI, 200 mild AD) recruited at around 50 sites in the United States and Canada

-   ADNI2

Cognitively normal, significant memory concern, early MCI, late MCI, mild AD

Inclusion criteria for rollover participants: originally diagnosed as either MCI (early or late) or CN

-   ADNI3

700 - 800 rollover participants

370 - 1200 newly enrolled

Three cohorts: CN, MCI, mild AD

-   ADNI4

Introducing the study, it's aims and the type of information gathered.

**IMPORTANT: REMEMBER WE NEED TO REFERENCE ADNI PROPERLY.**

## Introducing variables... biomarkers

Information about the variables we will use and why, what they can show, how are they recorded

# Pre-processing

ADNI documentation: <https://adni.loni.usc.edu/quick-start-guide-asset/anatomy2.html>

Some bullet points to add to the final paper about things we thought about/did:

-   Merging the diagnosis data with biomarker/amyloid data by RID (unique identifier for patients)

-   Selecting relevant variables from diagnosis and biomarker/amyloid data set

-   What variables do we need for longitudinal sub-model? (Repeated measurement, time followed?)

-   What variables do we need for survival sub-model? (Event, time followed?)

-   Joint model: time-var: a character string indicating the time variable in the mixed-effects model

-   How many observations going from CN/MCI -\> AD? Do we need a multi-state survival model?

# Results

Present the results from our analyses. Do not interpret results here! That should be done in the discussion.

# Discussion

## Main findings/summary

Interpret the results. What did we find?

## Discussion - Relating to the research question(s)/previous literature

How do our results help to answer the research question? Anything unexpected/not in accordance with previous studies?

## Further research

Suggest what kinds of questions this analysis leaves us with? What more can be done/looked at?

# Conclusions

What we conclude based on our findings.

# Bibliography

Andrinopoulou, E. R., & Rizopoulos, D. (2016). Bayesian shrinkage approach for a joint model of longitudinal and survival outcomes assuming different association structures. Statistics in medicine, 35(26), 4813-4823. <https://doi.org/10.1002/sim.7027>

Carrillo, M. C., Dean, R. A., Nicolas, F., Miller, D. S., Berman, R., Khachaturian, Z., Bain, L. J., Schindler, R., Knopman, D., & Alzheimer's Association Research Roundtable (2013). Revisiting the framework of the National Institute on Aging-Alzheimer's Association diagnostic criteria. *Alzheimer's & dementia : the journal of the Alzheimer's Association*, 9(5), 594-601. <https://doi.org/10.1016/j.jalz.2013.05.1762>

Ghahremani M, Wang M, Chen HY, Zetterberg H, Smith E, Ismail Z; Alzheimer's Disease Neuroimaging Initiative\*. Plasma Phosphorylated Tau at Threonine 181 and Neuropsychiatric Symptoms in Preclinical and Prodromal Alzheimer Disease. *Neurology*. 2023 Feb 14;100(7):e683-e693. doi: 10.1212/WNL.0000000000201517

Huang, Y., Chen, J., Xu, L., & Tang, N. S. (2022). Bayesian Joint Modeling of Multivariate Longitudinal and Survival Data With an Application to Diabetes Study. *Frontiers in big data*, 5, 812725. <https://doi.org/10.3389/fdata.2022.812725>

Peng. R.E. (2022). *Advanced Statistical Computing*. Chapter 7.4: Monitoring Convergence. URL: <https://bookdown.org/rdpeng/advstatcomp/monitoring-convergence.html> \[Last updated 27 may 2022\].

Wong, P.C., Savonenko, A., Li, T., Price, DL. (2012). 'Neurobiology of Alzheimer’s Disease' in Brady, S.T., Siegel, G.J., Albers, RW., Price, D.L (8 ed.) *Basic neurochemistry : principles of molecular, cellular, and medical neurobiology*. Waltham, Massachusetts; Oxford: Academic Press / Elsevier.

World Health Organization (2025). *Dementia*. URL: [https://www.who.int/news-room/fact-sheets/detal/dementia](https://www.who.int/news-room/fact-sheets/detail/dementia) \[Last updated 31 march 2025\].

# Pre-processing

## Checking RID with more than one measurement recorded on the same exam date

-   There are 52 individuals with two measurements recorded on the same exam date.

-   How should we deal with these? Average the measurements? Remove the observations?

-   Some of the values are quite different, what do we assume by averaging them? Do we introduce bias but just just removing them?

-   For now, remove them:

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
# ptau_data %>%
#   filter(RID %in% ptau_double$RID) %>% 
#   filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

```{r}
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", diagnosis_data$PTID)
```

## We select only the required variables and merge them

```{r, warning=FALSE}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# We merge NFL, ptau och diagnosis
merged_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID", "VISCODE2"))
```

## We check the correlation between the biomarkers

```{r}
# Correlation between the biomarkers 
cor(merged_data$PLASMAPTAU181, merged_data$PLASMA_NFL, use = "complete.obs")
```

## We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We merge with merged_data
merged_data <- merged_data %>%
  left_join(demographics_small, by = "RID")

# We make sure that gender is a factor 
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# We create a variable for apoe4 status
apoe$APOE4_status <- ifelse(apoe$GENOTYPE %in% c("2/4","3/4","4/4"), 1, 0)

# We merge apoe with merged_data 
merged_data <- merge(merged_data, apoe[, c("RID", "APOE4_status")], by="RID")
```

## We filter participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

## We extract baseline visits, and create an event variable and a time-to-event variable

```{r}
# checking which observations have the same examdate for nfl and ptau
common = which(merged_data$EXAMDATE_nfl %in% merged_data$EXAMDATE_ptau)

# checking the observations with different exam dates
merged_data[-common, ]

# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
merged_data <- merged_data %>%
group_by(RID) %>%
mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
ungroup()
```

```{r}
# We create a baseline-date
# merged_data <- merged_data %>%
#   group_by(RID) %>%
#   mutate(baseline_date = min(EXAMDATE_ptau)) %>%
#   ungroup()

# We create event and time_to_event (where dementia is the event) and add age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
  summarise(
    age_at_baseline = as.numeric(min(baseline_date), "%Y") - min(PTDOBYY),
    PTGENDER = PTGENDER[which.min(EXAMDATE_ptau)],
    APOE4_status = APOE4_status[which.min(EXAMDATE_ptau)],
    
    time_to_event = case_when(

    any(DIAGNOSIS_dx == "Dementia") ~
      as.numeric(min(EXAMDATE_ptau[DIAGNOSIS_dx == "Dementia"]) -
                   min(c(EXAMDATE_ptau, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_ptau, EXAMDATE_dx)) -
                    min(c(EXAMDATE_ptau, EXAMDATE_dx)))) / 365,

  event = as.integer(any(DIAGNOSIS_dx == "Dementia"))
) %>% ungroup()

# We put the survival data back into the merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## We create a time variable (time from baseline)

```{r}
# We create time_years (time from baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_ptau - baseline_date) / 365.25)
```

## We check the distribution of each relevant variable for multivariate joint model and transform

```{r}
# We check the distribution of each biomarker
hist(merged_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(merged_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# We log transform the biomarkers 
merged_data$log_PLASMAPTAU181 <- log(merged_data$PLASMAPTAU181)
merged_data$log_PLASMA_NFL <- log(merged_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(merged_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(merged_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

```{r}
# cube-root transformed
ggplot(merged_data, aes(sqrt(sqrt(PLASMA_NFL)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()

# cube-root transformed
ggplot(merged_data, aes(sqrt(sqrt(PLASMAPTAU181)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()
```

## We check for outliers

We still keep them, as we assume they are correct observations.

-   Checking for observations either side of the mean of the standardised data shows there are no values 3 sds below the mean, only above.

```{r}
merged_data$ptau_z <- scale(merged_data$PLASMAPTAU181)

merged_data$nfl_z <- scale(merged_data$PLASMA_NFL)

# We identify observations that are 3 standard deviations above or below the mean (z > 3 or < -3)
large_outliers_ptau <- merged_data[merged_data$ptau_z > 3, ]
small_outliers_ptau <- merged_data[merged_data$ptau_z < -3, ]
large_outliers_ptau
small_outliers_ptau

large_outliers_nfl <- merged_data[merged_data$nfl_z > 3, ]
small_outliers_nfl <- merged_data[merged_data$nfl_z < -3, ]
large_outliers_nfl
small_outliers_nfl
```

## We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# We remove missing values 
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(merged_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## We make sure that merged_data and surv_data have the same length (for the joint model)

```{r}
# We check the length of each dataset 
length(unique(merged_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

# Longitudinal modelFitted LME model for P-TAU

## Comparing longitudinal models

Suggested models:

-   Gamma: raw data could be modelled well with the gamma distribution since takes no negative values, variance increases with mean, and there is a right-skew to the raw distributions

-   Linear mixed effects model with log-transformed data: approximately normal, yet still some skew

-   Linear mixed effects model with cube root transformed data: again, approximately normal, still a bit of skew (why such a low AIC? seems wrong?)

-   GLMM with log-normal distribution

-   GLMM with Student-t distribution: longer tails

### pTAU-181

```{r}
# checking the mean and variance for ptau181
mean(merged_data$PLASMAPTAU181)
var(merged_data$PLASMAPTAU181)
```

```{r}
# GLMM
fit_gamma_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID), 
                   data = merged_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_ptau <- lmer(log(PLASMAPTAU181) ~ time_years + (1|RID), 
                data = merged_data)

# LME approach with log transformed data
fit_lme_cubed_ptau <- lmer(sqrt(sqrt(PLASMAPTAU181)) ~ time_years + (1|RID), 
                           data = merged_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                              data = merged_data,
                              family = lognormal(link = "log"))

# Student-t
fit_student_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                            data = merged_data,
                            family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_ptau <- simulateResiduals(fit_gamma_ptau)
testResiduals(gamma_residuals_ptau)

lme_residuals_log_ptau <- simulateResiduals(fit_lme_log_ptau)
testResiduals(lme_residuals_log_ptau)

lme_residuals_cubed_ptau <- simulateResiduals(fit_lme_cubed_ptau)
testResiduals(lme_residuals_cubed_ptau)

lognormal_residuals_ptau <- simulateResiduals(fit_lognormal_ptau)
testResiduals(lognormal_residuals_ptau)

student_residuals_ptau <- simulateResiduals(fit_student_ptau)
testResiduals(student_residuals_ptau)

AIC(fit_lme_log_ptau, fit_lme_cubed_ptau, fit_gamma_ptau, fit_lognormal_ptau, fit_student_ptau)
```

### NfL

-   Negative AIC for cube root transformed -\> take the absolute value

```{r}
# checking the mean and variance for nfl
mean(merged_data$PLASMA_NFL)
var(merged_data$PLASMA_NFL)
```

```{r}
# GLMM
fit_gamma_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID), 
                   data = merged_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_nfl <- lmer(log(PLASMA_NFL) ~ time_years + (1|RID), 
                data = merged_data)

# LME approach with log transformed data
fit_lme_cubed_nfl <- lmer(sqrt(sqrt(PLASMA_NFL)) ~ time_years + (1|RID), 
                          data = merged_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                             data = merged_data,
                             family = lognormal(link = "log"))

# Student-t
fit_student_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                           data = merged_data,
                           family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_nfl <- simulateResiduals(fit_gamma_nfl)
testResiduals(gamma_residuals_nfl)

lme_residuals_log_nfl <- simulateResiduals(fit_lme_log_nfl)
testResiduals(lme_residuals_log_nfl)

lme_residuals_cubed_nfl <- simulateResiduals(fit_lme_cubed_nfl)
testResiduals(lme_residuals_cubed_nfl)

lognormal_residuals_nfl <- simulateResiduals(fit_lognormal_nfl)
testResiduals(lognormal_residuals_nfl)

student_residuals_nfl <- simulateResiduals(fit_student_nfl)
testResiduals(student_residuals_nfl)

AIC(fit_lme_log_nfl, fit_lme_cubed_nfl, fit_gamma_nfl, fit_lognormal_nfl, fit_student_nfl)
```

## 1. Fitted LME model for P-TAU

```{r}
# Longitudinal model with t-distribution (identity link) 
library(GLMMadaptive)
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER.y + APOE4_status.y,
  random = ~ time_years | RID,                    
  data = merged_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

## Fitted LME model for NFL

```{r}
# Longitudinal model, with chosen covariates 
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER.y + APOE4_status.y, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model_nfl)
```

## Removing gender from NFL LME model

We see that gender is not significant in the LME model for NFL, so we remove it:

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + APOE4_status.y, 
  random = ~ time_years | RID, # random intercept + slope
  data = merged_data
  )

summary(long_model_nfl)
```

## Removing APOE4 status from ptau LME model

Even though APOE4 status was significant in the LME model for ptau, including it in both the LME model and the survival model led to poor MCMC convergence (high Rhat) in the joint model, probably due to difficulties in separating its direct effect on survival from its indirect effect through the biomarker. Therefore, to ensure stable estimation, we keep the APOE4 status only in the survival model, removing it from the LME modell:

```{r}
long_model_ptau <- mixed_model(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER.y,
  random = ~ time_years | RID,                    
  data = merged_data,
family = students.t(df = 3, link = "identity")
)

summary(long_model_ptau)
```

## Residuals

```{r}
# residuals 
res_ptau <- residuals(long_model_ptau, type = "subject_specific")

# fitted values
fitted_vals_ptau <- fitted(long_model_ptau, type = "subject_specific")

# plot
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for P-TAU")
abline(h = 0, col = "red", lty = 2)
```

```{r}
# residuals
res_nfl <- residuals(long_model_nfl, type = "pearson")  

# fitted values
fitted_vals_nfl <- fitted(long_model_nfl)  

# plot
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NFL")
abline(h = 0, col = "red", lty = 2)
```

## 6. QQ plot

```{r}
# For P-tau
qqnorm(res_ptau)
qqline(res_ptau, col = "red")

# For NFL 
qqnorm(res_nfl)
qqline(res_nfl, col = "red")
```

# Survival model

## We create a survival object

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

## Kaplan-Meier curve

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

## Fitted Coxph model

```{r}
# Cox survival-model, with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

## Removing gender

Gender is not significant in the survival model, so we remove it:

```{r}
# Cox survival-model, without gender 
surv_model <- coxph(
  surv_object ~ age_at_baseline + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

## Testing for the assumption of proportional hazards

```{r}
cox.zph(surv_model)
```

# Joint-model

## Fitted joint model for P-tau

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )
summary(joint_model)
```

## Fitted joint model for NFL

```{r}
# Joint model 
joint_model_nfl <- jm(
  surv_model,
  long_model_nfl,
  time_var = "time_years",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

summary(joint_model_nfl)
```

# Multivariate joint model

## We create a mixed object

```{r}
Mixed <- list(long_model_ptau, long_model_nfl)
```

## We check wether the models have stable random slopes

```{r}
summary(long_model_ptau)
VarCorr(long_model_nfl)

```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

## fForms

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )      
```

## Fitted multivariate joint model

```{r}
multi_joint_model <- jm(surv_model, Mixed,
                  time_var = "time_years",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 20000L,
                  n_burnin = 2000L
                  )
summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival
```
