---
format:
  pdf:
    include-before-body:
      text: |
        \begin{titlepage}
        \raggedleft
        \vspace*{2cm}
        
        {\Huge\bfseries Kandidatuppsats\par}
        \vspace{1.5cm}
        
        {\Large Statistiska Instiutionen\par}
        \vspace{3cm}
        
        {\Huge\bfseries Nr 2026:x\par}
        \vspace{0.5cm}
        
        {\LARGE \bfseries Title in English\par}

        {\large Title in Swedish\par}
        \vspace{0.5cm}
        
        {\large \today\par}
        \vspace{3cm}
        
        {\Huge Charlotte Emily Iszatt and Milou Rossi\par}
        
        \vfill
        
        \end{titlepage}
        
        \begin{titlepage}
        \raggedleft
        
        {\normalsize Självständigt arbete 15 högskolepoäng inom Statistik III, VT2026\par}
        {\normalsize Handledare: Hans Nyqvist\par}
        \vspace{2cm}
        
        \raggedright
        {\huge\bfseries Abstract\par}
        \vspace{1cm}
        
        {\Large The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. The abstract text can go here. Keywords: XXX, XXX, XXX\par}
        \vspace{3cm}
        
        \vfill
        
        \end{titlepage}
    toc: false
    colorlinks: false
    number-sections: true
editor: visual
---

```{=latex}
\tableofcontents

\newpage
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, output = TRUE)
```

```{r, warning=FALSE}
# loading libraries
library(dplyr)
library(tidyr)
library(JMbayes2)
library(JM)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) # for running residual diagnostics on GLMM
library(lcmm)
library(timeROC)
```

```{r}
# Loading data
#Milou
# # NfL data
 load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
 nfl_data <- BLENNOWPLASMANFLLONG
# # P-tau181 data
 load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
 ptau_data <- UGOTPTAU181
# # Diagnosis data
 load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
 diagnosis_data <- DXSUM
# # Demographics data
 load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
 demographics <- PTDEMOG
# # Apoe data
 load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
 apoe <- APOERES
```

```{r}
# Charley
library(ADNIMERGE2)
nfl_data = BLENNOWPLASMANFLLONG

ptau_data = UGOTPTAU181

diagnosis_data = DXSUM

demographics = PTDEMOG

apoe = APOERES
```

# Introduction

## Background

*General background information. Introducing some domain specific information about AD diagnosis and pathology, so that the reader is well prepared.*

**AD**

According to World Health Organization (2025), Alzheimer's disease (AD) is the most common neurodegenerative disorder and a leading cause of dementia around the world. First described by Dr. Alois Alzheimer in 1901, the disease continues to present significant challenges for the medical field, and no definitive cure exists to date (WHO, 2025). The 2011 National Institute on Aging-Alzheimer’s Association (NIA-AA) diagnostic guidelines describe the disease as a continuum with three stages: a symptom-free stage (preclinical AD), a predementia stage (mild cognitive impairment (MCI) due to AD) and a dementia stage (dementia due to AD) (Carrillo et al. 2013, p. 596). Pathophysiological changes in AD can begin up to 20 years before clinical symptoms appear, highlighting the crucial importance of early diagnosis and timely intervention to potentially slow the progression of the disease (Carrillo et al., p. 595).

**BIOMARKERS**

In the context of Alzheimer's disease, as well as diseases in general, the term "biomarker" is commonly used. According to the Biomarkers Definitions Working Group, a biomarker is defined as "a characteristic that is objectively measured and evaluated as an indicator of normal biological processes, pathogenic processes, or pharmacologic responses to a therapeutic intervention." (2001, p. 91). Similarly, the International Programme on Chemical Safety, is a joint venture on chemical safety led by WHO in coordination with the United Nations and the International Labor Organization, defines it as "any substance, structure, or process that can be measured in the body or its products and influence or predict the incidence of outcome or disease" (2001, p. 1).

**AD and BIOMARKERS**

AD pathophysiology has typically been detected by positron emission topography (PET) to confirm the presence and degree of amyloid plaques in the brain or by quantification of biomarkers implicated in AD pathology in cerbrospinal fluid (CSF). The former method involves radiation and the latter requires a lumbar puncture in order to extract the CSF. Recent advances in research have made it possible to replace these invasive procedures with blood tests. WHO published a report in 2024 highlighting the promising progress made in recent years regarding the use of blood-based biomarkers to support AD diagnosis. **\*\*Behövs källa i parentes väl**

**SPECIFIC AD BIOMARKERS**

**TAU/PTAU181**

The NIA-AA (Jack et al.) updated their criteria for the clinical diagnosis and staging of AD in 2024 to reflect the importance of blood-based biomarkers in the diagnosis process. According to the NIA-AA, there are several core biomarkers used in the diagnosis and staging of AD (Jack et al., 2024). One such core biomarker is tau protein phosphorylated at the threonine-181 position (p-tau181), where concentrations are expected to be higher in individuals with AD pathology (Wong et al., 2012). Plasma p-tau has had good performance as a standalone biomarker, compared with CSF p-tau which is usually used in conjunction with other biomarkers, suggesting that plasma p-tau assays perform better than CSF p-tau assays in AD diagnosis (Jack et al., p. 5151).

Från: <https://journals.sagepub.com/doi/10.1177/25424823261432112>

Tau proteins, which are microtubule-binding proteins, play an essential role in maintaining neuronal stability and morphology under normal conditions. The regional progression of brain atrophy in AD is closely associated with the accumulation of tau proteins. Hyperphosphorylated tau (p-tau), in its full-length or truncated forms, represents a key pathological hallmark of AD, exhibiting over 70 post-translational modifications, including more than 40 phosphorylation sites. Various forms of p-tau can be measured in CSF and plasma

Among the ones currently being explored and used in research settings, are mentioned p-tau181, a plasma phosphorylated tau, and NfL, a plasma neurofilament light (2024: p. 7).

<https://iris.who.int/server/api/core/bitstreams/43c83ef7-7a34-4b56-ae17-b584f2fc0798/content>

-   Among these, p-tau181 is significantly elevated in the CSF of AD patients and is considered a critical biomarker for the disease. The ability of p-tau181 to differentiate AD from other neurodegenerative conditions, such as frontotemporal lobar degeneration and chronic traumatic encephalopathy, is crucial for the accurate diagnosis of AD. Furthermore, p-tau181 levels correlate with AD severity and predict disease progression, providing valuable insights for evaluating therapeutic efficacy and guiding treatment strategies. Recent findings indicate that p-tau181 have excellent discriminative performance in plasma, paving the way for non-invasive monitoring of AD. **\*\*source?\*\***

**NFL**

Another core biomarker of interest in the diagnosis of AD pathology is neurofilament light chain (NfL), which is instead a marker of neuronal injury not specific to AD (Jack et al.). NfL is a molecule usually present inside neurons, and therefore

Analysis of the concentration of this molecule OUTSIDE of the cell (in CSF and blood) is a measure of neuronal damage.

**SUMMARY**

The abnormal concentrations of these biomarkers are the result of the neuropathological changes symptomatic of AD.

These two biomarkers, along with various other biomarkers and clinical measures related to Alzheimer’s disease, are the main focus of the *Alzheimer’s Disease Neuroimaging Initiative* (ADNI). Since 2004, researchers across more than 60 clinical sites in the USA and Canada have continuously collected and analyzed data throughout the different phases of the study. The current phase, ADNI4, began in 2022 and is expected to be completed in 2027. Its main goals are to provide open-access data to researchers worldwide and to improve the diagnosis of Alzheimer’s disease (ADNI, n.d.)

## Research Context

*Assessing the current landscape in terms of the scientific literature. What is known? What has been done Where are there gaps??? (This leads on to the next section, why are we doing this research, how do we hope to answer some remaining question in the AD research, orient this towards the statistical analysis/approaches!)*

Recent research has applied a range of prognostic models to estimate disease progression and related adverse outcomes in Alzheimer’s disease using biomarker data. One example is an article from 2020, in which the researchers used baseline plasma p-tau181 as a continuous predictor in a Cox proportional hazards model (Janelidze et al., 2020: 388) the most widely used procedure for modeling survival outcomes in relation to covariates (Therneau & Grambsch, 2000, p. 39). Additionally, they covaried for other plasma biomarkers, among which NfL. While adjusting the models for age, sex and education, the study found that higher baseline plasma P-tau181 levels were strongly associated with increased risk of progression to Alzheimer’s dementia, with hazard ratios around 3-4 per standard deviation increase. In the model including only P-tau181, it was a strong predictor of conversion across groups, with hazard ratios ranging from approximately 2.5 to 3.8 depending on cognitive status. When additional plasma biomarkers were included in the multivariate model, P-tau181 remained significant with only a slight reduction in effect size, whereas NfL showed no independent association with disease progression, with hazard ratios close to 1, indicating no meaningful contribution to risk prediction (Janelidze et al., pp. 385, 388).

Another article from 2024 further extended the approach by using a joint modeling framework to analyze multiple longitudinal biomarkers, including plasma p-tau181 and NfL (Yuan et al., p. 1). The joint model distinguishes itself from other models focusing on survival outcome by accounting for the effect of an endogenous time-dependent covariate measured with error, such as a biomarker (Rizopoulos, 2012, p. vii). More precisely, it couples the survival model, which is of primary interest, with an appropriate model for the repeated measurements of the endogenous covariate, thereby accounting for its specific characteristics (Rizopoulos, 2012, p. 51-52). Several studies have examined these biomarkers using only a longitudinal model (Mattsson et al., 2019; Chen et al., 2021). The study using a joint model found that higher longitudinal p-tau181 levels were associated with a significantly increased risk of progression, with the joint model estimating a hazard ratio of approximately 5.83 for p-tau181 and 4.18 for NfL (Yuan et al., p. 1). Additionally, the two biomarkers were combined into a multivariate joint model, leading to improved performance (Yuan et al., pp. 1, 3).

Additionally, for understanding the natural history of a disease, researchers in an article from 2023 describe it as crucial to simultaneously account for the assessment of multivariate markers over time, the occurrence of clinical endpoints, and a highly suspected heterogeneity between patients (Proust-Lima et al., p. 3996). Subsequently, they proposed a latent class approach modeling multiple repeated markers to help describe complex disease progression and identify subphenotypes for exploring new pathological hypotheses; in other terms, a joint latent class model. Through the example of progression to multiple system atrophy (MSA), characterized by various combinations of progressive autonomic failure and motor dysfunction, the researchers identified five subphenotypes of MSA that differ by the sequence and shape of marker degradation, as well as the associated risk of death (Proust-Lima et al., p. ?).

Similarly, in an article from 2026, researchers applied a multi-dimensional joint latent class model to capture heterogeneity in the trajectories of three cognitive domains from the time of Alzheimer’s dementia diagnosis: memory, language, and executive functioning (Scollard et al.: p. 383). They identified five latent classes differentiated by overall level of impairment, ranging from high to low, and rate of decline, from slow to fast. Within each latent class, the pattern of decline was relatively consistent across cognitive domains, suggesting parallel deterioration across functions rather than domain-specific divergence. In addition, class membership was associated with several key variables, such as APOE genotype, sex, race, education and neuropathological burden (Scollard et al., 2026: p. 383). Despite these advances, such modeling approaches have rarely been applied to longitudinal plasma biomarkers in Alzheimer’s disease research. Since there is evidence to support AD as a heterogeneous disease with differing disease trajectories (**SOURCE**), a latent class joint model could be employed to assess whether individuals can be categorised into meaningful classes based on NfL and ptau-181 trajectories alongside survival data.

Thus, the question arises as to which modeling approach is best suited for predicting progression to Alzheimer’s disease dementia when considering the added value of longitudinal biomarker information. Few studies have systematically compared the previously mentioned modeling approaches within a unified framework in Alzheimer’s disease, specifically focusing on plasma biomarkers such as NfL and p-tau.

Additionally, an important consideration is whether to include all participants with repeated biomarker measurements. Although the two previously mentioned studies included all participants regardless of the number of measurements per individual, this is not always the case. For example, a study examining different cognitive biomarkers to predict progression to Alzheimer’s disease explicitly states: "Participants were included if they had a baseline diagnosis of MCI and at least three follow-up visits with complete longitudinal cognitive assessments" (Guo et al., 2026, p. ? ). Similarly, in a study on type 1 diabetes and primary biliary cirrhosis (PBC), the joint model analysis was restricted to participants with recorded height and weight and at least three biomarker measurements (Xu, 2021, p. 9). Another study, focusing on cardiovascular events, restricted their joint model analysis to the cohort of 3784 patients who had at least 10 measurements of the selected biomarker (Wang et al., 2024, p. 591-592). This was done to ensure that the approximation method did not lead to "non-ignorable biases" (Wang et al., p. 591).

**Vet inte om detta är relevant \*\*\*\*** At the same time, this method signifies a loss of precision, as analyses restricted to subjects with complete data will generally be less efficient than methods that use all available data (Fitzmaurice, 2011, p. 491). The location of the missing data, such as missingness spread sporadically over many subjects, or concentrated at a specific set of time points in a few subjects, and how highly correlated the missing data are with the observed data, will also affect loss of precision. Finally, under certain circumstances missing data can introduce bias and thereby lead to misleading inferences about changes in the mean response. The authours of the book continues by writing: "It is this last factor, the potential for serious bias, that complicates the analysis of partially missing longitudinal data." (Fitzmaurice, p. 491). As a result, the reasons for any missing data, often referred to as the missing datamechanism, must be carefully considered (Fitzmaurice, p. 491).

Fitzmaurice, G.M., u, Laird, N.M., & Ware, J.H. (2011). *Applied Longitudinal Analysis*. John Wiley & Sons, Inc. Online ISBN:9781119513469. DOI:10.1002/97811

In other words, some studies include all available participants, whereas others choose to include only participants with a minimum number of measurements, meaning that those with only a single measurement are left excluded from the modeling process.

**Behöver införa källan** –\> On one hand, the longitudinal effects are only estimated from participants with more than one measurement. On the other hand, the excluded ones still contribute in estimating the cross-sectional effects of covariates, such as time effects. Additionally, there is the aspect of missing data, which means that a likelihood-based repeated measurements analysis will be valid under the missing at random assumption if they are in. It is thus recommended to still include the participants with only one measurement.

## Research Aims

*What are our research questions?*

This study investigates the prediction of Alzheimer’s disease/dementia progression using longitudinal biomarker data, comparing traditional and advanced statistical approaches. More precisely, the research questions are:

1.  Which modeling approach provides the best prediction of progression to Alzheimer’s disease/dementia when using longitudinal biomarker data (Including baseline levels, average levels over time, single-biomarker joint models, multivariate joint models, univariate latent class joint models and multivariate latent class joint models)?
2.  How do these associations differ between p-tau181 and NfL? (Comparison across biomarkers)
3.  How does the precision of the estimated longitudinal trajectories and their associations with progression (survival outcomes) change when including individuals with different numbers of repeated measurements? (Precision analysis)

# Methodology

*How will we accomplish this? –\> Introduce the methods we will use.*

To address the research questions, a series of models will be implemented and compared. First, Cox proportional hazards (Cox PH) models will be fitted using baseline biomarker levels as covariates to assess their association with progression to Alzheimer’s disease/dementia. Second, single-biomarker joint models will be applied to simultaneously model the longitudinal trajectories of each biomarker and their association with the time-to-event outcome. Third, multivariate joint models will be used to jointly analyse multiple biomarkers, allowing for the estimation of their combined and potentially correlated effects on disease progression. Finally, a latent class joint modelling approach will be implemented to account for unobserved heterogeneity in longitudinal biomarker trajectories and risk of progression by identifying distinct subgroups. A univariate joint latent class model will be fit for NfL and ptau-181 trajectories seperately, which is then compared with a multivariate joint latent class model which models both NfL and ptau-181 simultaneously. All models will be adjusted for relevant covariates.

To investigate differences between p-tau181 and NfL, the modelling frameworks will be applied separately for each biomarker, followed by a comparative evaluation of their predictive performance and estimated associations with clinical progression. Model performance will be evaluated using time-dependent AUC for dynamic prediction. The significance level will be set at 0.05 for all analyses.

In addition, a precision analysis will be conducted to assess how the number of repeated measurements per individual influences the estimation of longitudinal trajectories and their associations with the survival outcome. This will be evaluated by comparing results between individuals with at least one repeated biomarker measurement and those with at least three repeated measurements, and by assessing differences in model stability and parameter precision.

There exists a wide range of biomarkers for Alzheimer’s disease that could be investigated in this context. However, given that plasma-based biomarkers have only relatively recently gained attention in research, their inclusion was considered both timely and relevant. Accordingly, the two biomarkers included in the analyses are p-tau181 and neurofilament light chain (NfL). They were selected due to their established relevance in the prediction of dementia and Alzheimer’s disease progression, as well as the availability of rich longitudinal data with repeated measurements and overlapping follow-up information. Even though other clinically relevant biomarkers are available, they were not included due to the limited availability of sufficiently large longitudinal datasets with repeated measurements. In this study, priority was given to data quality and longitudinal depth rather than the number of biomarkers considered. In addition, the selected biomarkers represent biologically distinct processes: p-tau181 is more specifically associated with Alzheimer’s disease pathology, while NfL reflects more general neuroaxonal damage and is not disease-specific. This contrast makes them particularly informative for comparative analyses of biomarker behaviour and predictive performance in dementia progression.

Additional modelling approaches could have been considered given the aims of the study. For instance, the Cox Proportional Hazard model can be extended for modelling the relationship between time-dependent covariates and survival time. Caution implementing these must be taken, however, as the model cannot handle endogenous time-dependent covariates such as biomarkers, only those which are exogenous (Rizopoulos, 2012: p.51). **For example, if the event is defined as death, endogenous variables can inherently effect the survival outcome only be Endogenous variables inherently carry information about survival times........**

**Consequently, this approach is not included in the present study.**

In addition, given the limited time and scope of this thesis, including a larger number of modelling frameworks would have reduced the ability to investigate each approach in sufficient depth. Therefore, a smaller set of models was selected in order to allow for a more detailed and meaningful comparison. The chosen models were selected based on their methodological relevance and their ability to capture different aspects of the data, thereby providing a broad overview of alternative approaches for modelling longitudinal biomarker data and survival outcomes.

The modelling approaches will be applied to data obtained from the *Alzheimer’s Disease Neuroimaging Initiative* (ADNI) and handled in accordance with their data use and access policies. In this context, it is also worth noting the limited availability of Alzheimer’s disease data in Sweden due to legal and regulatory constraints governing access to health data. This limitation further increases the importance of utilizing well-characterized international datasets such as ADNI, as they provide the necessary longitudinal and high-quality data required to conduct the types of analyses performed in this study.

# Statistical theory

*Describe the statistical theories behind the chosen methods*

*Måste förkortas och göras mer sammanhängande*

## Survival analysis

Survival analysis, also known as time-to-event analysis, focuses on both whether an event occurs and the time from the start of follow-up until its occurrence (Ranganathan, Deo & Pramesh, 2025, p. ?). The start of follow-up, also referred to as baseline, is defined as an initial measurement taken at an early time point that serves as a reference for assessing changes over time (NCI, n.d.). The event of interest may include death, disease onset, relapse, recovery or other outcomes depending on the research question. However, this type of data analysis faces two main challenges. First, survival times are often skewed rather than normally distributed. Second, for some individuals the exact event time is unknown because the event has not occurred by the end of follow-up or because participants are lost to follow-up. This is referred to as censoring. Therefore, specialized statistical methods are required in order to appropriately incorporate both complete and censored observations when estimating survival-related quantities (Ranganathan, Deo & Pramesh, p. ?)

With respect to the temporal relationship between the event and censoring time, censoring is typically classified as either right or left censoring (Rizopoulos, 2012, p. 36). Right censoring occurs when, for some individuals, the event of interest is not observed during the study period, although it is known to occur after the last recorded observation. This may arise if follow-up ends before the event occurs, if the study is terminated after a pre-specified number of events, or if participants are lost to follow-up due to relocation or similar reasons. In all such cases, the exact event time is unknown but is known to exceed the last observed time point (Rizopoulos, p. 36). In studies of Alzheimer’s disease, not all participants develop the disease during the follow-up period; for these individuals, the exact time of onset is unknown, and they are therefore considered right-censored observations.

However, when longitudinal data are collected at different time points, issues such as left truncation (also referred to as delayed entry) may arise. Left truncation occurs when individuals enter the study after the origin of the underlying time scale, meaning that only individuals who have survived or remained event-free up to the point of entry are included in the analysis (källa?).

## Proportional Hazard Model

**Theory about the model... (specifically why we should use it)**

-   Time-to-diagnosis process

-   Parametric baseline hazard: **flexible parametrics, splines??**

-   OR leave undefined -\> Cox PH model

Assuming a Cox PH survival sub-model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha m_i(t) \right\}
$$

$\mathbf{v_i}:$ vector of time-independent co-variates

$\phi:$ vector of log-hazard ratios

$\alpha:$ the association parameter

$M_i = \left\{m_i, \quad 0 \le s \le t\right\}:$ the **true unobserved biomarker profile** up to time $t$

### Assumptions!!

-   What do we assume about the censoring process?

-   Proportionality

### Cox PH model using baseline values of each biomarker

## Joint models

A joint model consists of a survival process and a longitudinal process, which can be specified by a proportional hazards model, such as a Cox PH model, and a linear mixed effects model, respectively.

### Longitudinal Data

Longitudinal data in this analysis is defined as $n$ repeated measurements of p-tau181 and NfL on $N$ individuals in the ADNI1, ADNIGO and ADNI2 cohorts. $T_{i, j}$ and $N_{i, j}$ denote the measured concentrations of p-tau181 and NfL respectively for the $i^{th}$ individual at the $j^{th}$ occasion. Assuming that the expected concentration of the biomarkers varies with both time and individual level covariates, the mean responses have the notation $\mu_{T:\text{ } i, j} = E(T_{i, j})$ and $\mu_{N:\text{ } i, j} = E(N_{i, j})$. Considering the data are made up of repeated responses on the same individuals, the assumption of independence between observations is illogical. Clearly, the baseline levels of p-tau181 and NfL for an individual are correlated with the measurements on subsequent occasions.

### Mixed Effects Models

Why???

We have grouped data, specifically longitudinal data where groups of observations are correlated, since we have repeated measurements on the same individuals. We expect data *within* the groups, that is, data collected form the same individual, to be more correlated than those *between* the groups. Mixed effects models are not only for longitudinal data, but for any data where there is non-independence between observations.

### Linear Mixed Effect Model

**Theory about the model... (specifically why we should use it)**

-   Biomarker process

-   Describes the trajectory of biomarkers at the individual level 

**Linear mixed effects models?** What are these - why use them here?

$$ y_i(t) = m_i(t) + \epsilon_i(t) \qquad \epsilon_i(t) \sim N(0, \sigma^2) $$

$$ m_i(t) = \mathbf{X_i^T}\beta + \mathbf{Z_i}^T\mathbf{b_i} \qquad \mathbf{b_i}(t) \sim N(0, \sigma^2) $$

$y_i(t):$ the observed longitudinal biomarker

$m_i(t):$ the trajectory function which represents the **true unobserved** **value of the biomarker** at time $t$ for patient $i$ - we assume that we do not observe the true value due to measurement errors or other reasons.

$\mathbf{X_i}^T(t)\beta:$ the fixed effects

$\mathbf{Z_i}^T\mathbf{b_i}:$the random effects

The full model becomes:

$$ y_i(t) = \mathbf{X_i^T}\boldsymbol{\beta} + \mathbf{Z_i}^T\mathbf{b_i} + \epsilon_i(t) \qquad \mathbf{b_i}(t) \sim N(0, \sigma^2), \quad \epsilon_i(t) \sim N(0, \sigma^2) $$

### Generalised Linear Mixed Effects Model

*Inte ens säkert att vi behöver denna del, om residualerna för standard-LME modellerna kan räknas som ok*

Ref: <https://stats.oarc.ucla.edu/other/mult-pkg/introduction-to-generalized-linear-mixed-models/>

An extension of GLMs, where the linear predictor contains random effects alongside the fixed effects. This is used to extend the linear mixed model to non-normal data.

Given random effects, $\mathbf{u}$ the dependent variable, $\mathbf{y}$, is distributed according to the exponential family, such that $g(E[\mathbf{y}|\mathbf{u}]) = \mathbf{X𝛽} + \mathbf{Zu}$. Some link function, $g(x)$, links the expected value of $\mathbf{y}$ given the random effects to the linear predictor which included both fixed effects and random effects.

Where:

$\mathbf{X}$ = the fixed effects design matrix

$\mathbf{\beta}$ = the fixed effects

$\mathbf{Z}$ = the random effects design matrix

$\mathbf{u}$ = the random effects

What do we model as random?

We can model random intercept and keep all other effects fixed. In that case, for $i$ biomarker measurements on $j$ individuals, the random effects design matrix - $\mathbf{Z}$ - is a $i \cdot j$ matrix containing 1's and 0's, where $z_{i,j} = 1$ if the $i^{th}$ biomarker measurements were taken from the $j^{th}$ individuals, and $z_{i,j} = 0$ if the $i^{th}$ biomarker measurements were not taken from the $j^{th}$ individual.

We do not tend to estimate $\mathbf{u}$, but instead assume that $u_i \sim N(0, \mathbf{G})$, where $\mathbf{G}$ is the variance-covariance matrix of the random effects. The fixed effects are estimated, and the random effects are seen as deviations from the fixed effects and have $E(u) = 0$.

For a model with only a random intercept, $\mathbf{G}$ becomes $G = \sigma^2_{intercept}$. If we extend the model to have random slope, we get:

$$ G = \begin{bmatrix} \sigma^2_{intercept} & \sigma_{intercept, slope} \\ \sigma_{intercept, slope} & \sigma^2_{slope} \end{bmatrix} $$

Like all variance-covariance matrices, some elements are redundant since they are repeated. Estimating instead some function of $\mathbf{G}$, where $\mathbf{G} = \sigma(\theta)$, means estimating $\frac{p(p + 1)}{2}$ elements instead. Assuming (**?????**) the random effects are independent, the variance-covariance matrix for the random effects becomes diagonal:

$$ G = \begin{bmatrix} \sigma^2_{intercept} & 0 \\ 0 & \sigma^2_{slope} \end{bmatrix} $$

### Assumptions!!

-   Random effects $\mathbf{b_i}$ are **time-invariant** - they are assumed to come from a normal distribution, $\mathbf{b_i} \sim N(0, \sigma^2)$, and represent stable, latent characteristics about the individuals disease, such as a generally fast declining disease trajectory but cannot represent other factors that effect disease trajectory over-time, e.g. medication. Any external time-varying process that effects the biomarker trajectory and the hazard cannot be represented by the random effects.

### Simple Joint Model

#### Frequentist vs bayesian approach

*Man kan skatta joint-modellen med paketet JM, på frekventistiskt sätt med maximum likelihood, eller med paketet JMbayes2, på bayeisanskt sätt.*

*Vår frekventistiska modell konvergerar inte alls dock, även utan kovariat, d.v.s. även enklaste versionen –\> Det beror nog på att maximum likelihood-metoden är jättekänslig mot skev eller konstig fördelning (som vi har), medan bayesian är mer flexibel –\> Måste hitta källa på det, men det kan användas som motivering för valet av bayesian approach. Bra att ta upp båda metoderna dock.*

*Viktigt att ta upp om priordelning, att det behövs för bayesiansk fördelning. Väljs det automatiskt i JM()-funktionen (någon sorts standard) eller måste man välja själv? Uitfrån vad i såna fall?*

*Skriva om MCMC och Convergence, vad det innebär inom bayesian approach, för att ens kunna prata o the Gelmain-Rubin statistic*

**Theory about the model... (specifically why we should use it)**

In the joint model, the hazard function at time $t$ is directly dependent on the estimate of the biomarker at time $t$.

The association structure of the joint model:

-   How is the longitudinal model linked to the survival model? 

-   Does this association structure change for different biomarkers?

-   What happens to time-to-diagnosis predictions if we assume the standard structure when another underlying association is a better fit?

If we include the current rate of change alongside the current value in the proportional-hazard model:

$$
h(t|M_i, \mathbf{v_i}) = h_0(t)exp\left\{\phi^T\mathbf{v_i} + \alpha_1 m_i(t) + \alpha_2 m_i'(t) \right\} \qquad where \qquad m_i'(t) = \frac{d}{dt} m_i(t)
$$

$\alpha_1 :$ links the current value of the biomarker - $m_i(t)$

$\alpha_2 :$ links the current gradient/rate of change - $m_i'(t)$

Joint-model options:

-   Longitudinal sub-model: model of class lme()

### Assumptions!!

-   Full conditional independence

    -   Random effects $\mathbf{b_i}$ explain all statistical dependence between (1) the longitudinal outcome and the survival outcome and (2) repeated measurements within the longitudinal process.

    -   This assumption can fail when: there are time-varying confounders not captured by the random effects (see the longitudinal model for more info about $\mathbf{b_i}$ assumptions), if the measurement error, $e_i(t)$ in the longitudinal model $y_i(t) = m_i(t) + e_i(t)$ is informative, or there are sub-populations (APOE genotype??? Maybe we need to split into populations based on genotype)

-   So: the longitudinal outcome is independent of the time-to-event outcome and the repeated measurements in the longitudinal outcome are independent of each other

-   $\mathbf{b_i}$ drives the **true (unobserved)** biomarker trajectory and hazard of AD diagnosis: the observed $y_i(t)$ is the observed biomarker state, $m_i(t)$, plus noise. Where $m_i(t)$ is itself determined by $\mathbf{b_i}$, \*\*since the others are fixed ????? \*\*, once we know $\mathbf{b_i}$, the observed biomarker measurement and the survival times are no longer dependent. The observed $y_i$'s are useful in determining $\mathbf{b_i}$ which is the useful information in estimating survival times.

**Frequentist vs bayesian approach**

"A simultaneous inferential method through a joint likelihood may be favorable, but the computational burden for proposed MVJ modelings can be very intensive, even sometimes infeasible, and may cause problems of algorithm convergence. (Brown et al., 2003, p. ?; Wu et al. 2010, p ?) <https://academic.oup.com/biometrics/article-abstract/59/2/221/7286178?redirectedFrom=fulltext&login=false>

<https://pubmed.ncbi.nlm.nih.gov/19673859/>

Brown, E.R. & Ibrahim, J.G. (2003). A Bayesian Semiparametric Joint Hierarchical Model for Longitudinal and Survival Data, *Biometrics*, 59(2), 221-228, [https://doi.org/10.1111/1541-04](#0)

Wu, L., Liu, W., & Hu, X. J. (2010). Joint inference on HIV viral dynamics and immune suppression in presence of measurement errors. *Biometrics*, *66*(2), 327-335. https://doi.org/10.1111/j.1541-0420.2009.01308.x

Bayesian joint modeling method shows the advantage. Thus, the parameters can be estimated simultaneously for the MVJ models implemented by Markov Chain Monte Carlo (MCMC) techniques for the skew-normal MVJ model. The simultaneous statistical inference on all unknown model parameters will capture the underlying association between the longitudinal exposures and the event time data.

<https://www.frontiersin.org/journals/big-data/articles/10.3389/fdata.2022.812725/full#cite>

Huang, Y., Chen, J., Xu, L., & Tang, N.S. (2022). Bayesian Joint Modeling of Multivariate Longitudinal and Survival Data With an Application to Diabetes Study. *Front. Big Data* 5:812725. 10.3389/fdata.2022.812725

**MCMC and convergence**

<https://www.sciencedirect.com/topics/social-sciences/markov-chain-monte-carlo>

### Multivariate Joint Model

However, as the joint model includes a linear mixed-effects model with the normal assumption for a single longitudinal outcome, joint model-based analyses may not provide robust inference when longitudinal measurements exhibit skewness and/or heavy tails (Huang et al, 2022, p.?). Consequently, joint models, besides with the traditional approach of maximum likelihood estimation, can also be estimated within a Bayesian framework using MCMC techniques (Rizopoulos, 2012, p. 62). This approach allows for more flexible and complex association structures (Andrinopoulou & Rizopoulos, 2016, p.?).

The data collected for the joint model-based analyses are also often featured by multivariate longitudinal outcomes which are significantly correlated, and ignoring their correlation may lead to biased estimation (Huang et al, 2022, p.?). The bayesian approach allows for multivariate joint modelling (Andrinopoulou & Rizopoulos, 2016, p.?).

#### The Gelman-Rubin statistic

Convergence can be evaluated by running multiple chains from different starting values (Peng, 2022). All chains are expected to eventually reach the stationary distribution and become indistinguishable. Convergence can be assessed by comparing variation between chains to variation within chains, with small between-chain variation indicating that the chains are similar (Peng).

The Gelman-Rubin statistic R formalizes this idea (Peng). Consider $J$ Markov chains, each with samples $x_1^{(j)}, x_2^{(j)}, \dots$ from the $j$-th chain. For each chain, remove the first $D$ samples as “burn-in” and retain the following $L$ samples: $x_D^{(j)}, x_{D+1}^{(j)}, \dots, x_{D+L-1}^{(j)}$. One common choice is to set $D = L$.

Then calculate the chain mean for each chain,

$$
\bar{x}_j = \frac{1}{L} \sum_{t=1}^{L} x_t^{(j)}
$$

the overall mean across all chains (the grand mean),

$$
\bar{x}_\cdot = \frac{1}{J} \sum_{j=1}^{J} \bar{x}_j
$$

the between-chain variance,

$$
B = \frac{L}{J-1} \sum_{j=1}^{J} \left(\bar{x}_j - \bar{x}_\cdot\right)^2
$$

the within-chain variance for each chain,

$$
s_j^2 = \frac{1}{L-1} \sum_{t=1}^{L} \left(x_t^{(j)} - \bar{x}_j\right)^2
$$

and finally, the average within-chain variance,

$$
W = \frac{1}{J} \sum_{j=1}^{J} s_j^2
$$

Subsequently, the Gelman-Rubin statistic is:

$$
R = \frac{\frac{L-1}{L} W + \frac{1}{L} B}{W}
$$

As the number of iterations L increases and the between-chain variation B decreases, R approaches 1. Thus, chains are typically run until R is near 1, often around 1.1 or 1.2. Being a unit-free ratio, R provides a convenient summary of convergence without requiring specification of a particular parameter, making it a practical tool for monitoring MCMC chains before any inference is made (Peng).

### Joint Latent Class Model

The joint latent class model assumes that the study population is heterogeneous and that within these subgroups, the observations have the same distribution (Kyheng, 2025, p.2). Therefore, this model is useful for studying the relationship between longitudinal data and survival data when there is suspected heterogeneity in the population.

-   Assume that the study population, $N$, is made up of $G$ homogenous, unobservable classes

-   $c_i$ is the class that individual $i, i = (1, ... , N)$ belongs to, where the multinomial logistic submodel models the probabilities for each individual of belonging to latent class $g$, given the predefined variables $X_{1,i}$. HOWEVER: no prior assumptions about the latent classes, no covariates included in the model

$$
\pi_{i, g} = P(c_i = g | X_{1,i})
$$

-   The linear submodel now becomes unique to each latent class

-   The survival submodel is defined for all latent classes

-   Assumption: conditional independence, correlation between biomarker trajectories and event occurrence are fully described by the latent class (Jacqmin-Gadda: score test for residual dependence between longitudinal data and event risk )

Latent class joint models offer a way to model the expected longitudinal biomarker measurements alongside time-to-diagnosis given that the population is made up of latent, unobserved classes that are not described by the observed covariates (Rizopoulos, 2012, p. 122).

-   Correlations between observations (biomarker measurements) are modelled by the random effects, $b_i$, in the longitudinal model

-   The relationship between the biomarker trajectory and the time-to-diagnosis is described by the latent class indicator, $c_i$

-   THESE ARE THE CONDITIONAL INDEPENDENCE ASSUMPTIONS

According to Rizopoulos, the multinomial logistic regression models the latent classes, such that:

$$
Pr(c_i = g) = \frac{e^{\mathbf{X}_i \boldsymbol{\beta}_g}}{\sum^{G}_{l = 1}e^{\mathbf{X}_i \boldsymbol{\beta}_l}}
\qquad where \qquad \overset{\mathbf{X}_i \text{ is the set of covariates for individual }i}{\boldsymbol{\beta}_l \text{ is the set of coefficients for latent group } g}
$$

If there are no prior assumptions about the latent classes, then there will be no covariates in the multinomial logistic regression model (Kyheng, p.2). This then simplifies the model to:e

$$
Pr(c_i = g) = \frac{e^{\boldsymbol{\beta}_{0,g}}}{\sum^{G}_{l = 1}e^{\boldsymbol{\beta}_{0,l}}}
$$

An advantage of this formulation of the latent class joint model is that the log-likelihood is much more tractable than for the regular joint model (Rizopoulos, p. 123) ... **MORE, WHY, SHOW**

### Information criterion, entropy

#### Shannon Entropy

-   Uncertainty of the variable itself

-   Entropy is a quantification of the (average) level of uncertainty associated with a random variable's potential outcomes (similary to entropy in thermodynamics) <https://en.wikipedia.org/wiki/Entropy_(information_theory)>

-   The expected amount of information needed to describe the state of the random variable, given the distribution of probabilities across all possible outcomes

-   "Information content" of an outcome is related to how probable it is. If a highly likely outcome occurs, the information content is small. The information/surprisal, $I(E)$ of an event approaches $0$ as $p(E)$ approaches $1$:

$$
I(E) = log \left( \frac{1}{p(E)} \right) = -log(p(E))
$$

-   Entropy $H$ calculates the average expected amount of information across all possible outcomes of an random variable, $X$, such that:

    $$ 
    H(X) = - \sum_{x \in X} p(x)log(p(x))
    $$

-   why 1 - p(x)log(p(x))

-   <https://www.youtube.com/watch?v=LBOVhy3ITA4>

-   <https://www.youtube.com/watch?v=-4y4X8ELcEM>

#### Assignment based entropy

-   Uncertainty of the class assignment!

-   Posterior probabilities:

$$
p_i(c_g) = P(C = c_g | \mathbf{X}_ii)
$$

...

### ROC analysis

ROC analysis is a valuable method for assessing how well diagnostic tests perform, and more generally for measuring the accuracy of statistical models, such as logistic regression and discriminant analysis, that assign individuals to one of two groups, typically classified as diseased or not diseased. Earlier research on ROC methods mainly focused on ordinal rating data, where responses take a limited set of ordered categories reflecting levels of confidence about disease status. More recently, ROC techniques have increasingly been used with continuous data as well (Zou et al., 2012, p. 3)

The traditional ROC curve method assumes that both the disease status and the marker value for each individual remain constant over time. In real-world settings, however, this assumption is often violated. Individuals who are disease-free at baseline may develop the disease during follow-up, and their diagnostic marker values may also change over time. To address this limitation, time-dependent ROC curves have been developed to incorporate the temporal nature of event occurrence and the presence of censored survival data, making them more appropriate for evaluating prognostic accuracy in survival settings (Kamarudin et al., 2017, p. 1).

When a time-dependent framework is adopted, the disease status is evaluated at each time point, meaning that sensitivity and specificity may change over the course of the study. Let $T_i$ denote the time of disease onset for individual $i$, and let $X_i$ represent a marker value (typically measured at baseline) for individual $i$, where $i = 1, \ldots, n$ (Kamarudin et al., p. 2).

The observed time is defined as $Z_i = \min(T_i, C_i)$, where $C_i$ is the censoring time. In addition, let $\delta_i$ be an indicator variable that takes the value 1 if the event (disease) occurs and 0 otherwise. Furthermore, let $D_i(t)$ denote the disease status at time $t$, where $D_i(t) = 1$ indicates presence of disease at time $t$, and $D_i(t) = 0$ otherwise (Kamarudin et al., p. 2).

In what follows, $X$ is referred to as a marker, although it may also represent a risk score derived from a regression model or another predictive model, including previously published scores (Kamarudin et al., 2017, p. 2).

For a given threshold $c$, the time-dependent sensitivity and specificity are defined as:

$$ Se(c,t) = P(X_i > c \mid D_i(t) = 1) $$

$$ Sp(c,t) = P(X_i \le c \mid D_i(t) = 0) $$

Based on the definitions above, the corresponding receiver operating characteristic (ROC) curve at a given time $t$, denoted by $ROC(t)$, can be constructed by plotting the time-dependent sensitivity $Se(c,t)$ against $1 - Sp(c,t)$ across all possible thresholds $c$ (Kamarudin et al., p. 2).

The time-dependent area under the curve (AUC) is then defined as:

$$ AUC(t) = \int_{-\infty}^{\infty} Se(c,t)\, d[1 - Sp(c,t)] $$

# Data

The data used in this analysis comes from the ADNI cohort under the ADNI1 and ADNI2 protocols. Measurements of the concentration of p-tau181 and NfL in blood plasma of $n = xxx$ individuals were analysed. In accordance with the ADNI leadership guidelines, which state that there were serious data quality concerns for individuals with a PTID in the format "381_S_10###" , checks were run to assure that no observations with these PTIDs were in the data set (ADNI, 2026).

The variables of interest were all retrieved from the ADNIMERGE2 package, found in the ADNI database, which include various datasets all containing different variables related to the study of Alzheimer's.

The inclusion criteria required participants to have recorded diagnostic visits and biomarker measurements of both p-tau181 and NfL. Participants who did not meet these criteria were excluded. This resulted in a sample size of 1190 individuals with three cohorts with a final diagnosis of cognitively normal (CN) participants (n = 358), participants with mild cogntive impairment (MCI, n = 421) and with dementia due to AD (n = 411). Additionally, when restricting the sample to participants with three or more biomarker measurements, the sample size was reduced to 768 with final diagnoses of CN (n = 241), MCI (290) or AD dementia (n = 237). In the reduced sample size, 67 individuals were included who had dementia diagnoses at baseline and 178 individuals transitioned from either CN or MCI to AD dementia.

**REPORT ON TRANSITIONS...**

## ADNI

*Explain the different protocols and what they mean for data analysis, if we can compare across protocols etc.*

All protocols allow enrollment for individuals aged 55-90.

-   ADNI1

800 subjects (200 controls, 400 MCI, 200 mild AD) recruited at around 50 sites in the United States and Canada

-   ADNI2

Cognitively normal, significant memory concern, early MCI, late MCI, mild AD

Inclusion criteria for rollover participants: originally diagnosed as either MCI (early or late) or CN

-   ADNI3

700 - 800 rollover participants

370 - 1200 newly enrolled

Three cohorts: CN, MCI, mild AD

-   ADNI4

Introducing the study, it's aims and the type of information gathered.

**IMPORTANT: REMEMBER WE NEED TO REFERENCE ADNI PROPERLY.**

## Introducing variables... biomarkers

*Information about the variables we will use and why, what they can show, how are they recorded*

Under this section, all variables used in the analysis will be described, along with the reasons for their importance.

### Diagnosis status

The variable of primary interest is an ordinal categorical variable representing the diagnostic status of the participants. More precisely, it includes three diagnostic states: NC, MCI, and dementia. Each participant has multiple examination dates, meaning that diagnostic status may change over time.

To investigate progression to dementia, the first recorded diagnosis of dementia for each participant is used to define the event of interest. Accordingly, the study focuses on time to incident dementia, where participants are classified as either non-demented or demented. Participants without a dementia diagnosis are censored at their last available follow-up visit.

The variable was retrieved from the dataset "DXSUM" in ADNIMERGE2, for which the data was collected by ???. Across the 1190 included participants, there are 29743, of which 10812 observations are of CN, 4282 of Dementia and 14649 of MCI. They are all ranging from 29 September 2005 to 7 November 2025. Per individual, the number of observed diagnosis visits range from one to 95.

*Skriva något om hur regelbundna besöken är.*

*Lägga till om värdena när man sedan reducerar till de individer med 3 eller fler mätningar.*

*Lägga till om hur vi kommer använda variablerna i våra modeller exakt.*

### P-tau181

P-tau181 is a continuous numerical variable representing one of the two plasma biomarkers used to potentially improve prediction of participants’ progression to dementia. It was retrieved from the dataset UGOPTAU181 in ADNIMERGE2, for which the data was collected by ????. In contrast to the diagnostic data, the p-tau181 measurements are only available between 1 February 2007 and 6 June 2016, meaning that there are diagnostic visits without corresponding p-tau181 measurements.

The biomarker measurements are recorded on a scale of ??? and are observed 3647 times across the 1,190 included participants. Values range from a minimum of 0.4 to a maximum of 451.4. The median is 16.3, with the first quartile at 11.1 and the third quartile at 23.6. Overall, the distribution of the variable is highly right-skewed and not normally distributed.

### NfL

NfL, a continuous numerical variable, represents the other plasma biomarker with potential to improve prediction of participants’ progression to dementia. It was retrieved from the dataset "BLENNOWPLASMANFLLONG" in ADNIMERGE2 and the data was collected by ???.

Similarly to the previous biomarker, there are 3647 measurements across the 1191 included individuals, ranging from 1 February 2007 to 6 June 2016.

The values are on the scale ??? and range from a minimum value of 6.4 to a maximum value of 408. The median is 37.9, with the first quartile at 27.8 and the third quartile at 52.5. Overall, both p-tau181 and NfL exhibit highly right-skewed distributions, although NfL appears to be slightly less skewed.

### Age

Additionally, the participants’ age is included as a covariate when predicting progression to dementia, to examine how the risk of dementia changes with age. This variable, together with the following variable, were retrieved from the dataset PTDEMOG, for which the data was collected by ???.

The variable is initially coded numerically as the year of birth for each participant, ranging from 1915 to 1971 among the 1190 included participants, and is then transformed into age at visit and age at baseline visit, which are more suitable for use in the analyses.

### Gender

Another covariate of interest when predicting progression to dementia is participants’ sex at birth, as recorded on their original birth certificate. This variable is binary, with two categories: male and female. Among the 1190 participants, 543 were female and 647 were male. Including this variable allows for the examination of whether progression to dementia differs by sex.

### APOE4-status

Finally, apolipoprotein ε4 status is another covariate used to explain progression to dementia. The ε4 apoliprotein-E (APOE) allele is a risk factor for AD, whereas the ε2 allele has been shown to have protective factor (Wong et al.). Retrieved from the APOERES dataset, the variable was initially coded numerically based on the number of ε4 alleles and was then recoded into three ordinal categories: 0 alleles (non-carriers), 1 allele (heterozygous carriers; 2/4 or 3/4), and 2 alleles (homozygous carriers; 4/4). This allows the examination of whether the number of ε4 alleles is associated with the risk of progression to dementia. Consequently, there are 672 participants without any ε4 alleles, 417 participants with one allele, and 101 participants with two alleles among the 1190 included participants.

# Pre-processing

*ADNI documentation: <https://adni.loni.usc.edu/quick-start-guide-asset/anatomy2.html>*

*Some bullet points to add to the final paper about things we thought about/did:*

-   *Merging the diagnosis data with biomarker/amyloid data by RID (unique identifier for patients)*

-   *Selecting relevant variables from diagnosis and biomarker/amyloid data set*

-   *What variables do we need for longitudinal sub-model? (Repeated measurement, time followed?)*

-   *What variables do we need for survival sub-model? (Event, time followed?)*

-   *Joint model: time-var: a character string indicating the time variable in the mixed-effects model*

-   *How many observations going from CN/MCI -\> AD? Do we need a multi-state survival model?*

# Results

*Present the results from our analyses. Do not interpret results here! That should be done in the discussion.*

# Discussion

## Main findings/summary

*Summarise the results.*

## Conclusions

*Interpret the results. What did we find?*

## Discussion - Relating to the research question(s)/previous literature

*How do our results help to answer the research question? Anything unexpected/not in accordance with previous studies?*

## Further research

*Suggest what kinds of questions this analysis leaves us with? What more can be done/looked at?*

# Bibliography (APA-style)

Alzheimer’s Disease Neuroimaging Initiative (ADNI) (n.d.). *About ADNI*. <https://adni.loni.usc.edu/about/>

Andrinopoulou, E. R., & Rizopoulos, D. (2016). Bayesian shrinkage approach for a joint model of longitudinal and survival outcomes assuming different association structures. *Statistics in medicine*, 35(26), 4813-4823. <https://doi.org/10.1002/sim.7027>

Biomarkers Definitions Working Group. (2001). Biomarkers and surrogate endpoints: preferred definitions and conceptual framework". *Clinical pharmacology and therapeutics*, 69(3), 89-95. <https://doi.org/10.1067/mcp.2001.113989>

Carrillo, M. C., Dean, R. A., Nicolas, F., Miller, D. S., Berman, R., Khachaturian, Z., Bain, L. J., Schindler, R., Knopman, D., & Alzheimer's Association Research Roundtable (2013). Revisiting the framework of the National Institute on Aging-Alzheimer's Association diagnostic criteria. *Alzheimer's & dementia : the journal of the Alzheimer's Association*, 9(5), 594-601. <https://doi.org/10.1016/j.jalz.2013.05.1762>

Chen, S. D., Huang, Y. Y., Shen, X. N., Guo, Y., Tan, L., Dong, Q., Yu, J. T., & Alzheimer’s Disease Neuroimaging Initiative (2021). Longitudinal plasma phosphorylated tau 181 tracks disease progression in Alzheimer's disease. *Translational psychiatry*, 11(1), 356. <https://doi.org/10.1038/s41398-021-01476-7>

Ghahremani, M., Wang, M., Chen, H.-Y., Zetterberg, H., Smith, E. & Ismail, Z. (2022). Plasma P-Tau181 and Neuropsychiatric Symptoms in Preclinical and Prodromal Alzheimer Disease. *Neurology*, p.10.1212. [10.1212/WNL.0000000000201517](https://doi.org/10.1212/WNL.0000000000201517)

Guo, G., Song, W., Wang, A., Cui, Q., Yang, X., Wang, Y., Ma, Y., Han, H., Li, Z., Zhang, Z., Meng, W., & Wang, S. (2026). Sensitivity comparison of longitudinal cognitive function indicators of Alzheimer’s disease after mild cognitive impairment: a prospective cohort study. *Sci Rep*. <https://doi.org/10.1038/s41598-026-44192-2>

Huang, Y., Chen, J., Xu, L., & Tang, N. S. (2022). Bayesian Joint Modeling of Multivariate Longitudinal and Survival Data With an Application to Diabetes Study. *Frontiers in big data*, 5(1), 812725. <https://doi.org/10.3389/fdata.2022.812725>

Jack, C. R., Andrews, J. S., Beach, T. G., Buracchio, T., Dunn, B., Graf, A., Hansson, O., Ho, C., Jagust, W., McDade, E., Molinuevo, J. L., Okonkwo, O. C., Pani, L., Rafii, M. S., Scheltens, P., Siemers, E., Snyder, H. M., Sperling, R., Teunissen, C. E., & Carrillo, M. C. (2024). Revised criteria for diagnosis and staging of Alzheimer’s disease: Alzheimer’s Association Workgroup. *Alzheimer’s & Dementia*, *20*(8). https://doi.org/10.1002/alz.13859

Kamarudin, A. N., Cox, T., & Kolamunnage-Dona, R. (2017). Time-dependent ROC curve analysis in medical research: Current methods and applications. BMC Medical Research Methodology, 17, 53. <https://doi.org/10.1186/s12874-017-0332-6>

Kyheng, M., Babykina, G., & Duhamel, A. (2025). Joint Latent Class Models: A Tutorial on Practical Applications in Clinical Research. *Stat Med*, 44(8-9):e70047. [10.1002/sim.70047](https://doi.org/10.1002/sim.70047). PMID: 40277369; PMCID: PMC12023844.

Mattsson, N., Cullen, N. C., Andreasson, U., Zetterberg, H., & Blennow, K. (2019). Association Between Longitudinal Plasma Neurofilament Light and Neurodegeneration in Patients With Alzheimer Disease. *JAMA neurology*, 76(7), 791-799. <https://doi.org/10.1001/jamaneurol.2019.0765>

National Cancer Institute (NCI) (n.d.). *Baseline*. <https://www.cancer.gov/publications/dictionaries/cancer-terms/def/baseline>

Peng, R.E. (2022). *Advanced Statistical Computing*. Chapter 7.4.2: Gelman-Rubin Statistic. <https://bookdown.org/rdpeng/advstatcomp/monitoring-convergence.html#gelman-rubin-statistic>

Proust-Lima, C., Saulnier, T., Philipps, V., Traon, A. P., Péran, P., Rascol, O., Meissner, W. G., & Foubert-Samier, A. (2023). Describing complex disease progression using joint latent class models for multivariate longitudinal markers and clinical endpoints. *Statistics in medicine*, 42(22), 3996-4014. <https://doi.org/10.1002/sim.9844> 

Ranganathan, P., Deo, V., & Pramesh, C. S. (2025). Time-to-event analysis. *Perspectives in clinical research*, 16(2), 102-105. <https://doi.org/10.4103/picr.picr_52_25>

Rizopoulos, D. (2012). *Joint Models for Longitudinal and Time-to-Event Data: With Applications in R* (1st ed.). Chapman and Hall/CRC. <https://doi.org/10.1201/b12208>

Scollard, P., Mukherjee, S., Choi, S. E., Lee, M. L., Klinedinst, B., Gibbons, L. E., Trittschuh, E. H., Mez, J., Saykin, A. J., James, B. D., Proust-Lima, C., & Crane, P. K. (2026). Heterogeneous patterns of cognitive decline in Alzheimer's disease across three domains of cognition. *Journal of Alzheimer's disease : JAD*, 110(1), 383-396. <https://doi.org/10.1177/13872877251414975> 

Sweeting, M. J., Barrett, J. K., Thompson, S. G., & Wood, A. M. (2017). The use of repeated blood pressure measures for cardiovascular risk prediction: a comparison of statistical models in the ARIC study. *Statistics in medicine*, 36(28), 4514-4528. <https://doi.org/10.1002/sim.7144>

Therneau, T.M. & Grambsch, P.M (2000). *Modeling Survival Data: Extending the Cox Model*. Springer, Berlin. <https://doi.org/10.1007/978-1-4757-3294-8>

Wang, C., Shen, J., Charalambous, C., & Pan, J. (2024). Modeling biomarker variability in joint analysis of longitudinal and time-to-event data. *Biostatistics* (Oxford, England), 25(2), 577-596. <https://doi.org/10.1093/biostatistics/kxad009>

Wolfe, R. A., & Strawderman, R. L. (1996). Logical and statistical fallacies in the use of Cox regression models. *American journal of kidney diseases : the official journal of the National Kidney Foundation*, 27(1), 124-129. <https://doi.org/10.1016/s0272-6386(96)90039-6>

Wong, P.C., Savonenko, A., Li, T., Price, DL. (2012). 'Neurobiology of Alzheimer’s Disease' in Brady, S.T., Siegel, G.J., Albers, RW., Price, D.L (8 ed.) *Basic neurochemistry : principles of molecular, cellular, and medical neurobiology*. Waltham, Massachusetts; Oxford: Academic Press / Elsevier.

World Health Organization (2025). *Dementia*. [https://www.who.int/news-room/fact-sheets/detal/dementia](https://www.who.int/news-room/fact-sheets/detail/dementia)

-   (2024). *Preferred product characteristics of blood-based biomarker diagnostics for Alzheimer disease*. <https://iris.who.int/server/api/core/bitstreams/43c83ef7-7a34-4b56-ae17-b584f2fc0798/content>

Xu, L. (2021). Bayesian Multivariate Joint Modeling for Skewed-longitudinal and Time-to-event Data. *USF Tampa Graduate Theses and Dissertations*. <https://digitalcommons.usf.edu/etd/9268>

Yuan, M., Lian, S., Li, X., Long, X., Fang, Y., & Alzheimer's Disease Neuroimaging Initiative (ADNI) (2024). Blood biomarkers in dynamic prediction of conversion to Alzheimer's disease: An application of joint modeling. *International journal of geriatric psychiatry*, 39(3), e6079. <https://doi.org/10.1002/gps.6079>

Zou, K. H., Liu, A., Bandos, A. I., Ohno-Machado, L., & Rockette, H. E. (2012). Statistical evaluation of diagnostic performance: Topics in ROC analysis (1st ed.). Chapman and Hall/CRC. <https://doi.org/10.1201/b11031>

# Appendix

# ...

# Pre-processing

-   *There are 52 individuals with two measurements recorded on the same exam date.*

-   *How should we deal with these? Average the measurements? Remove the observations?*

1.  *Describe the measurements, their differences and averages*

-   *Some of the values are quite different, what do we assume by averaging them? Do we introduce bias but just just removing them?*

2.  *Show how different these values are from the nfl and ptau averages (report SD differences in log transformations)*
3.  

```{r}
# checking whether there are any observations that should be removed as per the ADNI leadership guidelines about observations with PTID in the format "381_S_10###"
grep("381_S_10###", diagnosis_data$PTID)
grep("381_S_10###", apoe$PTID)
```

## Selection of required variables

```{r, warning=FALSE}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  dplyr::select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required genotype variables
apoe <- apoe %>%
  dplyr::select(RID, GENOTYPE)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  dplyr::select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))
```

## Checking RID with more than one measurement recorded on the same exam date

For participants with duplicate measurements on the same examination date, differences were observed depending on how they were handled: using the first measurement, the mean, or the last measurement. The mean value was used as the primary approach, and sensitivity analyses will be performed to evaluate the robustness of the findings.

```{r}
## Checking which obs have double VISCODE2 codes
# for ptau data
ptau_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# for nfl data
nfl_data %>% group_by(RID, VISCODE2) %>% filter(n() > 1)

# changing VISCODE2 for RID 1097 to m60_1 and m60_2
ptau_data[ptau_data$RID == 1097, "VISCODE2"] <- c("m48", "m60_2", "m60_1")

nfl_data[nfl_data$RID == 1097, "VISCODE2"] <-  c("m48", "m60_1", "m60_2")
```

```{r}
# Replacing visits with double measurements with visits with mean-value of the measurements as well as saving both measurements for future sensitivity analysis
# Naming the mean as the original variable to do analysis first with this method and then sensitivity analysis with the first and last measurements

# Adding columns for each ptau measurement and their mean and then collapsing down to unique RID and EXAMDATE
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_ptau = first(PLASMAPTAU181),
    last_ptau = last(PLASMAPTAU181),
    PLASMAPTAU181 = mean(PLASMAPTAU181),
    ) %>%
  ungroup() %>%
  distinct()

# Adding columns for each nfl measurement and their mean and then collapsing down to unique RID and EXAMDATE
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  arrange(RID, EXAMDATE) %>%
  mutate(
    first_nfl = first(PLASMA_NFL),
    last_nfl = last(PLASMA_NFL),
    PLASMA_NFL = mean(PLASMA_NFL)
    ) %>%
  ungroup() %>%
  distinct()
```

```{r}
# checking the difference in method for those with 2 obs on the same examdate
# for ptau
head(ptau_data %>%
       filter(PLASMAPTAU181 != first_ptau) %>%
       dplyr::select(PLASMAPTAU181, first_ptau, last_ptau)
     )

# for nfl
head(nfl_data %>%
       filter(PLASMA_NFL != first_nfl) %>%
       dplyr::select(PLASMA_NFL, first_nfl, last_nfl)
     )
```

## Merging nfl and ptau data for longitudinal data set

```{r}
# We merge NFL and ptau data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))
```

## Creating a common EXAMDATE and EXAMDATE_baseline variable

-   Do this now in order to remove observations from diagnosis data which occur long before first biomarker measurements

```{r}
# checking how many observations where the EXAMDATE is different for nfl and ptau
diff <- long_data %>% mutate(diff = as.numeric(EXAMDATE_nfl != EXAMDATE_ptau))
table(diff$diff)
```

```{r}
# filtering out the observations with different EXAMDATE
diff <- diff %>% filter(diff == TRUE)

# checking how many of these are baseline visits
table(diff$VISCODE2)

# # checking which observations have the same examdate for nfl and ptau
# common <- which(long_data$EXAMDATE_nfl %in% long_data$EXAMDATE_ptau)

# # checking the observations with different exam dates
# long_data[-common, ]
```

-   There are 5 observations with time differences greater than 30 days

-   231/115 days: clearly we can't say these are the same date... BUT with setting a new EXAMDATE as the midpoint between dates means the new EXAMDATE is 2 months out from EXAMDATE_nfl and EXAMDATE_ptau

-   Less clear what the case is with a time difference of 31 days, even 49 days

```{r}
# calculating the difference in time between EXAMDATE_nfl and EXAMDATE_ptau
diff <- diff %>% mutate(time_diff = EXAMDATE_nfl - EXAMDATE_ptau)

# fetching the time differences
diff$time_diff 

# checking how many observations have a time difference over 30 days (1 month)
diff %>% filter(abs(time_diff) > 30) %>% dplyr::select(time_diff)
```

### EXAMDATE is EXAMDATE_nfl or EXAMDATE_nfl +- time difference

-   Can create a new variable for a common EXAMDATE in order to define the baseline date from this

```{r}
# testing the method
diff_examdate <- long_data %>%
  mutate(
    time_diff = EXAMDATE_nfl - EXAMDATE_ptau,
    EXAMDATE_new = case_when(
      time_diff == 0 ~ EXAMDATE_nfl,
      TRUE ~ EXAMDATE_ptau + 0.5 * time_diff
    )
  ) %>%
  filter(time_diff != 0) %>%
  relocate(where(~ inherits(.x, "Date")), time_diff, .after = EXAMDATE_new)
```

```{r}
# creating a common EXAMDATE from two cases
# 1: EXAMDATE_nfl == EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl
# 2: EXAMDATE_nfl != EXAMDATE_ptau -> EXAMDATE == EXAMDATE_nfl + the difference between the two dates
# then defining the baseline date from the earliest EXAMDATE 
long_data <- long_data %>%
  group_by(RID) %>%
  mutate(
    EXAMDATE = EXAMDATE_ptau + 0.5 * (EXAMDATE_nfl - EXAMDATE_ptau),
    EXAMDATE_baseline = min(EXAMDATE),
    time_years = as.numeric(EXAMDATE - EXAMDATE_baseline) / 365.25) %>%
  ungroup()


```

### Baseline date is the earliest of EXAMDATE_nfl and EXAMDATE_ptau

```{r}
# dates aren't too different, and only one observation at baseline, but to be sure create the baseline date out of the earlier of the two dates
# long_data <- long_data %>%
#   group_by(RID) %>%
#   mutate(baseline_date = min(EXAMDATE_ptau, EXAMDATE_nfl)) %>%
#   ungroup()
```

## Removing diagnosis observations before biomarker measurements

1.  Use diagnosis observations no longer than 50 days before first biomarker measurements – we can change this number and see what difference it makes

```{r}
# filtering the observations out of the diagnosis data set where EXAMDATE is much before baseline biomarker EXAMDATE
# creating a baseline_dx variable, so we can set time_to_event = 0 if baseline_dx == DEMENTIA
diagnosis_data <- diagnosis_data %>%
  merge(
    long_data %>% dplyr::select(RID, EXAMDATE_baseline), by = "RID"
    ) %>%
  filter(EXAMDATE_dx - EXAMDATE_baseline >= -50) %>%
  distinct() %>%
  group_by(RID) %>%
  arrange(EXAMDATE_dx) %>%
  mutate(
    baseline_dx = first(DIAGNOSIS_dx),
    last_dx = last(DIAGNOSIS_dx)
  ) %>%
  dplyr::select(!EXAMDATE_baseline)
```

## Merging longitudinal and diagnosis data for survival data set

```{r}
# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- long_data %>% left_join(diagnosis_data, by = c("RID"))
```

## Count of missing values

```{r}
# We count missing values per column in long_data: all these are variables we will use in the model, so we can remove with na.omit()
sapply(long_data, function(x) sum(is.na(x)))
```

```{r}
# We count missing values per column in surv_data: VISCODE2.y will not be used in the model, so we need to see how these observations look
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
# There are no rows with missing VISCODE2.y with full data for the model variables, so we can remove all with na.omit()
surv_data %>% filter(is.na(VISCODE2.y))
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

## Length check of longitudinal and survival data

```{r}
# We check the number of individuals in each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data

# # removing the two individuals who are not in the survival data set from the longitudinal data set
long_data <- long_data %>% filter(RID %in% unique(surv_data$RID))

# Rechecking the number of individuals in each dataset 
length(unique(long_data$RID)) # longitudinal data
length(unique(surv_data$RID)) # survival data
```

## Correlation check between biomarkers

```{r}
# Correlation between the biomarkers 
cor(long_data$PLASMAPTAU181, long_data$PLASMA_NFL, use = "complete.obs")
```

## Addition of selected demographic variables (age and gender) and APOE4 status

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  dplyr::select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We define APOE4 status based on number of E4 alleles
apoe <- apoe %>%
  mutate(APOE4_status = case_when(
    GENOTYPE == "4/4" ~ "2 alleles",
    GENOTYPE %in% c("2/4", "3/4") ~ "1 allele",
    TRUE ~ "0 alleles"
  ),
  genotype = as.factor(GENOTYPE)
  )
```

```{r}
# We merge gender with long_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We merge apoe with long_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status", "genotype")], by="RID")
surv_data <- merge(surv_data, apoe[, c("RID", "APOE4_status", "genotype")], by="RID")

# We make sure that gender and apoe4-status are factors
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)
long_data$APOE4_status <- factor(long_data$APOE4_status,
                                 levels = c("0 alleles", "1 allele", "2 alleles"))
surv_data$APOE4_status <- factor(surv_data$APOE4_status,
                                 levels = c("0 alleles", "1 allele", "2 alleles"))

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

### Checking individuals final diagnoses for the full data set

```{r}
# final diagnoses for the full sample size
table(
  (surv_data %>%
     group_by(RID) %>%
     filter(EXAMDATE_dx == last(EXAMDATE_dx)) %>%
     dplyr::select(DIAGNOSIS_dx) %>%
     distinct()
   )$DIAGNOSIS_dx
  )
```

## Selection of participants with at least 3 measurements

```{r}
# We filter individuals with at least 3 measurements of ptau_data
keep <- unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
long_data <- long_data %>% filter(RID %in% keep)
surv_data <- surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

### Checking individuals final diagnoses for the reduced data set

```{r}
# final diagnoses for the reduced sample size
table(
  (surv_data %>%
     group_by(RID) %>%
     filter(EXAMDATE_dx == last(EXAMDATE_dx)) %>%
     dplyr::select(DIAGNOSIS_dx) %>%
     distinct()
   )$DIAGNOSIS_dx
  )
```

## Preparation for Cox PH model

### Construction of global (common) baseline, age at baseline, event and time-to-event variables

```{r}
surv_data <- surv_data %>%
  group_by(RID) %>%
  summarise(
#     carrying these forward to check events and event times are coded correctly
    baseline_dx = unique(baseline_dx),
    last_dx = unique(last_dx),
    
#     baseline EXAMDATE defined as the earlier biomarker EXAMDATE (nfl or ptau)
    EXAMDATE_baseline = min(EXAMDATE_baseline),
    
#     age at baseline visit
    age_at_baseline = age_at_visit[
      EXAMDATE_ptau == min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)]) |
      EXAMDATE_nfl == min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])
      ][1],

#     last dx/nfl/ptau exam
      last_followup = max(c(EXAMDATE_ptau, EXAMDATE_nfl, EXAMDATE_dx), na.rm = TRUE),

#     1 = dementia, 0 = otherwise
      event = as.integer(any(DIAGNOSIS_dx == "Dementia")),

#     0 if dementia dx at baseline, otherwise time to dementia dx or date of censuring
      time_to_event = case_when(
        event == 1 & unique(baseline_dx) == "Dementia" ~ 0,
        event == 1 & unique(baseline_dx) != "Dementia" ~
          as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - EXAMDATE_baseline)/365,
        TRUE ~
          as.numeric(last_followup - EXAMDATE_baseline)/365))
```

```{r}
# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

## Removing individuals with dementia at start

```{r}
# Removing individuals who had a dementia dx at time = 0
surv_data <- surv_data %>% filter(time_to_event > 0)
long_data <- long_data %>% filter(RID %in% unique(surv_data$RID))
```

### Construction of baseline values of each biomarker, to include in the Cox PH models as covariates

```{r}
# Creating a data frame with the first biomarker measurements
baseline <- long_data %>% 
  group_by(RID) %>%
  arrange(RID, EXAMDATE_nfl) %>% 
  mutate(baseline_value_nfl = first(PLASMA_NFL)) %>%
  arrange(RID, EXAMDATE_ptau) %>%
  mutate(baseline_value_ptau = first(PLASMAPTAU181)) %>%
  dplyr::select(RID, baseline_value_ptau, baseline_value_nfl) %>%
  ungroup() %>%
  distinct()
```

```{r}
# merging the survival data with demographics and the baseline biomarker values
surv_data <- surv_data %>%
  merge(demographics_small, by = "RID") %>% 
  merge(apoe %>% dplyr::select(RID, APOE4_status), by = "RID") %>%
  merge(baseline, by = "RID")
```

## Recheck of longitudinal and survival data lengths

```{r}
# We check the number of individuals in each dataset 
length(unique(long_data$RID)) # longitudinal data for ptau
length(unique(surv_data$RID)) # survival data, global test version
```

# Distribution check of each biomarker

1.  **Since cube root transform gives a better Gaussian approximation for ptau, should we consider using this? The downside is that values cannot take negative values, which is possible with the natural logarithm transformation: is this true though - sqrt(4) = +- 2 ... but how does this work in practice?**

```{r}
# We check the distribution of each biomarker
hist(long_data$PLASMAPTAU181, main="PLASMAPTAU181")
hist(long_data$PLASMA_NFL, main="PLASMA_NFL")
```

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)
```

```{r}
# We re-check the distribution of each biomarker  
hist(long_data$log_PLASMAPTAU181, main="log_PLASMAPTAU181")
hist(long_data$log_PLASMA_NFL, main="log_PLASMA_NFL")
```

```{r}
# cube-root transformed
ggplot(long_data, aes(sqrt(sqrt(PLASMA_NFL)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()

# cube-root transformed
ggplot(long_data, aes(sqrt(sqrt(PLASMAPTAU181)))) +
  geom_histogram(binwidth = 0.1, colour = "white") +
  theme_classic()
```

```{r}
# Transforming the baseline values in the survival data
surv_data$baseline_value_ptau <- log(surv_data$baseline_value_ptau)
surv_data$baseline_value_nfl <- log(surv_data$baseline_value_nfl)
```

## Check for outliers

We still keep them, as we assume they are correct observations.

-   Checking for observations either side of the mean of the standardised data shows there are no values 3 sds below the mean, only above.

```{r}
# We identify "extreme" values

# Values greater than 3 sd above ptau mean
long_data[scale(long_data$PLASMAPTAU181) > 3, ]
# Values smaller than 3 sd below ptau mean
long_data[scale(long_data$PLASMAPTAU181) < -3, ]

# Values greater than 3 sd above nfl mean
long_data[scale(long_data$PLASMA_NFL) > 3, ]
# Values smaller than 3 sd below nfl mean
long_data[scale(long_data$PLASMA_NFL) < -3, ]
```

# ...

```{r}
# extracting number of events with time_to_event != 0
table((surv_data %>% filter(time_to_event > 0))$event)
table((surv_data %>% filter(time_to_event == 0))$event)
```

# Longitudinal models

## Comparing longitudinal models

Suggested models:

-   Gamma: raw data could be modelled well with the gamma distribution since takes no negative values, variance increases with mean, and there is a right-skew to the raw distributions

-   Linear mixed effects model with log-transformed data: approximately normal, yet still some skew

-   Linear mixed effects model with cube root transformed data: again, approximately normal, still a bit of skew (why such a low AIC? seems wrong?)

-   GLMM with log-normal distribution

-   GLMM with Student-t distribution: longer tails

### p-tau181

```{r}
# checking the mean and variance for ptau181
mean(long_data$PLASMAPTAU181)
var(long_data$PLASMAPTAU181)
```

```{r}
# GLMM
fit_gamma_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID), 
                   data = long_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_ptau <- lmer(log(PLASMAPTAU181) ~ time_years + (1|RID), 
                data = long_data)

# LME approach with log transformed data
fit_lme_cubed_ptau <- lmer(sqrt(sqrt(PLASMAPTAU181)) ~ time_years + (1|RID), 
                           data = long_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                             data = long_data,
                             family = lognormal(link = "log"))

# Student-t
fit_student_ptau <- glmmTMB(PLASMAPTAU181 ~ time_years + (1|RID),
                            data = long_data,
                            family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_ptau <- simulateResiduals(fit_gamma_ptau)
testResiduals(gamma_residuals_ptau)

lme_residuals_log_ptau <- simulateResiduals(fit_lme_log_ptau)
testResiduals(lme_residuals_log_ptau)

lme_residuals_cubed_ptau <- simulateResiduals(fit_lme_cubed_ptau)
testResiduals(lme_residuals_cubed_ptau)

lognormal_residuals_ptau <- simulateResiduals(fit_lognormal_ptau)
testResiduals(lognormal_residuals_ptau)

student_residuals_ptau <- simulateResiduals(fit_student_ptau)
testResiduals(student_residuals_ptau)

AIC(fit_lme_log_ptau, fit_lme_cubed_ptau, fit_gamma_ptau, fit_lognormal_ptau, fit_student_ptau)
```

### NfL

-   Negative AIC for cube root transformed -\> take the absolute value

```{r}
# checking the mean and variance for nfl
mean(long_data$PLASMA_NFL)
var(long_data$PLASMA_NFL)
```

```{r}
# GLMM
fit_gamma_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID), 
                   data = long_data,
                   family = Gamma(link = "log"))

# LME approach with log transformed data
fit_lme_log_nfl <- lmer(log(PLASMA_NFL) ~ time_years + (1|RID), 
                data = long_data)

# LME approach with log transformed data
fit_lme_cubed_nfl <- lmer(sqrt(sqrt(PLASMA_NFL)) ~ time_years + (1|RID), 
                          data = long_data)

# Lognormal - similar to log-transforming then fitting LME
fit_lognormal_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                             data = long_data,
                             family = lognormal(link = "log"))

# Student-t
fit_student_nfl <- glmmTMB(PLASMA_NFL ~ time_years + (1|RID),
                           data = long_data,
                           family = students.t(df = 3, link = "identity"))

# residual tests for each model
gamma_residuals_nfl <- simulateResiduals(fit_gamma_nfl)
testResiduals(gamma_residuals_nfl)

lme_residuals_log_nfl <- simulateResiduals(fit_lme_log_nfl)
testResiduals(lme_residuals_log_nfl)

lme_residuals_cubed_nfl <- simulateResiduals(fit_lme_cubed_nfl)
testResiduals(lme_residuals_cubed_nfl)

lognormal_residuals_nfl <- simulateResiduals(fit_lognormal_nfl)
testResiduals(lognormal_residuals_nfl)

student_residuals_nfl <- simulateResiduals(fit_student_nfl)
testResiduals(student_residuals_nfl)

AIC(fit_lme_log_nfl, fit_lme_cubed_nfl, fit_gamma_nfl, fit_lognormal_nfl, fit_student_nfl)
```

## For separate joint models

### Fitted LME models

```{r}
# Longitudinal model for p-tau181 with normal distribution 
long_model_ptau <- lme(
  fixed = log_PLASMAPTAU181 ~ time_years + PTGENDER + APOE4_status,
  random = ~ time_years | RID,                    
  data = long_data
)

summary(long_model_ptau)

# Longitudinal model for NfL with normal distribution 
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years + PTGENDER + APOE4_status, 
  random = ~ time_years | RID, # random intercept + slope
  data = long_data
  )

summary(long_model_nfl)

# Mixed object for multivariate joint model
mixed <- list(long_model_ptau, long_model_nfl)
```

```{r}
# Longitudinal model for p-tau181 with t-distribution (identity link)
#library(GLMMadaptive)
#long_model_ptau <- mixed_model(
  #fixed = log_PLASMAPTAU181 ~ time_years_ptau + PTGENDER + APOE4_status,
  #random = ~ time_years_ptau | RID,                    
  #data = long_data_ptau,
      
#family = students.t(df = 3, link = "identity")
#)

#summary(long_model_ptau)
```

### Residual check

```{r}
# residuals of the p-tau181 LME model
res_ptau <- residuals(long_model_ptau, type = "pearson")  

# fitted values of the p-tau181 LME model
fitted_vals_ptau <- fitted(long_model_ptau)  

# plot of residuals vs fitted values 
plot(fitted_vals_ptau, res_ptau,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for p-tau181")
abline(h = 0, col = "red", lty = 2)
```

```{r}
# residuals of the NfL LME model
res_nfl <- residuals(long_model_nfl, type = "pearson")  

# fitted values of the NfL LME model
fitted_vals_nfl <- fitted(long_model_nfl)  

# plot of residuals vs fitted values 
plot(fitted_vals_nfl, res_nfl,
     xlab = "Fitted values",
     ylab = "Residuals",
     main = "Residuals vs Fitted for NFL")
abline(h = 0, col = "red", lty = 2)
```

### QQ plots

```{r}
# For P-tau
qqnorm(res_ptau, main = "QQ plot: p-tau181 residuals")
qqline(res_ptau, col = "red")

# For NFL 
qqnorm(res_nfl, main = "QQ plot: NfL residuals")
qqline(res_nfl, col = "red")
```

# Survival models

## For the different cox PH models and the separate joint models

### Fitted standard Cox PH models (for the joint models)

```{r}
# Cox survival-models, with covariates
surv_model <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + APOE4_status,
  data = surv_data,
  x = TRUE
)

summary(surv_model)
```

### Fitted Cox PH models using baseline values of each biomarker

```{r}
# Cox survival-models, with covariates including baseline values of each biomarker 
# For p-tau181
surv_model_ptau_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau,
  data = surv_data
)

summary(surv_model_ptau_bl)
```

```{r}
# For NfL
surv_model_nfl_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_nfl,
  data = surv_data
)

summary(surv_model_nfl_bl)
```

```{r}
# Inlcuding both p-tau181 and NfL
surv_model_both_bl <- coxph(
  Surv(time_to_event, event) ~
    age_at_baseline + PTGENDER + APOE4_status + baseline_value_ptau + baseline_value_nfl,
  data = surv_data
)

summary(surv_model_both_bl) 
```

### Test of the assumption of proportional hazards

```{r}
##### For the standard Cox PH model
cox.zph(surv_model)

##### For the Cox PH models with baseline values of each biomarker 
# With p-tau181 baseline values
cox.zph(surv_model_ptau_bl)
# With NfL baseline values
cox.zph(surv_model_nfl_bl)
# With both p-tau181 and NfL baseline values
cox.zph(surv_model_both_bl)
```

### Test of time-varying coefficient Cox models for APOE4-status due to evidence of non-proportional hazards

```{r}
# Making apoe4-status numeric
surv_data$APOE4_num <- dplyr::recode(
  surv_data$APOE4_status,
  "0 alleles" = 0,
  "1 allele"  = 1,
  "2 alleles" = 2
)

# We fit a time-varying coefficient Cox model where APOE4 status is modeled as time-varying
surv_model_tv <- coxph(
  Surv(time_to_event, event) ~ age_at_baseline + PTGENDER + 
    APOE4_num + tt(APOE4_num),
  data = surv_data,
  tt = function(x, t, ...) x * log(t + 1)
)

summary(surv_model_tv)
```

```{r}
AIC(surv_model, surv_model_tv)

logLik(surv_model)
logLik(surv_model_tv)
```

Time-varying apoe4-status is not significant in the time-varying Cox PH model (p-value = 0.105). Additionally, It barely gives better model fit (AIC values: 2077 vs 2075. Log likelihood: -1034 vs -1033) –\> Given these results, and to keep interpretation simple, we'll continue with the standard Cox PH model.

# Two separate joint models

## Fitted joint model for p-tau181

```{r}
# Joint model for p-tau181 using package JM (frequentist approach)
joint_model_ptau <- jointModel(
  long_model_ptau,
  surv_model,
  timeVar = "time_years"
)

summary(joint_model_ptau)
```

```{r}
# Joint model for p-tau181 using package JMbayes2 (bayesian approach)
joint_model_ptau <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years",
  n_chains = 3L,
  n_iter = 11000L,
  n_burnin = 1000L
  )

summary(joint_model_ptau)
```

## Fitted joint model for NfL

1.  High rhat for log(shape) with JM

2.  High rhat for sigma with JMBayes2

```{r}
# Joint model for NfL using package JM (frequentist approach)
joint_model_nfl <- jointModel(
  long_model_nfl,
  surv_model,
  timeVar = "time_years"
)

summary(joint_model_nfl)
```

```{r}
# Joint model for NfL using package JMbayes2 (bayesian approach)
joint_model_nfl <- jm(
  surv_model,
  long_model_nfl,
  time_var = "time_years",
  n_chains = 3L,
  n_iter = 11000L,
  n_burnin = 1000L
  )
```

```{r}
summary(joint_model_nfl)
```

### Traceplots

```{r}
traceplot(joint_model_ptau)
traceplot(joint_model_nfl)
```

```{r}
traceplot(joint_model_ptau, parm = "betas")
traceplot(joint_model_ptau, parm = "alphas")
traceplot(joint_model_ptau, parm = "sigmas")
```

### Density plots

```{r}
densplot(joint_model_ptau)
densplot(joint_model_nfl)
```

```{r}
densplot(joint_model_ptau, parm = "betas")
densplot(joint_model_ptau, parm = "alphas")
densplot(joint_model_ptau, parm = "sigmas")
```

# Multivariate joint model

## To check if the models have stable random slopes

```{r}
VarCorr(long_model_ptau)
VarCorr(long_model_nfl)
```

For both models, the slope variance is very low (lme: 0.0011, mixed_model: 0.036), which indicates that individual biomarker trajectories over time show minimal variation --\> We skip the random slope in fForms:

## fForms

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )      
```

## Fitted multivariate joint model

```{r}
multi_joint_model <- jm(surv_model, mixed ,
                        time_var = "time_years",
                        functional_forms = fForms,
                        n_chains = 3L,
                        n_iter = 20000L,
                        n_burnin = 2000L
                        )

summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival
```

# ...

# Latent Class Mixed Models

Choice of p-value <https://pmc.ncbi.nlm.nih.gov/articles/PMC5059270/>

# LCMM: NFL

Latent class mixed models for NFL

## M0 (1 latent class)

First, a one class linear mixed effects model with a random intercept was estimated. This model has statistically significant coefficients for the intercept and slope for the fixed effects ($p < 0.05$) and the variance for the intercept was given to be $0.196$.

A model with both a random slope and intercept on the longitudinal NfL measurements gave statistically significant coefficients for the fixed effects (p \< 0.05) and variances for the intercept and slope of $0.184$ and $0.001$ respectively. The slope variance was small but non-zero.

Comparison of $BIC$ and the log-likelihoods of these two models show a small but

```{r}
hlme_m0_1 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_1)
```

```{r}
hlme_m0_2 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_2)
```

### Determining the random effects

The $BIC$ is reduced and the log-likelihood increases when adding a random slope to the linear mixed effects model. Additionally, testing whether the improvement in the model with the random slope is statistically significant with a likelihood ratio tests gave $\lambda_{LR} = 16.09$ on $2$ degrees of freedom ($p < 0.05$).

```{r}
summarytable(hlme_m0_1, hlme_m0_2)
```

$$
\lambda_{LR} = - 2 \left\{ \ell(\theta_0) - \ell(\widehat\theta) \right\} \quad \text{where, under } H_0 \quad \lambda_{LR} \sim \chi^2
$$

```{r}
# log likelihood for the full model
l = hlme_m0_2$loglik

# log likelihood for the reduced model
l_0 = hlme_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M2 (2 latent classes)

The linear mixed effects model was extended to model 2 classes with a latent class mixed effects model. One assumption of latent class mixed models is that the sample population, $N$ is comprised of $G$ latent, unobserved heterogenous classes. Modelling the longitudinal NfL measurements with a 2 class linear mixed effects model produced insignificant coefficients ($p > 0.05$). From this, it can be concluded that there is no evidence for 2 latent classes. Comparing $BIC$ for the 1 class ($1638$) and 2 class ($1658$) models supports the conclusion that there are no latent classes in the data. Additionally, the entropy of the 2 class model was calculated to be $0.0003$, which is further evidence against 2 latent classes. The posterior probabilities were extracted from the 2 class model and the distributions of the probabilities for class 1 and class 2 are visualised below. The posterior probabilities cluster around $0.50$, showing indistinguishable classification between class 1 and class 2.

```{r}
hlme_m1 <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ time_years,
    random = ~ time_years,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = long_data,
    B = random(hlme_m0_2)
  )

summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

## Comparing M0 and M1

```{r}
summarytable(hlme_m0_2, hlme_m1, which = c("G", "loglik", "BIC", "entropy"))
```

# LCMM: PTAU

Latent class mixed models for PTAU

## M0 (1 latent class)

```{r}
hlme_m0_1 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_1)
```

```{r}
hlme_m0_2 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0_2)
```

### Determining the random effects

Comparing the $BIC$ from a model with random intercept versus a model with both random intercept and slope and the results of a log-likelihood test give support to the reduced model with only a random intercept in the longitudinal model. Therefore, only a random intercept will be used when modelling the longitudinal ptau-181 data.

```{r}
summarytable(hlme_m0_1, hlme_m0_2)
```

```{r}
# log likelihood for the full model
l = hlme_m0_2$loglik

# log likelihood for the reduced model
l_0 = hlme_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M2 (2 latent classes)

A latent mixed effects model with 2 classes was estimated on the longitudinal ptau181 data which produced significant coefficients for the longitudinal model but not for the class-membership model. The $BIC$ increased when compared to the 1 class model and the posterior probabilities show no clear differentiation between class 1 and class 2, with the average posterior probability being $0.50$ for both classes.

```{r}
hlme_m1 <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ 1,
    ng = 2,
    data = long_data,
    B = random(hlme_m0_1)
  )
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
mean(pp$prob1)
mean(pp$prob2)
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

## Comparing M0 and M1

```{r}
summarytable(hlme_m0_1, hlme_m1, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

# LCMM: NFL (including classmb covariates)

Latent class mixed models for NFL including covariates in the class model

## M0 (1 latent class)

```{r}
# updating the 1 class model
hlme_m0 <- hlme(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0)
```

## M2 (2 latent classes)

The 2 class model with covariates $APOE4\_status$ and $age\_at\_visit$ is a significantly better fit than the 1 class model according to a likelihood ratio test and comparison of $BIC$ for the two models.

### Including PTGENDER

```{r}
hlme_m1_PTGENDER <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ time_years,
    random = ~ time_years,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit + PTGENDER,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1_PTGENDER)
```

### Excluding PTGENDER

```{r}
hlme_m1_nfl_classmb <- hlme(
    fixed = log_PLASMA_NFL ~ time_years,
    mixture = ~ time_years,
    random = ~ time_years,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1_nfl_classmb)
```

### Assessment of covariate PTGENDER

A likelihood ratio test for the two models shows that there is a statistically significant increase in performance for the reduced model without gender as a covariate. It is therefore the reduced 2 class model which we compare with the 1 class model henceforth.

```{r}
# log likelihood for the full model
l = hlme_m1_PTGENDER$loglik

# log likelihood for the reduced model
l_0 = hlme_m1_nfl_classmb$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

### Posterior probabilities

```{r}
postprob(hlme_m1_nfl_classmb)
```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1_nfl_classmb$pprob
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

## Comparing M0 and M1

```{r}
summarytable(hlme_m0, hlme_m1_nfl_classmb, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

# LCMM: PTAU (including classmb covariates)

Latent class mixed models for PTAU including covariates in the class model

## M0 (1 latent class)

```{r}
# updating the 1 class model
hlme_m0 <- hlme(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(hlme_m0)
```

## M2 (2 latent classes)

TO CHECK: difference between genotype in latent class file

```{r}
levels(long_data$APOE4_status)
```

### Including PTGENDER

```{r}
hlme_m1_PTGENDER <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit + PTGENDER,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1_PTGENDER)
```

### Excluding PTGENDER

```{r}
hlme_m1 <- hlme(
    fixed = log_PLASMAPTAU181 ~ time_years,
    mixture = ~ 1,
    random = ~ 1,
    subject = "RID",
    classmb = ~ APOE4_status + age_at_visit,
    ng = 2,
    data = long_data,
    B = random(hlme_m0)
  )
```

```{r}
summary(hlme_m1)
```

### Posterior probabilities

```{r}
postprob(hlme_m1)
```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- hlme_m1$pprob
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 2") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

## Comparing M0 and M1

```{r}
summarytable(hlme_m0, hlme_m1, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

# MLCMM: NFL & PTAU

Multivariate latent class mixed models with both NFL and PTAU

## M0 (1 latent class)

Some support from the likelihood ratio test that the full model is the better fit for the multivariate mixed effects model. However, there is only a very modest decrease in $BIC$ for the model with both random intercept and slope. Considering these two points, a 2 class model with only random intercept will be compared against a model with a random intercept and slope in order to better assess which random effects structure is best.

```{r}
mlcmm_m0_1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(mlcmm_m0_1)
```

```{r}
mlcmm_m0_2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = long_data
)
```

```{r}
summary(mlcmm_m0_2)
```

### Determining the random effects

```{r}
summarytable(mlcmm_m0_1, mlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = mlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = mlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M2 (2 latent classes)

```{r}
mlcmm_m1_1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = long_data,
  B = random(mlcmm_m0_1)
)
```

```{r}
summary(mlcmm_m1_1)
```

```{r}
mlcmm_m1_2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = long_data,
  B = random(mlcmm_m0_2)
)
```

```{r}
summary(mlcmm_m1_2)
```

### Determining the random effects

Interesting: log likelihood is increased for the full model, but $BIC$ increases. Even more uncertain results -\> choose the simpler model (parsimony - look into this)

```{r}
summarytable(mlcmm_m1_1, mlcmm_m1_2)
```

```{r}
# log likelihood for the full model
l = mlcmm_m1_2$loglik

# log likelihood for the reduced model
l_0 = mlcmm_m1_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

### Posterior probabilities

```{r}
postprob(mlcmm_m1_1)
```

## M2 (2 latent classes) (including class covariates)

```{r}
mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  subject = "RID",
  classmb = ~ APOE4_status + age_at_visit,
  ng = 2,
  data = long_data,
  B = random(mlcmm_m0_1)
)
```

```{r}
summary(mlcmm_m1)
```

### Posterior probabilities

```{r}
postprob(mlcmm_m1)
```

## Visualisation

```{r}
# saving the posterior probabilities into a data frame
pp <- mlcmm_m1$pprob
```

### Distribution of posterior probabilities (class 1)

```{r}
# plotting posterior probabilities for class 1
ggplot(pp, aes(x = prob1)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 1") +
  theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
ggplot(pp, aes(x = prob2)) + 
  geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
  labs(title = "Posterior Probabilities for Class 2") +
  theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
ggplot(pp, aes(prob1, prob2)) +
  geom_point(alpha = .8) +
  geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
  labs(x = "P(Class 1)", y = "P(Class 2)") +
  theme_classic()
```

## Comparing M0 and M1

cannot run LT tests on different class models (1 class vs 2 class) as they are not nested in the typical sense

```{r}
summarytable(mlcmm_m0_1, mlcmm_m1, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

# JLCMM: NFL

1.  Posterior probabilities show good classification

2.  Score test statistic for conditional independence is significant with 1 class and not significant with 2 classes: evidence for 2 classes

3.  No covariates in the survival model OR the class membership model, evidence for latent classes?

```{r}
# merging the longitudinal and survival data for jlcmm
jlcmm_data <- surv_data %>%
  dplyr::select(RID, event, time_to_event) %>%
  merge(long_data, by = "RID")
```

## M0 (1 latent class)

```{r}
jlcmm_m0_1 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_1)
```

```{r}
jlcmm_m0_2 <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1 + time_years,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0_2)
```

### Determining the random effects

```{r}
summarytable(jlcmm_m0_1, jlcmm_m0_2)
```

```{r}
# log likelihood for the full model
l = jlcmm_m0_2$loglik

# log likelihood for the reduced model
l_0 = jlcmm_m0_1$loglik

# computing the LR test statistic
LR = - 2 * (l_0 - l)

# computing the p-value given the null hypothesis is true (the reduced model is the true model)
pchisq(LR, 2, lower.tail = FALSE)
LR
```

## M1 (2 latent classes)

```{r}
jlcmm_m1_nfl <- jlcmm(
  fixed = log_PLASMA_NFL ~ time_years,
  random = ~ 1 + time_years,
  mixture = ~ time_years,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0_2)
)
```

```{r}
summary(jlcmm_m1_nfl)
```

### Posterior probabilities

```{r}
postprob(jlcmm_m1_nfl)
```

## Visualisation

1.  Need to check probYT1 and probYT2

```{r}
# saving the posterior probabilities into a data frame
# pp <- jlcmm_m1$pprob
# pp1 <- pp %>% filter(class == 1)
# pp2 <- pp %>% filter(class == 2)
```

### Distribution of posterior probabilities (class 1)

1.  Check that YT1 is for the joint model, not the longitudinal model only

```{r}
# plotting posterior probabilities for class 1
# ggplot(pp1, aes(x = probYT1)) + 
#   geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
#   labs(title = "Posterior Probabilities for Class 1") +
#   theme_classic()
```

### Distribution of posterior probabilities (class 2)

```{r}
# plotting posterior probabilities for class 2
# ggplot(pp2, aes(x = probYT1)) + 
#   geom_histogram(bins = 30, colour = "white", fill = "cornflowerblue") +
#   labs(title = "Posterior Probabilities for Class 2") +
#   theme_classic()
```

### Class certainty scatterplot

```{r}
# scatterplot of class certainty
# ggplot(pp, aes(prob1, prob2)) +
#   geom_point(alpha = .8) +
#   geom_abline(slope = -1, intercept = 1, linetype = "dashed") +
#   labs(x = "P(Class 1)", y = "P(Class 2)") +
#   theme_classic()
```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m1_nfl, which = "survival")
```

### Survival curves by class

```{r}
# longitudinal trajectory by class
plot(jlcmm_m1_nfl, var.time = "time_years", which = "fit")
```

## Comparing M0 and M1

```{r}
summarytable(jlcmm_m0_2, jlcmm_m1_nfl, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

# JLCMM: PTAU

## M0 (1 latent class)

```{r}
jlcmm_m0 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data
)
```

```{r}
summary(jlcmm_m0)
```

## M1 (2 latent classes)

```{r}
jlcmm_m1 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 2,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0)
)
```

```{r}
summary(jlcmm_m1)
```

### Posterior probabilities

```{r}
postprob(jlcmm_m1)
```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}

```

### Distribution of posterior probabilities (class 2)

```{r}

```

### Class certainty scatterplot

```{r}

```

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m1, which = "survival")
```

### Longitudinal trajectories by class

```{r}
# longitudnal trajectories by class
plot(jlcmm_m1, var.time = "time_years", which = "fit")
```

## Comparing M0 and M1

```{r}
summarytable(jlcmm_m0, jlcmm_m1, which = c("G", "loglik", "BIC", "entropy"))
```

### Adjusted LRT

```{r}

```

## M2 (3 latent classes)

*Things to assess:*

1.  *Fit 4 class model -\> does BIC improve any more or plateau?*

2.  *Score test now not significant compared with the two class model*

3.  *VERY small class 1*

4.  *CLass 2 and 3 are not cleanly separated (look at posterior probabilities)*

5.  Weibull class 3 parameters not significant at 0.05 level

6.  *Plot survival curves (how?)*

### gridsearch()

```{r}
# gridsearch(m = jlcmm(
#   fixed = log_PLASMAPTAU181 ~ time_years_ptau,
#   random = ~ 1,
#   mixture = ~ 1,
#   classmb = ~ 1,
#   ng = 2,
#   subject = "RID",
#   survival = Surv(time_to_event_ptau, event_ptau) ~ 1,
#   data = jlcmm_data_ptau,
#   B = random(jlcmm_m0)),
#   minit = jlcmm_m1,
#   rep = 50,
#   maxiter = 30)

```

```{r}
jlcmm_m2 <- jlcmm(
  fixed = log_PLASMAPTAU181 ~ time_years,
  random = ~ 1,
  mixture = ~ 1,
  classmb = ~ 1,
  ng = 3,
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  data = jlcmm_data,
  B = random(jlcmm_m0)
  )
```

```{r}
summary(jlcmm_m2)
```

```{r}
postprob(jlcmm_m2)
```

## Visualisation

### Survival curves by class

```{r}
# survival curves by class
plot(jlcmm_m2, which = "survival")
```

### Longitudinal trajectories by class

```{r}
# longitudnal trajectories by class
plot(jlcmm_m2, var.time = "time_years", which = "fit")
```

## Comparing M0, M1, M3

```{r}
summarytable(jlcmm_m0, jlcmm_m1, jlcmm_m2,
             which = c("G", "loglik", "BIC", "entropy"))
```

```{r}
# saving which class each individual was predicted to be in
class_preds <- jlcmm_m2$pprobY

# merging with genotype data
class_preds <- class_preds %>%
  merge(apoe %>% dplyr::select(RID, APOE4_status), by = "RID")

# checking the distribution of genotype by class
table(class_preds$class, class_preds$APOE4_status)
```

# MJLCMM: NFL & PTAU

TWO CLASS MULTIVARIATE LATENT CLASS MIXED MODEL -\> significant

```{r}
# # merging the longitudinal and survival data for NfL
# jlcmm_data_multi <- surv_test_1 %>%
#   dplyr::select(RID, event, time_to_event) %>%
#   merge(long_data, by = "RID") %>%
#   filter(time_to_event != 0)
```

```{r}
# fitting multivariate latent class model with 1 class and 2 classes with jlcmm_data_multi

mlcmm_m0 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  subject = "RID",
  data = jlcmm_data
)

mlcmm_m1 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)

mlcmm_m2 <- multlcmm(
  fixed = log_PLASMA_NFL + log_PLASMAPTAU181 ~ time_years,
  random = ~ time_years,
  mixture = ~ time_years,
  subject = "RID",
  classmb = ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mlcmm_m0)
)
```

## M0 (1 latent class)

```{r}
mult_jlcmm_m0 <- mpjlcmm(
  longitudinal = list(mlcmm_m0),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 1,
  data = jlcmm_data
)
```

```{r}
summary(mult_jlcmm_m0)
```

## M1 (2 latent classes)

```{r}
mult_jlcmm_m1 <- mpjlcmm(
  longitudinal = list(mlcmm_m1),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 2,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
)
```

```{r}
summary(mult_jlcmm_m1)
```

### Posterior probabilities

```{r}
postprob(mult_jlcmm_m1)
```

## Visualisation

### Distribution of posterior probabilities (class 1)

```{r}

```

### Distribution of posterior probabilities (class 2)

```{r}

```

### Class certainty scatterplot

```{r}

```

### Survival curves by class

```{r}
plot(mult_jlcmm_m1, which = "survival")
```

```{r}
# plot(mult_jlcmm_m1, var.time = "time_years", which = "fit")
```

## Comparing M0 and M1

```{r}
summarytable(mult_jlcmm_m0, mult_jlcmm_m1, which = c("G", "loglik", "BIC", "entropy"))
```

## M2 (3 latent classes)

```{r}
mult_jlcmm_m2 <- mpjlcmm(
  longitudinal = list(mlcmm_m2),
  subject = "RID",
  survival = Surv(time_to_event, event) ~ 1,
  ng = 3,
  data = jlcmm_data,
  B = random(mult_jlcmm_m0)
  )
```

```{r}
summary(mult_jlcmm_m2)
```

### Posterior probabilities

```{r}
postprob(mult_jlcmm_m2)
```

## Visualisation

### Survival curves by class

```{r}
# survival curves by class
plot(mult_jlcmm_m2, which = "survival")
```

```{r}
# # longitudnal trajectories by class
# plot(mult_jlcmm_m2, outcome = 1, which = "fit")
```

## TO DO

Check sensitivity to starting values by using gridsearch()

External validation: what do these classes look like? Do they just represent different genotypes? How do we check this and are they truly valid LATENT classes?

Adjusted LRT (Lo-Mendell-Rubin test)

Bootstrap LRT???

# Comparing models using time-dependent AUC

*For the Cox PH models = using timeROC, from the package timeROC*

*For the joint models = using tvAUC, from the package JMbayes2*

## Defining evaluation time points for time-dependent AUC comparison

```{r}
# We define evaluation time points for time-dependent AUC comparison
times <- seq(1, 10, by = 0.5)
```

## Time-dependent AUC for fitted Cox PH models

```{r}
# We define a function to compute time-dependent AUC for Cox models using timeROC
# We use the linear predictor as the risk score and survival outcome from the dataset
#get_cox_auc <- function(model, data, times, time_var, event_var) {
#  
#  # Extracting survival outcome variables dynamically
#  T <- data[[time_var]]
#  delta <- data[[event_var]]
#  
#  # Linear predictor
#  lp <- predict(model, newdata = data, type = "lp")
#  
#  # Time-dependent AUC
#  auc <- timeROC(
#    T = T,
#    delta = delta,
#    marker = lp,
#    cause = 1,
#    times = times,
#    iid = TRUE
#  )$AUC
#  
#  return(auc)
#}
```

```{r}
# We compute time-dependent AUC for Cox model with p-tau181
#auc_ptau_bl <- get_cox_auc(
#  surv_model_ptau_bl,
#  surv_data_ptau,
#  times,
#  time_var = "time_to_event_ptau",
#  event_var = "event_ptau"
#)
#
# We compute time-dependent AUC for Cox model with NfL
#auc_nfl_bl <- get_cox_auc(
#  surv_model_nfl_bl,
#  surv_data_nfl,
#  times,
#  time_var = "time_to_event_nfl",
#  event_var = "event_nfl"
#)
#
# We compute time-dependent AUC for Cox model with both p-tau181 and NfL (combined model)
#auc_both1_bl <- get_cox_auc(
#  surv_model_both1_bl,
#  surv_data_ptau,
#  times,
#  time_var = "time_to_event_ptau",
#  event_var = "event_ptau"
#)
#
#auc_both2_bl <- get_cox_auc(
#  surv_model_both2_bl,
#  surv_data_nfl,
#  times,
#  time_var = "time_to_event_nfl",
#  event_var = "event_nfl"
#)
```

## Time-dependent AUC for fitted joint models

*From: <https://cloud.r-project.org/web/packages/JMbayes2/JMbayes2.pdf>*

*The function tvAUC requires the following:*

-   *"newdata = a data.frame that contains the longitudinal and covariate information for the subjects for which prediction of survival probabilities is required. The names of the variables in this data.frame must be the same as in the data.frames that were used to fit the linear mixed effects and the event process model that were supplied as the two first argument of jm."*

```{r}
# We construct prediction datasets for each joint model by merging the corresponding longitudinal and survival data
#newdata_ptau <- long_data_ptau %>%
#  left_join(
#    surv_data_ptau %>%
#      dplyr::select(RID, age_at_baseline_ptau, time_to_event_ptau, event_ptau),
#    by = "RID"
#  )
#
#newdata_nfl <- long_data_nfl %>%
#  dplyr::left_join(
#    surv_data_nfl %>%
#      dplyr::select(RID, age_at_baseline_nfl, time_to_event_nfl, event_nfl),
#    by = "RID"
#  )
#
#newdata_multi <- long_data_multi %>%
#  dplyr::left_join(
#    surv_data_multi %>%
#      dplyr::select(RID, age_at_baseline, time_to_event, event),
#    by = "RID"
#  )
```

```{r}
# We define a function to compute time-dependent AUC for joint models using tvAUC
#We extract the AUC value from the tvAUC output for each horizon time point
#get_jm_auc <- function(model, newdata, times) {
#  
#  aucs <- sapply(times, function(t) {
#    
#    tmp <- tvAUC(
#      model,
#      newdata = newdata,
#      Tstart = 0,
#      Thoriz = t
#    )
#    
#    tmp$auc   
#  })
#  
#  return(as.numeric(aucs))
#}
```

```{r}
# We compute time-dependent AUC for the joint model including p-tau181 using tvAUC
#auc_ptau_jm <- get_jm_auc(
#  joint_model_ptau,
#  newdata_ptau,
#  times
#)
#
# We compute time-dependent AUC for the joint model including NfL using tvAUC
#auc_nfl_jm <- get_jm_auc(
#  joint_model_nfl,
#  newdata_nfl,
#  times
#)
#
# We compute time-dependent AUC for the multivariate joint model using tvAUC
#auc_multi_jm <- get_jm_auc(
#  multi_joint_model,
#  newdata_multi,
#  times
#)
```

## Time dependent AUC for latent class joint models

Landmark: the time at which predictions are made

Horizon: the predict event within $h$ years

Threshold: e.g if risk \> 0.5 -\> conversion

ROC evaluates the model (sensitivity and specificity) at all possible thresholds, plotting sensitivity (y axis) vs 1 - specificity (x axis)

We can fix the landmark, or the horizon

Fixing the landmark, varying the horizon: how well does the model predict over different horizons? Does prediction accuracy change with horizon, e.g model predicts well over short horizons but not long.

Fiing the horizon, varying the landmark: how well does the model predict conversion within $h$ years as more longitudinal data is gathered?

#### Function for extracting time-dependent AUC

1.  Confidence intervals

2.  Overall model discrimination

3.  Class specific discrimination

## Table of the results

```{r}
# Time-dependent AUC for Cox and joint models summarized at clinically relevant time points
#results <- data.frame(
#  Model = c("Cox p-tau181", "Cox NfL", "Cox both1", "Cox both2",
#            "JM p-tau181", "JM NfL", "JM both"),
#  
#  AUC_1y  = c(
#    auc_ptau_bl[2],
#    auc_nfl_bl[2],
#    auc_both1_bl[2],
#    auc_both2_bl[2],
#    auc_ptau_jm[2],
#    auc_nfl_jm[2],
#    auc_multi_jm[2]
#  ),
#  
#  AUC_3y  = c(
#    auc_ptau_bl[6],
#    auc_nfl_bl[6],
#    auc_both1_bl[6],
#    auc_both2_bl[6],
#    auc_ptau_jm[6],
#    auc_nfl_jm[6],
#    auc_multi_jm[6]
#  ),
#  
#  AUC_5y  = c(
#    auc_ptau_bl[10],
#    auc_nfl_bl[10],
#    auc_both1_bl[10],
#    auc_both2_bl[10],
#    auc_ptau_jm[10],
#    auc_nfl_jm[10],
#    auc_multi_jm[10]
#  ),
#  
#  AUC_10y = c(
#    auc_ptau_bl[19],
#    auc_nfl_bl[19],
#    auc_both1_bl[19],
#    auc_both2_bl[19],
#    auc_ptau_jm[19],
#    auc_nfl_jm[19],
#    auc_multi_jm[19]
#  )
#)
#
#results[, -1] <- round(results[, -1], 3)
#
#results
```

## Plot of the results

```{r}
# We reshape all AUC results into long format for plotting
#plot_data <- data.frame(
#  Time = rep(times, times = 7),
#  Model = rep(c("Cox p-tau", "Cox NfL", "Cox both1", "Cox both2", "JM p-tau", "JM NfL", "JM multi"), each = length(times)),
#  AUC = c(
#    auc_ptau_bl,
#    auc_nfl_bl,
#    auc_both1_bl,
#    auc_both2_bl,
#    auc_ptau_jm,
#    auc_nfl_jm,
#    auc_multi_jm
#  )
#)
#
#ggplot(plot_data, aes(x = Time, y = AUC, color = Model)) +
#  geom_line(linewidth = 1) +
#  geom_point(size = 2) +
#  ylim(0, 1) +
#  labs(
#    title = "Time-dependent AUC comparison",
#    x = "Time (years)",
#    y = "AUC"
#  )
```

Sensitivity anaysis: how does including those with only one measurement affect the precision?
