---
title: "Untitled"
format: pdf
editor: visual
---

# UPENN-PLASMA 12 MARS 2026, men inklusive de med bara 1 measurement

```{r, warning=FALSE}
# loading libraries
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(corrr) 
```

```{r}
# read data Milou
library(readr)
plasma_data <- read.csv("UPENN_PLASMA_FUJIREBIO_QUANTERIX_27Mar2026.csv")

# # We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
# 
# # We load demographics data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG

# We load APOE data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe_data <- APOERES
```

```{r}
# read data Charley
library(ADNIMERGE2)

# dx data
diagnosis_data = DXSUM

# demographics
demographics = PTDEMOG
```

## 1. We select only the required variables and merge them

```{r}
library(dplyr)
# selecting the relevant variables -> why these ones???
plasma_data = plasma_data %>%
  select("RID", "VISCODE2", "EXAMDATE", "pT217_F",
         "AB42_F", "AB40_F", "AB42_AB40_F", "pT217_AB42_F", "NfL_Q", "GFAP_Q")

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = plasma_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx")) %>%
  mutate(dx = case_when(DIAGNOSIS == "Dementia" ~ 1, TRUE ~ 0))
```

## 2. We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# We merge age and gender into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
)

# We make sure that gender is categorical (factor)
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# 5. Konvertera EXAMDATE_plasma till Date (hanterar både character och factor)
if (is.factor(merged_data$EXAMDATE_plasma)) {
  merged_data$EXAMDATE_plasma <- as.character(merged_data$EXAMDATE_plasma)
}
merged_data$EXAMDATE_plasma <- as.Date(merged_data$EXAMDATE_plasma, format="%Y-%m-%d")

# 6. Skapa age_at_visit
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_plasma, "%Y")) - merged_data$PTDOBYY
```

## 3. We filter participants with at least one measurement

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((plasma_data %>% group_by(RID) %>% filter(n() > 0))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 2 measurements 
length(keep)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_plasma)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event) + lägger till age och gender 
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_plasma)],
    PTGENDER = PTGENDER[which.min(EXAMDATE_plasma)],
    time_to_event = case_when(

    any(DIAGNOSIS == "Dementia") ~
      as.numeric(min(EXAMDATE_plasma[DIAGNOSIS == "Dementia"]) -
                   min(c(EXAMDATE_plasma, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_plasma, EXAMDATE_dx)) -
                    min(c(EXAMDATE_plasma, EXAMDATE_dx)))) / 365,

  event = as.integer(any(DIAGNOSIS == "Dementia"))
) %>% ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_plasma - baseline_date) / 365.25)
```

## 6. We check the distribution of each relevant variable for multivariate joint model and transform

```{r}
# Vi kollar efter outliers 
boxplot(merged_data$pT217_F, main="pT217_F")
boxplot(merged_data$AB42_F, main="AB42_F")
boxplot(merged_data$AB40_F, main="AB40_F")
boxplot(merged_data$pT217_AB42_F, main="pT217_AB42_F")
boxplot(merged_data$AB42_AB40_F, main="AB42_AB40_F")
boxplot(merged_data$NfL_Q, main="NfL_Q")
boxplot(merged_data$GFAP_Q, main="GFAP_Q")
```

```{r}
# Vi transformerar dem
merged_data$log_pT217_F <- log(merged_data$pT217_F)
merged_data$log_AB42_F <- log(merged_data$AB42_F)
merged_data$log_AB40_F <- log(merged_data$AB40_F)
merged_data$log_pT217_AB42_F <- log(merged_data$pT217_AB42_F)
merged_data$log_AB42_AB40_F <- log(merged_data$AB42_AB40_F)
merged_data$log_NfL_Q <- log(merged_data$NfL_Q)
merged_data$log_GFAP_Q <- log(merged_data$GFAP_Q)
```

## 7. We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
length(unique(merged_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

```{r}
# Gör RID till character (säkerhet)
merged_data$RID <- as.character(merged_data$RID)
surv_data$RID   <- as.character(surv_data$RID)

# Hitta gemensamma ID:n
common_IDs <- intersect(unique(merged_data$RID), unique(surv_data$RID))

# Behåll endast gemensamma individer
merged_data <- merged_data %>% filter(RID %in% common_IDs)
surv_data   <- surv_data %>% filter(RID %in% common_IDs)

# Kontroll
length(unique(merged_data$RID))  # ska matcha
length(unique(surv_data$RID))    # ska matcha
```

# Longitudinal model

## We check correlation between time_years and age_at_visit

```{r}
cor(merged_data$age_at_baseline, merged_data$age_at_visit)
```

--\> Using both time_years + age_at_visit is redundant since we have age_at_baseline in the coxph model: if we omit age_at_visit variable from the lme model, the joint model will be stable.

```{r}
# Longitudinal model
long_model <- lme(
  log_pT217_F ~ time_years + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline plasma", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

# Joint model

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```

# Multivariate model

```{r}
library(nlme)

fm_pT217 <- lme(log_pT217_F ~ time_years + PTGENDER.x,
                random = ~ time_years | RID,
                data = merged_data)

fm_ABratio <- lme(log_AB42_AB40_F ~ time_years + PTGENDER.x,
                  random = ~ time_years | RID,
                  data = merged_data)

fm_NfL <- lme(log_NfL_Q ~ time_years + PTGENDER.x,
              random = ~ time_years | RID,
              data = merged_data)
```

```{r}
Mixed <- list(fm_pT217, fm_ABratio, fm_NfL)
```

```{r}
# Har modellen stabila random slopes?
VarCorr(fm_pT217)
VarCorr(fm_ABratio)
VarCorr(fm_NfL)
# Om slope-variansen är runt 0 → den biomarkören bidrar nästan inget i en joint model
```

```{r}
fForms <- list(
  "log_pT217_F"   = ~ value(log_pT217_F) + slope(log_pT217_F),   # tau
  "log_AB42_AB40_F" = ~ value(log_AB42_AB40_F),                  # amyloid 
  "log_NfL_Q"     = ~ value(log_NfL_Q) + slope(log_NfL_Q))      # neurodegeneration
```

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)

# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

```{r}
multi_joint_model <- jm(surv_model, Mixed,
                  time_var = "time_years",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 10000L,
                  n_burnin = 2000L
                  )
summary(multi_joint_model)
```

```{r}
summary <- summary(multi_joint_model)
summary$Survival 
```

```{r}
plot(Mixed[[1]])
plot(Mixed[[2]])
plot(Mixed[[3]])
```

P-värden och Rhat-värden vid 10000L i n_iter:

```         
                                 P     Rhat
value(log_pT217_F)     0.000000000 1.043781 
slope(log_pT217_F)     0.007555556 1.136719 
value(log_AB42_AB40_F) 0.802666667 1.021286 
value(log_NfL_Q)       0.138666667 1.057651 
slope(log_NfL_Q)       0.000000000 1.075836
```

P-värden och Rhat-värden vid 15000L i n_iter:

```         
            Mean     StDev        2.5%      97.5%

value(log_pT217_F) 1.1645012 0.1377146  0.90610472  1.4276549
slope(log_pT217_F) -2.2869174 0.8908001 -3.94984964 -0.5624709
value(log_AB42_AB40_F) -0.1201799 0.4556441 -0.97080770  0.7763266
value(log_NfL_Q)   0.2368298 0.1606073 -0.07445107  0.5634655
slope(log_NfL_Q)  4.1828948 1.3029137  1.86158226  6.9034204

                                 P     Rhat
value(log_pT217_F)     0.000000000 1.043781 
slope(log_pT217_F)     0.007555556 1.136719
value(log_AB42_AB40_F) 0.802666667 1.021286
value(log_NfL_Q)       0.138666667 1.057651
slope(log_NfL_Q)       0.000000000 1.075836
```

--\> Ändrade inte

Multivariate joint-modellen har dock stora CI –\> Beror förmodligen på lilla n-storleken eller det låga antalet events (ska typ vara +10 events per parameter) –\> Bättre med 3 separata joint-modeller och jämföra istället? –\>

# 3 separata joint models

```{r}
lme_log_pT217_F <- lme(
  log_pT217_F ~ time_years + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

lme_log_AB42_AB40_F <- lme(
  log_AB42_AB40_F ~ time_years + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

lme_log_NfL_Q <- lme(
  log_NfL_Q ~ time_years + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

```

```{r}
# Joint model 1 
joint_model1 <- jm(
  surv_model,
  lme_log_pT217_F,
  time_var = "time_years"
  )

summary(joint_model1)
```

```{r}
# Joint model 2
joint_model2 <- jm(
  surv_model,
  lme_log_AB42_AB40_F,
  time_var = "time_years"
  )

summary(joint_model2)
```

```{r}
# Joint model 3 
joint_model3 <- jm(
  surv_model,
  lme_log_NfL_Q,
  time_var = "time_years"
  )

summary(joint_model3)
```

Alla 3 modeller har små CI –-\> Bättre säkerhet i skattningarna, mer stabila modeller - MEN multivariat joint modell har sina egna fördelar som separata modeller inte fångar –\> Vi kan göra 3 separata för uppsatsen och samtidigt göra en multivariat version i Appendix, och visar då ändå kunskapen.
