---
title: "test"
format: pdf
editor: visual
---

```{r, warning=FALSE}
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
library(ggplot2)
library(glmmTMB)
library(DHARMa) # for running residual diagnostics on GLMM
```

```{r}
# NfL data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/BLENNOWPLASMANFLLONG.rda")
nfl_data <- BLENNOWPLASMANFLLONG
# P-tau181 data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/UGOTPTAU181.rda")
ptau_data <- UGOTPTAU181
# Diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM
# Demographics data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
# Apoe data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/APOERES.rda")
apoe <- APOERES
```

```{r}
# checking observations with more than one measurement recorded on the same exam date for nfl
nfl_double <- nfl_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L) %>%
  select(RID, EXAMDATE)

# removing the doubled observations
nfl_data <- nfl_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()

# checking observations with more than one measurement recorded on the same exam date for ptau
ptau_double <- ptau_data %>%
  summarise(n = n(), .by = c(RID, EXAMDATE)) %>%
  filter(n > 1L)  %>%
  select(RID, EXAMDATE)

# extracting the doubled observations
ptau_data %>%
   filter(RID %in% ptau_double$RID) %>% 
   filter(EXAMDATE %in% ptau_double$EXAMDATE)

# removing the doubled observations
ptau_data <- ptau_data %>%
  group_by(RID, EXAMDATE) %>%
  filter(n() == 1) %>%
  ungroup()
```

```{r}
# We select the required nfl variables
nfl_data <- nfl_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMA_NFL)

# We select the required ptau variables
ptau_data <- ptau_data %>%
  select(RID, VISCODE2, EXAMDATE, PLASMAPTAU181)

# We select the required diagnosis variables
diagnosis_data <- diagnosis_data %>%
  select(ORIGPROT, COLPROT, RID, VISCODE2, EXAMDATE, DIAGNOSIS)

# We add _dx on the diagnosis variables 
diagnosis_data_dx <- diagnosis_data %>%
  rename_with(~paste0(.x, "_dx"), -c(RID, VISCODE2))

# We merge NFL and ptau data for longitudinal data by RID and VISCODE2
long_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau"))

# We merge NFL, ptau and diagnosis data for survival data, only by RID 
surv_data <- nfl_data %>%
  left_join(ptau_data, by = c("RID", "VISCODE2"), suffix = c("_nfl", "_ptau")) %>%
  left_join(diagnosis_data_dx, by = c("RID"))
```

```{r}
# We only keep the demographics variables we need and one row per person 
demographics_small <- demographics %>%
  select(RID, PTDOBYY, PTGENDER) %>%
  distinct(RID, .keep_all = TRUE)

# We create a variable for apoe4 status
apoe$APOE4_status <- ifelse(apoe$GENOTYPE %in% c("2/4","3/4","4/4"), 1, 0)

# We merge gender with merged_data and surv_data
long_data <- long_data %>%
  left_join(demographics_small, by = "RID")
surv_data <- surv_data %>%
  left_join(demographics_small, by = "RID")

# We make sure that gender is a factor 
long_data$PTGENDER <- as.factor(long_data$PTGENDER)
surv_data$PTGENDER <- as.factor(surv_data$PTGENDER)

# We merge apoe with merged_data and surv_data
long_data <- merge(long_data, apoe[, c("RID", "APOE4_status")], by="RID")
surv_data <- merge(surv_data, apoe[, c("RID", "APOE4_status")], by="RID")

# We create age_at_visit 
long_data$age_at_visit <- as.numeric(format(long_data$EXAMDATE_ptau, "%Y")) - long_data$PTDOBYY
surv_data$age_at_visit <- as.numeric(format(surv_data$EXAMDATE_ptau, "%Y")) - surv_data$PTDOBYY
```

```{r}
# We filter individuals with at least 3 measurements of ptau_data
keep = unique((ptau_data %>% group_by(RID) %>% filter(n() > 2))$RID)
long_data <- long_data %>% filter(RID %in% keep)
surv_data <- surv_data %>% filter(RID %in% keep)

# We check number of individuals with at least 3 measurements 
length(keep)
```

# Event data

## För enkel cox modell 

```{r}
# We calculate baseline date, event, and time_to_event per RID
surv_data <- surv_data %>%
  group_by(RID) %>%
  mutate(
    # We assign baseline_date as first non-dementia exam; fallback to first exam if none
    baseline_date = ifelse(
      any(DIAGNOSIS_dx != "Dementia"),
      min(EXAMDATE_dx[DIAGNOSIS_dx != "Dementia"]),
      min(EXAMDATE_dx)  # We fallback to first available exam
    ),    
    
    # Age at baseline (age at the participant's baseline visit)
    age_at_baseline = age_at_visit[EXAMDATE_dx == baseline_date][1],

    # We indicate whether this participant ever developed dementia
    event = as.integer(any(DIAGNOSIS_dx == "Dementia")),
    # We calculate time_to_event in years
    time_to_event = case_when(
      event == 1 ~ as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - baseline_date)/365,  # We calculate time to first dementia
      TRUE ~ as.numeric(max(EXAMDATE_dx) - baseline_date)/365  # We calculate time to last follow-up if censored
    )
  ) %>%
  filter(EXAMDATE_dx == baseline_date) %>%  # We keep only baseline row
  slice(1) %>%  # We take first row if multiple baseline rows
  ungroup()

# We check the number of events
table(surv_data$event)

# Example
surv_data %>%
  filter(RID == 61)
```

## För separata joint modeller 

```{r}
# We calculate baseline for each biomarker per participant
surv_data <- surv_data %>%
  group_by(RID) %>%
  mutate(
    # We calculate baseline for pTau using the first non-missing measurement
    baseline_pTau = ifelse(all(is.na(PLASMAPTAU181)), NA, min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)])),
    
    # We calculate baseline for NFL using the first non-missing measurement
    baseline_NFL  = ifelse(all(is.na(PLASMA_NFL)), NA, min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])),
    
    # We define a common baseline date as the earliest biomarker measurement
    baseline_date = pmin(baseline_pTau, baseline_NFL, na.rm = TRUE),
    
    # Age at baseline (age at the participant's baseline visit)
    age_at_baseline = age_at_visit[EXAMDATE_ptau == baseline_date | EXAMDATE_nfl == baseline_date][1],
    
    # We indicate whether the participant ever developed dementia
    event = as.integer(any(DIAGNOSIS_dx == "Dementia")),
    
    # We calculate time-to-event from the baseline date
    time_to_event = case_when(
      event == 1 ~ as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - baseline_date)/365,
      TRUE      ~ as.numeric(max(EXAMDATE_dx) - baseline_date)/365
    )
    
  ) %>%
  ungroup()

# We only keep one baseline row per participant
surv_data <- surv_data %>%
  group_by(RID) %>%
  # We keep rows where either pTau or NFL measurement matches the baseline
  filter(EXAMDATE_ptau == baseline_date | EXAMDATE_nfl == baseline_date) %>%
  slice(1) %>%  # We take the first row if multiple rows match
  ungroup()

# We count how many participants experienced the event
table(surv_data$event)

# Example
surv_data %>%
    filter(RID == 61)
```

## För multivariat joint model 

```{r}
surv_data <- surv_data %>%
  group_by(RID) %>%
  mutate(
    # We assign baseline_common as the first available measurement of either biomarker
    baseline_common = pmin(
      ifelse(all(is.na(PLASMAPTAU181)), NA, min(EXAMDATE_ptau[!is.na(PLASMAPTAU181)])),
      ifelse(all(is.na(PLASMA_NFL)), NA, min(EXAMDATE_nfl[!is.na(PLASMA_NFL)])),
      na.rm = TRUE
    ),
    
    # Age at baseline (age at the participant's baseline visit) 
    age_at_baseline = age_at_visit[EXAMDATE_ptau == baseline_common | EXAMDATE_nfl == baseline_common][1],
    
    # We indicate whether this participant ever developed dementia
    event_multi = as.integer(any(DIAGNOSIS_dx == "Dementia")),
    
    # We calculate time-to-event in years from the common baseline
    time_to_event_multi = case_when(
      event_multi == 1 ~ as.numeric(min(EXAMDATE_dx[DIAGNOSIS_dx == "Dementia"]) - baseline_common)/365,
      TRUE             ~ as.numeric(max(EXAMDATE_dx) - baseline_common)/365
    )
    
  ) %>%
  ungroup()

# We keep only one baseline row per participant for the joint model
surv_data <- surv_data %>%
  group_by(RID) %>%
  filter(EXAMDATE_ptau == baseline_common | EXAMDATE_nfl == baseline_common) %>%
  slice(1) %>%  # We take the first row if multiple rows match
  ungroup()

# We check the number of events in the multivariate dataset
table(surv_data$event_multi)

# Example
surv_data %>%
  filter(RID == 61)

```

# Time variable 

## För separata joint modeller 

```{r}
long_data <- long_data %>%
  group_by(RID) %>%
  # We create baseline for PTau
  mutate(
    baseline_date_ptau = min(EXAMDATE_ptau, na.rm = TRUE),
    time_years_ptau = as.numeric(EXAMDATE_ptau - baseline_date_ptau) / 365.25
  ) %>%
  # We create baseline for NFL
  mutate(
    baseline_date_nfl = min(EXAMDATE_nfl, na.rm = TRUE),
    time_years_nfl = as.numeric(EXAMDATE_nfl - baseline_date_nfl) / 365.25
  ) %>%
  ungroup()
```

## För multivariat joint modell

```{r}
long_data <- long_data %>%
  group_by(RID) %>%
  mutate(
    # We create a common baseline for multivariate model
    baseline_common = pmin(min(EXAMDATE_ptau, na.rm = TRUE),
                           min(EXAMDATE_nfl, na.rm = TRUE)),
    time_from_baseline_ptau = as.numeric(EXAMDATE_ptau - baseline_common)/365.25,
    time_from_baseline_nfl  = as.numeric(EXAMDATE_nfl - baseline_common)/365.25
  ) %>%
  ungroup()
```

# More preprocessing

```{r}
# We log transform the biomarkers 
long_data$log_PLASMAPTAU181 <- log(long_data$PLASMAPTAU181)
long_data$log_PLASMA_NFL <- log(long_data$PLASMA_NFL)
```

```{r}
# We remove missing values 
long_data <- na.omit(long_data)
surv_data <- na.omit(surv_data)

# We re-count missing values
sapply(long_data, function(x) sum(is.na(x)))
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
length(unique(long_data$RID))   # longitudinala data
length(unique(surv_data$RID))     # survival data
```

```{r}
# Hitta gemensamma ID:n
common_IDs <- intersect(unique(long_data$RID), unique(surv_data$RID))

# Behåll endast gemensamma individer
long_data <- long_data %>% filter(RID %in% common_IDs)
surv_data   <- surv_data %>% filter(RID %in% common_IDs)

# Kontroll
length(unique(long_data$RID))  # ska matcha
length(unique(surv_data$RID))    # ska matcha
```

# Survival modell

## För enkel cox modell 

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline amyloid", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

## För separata joint modeller 

```{r}
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit_ptau <- survfit(surv_object ~ 1, data = surv_data)
plot(
  km_fit_ptau, 
  xlab = "Days since baseline amyloid", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

```{r}
surv_model <- coxph(
  surv_object ~ age_at_baseline + APOE4_status,
  data <- surv_data
)

summary(surv_model)
```

## För multivariat joint modell

```{r}
# We create a Surv object: time and event
surv_object_multi <- Surv(
  time = surv_data$time_to_event_multi, 
  event = surv_data$event_multi
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object_multi ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline amyloid", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
surv_model_multi <- coxph(
  surv_object ~ age_at_baseline + PTGENDER + APOE4_status,
  data <- surv_data
)

summary(surv_model_multi)
```

# Longitudinal model

## För separata joint modeller 

```{r}
long_model_ptau <- lme(
  log_PLASMAPTAU181 ~ time_years_ptau + PTGENDER + APOE4_status, 
  random = ~ time_years_ptau | RID,   # random intercept + slope
  data = long_data
  )

summary(long_model_ptau)
```

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years_nfl + PTGENDER + APOE4_status, 
  random = ~ time_years_nfl | RID,   # random intercept + slope
  data = long_data
  )

summary(long_model_nfl)
```

```{r}
long_model_nfl <- lme(
  log_PLASMA_NFL ~ time_years_nfl + APOE4_status, 
  random = ~ time_years_nfl | RID,   # random intercept + slope
  data = long_data
  )

summary(long_model_nfl)
```

## För multivariat joint modell

```{r}
long_model_ptau_multi <- lme(
  log_PLASMA_PTAU181 ~ time_from_baseline_ptau + PTGENDER + APOE4_status, 
  random = ~ time_from_baseline_ptau | RID,   # random intercept + slope
  data = long_data
  )

summary(long_model_ptau_multi)
```

```{r}
long_model_nfl_multi <- lme(
  log_PLASMA_NFL ~ time_from_baseline_nfl + PTGENDER + APOE4_status, 
  random = ~ time_from_baseline_nfl | RID,   # random intercept + slope
  data = long_data
  )

summary(long_model_nfl_multi)
```

# Separata joint modeller 

```{r}
# Joint model 
joint_model_ptau <- jm(
  surv_model,
  long_model_ptau,
  time_var = "time_years_ptau",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

summary(joint_model_ptau)
```

```{r}
# Joint model 
joint_model_nfl <- jm(
  surv_model,
  long_model_nfl,
  time_var = "time_years_nfl",
  n_chains = 3L, 
  n_iter = 11000L, 
  n_burnin = 1000L
  )

summary(joint_model_nfl)
```

# Multivariat joint modell

```{r}
Mixed <- list(long_model_ptau_multi, long_model_nfl_multi)
```

```{r}
# Har modellen stabila random slopes?
summary(long_model_ptau_multi)
VarCorr(long_model_nfl_multi)
# Om slope-variansen är runt 0 → den biomarkören bidrar nästan inget i en joint model
```

```{r}
fForms <- list(
  "log_PLASMAPTAU181" = ~ value(log_PLASMAPTAU181), 
  "log_PLASMA_NFL" = ~ value(log_PLASMA_NFL) 
  )    
```

```{r}
multi_joint_model <- jm(surv_model, Mixed,
                  time_var = "baseline_common",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 20000L,
                  n_burnin = 2000L
                  )
summary(multi_joint_model)
```
