---
title: "Untitled"
format: pdf
editor: visual
---

```{r}
library(readr)
biomarkers <- read.csv("UPENN_PLASMA_FUJIREBIO_QUANTERIX_27Mar2026.csv")
```

```{r}
rid_vis_counts <- biomarkers %>%
  select(RID, VISCODE) %>%        # plocka ut kolumner
  distinct() %>%                  # ta bort dubbletter
  group_by(RID) %>%               # gruppera per deltagare
  summarise(n_measurements = n()) %>%  # räkna antal unika mätningar
  arrange(desc(n_measurements))   # sortera efter flest mätpunkter

# Visa resultatet
print(rid_vis_counts)
```

```{r}
# Räkna antal mätpunkter per individ
individual_counts <- biomarkers %>%
  group_by(RID) %>%             # gruppera per deltagare
  summarise(n_measurements = n_distinct(VISCODE)) %>%  # räkna unika tidpunkter
  ungroup() %>%
  filter(n_measurements > 2)    # bara de med >2 mätningar

# Antal individer med fler än 2 mätpunkter
n_individuals_gt2 <- nrow(individual_counts)
print(n_individuals_gt2)
```

```{r}
library(dplyr)

# Räkna antal mätpunkter per individ
measurement_distribution <- biomarkers %>%
  group_by(RID) %>%
  summarise(n_measurements = n_distinct(VISCODE)) %>%
  ungroup() %>%
  group_by(n_measurements) %>%
  summarise(n_individuals = n()) %>%
  arrange(n_measurements)

# Visa resultatet
print(measurement_distribution)
```

# Analysis

```{r, warning=FALSE}
# loading libraries
library(dplyr)
library(JMbayes2)
library(lme4) # decide which package will we use for lme
library(nlme) # decide which package will we use for lme
library(survival)
```

```{r}
plasma_data <- read.csv("UPENN_PLASMA_FUJIREBIO_QUANTERIX_27Mar2026.csv")

# We load diagnosis data
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/DXSUM.rda")
diagnosis_data <- DXSUM

# We load demographics data 
load("ADNIMERGE2_unpacked/ADNIMERGE2/data/PTDEMOG.rda")
demographics <- PTDEMOG
```

## 1. We select only the required variables and merge them

```{r}
# selecting the relevant variables -> why these ones???
plasma_data = plasma_data %>%
  select("RID", "VISCODE2", "EXAMDATE", "pT217_F",
         "AB42_F", "AB40_F", "AB42_AB40_F", "pT217_AB42_F", "NfL_Q", "GFAP_Q")

diagnosis_data = diagnosis_data %>%
  select("ORIGPROT", "COLPROT", "RID", "VISCODE2", "EXAMDATE", "DIAGNOSIS")

# merging the plasma data set and the diagnosis data set by RID (unique identifier) and recoding the diagnosis variable
merged_data = plasma_data %>%
  merge(diagnosis_data, by = "RID", suffix = c("_plasma", "_dx")) %>%
  mutate(dx = case_when(DIAGNOSIS == "Dementia" ~ 1, TRUE ~ 0))
```

## 2. We add selected demographics variables (age + gender), to later use in survival model and longitudinal model

```{r}
# We keep only the variables we need (RID, birth year, gender)
demographics_small <- demographics[, c("RID", "PTDOBYY", "PTGENDER")]

# We make sure we only have one row per person
demographics_small <- demographics_small[!duplicated(demographics_small$RID), ]

# We merge age and gender into merged_data
merged_data <- merge(
  merged_data,
  demographics_small,
  by = "RID"
)

# We make sure that gender is categorical (factor)
merged_data$PTGENDER <- as.factor(merged_data$PTGENDER)

# 5. Konvertera EXAMDATE_plasma till Date (hanterar både character och factor)
if (is.factor(merged_data$EXAMDATE_plasma)) {
  merged_data$EXAMDATE_plasma <- as.character(merged_data$EXAMDATE_plasma)
}
merged_data$EXAMDATE_plasma <- as.Date(merged_data$EXAMDATE_plasma, format="%Y-%m-%d")

# 6. Skapa age_at_visit
merged_data$age_at_visit <- as.numeric(format(merged_data$EXAMDATE_plasma, "%Y")) - merged_data$PTDOBYY
```

## 3. We filter participants with at least 2 measurements

```{r}
# 1) Filtrera personer med minst 2 mätningar
keep = unique((plasma_data %>% group_by(RID) %>% filter(n() > 2))$RID)
merged_data <- merged_data %>% filter(RID %in% keep)

# We check number of individuals with at least 2 measurements 
length(keep)

# We check exact number of measurements per individual
#table(merged_data$RID)
```

## 4. We extract baseline visits, and create an event and time-to-event variable

```{r}
# Skapar baseline-datum
merged_data <- merged_data %>%
  group_by(RID) %>%
  mutate(baseline_date = min(EXAMDATE_plasma)) %>%
  ungroup()

# Skapar event + time_to_event (Dementia räknas som event)
surv_data <- merged_data %>%
group_by(RID) %>%
summarise(
    age_at_baseline = age_at_visit[which.min(EXAMDATE_plasma)],
    PTGENDER = PTGENDER[which.min(EXAMDATE_plasma)],
    time_to_event = case_when(

    any(DIAGNOSIS == "Dementia") ~
      as.numeric(min(EXAMDATE_plasma[DIAGNOSIS == "Dementia"]) -
                   min(c(EXAMDATE_plasma, EXAMDATE_dx))),

    TRUE ~ as.numeric(max(c(EXAMDATE_plasma, EXAMDATE_dx)) -
                    min(c(EXAMDATE_plasma, EXAMDATE_dx)))) / 365,

  event = as.integer(any(DIAGNOSIS == "Dementia"))
) %>% ungroup()

# Lägger survival-data tillbaka i merged_data
merged_data <- merged_data %>%
  left_join(surv_data, by = "RID")

# We check the number of events 
table(surv_data$event)
```

## 5. We create a time variable (time from baseline)

```{r}
# 5) Skapa time_years (tid från baseline)
merged_data <- merged_data %>%
  mutate(time_years = as.numeric(EXAMDATE_plasma - baseline_date) / 365.25)
```

## 6. We check for missing values in both merged_data and long_data

```{r}
# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)
surv_data <- na.omit(surv_data)

# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))

# We count missing values per column in surv_data
sapply(surv_data, function(x) sum(is.na(x)))
```

# Longitudinal model

```{r}
# Longitudinal model
long_model <- lme(
  pT217_F ~ time_years + age_at_visit + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

summary(long_model)
```

# Survival model 

```{r}
# We create a Surv object: time and event
surv_object <- Surv(
  time = surv_data$time_to_event, 
  event = surv_data$event
)
```

```{r}
# We fit Kaplan-Meier curve
km_fit <- survfit(surv_object ~ 1, data = surv_data)

# Plot
plot(
  km_fit, 
  xlab = "Days since baseline plasma", 
  ylab = "Survival probability (no dementia)", 
  main = "Kaplan-Meier Curve for Dementia"
)
```

```{r}
# Cox survival-model with covariates
surv_model <- coxph(
  surv_object ~ age_at_baseline + PTGENDER,
  data = surv_data
)

summary(surv_model)
```

```{r}
cox.zph(surv_model)
```

# Joint model

```{r}
# Joint model 
joint_model <- jm(
  surv_model,
  long_model,
  time_var = "time_years"
  )

summary(joint_model)
```

# Multivariate model

```{r}
# Vi kollar efter outliers 
boxplot(merged_data$pT217_AB42_F, main="pT217_AB42_F")
boxplot(merged_data$AB42_AB40_F, main="AB42_AB40_F")
boxplot(merged_data$NfL_Q, main="NfL_Q")
```

```{r}
# Vi transformerar dem
merged_data$log_pT217_AB42_F <- log(merged_data$pT217_AB42_F)
merged_data$log_AB42_AB40_F <- log(merged_data$AB42_AB40_F)
merged_data$log_NfL_Q <- log(merged_data$NfL_Q)
```

```{r}
sapply(merged_data, function(x) sum(is.na(x)))
```

```{r}
merged_data <- na.omit(merged_data)

# We count missing values per column in merged_data
sapply(merged_data, function(x) sum(is.na(x)))
```

```{r}
library(nlme)

fm_pT217 <- lme(log_pT217_AB42_F ~ time_years + age_at_visit + PTGENDER.x,
                random = ~ time_years | RID,
                data = merged_data)

fm_ABratio <- lme(log_AB42_AB40_F ~ time_years + age_at_visit + PTGENDER.x,
                  random = ~ time_years | RID,
                  data = merged_data)

fm_NfL <- lme(log_NfL_Q ~ time_years + age_at_visit + PTGENDER.x,
              random = ~ time_years | RID,
              data = merged_data)
```

```{r}
Mixed <- list(fm_pT217, fm_ABratio, fm_NfL)
```

```{r}
fForms <- list(
  "log_pT217_AB42_F" = ~ value(log_pT217_AB42_F) + slope(log_pT217_AB42_F),
  "log_AB42_AB40_F" = ~ value(log_AB42_AB40_F),
  "log_NfL_Q" = ~ value(log_NfL_Q) + slope(log_NfL_Q)
)
```

```{r}
joint_model <- jm(surv_model, Mixed,
                  time_var = "time_years",
                  functional_forms = fForms,
                  n_chains = 3L,    
                  n_iter = 20000L,
                  n_burnin = 2000L
                  )
summary(joint_model)
```

```{r}
summary <- summary(joint_model)
summary$Survival 
```

```{r}
plot(Mixed[[1]])
plot(Mixed[[2]])
plot(Mixed[[3]])
```

Instabil modell dock (för höga Rhat-värden, d.v.s. \> 1) –\> Går ej att tolka

Förklaring = För få events (19 st) - Man bör ha \> 10 events per parameter!!!

Lösning = Göra joint modell för varje biomarkör separat och jämföra –\>

```{r}
lme_log_pT217_AB42_F <- lme(
  log_pT217_AB42_F ~ time_years + age_at_visit + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

lme_log_AB42_AB40_F <- lme(
  log_AB42_AB40_F ~ time_years + age_at_visit + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )

lme_log_NfL_Q <- lme(
  log_NfL_Q ~ time_years + age_at_visit + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )
```

```{r}
log_NfL_Q
lme_log_NfL_Q <- lme(
  log_NfL_Q ~ time_years + age_at_visit + PTGENDER.x, 
  random = ~ time_years | RID,   # random intercept + slope
  data = merged_data
  )
```

```{r}
# Joint model 1 
joint_model1 <- jm(
  surv_model,
  lme_log_pT217_AB42_F,
  time_var = "time_years"
  )

summary(joint_model1)
```

```{r}
# Joint model 2
joint_model2 <- jm(
  surv_model,
  lme_log_AB42_AB40_F,
  time_var = "time_years"
  )

summary(joint_model2)
```

```{r}
# Joint model 3 
joint_model3 <- jm(
  surv_model,
  lme_log_NfL_Q,
  time_var = "time_years"
  )

summary(joint_model3)
```
